### 2021-02-14 rad-icfp-2021/abstract.tex
initial commit ifcp paper

\begin{abstract}
Our work aims at bridging the gap between high-level functional reverse-mode differentiation and efficient imperative implementations.
We define a new source-to-source transformation for reverse-mode differentiation on a standard higher-order purely functional language. 
Using a logical relations argument, we prove our transformation correct using diffeological spaces as a denotational semantics for our language. 
We then show that combining our macro with some classical optimisations lead to low-level SSA-like code, which is as efficient as low-level imperative implementations. 
We prove some theoretical complexity guarantees and provide experimental results which showcase the efficiency of our method. 
\end{abstract}

### 2021-02-14 rad-icfp-2021/background.tex
initial commit ifcp paper

\section{Pure $\lambda$-calculus for differentiable programming} % (fold)
\label{sec:pure_lambda_calculus_for_differentiable_programming}

\subsection{Source language} % (fold)
\label{sub:source_language}

\subsection{Target Language} % (fold)
\label{sub:target_language}

% subsection target_language (end)

### 2021-02-14 rad-icfp-2021/correctness.tex
initial commit ifcp paper

\section{Correctness}

%decompose the macro from section 4 into unary form transfo, then principled macro, then opti
\MH{Question: when is it best to give the idea about our macro? that it's UNF+macro+easy-opti?}
\MH{probably won't use UNF in this paper}

\subsection{Unary Form} % (fold)
\label{sub:unary_form}

\subsection{Principled Macro} % (fold)
\label{sub:principled_macro}

\subsection{Correctness via subsconing} % (fold)
\label{sub:correctness_via_subsconing}

\subsection{Optimizations} % (fold)
\label{sub:optimizations}
%the ones in the macro of Sec4 then the further classical ones to ANF

\subsection{Correctness of the macro from section \ref{}} % (fold)
\label{sub:correctness_of_the_macro_from_section_ref}

### 2021-02-14 rad-icfp-2021/intro.tex
initial commit ifcp paper

\section{Introduction}

\subsection{Problem} % (fold)
\label{sub:problem}

\TODO{find simple yet telling concrete example here}
Consider the following 
TODO: Jacobian, chain rule in maths
It is easy to write a purely functional implementation of it
TODO: code using FP
However, if we run it naively it will take 
TODO: time or complexity
A good implementation would instead take
TODO: time or complexity
and looks like
TODO: imperative code

There is a huge difference between the purely functional yet inefficient implementation and the efficient imperative one. The first one is also easy to prove correct, and purely functional code can benefit from a lot of optimisations \cite{}. The imperative version is essentially a manually optimised representation, yet it is error prone. In addition, the imperative code will usually be transformed to SSA-form \cite{} for further optimisations by the compiler. It is often believed \cite{} that a purely functional implementation of reverse-mode differentiation will be unefficient and that there is a need for imperative features such as references or the use of intermediate representations \cite{} to reach good performance.
\TODO{Add here we want to investigate this for rich HO languages, which offer more possibilities for optimisations and cleaner to write. detail more in problem section that this unified way would be good for ||, opti, correctness, etc.}

\subsection{Contribution} % (fold)
\label{sub:contribution}

\MH{Be clearer and more precise here as to why our contribution solves the problem above.}

This work aims at bridging the gap between proved-correct high-level idealized purely functional differentiation and efficient low-level imperative implementations.  We do this by providing a correct, efficient, and purely functional implementation. This allows for more direct optimisations and paves the way for future work. In short, after detailing more the problem (Sec. \ref{}) and our approach (Sec. \ref{}) we provide a source and target language for differentiation (Sec. \ref{}) which we see as a source-to-source transformation \cite{}. We then propose a macro for differentiation (Sec. \ref{}). To prove our transformation correct, we give our language a denotational semantics using diffeological spaces (Sec. \ref{}) and show correctness via a categorical version of logical relations (Sec. \ref{}). We apply some standard optimisation techniques on the code.
We finally explore some theoretical guarantees of the resulting code (Sec. \ref{}) and give some experimental results (Sec. \ref{}).

% \subsubsection{Brief overview} % (fold)
% \label{sub:brief_overview}
%of the problem: code example

% % diff prog, what state of the art is, what we propose, roughly our contribution
% Under the slogan of differentiable programming, Automatic Differentiation has seen a renewal in the machine learning community over the past few years. The need for fast gradient computations of code implementing functions is ubiquitous and core in deep learning. Recent work has either focused on correct differentiation on high-level functional languages or about fast, usually low-level, imperative and unpure implementations. This work aims to bridge the gap between the two, first by providing a high-level source-to-source differentation on a high-level language, which is then optimised in several ways. Doing so, we reach state-of-the-art performance and ultimately produce code that is similar to the manually-optimised imperative code that one could get. 
% %maybe add that it's principled, explains more things, is correct, and great for generalisation to other languages/libraries/etc.

\subsection{Our approach} % (fold)
\label{sub:our_approach}

\MH{start from pure HO functional language with smooth primitives, use an optimised (principled) macro, then further classical optimisations. forward def to things in more detail. list of key insights.}

%rename Context?

\subsection{Differentiable programming} % (fold)
\label{sub:differentiable_programming}

%as a "solution" people found? maybe skip this entirely

%define AD
%mostly from Matthijs' POPL paper
(Source-to-source) Automatic Differentiation (AD) is a technique for transforming code that implements 
a function $f$ into code that computes $f$'s derivative, essentially by using the chain rule for derivatives. 
Due to its efficiency and numerical stability, AD is the technique of choice whenever derivatives need to be computed 
of functions that are implemented as programs, particularly in high dimentional settings. 
Optimisations and Monte Carlo integration algorithms, such as Gradient Descent and Hamiltonian Monte-Carlo methods, 
rely crucially on the calculation of derivatives. These algorithms are used in virtually everywhere in machine learning 
and computational statistics, and the calculation of derivatives is usually the computational bottle-neck. 
AD, roughly speaking, comes in two modes; forward-mode and reverse-mode. 
When differentiating a function $\RR^n\to\RR^m$, forward-mode tends to be more efficient if $m \gg n$, while reverse-mode 
is generally more performant if $n \gg m$. As most applications reduce to optimization of Monte-Carlo 
integration of an objective function $\RR^n\to\RR$ with $n$ very large ($10^4-10^7$), reverse-mode AD is 
in many ways the more interesting algorithm.

%define diff prog beyond AD, be specific with references/different define-then-run vs Pytorch/Swift etc.
%Diff Curry has OK intro

### 2021-02-14 rad-icfp-2021/macrodiff.tex
initial commit ifcp paper

\section{A macro for source-to-source reverse-mode differentiation}

### 2021-02-14 rad-icfp-2021/relwork.tex
initial commit ifcp paper

\section{Conclusion}

\subsection{Summary} % (fold)
\label{sub:summary}

\subsection{Related Work} % (fold)
\label{sub:related_work}

\noindent \textbf{Automatic Differentiation.}

\noindent \textbf{Array Languages and Fusion.}

\noindent \textbf{Numerical DSLs.}

\noindent \textbf{Correctness of AD in functional languages.}

\noindent \textbf{Comparison to other recent papers.}

\subsection{Discussion and future work} % (fold)
\label{sub:discussion_and_future_work}

\noindent \textbf{Other Language features.}
%more primitives, proved correct and more efficient. extend library

\noindent \textbf{Probabilistic Programming.}
%find something simple and precise, o.w. useless section

\noindent \textbf{More dynamic language}
%TF is losing ground to Pytorch, Swift tries to renew with Pytorch-like dynamic features.

\noindent \textbf{More optimisations.}
%maybe?

\noindent \textbf{Limitations.}
%maybe?

\MH{useful links (Swift is close to our project, and the doc is very readable):}

\begin{verbatim}https://docs.google.com/document/d/1_BirmTqdotglwNTOcYAW-ib6mx_jl-gH9Dbg4WmHZh0/edit# \end{verbatim}
\begin{verbatim}

	https://github.com/apple/swift/blob/master/docs/DifferentiableProgramming.md#approaches-to-automatic-differentiation
\end{verbatim}

\subsection{State-of-the-art} % (fold)
\label{sub:state_of_the_art}

All the existing work have either focused on efficiency or correctness.

We present \system{}, a correct and efficient functional differentiable programming framework. This is the first work that propose a pure reverse-mode AD on a higher-order functional language with a correctness proof and complexity guarantees that is implemented and achieves good practical performance. In Figure\ref{fig:comparison-table} we compare our work to the recent papers in the field.

\MH{Separate complexity into complexity and performent.}
\MH{reorder topics: theory on the left, practice on the right, to make the dichotomy more apparent}
\MH{most refs here are pure, which is not good for our emphasis on pureness, need to add more practical papers}
\begin{table}
\label{fig:comparison-table}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|}
\hline
 & Reverse  & Complexity & Impl. & Pure & Correct & HO Fun. & Tensors & Rec. & Cond. & HO $\partial$ ? \\
\hline
\system{} (This Paper) &
\supfull & \supfull & \supfull & \supfull & \supfull & \supfull  & \supfull  & \supfull & \supfull & \supfull \\ 
\hline
Lantern~\cite{lantern_icfp} & 
\supfull & \suphalf & \supfull & \supnone & \supnone & \supfull & \supfull & \supfull & \supfull &? \\ 
\hline
\dfsmooth{}~\cite{shaikhha2019efficient} 
& 
\supnone & \supnone & \supfull & \supfull & \supnone & \supfull & \supfull  & \supnone  & \supnone &? \\ 
\hline
\cite{huot2020correctness} &
\supfull & \supnone & \supnone & \supfull & \supfull & \supfull & \supnone & \suphalf & \supfull &?\\ 
\hline
\cite{brunel2019backpropagation} &
\supfull & \suphalf ? & \supnone & \supfull & \supfull & \supfull & \supnone & \supnone & \supnone &?\\ 
\hline
\cite{abadi2019simple} &
\supfull & \supfull? & \supnone & \supfull & \supfull & \supnone & \supnone? & \supfull & \supfull &?\\ 
\hline
\cite{barthe2020versatility} &
\supnone & \supnone & \supnone & \supfull & \supfull & \supnone? & \supnone & \supnone & \supfull &?\\ 
\hline
\cite{pearlmutter2008reverse} &
\supfull & \supfull (not proved) & \supfull & \supnone & \supnone(not proved) & \supfull & \supnone? & \supfull & \supfull &?\\ 
\hline
\cite{Elliott:2018:SEA:3243631.3236765} &
\supfull & \suphalf & \suphalf & \supfull & \supnone & \supnone & \supnone? & \supnone & \supnone? &?\\ 
\hline
\cite{sherman2020lambda_s} & \suphalf or \supnone? & \supnone & \supfull & \supfull & \supfull? & \supfull & \supnone? & \supnone & \supnone &?\\ 
\hline
\cite{vytiniotis2019differentiable} &
\supfull & \suphalf & \suphalf & \supfull & \supnone & \supfull & \supnone? & \supnone & \supnone? &?\\ 
\hline
\cite{mak2020differential} & \supfull & \suphalf ? & \supnone & \supfull & \supfull & \supfull & \supnone & \supnone & \supnone &?\\ 
\hline
\cite{vakar2020reverse} & \supfull & \suphalf ? & \supnone & \supfull & \supfull & \supfull & \supnone & \supnone & \supnone &?\\ 
\hline
\cite{Manzyuk2012} & \supnone & \supnone & \supnone & \supfull & \supfull? & \supfull & \supnone & \supnone & \supnone &?\\ 
\hline 
\cite{gallagher-sdg} & \supnone & \supnone & \supnone & \supfull & \supfull ? & \supfull & \supnone & \supfull & \supfull & ?  \\ 
\hline
\end{tabular}
\caption{Comparison of different differentiable programming frameworks.}
A $\supfull$ means the property is verified, and $\supnone$ means its absent from the work. Reverse stands for reverse-mode differentiation, complexity to whether there is a proof of complexity and whether its competitive, impl. stands for running implementation, pure for whether differentiation is pure or effectful, correct to whether itis proved correct, HO Fun. stands for Higher-order functions, Tensors to efficient array like structures, Rec. for term-recurion, Cond. for conditionals, and finally HO $\partial$ for higher-order derivatives.
\end{table}
%Actually very much in the spirit of Swift and Julia Zygote (see Diff.Curry intro), probably refer to this somewhere

\MH{we only have a limited form of recursion, maybe make that clearer. and we don't really have tensors, we have arrays. we don't have conditionals yet but that's something we can think of. we only have limited HO too for performance reason. maybe we should heavily emphasize we can add more but then it's harder to get guarantees and maybe not a great idea for speed...}

\MH{understand related work sentence from Diff. Curry paper: The idea of using closures as back-propagators is receiving recent attention. For example Julia Zy- gote [15] and Swift AD adopts this design. Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.}

\clearpage

### 2021-02-14 rad-icfp-2021/background.tex
some comments

@@
-\section{Pure $\lambda$-calculus for differentiable programming} % (fold)
+\section{Pure $\lambda$-calculus for differentiable programming} 
 \label{sec:pure_lambda_calculus_for_differentiable_programming}
@@
 
+\MH{is this useful to present both languages?
+not sure.
+also not sure when to talk about the restrictions for efficiency.
+but again it's a theory paper mostly so we don't want to talk too much about that.
+just a ref to some papers that do that too.}
+\MH{will probably just be a subsection}
+
 \subsection{Target Language} % (fold)

### 2021-02-14 rad-icfp-2021/intro.tex
minor

@@
 
-\subsection{Problem} % (fold)
-\label{sub:problem}
+\MH{might get some inspiration from older notes}
 
-\TODO{find simple yet telling concrete example here}
-Consider the following 
-TODO: Jacobian, chain rule in maths
-It is easy to write a purely functional implementation of it
-TODO: code using FP
-However, if we run it naively it will take 
-TODO: time or complexity
-A good implementation would instead take
-TODO: time or complexity
-and looks like
-TODO: imperative code
+% \subsection{Problem} % (fold)
+% \label{sub:problem}
 
-There is a huge difference between the purely functional yet inefficient implementation and the efficient imperative one.
+% \TODO{find simple yet telling concrete example here}
+% Consider the following 
+% TODO: Jacobian, chain rule in maths
+% It is easy to write a purely functional implementation of it
+% TODO: code using FP
+% However, if we run it naively it will take 
+% TODO: time or complexity
+% A good implementation would instead take
+% TODO: time or complexity
+% and looks like
+% TODO: imperative code
+
+% There is a huge difference between the purely functional yet inefficient implementation and the efficient imperative one.
 The first one is also easy to prove correct, and purely functional code can benefit from a lot of optimisations \cite{}.
@@
 It is often believed \cite{} that a purely functional implementation of reverse-mode differentiation will be unefficient and that there is a need for imperative features such as references or the use of intermediate representations \cite{} to reach good performance.
-\TODO{Add here we want to investigate this for rich HO languages, which offer more possibilities for optimisations and cleaner to write.
+% \TODO{Add here we want to investigate this for rich HO languages, which offer more possibilities for optimisations and cleaner to write.
 detail more in problem section that this unified way would be good for ||, opti, correctness, etc.}
 
-\subsection{Contribution} % (fold)
-\label{sub:contribution}
+% \subsection{Contribution} % (fold)
+% \label{sub:contribution}
 
-\MH{Be clearer and more precise here as to why our contribution solves the problem above.}
+% \MH{Be clearer and more precise here as to why our contribution solves the problem above.}
 
-This work aims at bridging the gap between proved-correct high-level idealized purely functional differentiation and efficient low-level imperative implementations.
+% This work aims at bridging the gap between proved-correct high-level idealized purely functional differentiation and efficient low-level imperative implementations.
 We do this by providing a correct, efficient, and purely functional implementation.
@@
 We apply some standard optimisation techniques on the code.
-We finally explore some theoretical guarantees of the resulting code (Sec. \ref{}) and give some experimental results (Sec. \ref{}).
+% We finally explore some theoretical guarantees of the resulting code (Sec. \ref{}) and give some experimental results (Sec. \ref{}).
 
-% \subsubsection{Brief overview} % (fold)
-% \label{sub:brief_overview}
-%of the problem: code example
+% % \subsubsection{Brief overview} % (fold)
+% % \label{sub:brief_overview}
+% %of the problem: code example
 
-% % diff prog, what state of the art is, what we propose, roughly our contribution
-% Under the slogan of differentiable programming, Automatic Differentiation has seen a renewal in the machine learning community over the past few years.
+% % % diff prog, what state of the art is, what we propose, roughly our contribution
+% % Under the slogan of differentiable programming, Automatic Differentiation has seen a renewal in the machine learning community over the past few years.
 The need for fast gradient computations of code implementing functions is ubiquitous and core in deep learning.
@@
 Doing so, we reach state-of-the-art performance and ultimately produce code that is similar to the manually-optimised imperative code that one could get.
-% %maybe add that it's principled, explains more things, is correct, and great for generalisation to other languages/libraries/etc.
+% % %maybe add that it's principled, explains more things, is correct, and great for generalisation to other languages/libraries/etc.
 
-\subsection{Our approach} % (fold)
-\label{sub:our_approach}
+% \subsection{Our approach} % (fold)
+% \label{sub:our_approach}
 
-\MH{start from pure HO functional language with smooth primitives, use an optimised (principled) macro, then further classical optimisations.
+% \MH{start from pure HO functional language with smooth primitives, use an optimised (principled) macro, then further classical optimisations.
 forward def to things in more detail.
@@
 
-%rename Context?
+% %rename Context?
 
-\subsection{Differentiable programming} % (fold)
-\label{sub:differentiable_programming}
+% \subsection{Differentiable programming} % (fold)
+% \label{sub:differentiable_programming}
 
-%as a "solution" people found? maybe skip this entirely
+% %as a "solution" people found? maybe skip this entirely
 
-%define AD
-%mostly from Matthijs' POPL paper
-(Source-to-source) Automatic Differentiation (AD) is a technique for transforming code that implements 
-a function $f$ into code that computes $f$'s derivative, essentially by using the chain rule for derivatives. 
-Due to its efficiency and numerical stability, AD is the technique of choice whenever derivatives need to be computed 
-of functions that are implemented as programs, particularly in high dimentional settings. 
-Optimisations and Monte Carlo integration algorithms, such as Gradient Descent and Hamiltonian Monte-Carlo methods, 
-rely crucially on the calculation of derivatives. These algorithms are used in virtually everywhere in machine learning 
-and computational statistics, and the calculation of derivatives is usually the computational bottle-neck. 
-AD, roughly speaking, comes in two modes; forward-mode and reverse-mode. 
-When differentiating a function $\RR^n\to\RR^m$, forward-mode tends to be more efficient if $m \gg n$, while reverse-mode 
-is generally more performant if $n \gg m$. As most applications reduce to optimization of Monte-Carlo 
-integration of an objective function $\RR^n\to\RR$ with $n$ very large ($10^4-10^7$), reverse-mode AD is 
-in many ways the more interesting algorithm.
+% %define AD
+% %mostly from Matthijs' POPL paper
+% (Source-to-source) Automatic Differentiation (AD) is a technique for transforming code that implements 
+% a function $f$ into code that computes $f$'s derivative, essentially by using the chain rule for derivatives. 
+% Due to its efficiency and numerical stability, AD is the technique of choice whenever derivatives need to be computed 
+% of functions that are implemented as programs, particularly in high dimentional settings. 
+% Optimisations and Monte Carlo integration algorithms, such as Gradient Descent and Hamiltonian Monte-Carlo methods, 
+% rely crucially on the calculation of derivatives. These algorithms are used in virtually everywhere in machine learning 
+% and computational statistics, and the calculation of derivatives is usually the computational bottle-neck. 
+% AD, roughly speaking, comes in two modes; forward-mode and reverse-mode. 
+% When differentiating a function $\RR^n\to\RR^m$, forward-mode tends to be more efficient if $m \gg n$, while reverse-mode 
+% is generally more performant if $n \gg m$. As most applications reduce to optimization of Monte-Carlo 
+% integration of an objective function $\RR^n\to\RR$ with $n$ very large ($10^4-10^7$), reverse-mode AD is 
+% in many ways the more interesting algorithm.
 
-%define diff prog beyond AD, be specific with references/different define-then-run vs Pytorch/Swift etc.
-%Diff Curry has OK intro
+% %define diff prog beyond AD, be specific with references/different define-then-run vs Pytorch/Swift etc.
+% %Diff Curry has OK intro

### 2021-02-16 rad-icfp-2021/abstract.tex
some very rough starting draft

@@
 \begin{abstract}
-Our work aims at bridging the gap between high-level functional reverse-mode differentiation and efficient imperative implementations.
-We define a new source-to-source transformation for reverse-mode differentiation on a standard higher-order purely functional language. 
-Using a logical relations argument, we prove our transformation correct using diffeological spaces as a denotational semantics for our language. 
-We then show that combining our macro with some classical optimisations lead to low-level SSA-like code, which is as efficient as low-level imperative implementations. 
-We prove some theoretical complexity guarantees and provide experimental results which showcase the efficiency of our method. 
+Recent work on automatic differentiation (AD) either focus on expressiveness and correctness, or focus on efficiency. 
+Reverse mode implementations are notoriously hard to optimize as they add unpurity to the transformed program, and harder to prove correct or efficient.
+We start from a high-level picture on AD where we can prove correctness and keep purity, and show how to turn it into efficient lower level code. 
+On a small functional language, we show how the optimized code after AD looks like static single assignment (SSA) code. 
+On a more expressive language with higher-order primitives, we introduce an intermediate representation, which we call unary form (UNF), 
+which enables us to perform efficient reverse-mode on these constructs while keeping purity and efficiency.
+We give a denotational semantics to our language, we show correctness of our reverse-mode transformation, and we prove complexity guarantees. 
 \end{abstract}

### 2021-02-16 rad-icfp-2021/background.tex
some very rough starting draft

--- 2021-02-14 rad-icfp-2021/background.tex
@@
-\section{Pure $\lambda$-calculus for differentiable programming} % (fold)
-\label{sec:pure_lambda_calculus_for_differentiable_programming}
+\section{Rudiments of AD}
 
-\subsection{Source language} % (fold)
-\label{sub:source_language}
+\subsection{Rudiments of differentiation and dual numbers.}
+Recall that the derivative of a function $f:\RR\to \RR$, if it exists, is a function
+$\nabla f:\RR\to \RR$ such that $\nabla f(x_0)=\frac {\dif f(x)}{\dif x}(x_0)$ is the gradient of $f$ at $x_0$.
 
-\subsection{Target Language} % (fold)
-\label{sub:target_language}
+To find $\nabla f$ in a compositional way, two generalizations are reasonable:
+\begin{itemize}
+\item We need both $f$ and $\nabla f$ when calculating $\nabla (f;g)$
+of a composition $f;g$, using the chain rule, so we are really interested in the pair $(f,\nabla f):\RR\to \RR\times \RR$;
+\item In building $f$ we will need to consider functions of multiple arguments, such as $+:\RR^2\to \RR$, and these functions should propagate derivatives.
+\end{itemize}
+Thus we are more generally interested in transforming a function $g:\RR^n\to \RR$ into a function
+$h:(\RR\times \RR)^n\to \RR\times \RR$ in such a way that for any
+$f_1\dots f_n:\RR\to\RR$, 
+\begin{equation}
+  \label{eqn:dualnumber}
+  (f_1,\nabla f_1,\dots, f_n,\nabla f_n);h
+  =
+  ((f_1,\dots, f_n);g,\nabla ((f_1, \dots, f_n);g))\text.
+\end{equation}
 
-% subsection target_language (end)
+An intuition for $h$ is often given in terms of dual numbers.
+The transformed function operates on pairs of numbers, $(x,x')$, and it is common
+to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
+But while this is a helpful intuition, the formalization of infinitesimals can be intricate, 
+and the development in this paper is focussed on the elementary formulation in~\eqref{eqn:dualnumber}.
+
+The reader may also notice that $h$ encodes all the partial derivatives of
+$g$. For example, 
+if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
+$h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
+and similarly 
+$h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
+And conversely, if $g$ is differentiable in each argument, then
+a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
+combinations of partial derivatives:
+\[\textstyle h(x_1,x_1',x_2,x_2')=(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))\text.\]
+
+In summary, the idea of differentiation with dual numbers is 
+to transform a differentiable function
+$g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives.
+We packaged this up in~\eqref{eqn:dualnumber} as a sort-of invariant which is useful for building derivatives of compound functions $\RR\to\RR$ in a compositional way.
+The idea of automatic differentiation is to perform this transformation at the source code level.
+
+The two main ways in which it is performed in practice is by operator overloading \cite{} or by source code transformation. We focus our work on the latter.
+
+\subsection{Reverse-mode AD}
+
+AD comes in two main flavors: forward mode and reverse mode. 
+Forward mode computes a directional derivative of the Jacobian while reverse-mode computes a directional derivative of the transpose Jaobian. 
+This has a significant impact in practice as this means reverse mode can compute the whole gradient of a function $f:\RR^n\to \RR$ in one pass, whereas forward mode needs $n$ passes. 
+Assuming that both have about the same complexity, this can have a huge impact in practice. 
+Such functions are expremely commonly seen as loss functions in the context of machine learning for instance.
+
+Reverse-mode computes the derivatives backward, using the symmetry of the chain rule.
+This means the compute flow is reverted after the first pass to compute the value of the function and its intermediate values. 
+These values appear in the chain rule so should be stored or recomputed.
+These things make reverse-mode harder to implement efficiently, and especially harder to analyse and optimise. 
+Traditional implementions use references.
+For instance, let 
+ \begin{align*}
+     \trm\defeq &\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
+ \end{align*}
+ The imperative version of the reverse differentiation of $\trm$ is
+	\begin{align*}
+		&w_1=\var_1 * \var_2 \\
+		&w_2=w_1 * \var_1 \\
+        &\var_1':=0~;~\var_2':=0 \\
+        &w_1':=0 \\
+		&w_2':=1 \\
+		&w_1'+= \var_1 * w_2'; \var_1'+= w_1*w_2' \\
+		&\var_1'+= \var_2 * w_1'; \var_2'+=\var_1 * w_1'\\
+		&\text{return }(\var_1',\var_2')  
+	\end{align*}
+The tangent part for every variable from the context $\var_i'$ and every intermediate variable $w_i'$ is initialized at $0$, 
+except for the return variable $w_2$ for which  $w_2':=1$.
+
+As noted in \cite{pearlmutter2008reverse}, in a functional language, one can have a purely functional reverse-mode using continuations. 
+This representation is notoriously inefficient and all the implementions of reverse mode in a functional setting have followed the lines of the imperative version above \cite{pearlmutter2008reverse, wang2018demystifying, baydin2016diffsharp}.
+
+In the next section, we show on a very simple language how we can use standard optimisations to turn this inefficient representation into a very efficient one which is still purely functional.
+The remaining of the paper will generalise this idea and explore some consequences of this new approach.

### 2021-02-16 rad-icfp-2021/intro.tex
some very rough starting draft

@@
 \section{Introduction}
-
-\MH{might get some inspiration from older notes}
 
@@
 % %Diff Curry has OK intro
+
+\[
+\begin{tikzcd}
+	& \TSSyn \ar[r,"{\Dsynrev[\rho]{-}}"] 
+	& \TTSynHO_1 \ar[d,"\UNF{-}^{-1}"] &  \\
+	& \SSSyn \ar[u,"\UNF{-}"]  \ar[r,"{\Dsynrev[\rho]{-}^*}"'] 
+	& \TTSynHO_2 \ar[d,"Jacob"] \\
+	& \SSSyn \ar[u,"ANF"] \ar[r,"{\Dsynrev[\rho]{-}^{**}}"'] 
+	&  \TTSynHO_3 \ar[dl,"\partial_2"] \\
+	\SSSynHO \ar[ur,"\partial_1"]  \ar[rrr,bend right=20,color=red,"Example\,1"]
+	& \TTSyn \ar[r,"OPT"'] 
+	& \TTSyn \ar[r,"\partial_3"']
+	& \TTSyn 
+\end{tikzcd}
+\]
+
+\begin{itemize}
+	\item $\SSSynHO$ is a higher-order language and the language the user uses to write a prgram.
+	\item $\partial_1$ denotes partial evaluation of all higher-order variables. We end up with a program whose only higher-order bits are primitive operators.
+	\item $\SSSyn$ is a first-order language with a few higher-order primitives like map,fold,etc.
+	\item $ANF$ is for (weak) A-normal form. Every operator $\op$ is only given variables as arguments. We do this by introducting additional let bindings. It is only used for the $Jacob$ optimisation.
+	\item $\UNF{-}$ is the unary form transformation presented in the next section. Its goal is to deal with the fanout problem before applying the reverse-mode macro.
+	\item $\TSSyn$ is the target language for $\UNF{-}$ and also the source for the reverse-mode macro. It has a bit of a special structure.
+	\item $\Dsynrev[\rho]{-}$ is a reverse-mode macro. It is efficient on unary operators but does not deal well with fanout. Fortunately, it is only given unary operators at this point.
+	\item $\TTSynHO_1$ is the target language of the reverse-mode macro. It is a higher-order language as it needs a continuation for the reverse derivatives.
+	\item $\TTSynHO_2,\TTSynHO_3$ are small variations of $\TTSynHO_1$.
+	\item $\UNF{-}^{-1}$ is an optimisation which removes the overhead caused by $\UNF{-}$. It reverts the primal part of the derivative to the original term.
+	\item $\Dsynrev[\rho]{-}^*$ is an efficient macro for reverse-mode which deals with the fanout problem.
+	\item $Jacob$ stands for an efficient representation of the Jacobian.
+It needs a weak form of ANF to compute the Jacobian with a single vector and does not need to use a matrix-vector multiplication.
+In particular, it does not need to store the matrix.
+	\item $\Dsynrev[\rho]{-}^{**}$ is the efficient macro for reverse mode that would really be implemented. It will be quicker to run than the composition of the transformations. 
+	\item $\partial_2$ stands for partial evaluation. We evaluate (CBN) the higher-order variables binding the lambdas we introduced and inlined with the reverse-mode macro
+	\item $\TTSyn$ is the target language of the previous partial evaluation and the final language we use.
+	\item $OPT$ is an optimisation. First, we remplace the previously inlined lambdas by let bindings, and then apply forward substitution.
+	\item $\partial_3$  is optionnal. We partially evaluate more and do algebraic simplifications, like $0+\trm\to\trm, 1*\trm\to \trm$.
+	\item $Example\,1$ is detailed below.
+\end{itemize}
+
+\MH{I could add a different name for the target of ANF, of Jacob, and of OPT.}
+
+\begin{example}
+	Let 
+\begin{align*}
+	\var_1,\var_2,\var_3:\RR\vdash \trm:=\big(\lambda f:\RR\to\RR. (f(\var_1))*\var_1\big)(\lambda \var[2]. \var[2]* \var_2)
+\end{align*}
+After $\partial_1$ we obtain 
+\begin{align*}
+	(\letin{\var[2]}{\var_1}{\var[2]*\var_2})*\var_1
+\end{align*}
+After $ANF$ we obtain
+\begin{align*}
+	&\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
+\end{align*}
+After $\Dsynrev[\rho]{-}^*$ we obtain
+	\begin{align*}
+		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Y\big(J^1(\var_1,\var_2,\var_3)^T(\var[2]_1,...,\var[2]_4)\big)} }{
+		\\
+		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'\big(J^2(\var_1,\var_2,\var_3,w_1)^T(\var[2]_1,...,\var[2]_5)\big)}}{\\ 
+		& w_2'}}
+	\end{align*}
+where $Y$ is a continuation variable of type $\RR^3\to\RR^3$, $J^1,J^2$ are Jacobian matrices for $*$ operators which also return their context, $\inllambda$ is just a normal $\lambda$ but inlined for future optimisation.
+
+After applying $Jacob$, we obtain
+	\begin{align*}
+		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Y(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
+		\\
+		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
+		& w_2'}}
+	\end{align*}
+To actually compute the reverse-derivative of the full term, we substitute the identity to $Y$ and apply the whole term to $(0,0,0,0,1)$, the functional equivalent to initialing the sensitivities in the imperative reverse-mode algorithm.
+We obtain
+
+\begin{align*}
+		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
+		\\
+		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
+		& w_2'(0,0,0,0,1)}}
+	\end{align*}
+
+After $\partial_2$, we get
+	\begin{align*}
+		&\letin{w_1}{\var_1 * \var_2}{\\
+		&\letin{w_2}{w_1 * \var_1}{\\
+		&\bigg[\inllambda \var[2]_1...\var[2]_5.\Big(\inllambda \var[2]_1'...\var[2]_4'.(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)\Big)(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)\bigg](0,0,0,0,1)
+		}
+		}
+	\end{align*}
+
+	Now let's apply the first part of $OPT$:
+	\begin{align*}
+		&\letin{w_1}{\var_1 * \var_2}{\\
+		&\letin{w_2}{w_1 * \var_1}{\\
+		&\letin{(\var[2]_1\ldots\var[2]_5)}{(0,0,0,0,1)}{\\
+		& \letin{(\var[2]_1'\ldots\var[2]_4')}{(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}{\\
+		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2'+\var_1*\var[2]_4',\var[2]_3')
+		}}}}
+	\end{align*}
+Let's unsugar the lets:
+	\begin{align*}
+		&\letin{w_1}{\var_1 * \var_2}{\\
+		&\letin{w_2}{w_1 * \var_1}{\\
+		&\letin{\var[2]_1}{0}{\\
+		&\letin{\var[2]_2}{0}{\\
+		&\letin{\var[2]_3}{0}{\\
+		&\letin{\var[2]_4}{0}{\\
+		&\letin{\var[2]_5}{1}{\\
+		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
+		&\letin{ {\color{red} \var[2]_2'} }{  {\color{red} \var[2]_2} }{\\
+		&\letin{ {\color{red} \var[2]_3'} }{ {\color{red} \var[2]_3}  }{\\
+		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
+		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2'+\var_1*\var[2]_4',\var[2]_3')
+		}}}}}}}}}}}
+	\end{align*}
+	We apply the second part of $OPT$ to remove the variables in red:
+	\begin{align*}
+		&\letin{w_1}{\var_1 * \var_2}{\\
+		&\letin{w_2}{w_1 * \var_1}{\\
+		&\letin{\var[2]_1}{0}{\\
+		&\letin{\var[2]_2}{0}{\\
+		&\letin{\var[2]_3}{0}{\\
+		&\letin{\var[2]_4}{0}{\\
+		&\letin{\var[2]_5}{1}{\\
+		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
+		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
+		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2+\var_1*\var[2]_4',\var[2]_3)
+		}}}}}}}}}
+	\end{align*}
+	We apply some algebraic simplifications in $\partial_3$
+	\begin{align*}
+		&\letin{w_1}{\var_1 * \var_2}{\\
+		&\letin{w_2}{w_1 * \var_1}{\\
+		& \letin{\var[2]_1'}{w_1}{\\
+		&\letin{\var[2]_4'}{\var_1}{\\
+		&(\var[2]_1'+\var_2*\var[2]_4',\var_1*\var[2]_4',0)
+		}}}}
+	\end{align*}
+	We can optimise some more with $\partial_3$, and we finally obtain
+	\begin{align*}
+		&\letin{w_1}{\var_1 * \var_2}{\\
+		&\letin{w_2}{w_1 * \var_1}{\\
+		&(w_1+\var_2*\var_1,\var_1*\var_1,0)
+		}}
+	\end{align*}
+\end{example}

### 2021-02-16 rad-icfp-2021/relwork.tex
some very rough starting draft

@@
 
-\subsection{State-of-the-art} % (fold)
-\label{sub:state_of_the_art}
-
-All the existing work have either focused on efficiency or correctness.
-
-We present \system{}, a correct and efficient functional differentiable programming framework.
-This is the first work that propose a pure reverse-mode AD on a higher-order functional language with a correctness proof and complexity guarantees that is implemented and achieves good practical performance.
-In Figure\ref{fig:comparison-table} we compare our work to the recent papers in the field.
-
-\MH{Separate complexity into complexity and performent.}
-\MH{reorder topics: theory on the left, practice on the right, to make the dichotomy more apparent}
-\MH{most refs here are pure, which is not good for our emphasis on pureness, need to add more practical papers}
-\begin{table}
-\label{fig:comparison-table}
-\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|}
-\hline
- & Reverse  & Complexity & Impl. & Pure & Correct & HO Fun. & Tensors & Rec. & Cond. & HO $\partial$ ? \\
-\hline
-\system{} (This Paper) &
-\supfull & \supfull & \supfull & \supfull & \supfull & \supfull  & \supfull  & \supfull & \supfull & \supfull \\ 
-\hline
-Lantern~\cite{lantern_icfp} & 
-\supfull & \suphalf & \supfull & \supnone & \supnone & \supfull & \supfull & \supfull & \supfull &? \\ 
-\hline
-\dfsmooth{}~\cite{shaikhha2019efficient} 
-& 
-\supnone & \supnone & \supfull & \supfull & \supnone & \supfull & \supfull  & \supnone  & \supnone &? \\ 
-\hline
-\cite{huot2020correctness} &
-\supfull & \supnone & \supnone & \supfull & \supfull & \supfull & \supnone & \suphalf & \supfull &?\\ 
-\hline
-\cite{brunel2019backpropagation} &
-\supfull & \suphalf ? & \supnone & \supfull & \supfull & \supfull & \supnone & \supnone & \supnone &?\\ 
-\hline
-\cite{abadi2019simple} &
-\supfull & \supfull? & \supnone & \supfull & \supfull & \supnone & \supnone? & \supfull & \supfull &?\\ 
-\hline
-\cite{barthe2020versatility} &
-\supnone & \supnone & \supnone & \supfull & \supfull & \supnone? & \supnone & \supnone & \supfull &?\\ 
-\hline
-\cite{pearlmutter2008reverse} &
-\supfull & \supfull (not proved) & \supfull & \supnone & \supnone(not proved) & \supfull & \supnone? & \supfull & \supfull &?\\ 
-\hline
-\cite{Elliott:2018:SEA:3243631.3236765} &
-\supfull & \suphalf & \suphalf & \supfull & \supnone & \supnone & \supnone? & \supnone & \supnone? &?\\ 
-\hline
-\cite{sherman2020lambda_s} & \suphalf or \supnone? & \supnone & \supfull & \supfull & \supfull? & \supfull & \supnone? & \supnone & \supnone &?\\ 
-\hline
-\cite{vytiniotis2019differentiable} &
-\supfull & \suphalf & \suphalf & \supfull & \supnone & \supfull & \supnone? & \supnone & \supnone? &?\\ 
-\hline
-\cite{mak2020differential} & \supfull & \suphalf ? & \supnone & \supfull & \supfull & \supfull & \supnone & \supnone & \supnone &?\\ 
-\hline
-\cite{vakar2020reverse} & \supfull & \suphalf ? & \supnone & \supfull & \supfull & \supfull & \supnone & \supnone & \supnone &?\\ 
-\hline
-\cite{Manzyuk2012} & \supnone & \supnone & \supnone & \supfull & \supfull? & \supfull & \supnone & \supnone & \supnone &?\\ 
-\hline 
-\cite{gallagher-sdg} & \supnone & \supnone & \supnone & \supfull & \supfull ? & \supfull & \supnone & \supfull & \supfull & ?  \\ 
-\hline
-\end{tabular}
-\caption{Comparison of different differentiable programming frameworks.}
-A $\supfull$ means the property is verified, and $\supnone$ means its absent from the work.
-Reverse stands for reverse-mode differentiation, complexity to whether there is a proof of complexity and whether its competitive, impl.
-stands for running implementation, pure for whether differentiation is pure or effectful, correct to whether itis proved correct, HO Fun.
-stands for Higher-order functions, Tensors to efficient array like structures, Rec.
-for term-recurion, Cond.
-for conditionals, and finally HO $\partial$ for higher-order derivatives.
-\end{table}
-%Actually very much in the spirit of Swift and Julia Zygote (see Diff.Curry intro), probably refer to this somewhere
-
-\MH{we only have a limited form of recursion, maybe make that clearer.
-and we don't really have tensors, we have arrays.
-we don't have conditionals yet but that's something we can think of.
-we only have limited HO too for performance reason.
-maybe we should heavily emphasize we can add more but then it's harder to get guarantees and maybe not a great idea for speed...}
-
 \MH{understand related work sentence from Diff.

### 2021-02-16 rad-icfp-2021/simplecase.tex
some very rough starting draft

\section{Simple case}

\subsection{Language}

We consider a simple first-order language. 
TODO

\subsection{Naive reverse-mode}

\subsection{Optimisations}

\subsection{Correctness}

\subsection{Cheap gradient principle}

\subsubsection*{Key insight:}

As is well known (see e.g. \cite{pearlmutter2008reverse}), duplication of a variable $\Delta$ is turned into $+$ by reverse mode.
TODO: explain this is the hard part to make pure and efficient, and shows up in the case of non-unary operators, and things like let binding.
In the inefficient representation, every operator was seen as a unary operator using all the available variables and returning them in addition to the extra return value. 
In essence, we are putting boxes locally around places where the compute-flow might be non-linear and turning everything into something linear.
This linearisation of the compute-flow is the key to efficient purely functional reverse-mode. We extend this idea formally in the next sections by introducing a unary form, 
on which it is easier to do naive reverse-mode that can be optimized.

### 2021-02-17 rad-icfp-2021/relwork.tex
some slow progress

@@
 \label{sub:related_work}
+
+Relatively recent work in AD (2008-2020):
+\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,wang2018demystifying,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing}
+
+Some work in ML where some kind of iteration/recursion is used.
+Dynamic neural networks: \cite{jin2017manipulability,wu2016deep}
+Recursive NN: \cite{socher2011parsing,biancofiore2017recursive}
+Reccurent NN: \cite{bahdanau2014neural,luong2015effective}
+Tree LSTM: \cite{tai2015improved,chen2016enhanced}
+Dynamic Recursive NN: \cite{guo2019dynamic}. It's recent with very few citations, but their point is close to ours. 
+Top-down Tree LSTM: \cite{zhang2015top}
+Recursion in DNN: \cite{jeong2018improving}
 

### 2021-02-17 rad-icfp-2021/simplecase.tex
some slow progress

@@
 
-We consider a simple first-order language. 
-TODO
+We first consider a very simple first-order language. The types and terms are as follows:
+% \input{simpletypesandterms}
+
+The type system is given in Figure \ref{fig:types1}.
+
+% \begin{figure}[b]
+%     \framebox{\scalebox{0.82}{\begin{minipage}{1.2\linewidth}\noindent\input{simpletypesystemsource}\end{minipage}}}
+%     \caption{Typing rules for the simple language.\label{fig:types1}}
+% \end{figure}
+
+Reverse-mode introduces pairs and continuations, so we need to enrich the target language of differentiation with these constructs.
+The target language is given in Figure \ref{fig:types2}.
+
+% \begin{figure}[b]
+%     \framebox{\scalebox{0.82}{\begin{minipage}{1.2\linewidth}\noindent\input{simpletypesystemtarget}\end{minipage}}}
+%     \caption{Typing rules for the simple language.\label{fig:types2}}
+% \end{figure}
 
@@
 
+We define the macro $\Dsynrevsymbol[][]{}$ by induction on the syntax of the source language as follows:
+
+% \begin{figure}[b]
+%     \framebox{\scalebox{0.82}{\begin{minipage}{1.2\linewidth}\noindent\input{simplenaivediffmacro}\end{minipage}}}
+%     \caption{Typing rules for the simple language.\label{fig:simplediff}}
+% \end{figure}
+
+We show that our macro $\Dsynrev[]{t}$ is well-typed on the restricted language above.
+Given $x:\RR^n\vdash t:\RR$, we define the type $\TyD{t}\defeq\TyDinter{x:\RR^n}{t}$ where $\TyDinter{L}{t}$ is defined by induction on the structure of $t$ as follows:
+\begin{itemize}
+	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{\letin{y}{s}{t}}\defeq \TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR,y:\RR}{t}$
+	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{\cnst}\defeq \RR\times (\RR^{n+k+1}\to \RR^{n})$
+	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{x_i}\defeq\RR\times (\RR^{n+k+1}\to \RR^{n})$
+	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{y_i}\defeq\RR\times (\RR^{n+k+1}\to \RR^{n})$
+	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{\Op_1(y_i)}\defeq \RR\times (\RR^{n+k+1}\to \RR^{n})$ and the same with $\Op_1(x_i)$
+	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{\Op_2(y_i,y_j)}\defeq \RR\times (\RR^{n+k+1}\to \RR^{n})$
+\end{itemize}
+
+In other words, most items can be summarised as $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{W}\defeq \RR\times (\RR^{n+k+1}\to \RR^{n+k})$ where $W$ is an almost immediate expression.
+
+\begin{lemma}
+	The macro $\Dsynrev[]{t}$ is well-typed. More precisely, if $x:\RR^n\vdash t:\RR$, then $(x,x'):\RR^n\times (\RR^n\to\RR^n)\vdash \Dsynrev[x]{t}:\TyD{t}$.
+\end{lemma}
+
+\subsection{Semantics} 
+\label{sub:Semantics}
+
+We give a standard denotational semantics $\sem{-}$ to our language in the category of Euclidean spaces and smooth maps.
+
 \subsection{Optimisations}
+
+With the current reverse mode transformation, the produced term is much bigger than the original term, by about a quadratic factor. 
+That is, if the original term has roughly $n$ reductions, the produced term will have $O(n^2)$ reductions. 
+For efficient reverse mode implementations, this should be $O(n)$ and the constant (usually at most 4) matters as well.
+By applying some well chosen partial evaluation, we can recover an efficient term.
+
+The simple insight in each function in the continuation, at most $2$ out of the $m$ current variables are not simply returned, so the continuation as each step looks like
+$\lambda x1,\ldots,xm.X(x1,\ldots,xm)$ except for at most 2 positions.
+Unrolling X one step further, we see that when beta-reducing, most substitutions will be of the form $[x/y]$ where $x,y$ are both variables, 
+and this gives a big opportunity for partial evaluation.
+
+Slighly more formally, the basic partial evaluation consist of:
+\begin{enumerate}
+	\item Remplacing the variables binding the $\inllambda$s by the actual function they represent (CBN evaluation)
+	\item Replace all $(\inllambda x.e_1)e2$s by $\letin{x}{e_2}{e_1}$
+	\item Use the following reduction $\letin{\var}{\var[2]}{\trm}\to\trm{}[\var[2]/\var{}]$ when $\var,\var[2]$ are variables
+\end{enumerate}
+
+Note that we didn't use anything about the tuple $(0,0,0,0,1)$ used to compute the reverse-derivative of the whole term. 
+At this point we can optimise further with partial evaluation. We could also use another tuple to compute the partial derivative of a subterm, 
+i.e. with the tuple $(0,0,0,1,0)$, we would compute the derivative of the subterm ending where $w_1$ is introduced (and returned). 
+The standard imperative form for reverse-mode is not as flexible and one would need to write a separate algorithm 
+or use a few tricks to compute the derivative of a subterm.
+
+Each inlined lambda $\inllambda$ is used exactly once and we are in a pure typed setting, so termination is easy to prove. These transformations are semantics preserving.
+
+\begin{proposition}
+	Steps 1)-3) terminate and are semantics preserving.
+\end{proposition}
 
@@
 
+Now we can phrase that our sequence of transformations is correct.
+
+\begin{proposition}
+    TODO
+\end{proposition}
+
 \subsection{Cheap gradient principle}
+
+TODO: recall what it is and that it's what we want. introduce simple cost model.
+
+\begin{proposition}
+    If $\Gamma\vdash e:\RR$ then $C(rad(e))\leq k C(e)$ for a constant $k$ independant of $e$, and $k\leq4$.
+\end{proposition}
 

### 2021-02-17 rad-icfp-2021/relwork.tex
adding some refs

@@
 
-\noindent \textbf{Other Language features.}
-%more primitives, proved correct and more efficient. extend library
-
-\noindent \textbf{Probabilistic Programming.}
-%find something simple and precise, o.w. useless section
-
 \noindent \textbf{More dynamic language}
@@
 \noindent \textbf{Limitations.}
-%maybe?
+\begin{itemize}
+    \item memory access for cost function 
+    \item parallelism
+    \item dedicated tensors for ML, and BLaS considerations
+    \item implementation and benchmark
+    \item conditionals
+    \item recursion
+    \item general HO function
+    \item more 2nd order functions, though can be derived by hand once and then it's ok
+    \item dynamic arrays
+    \item more dynamic control flow
+    \item all store vs all recompute reverse-mode, and advanced checkpointing
+    \item numerical accuracy
+    \item fancier types and datatypes (eg dependent, adt) 
+    \item iterated ad and higher derivatives
+    \item use in Probabilistic programming
+    \item non smooth and partial functions, although by simply following the ref book it's as usual in AD
+    \item sparsity
+    \item other optimisations like implicit function theorem
+\end{itemize}
 
 \MH{useful links (Swift is close to our project, and the doc is very readable):}
-
 \begin{verbatim}https://docs.google.com/document/d/1_BirmTqdotglwNTOcYAW-ib6mx_jl-gH9Dbg4WmHZh0/edit# \end{verbatim}
 \begin{verbatim}
-
 	https://github.com/apple/swift/blob/master/docs/DifferentiableProgramming.md#approaches-to-automatic-differentiation

### 2021-02-17 rad-icfp-2021/macrodiff.tex
some comments
some more basic writing

@@
 \section{A macro for source-to-source reverse-mode differentiation}
+
+\subsubsection{Reverse AD for unary operators}
+
+We define reverse-mode by the following inductively defined macro $\Dsynrevsymbol[\rho]:\TSSyn\to \TTSynHO$, where $\rho$ is some type of our language, on both types and terms.
+
+\[
+\begin{array}{l}
+\Dsynrev[\rho]{A} \defeq A\times (A\To \rho)
+\end{array}    
+\]
+
+\[
+\begin{array}{l}
+	 \Dsynrev[\rho]{var_{i,\Gamma}}:= \var,Y\vdash \tPair{var_{i,\Gamma}}{\lambda \var[2]. Y(Jvar_{i,\Gamma}^T(\var[2]))}\\
+	 \Dsynrev[\rho]{\op_{n,\Gamma}}:= \var,Y\vdash \tPair{\op_{n,\Gamma}}{\lambda \var. Y(J\op_{n,\Gamma}^T(\var))}\\
+	 \Dsynrev[\rho]{\trm;\trm[2]}:= \var,Y\vdash \Dsynrev[\rho]{\trm[2]}(\Dsynrev[\rho]{\trm}(\var,Y))\\
+	 \Dsynrev[\rho]{swap_\sigma}:= \var,Y\vdash \tPair{swap_\sigma}{\lambda \var[2].Y(Jswap_\sigma^T(\var[2]))} \\
+	 \Dsynrev[\rho]{b:\BB\text{ or }\NN}:= \var,Y\vdash \tPair{b}{\lambda \var[2].Y(\underline{0})}\\
+	 \Dsynrev[\rho]{\widehat{\trm}}:= \var,Y\vdash \tMatch{\Dsynrev[\rho]{\trm}}{\var[2],Y'}{\tPair{\widehat{\trm}}{\lambda \var[2].\tMatch{\var[2]}{\var[2]_1,\var[2]_2}{Y'(\var[2]_1)}}}\\
+	 \Dsynrev[\rho]{\ifelse{\trm}{\trm_1}{\trm_2}}:=\var,Y\vdash\ifelse{\trm}{\Dsynrev[\rho]{\trm_1}}{\Dsynrev[\rho]{\trm_2}}\\
+\end{array}
+\]
+
+Observe that we use the transposed matrix of partial derivatives in the differentiation of an operation.
+Moreover, observe that we accumulate such transposed derivatives in reverse order, by making use of continuation passing style.

### 2021-02-21 rad-icfp-2021/background.tex
minor

@@
 These values appear in the chain rule so should be stored or recomputed.
-These things make reverse-mode harder to implement efficiently, and especially harder to analyse and optimise. 
-Traditional implementions use references.
+These things make reverse-mode harder to implement efficiently, and especially harder to analyze and optimize. 
+Traditional implementations use references.
 For instance, let 
@@
 As noted in \cite{pearlmutter2008reverse}, in a functional language, one can have a purely functional reverse-mode using continuations. 
-This representation is notoriously inefficient and all the implementions of reverse mode in a functional setting have followed the lines of the imperative version above \cite{pearlmutter2008reverse, wang2018demystifying, baydin2016diffsharp}.
+This representation is notoriously inefficient, and the existing implementations of reverse mode in a functional setting have followed the lines of the imperative version above \cite{pearlmutter2008reverse, wang2018demystifying, baydin2016diffsharp}.
 
-In the next section, we show on a very simple language how we can use standard optimisations to turn this inefficient representation into a very efficient one which is still purely functional.
-The remaining of the paper will generalise this idea and explore some consequences of this new approach.
+In the next section, we first introduce a very simple first-order language. 
+Then we define reverse-mode on this language as a program transformation. 
+This first transformation is quite inefficient and we how we can use standard optimizations on functional languages to obtain a provably efficient pure program.
+The key idea behind that will be generalized and explored in detail in the rest of the paper is that of turning every operator into a unary one.     
+The remaining of the paper will generalize this idea and explore some consequences of this new approach.

### 2021-02-22 rad-icfp-2021/complexity.tex
change outline

--- 2021-02-16 rad-icfp-2021/simplecase.tex
@@
-\section{Simple case}
+\section{Complexity}
 
-\subsection{Language}
-
-We consider a simple first-order language. 
-TODO
-
-\subsection{Naive reverse-mode}
-
-\subsection{Optimisations}
-
-\subsection{Correctness}
+\subsection{Cost model}
 
 \subsection{Cheap gradient principle}
-
-\subsubsection*{Key insight:}
-
-As is well known (see e.g. \cite{pearlmutter2008reverse}), duplication of a variable $\Delta$ is turned into $+$ by reverse mode.
-TODO: explain this is the hard part to make pure and efficient, and shows up in the case of non-unary operators, and things like let binding.
-In the inefficient representation, every operator was seen as a unary operator using all the available variables and returning them in addition to the extra return value. 
-In essence, we are putting boxes locally around places where the compute-flow might be non-linear and turning everything into something linear.
-This linearisation of the compute-flow is the key to efficient purely functional reverse-mode. We extend this idea formally in the next sections by introducing a unary form, 
-on which it is easier to do naive reverse-mode that can be optimized.

### 2021-02-22 rad-icfp-2021/efficientdiff.tex
change outline

\section{Efficient reverse-mode}

\subsection{Target Language}

\subsection{Efficient representation}

\subsection{Optimisations via partial evaluation}

### 2021-02-22 rad-icfp-2021/generalisation.tex
change outline

\section{Generalisations}

\subsection{Arrays}

\subsection{Conditionals}

\subsection{Partial operations}

\subsubsection{Higher-order functions}

### 2021-02-22 rad-icfp-2021/intro.tex
change outline

--- 2021-02-14 rad-icfp-2021/intro.tex
@@
 
-\subsection{Problem} % (fold)
-\label{sub:problem}
+% % \subsection{Problem} % (fold)
+% % \label{sub:problem}
 
-\TODO{find simple yet telling concrete example here}
-Consider the following 
-TODO: Jacobian, chain rule in maths
-It is easy to write a purely functional implementation of it
-TODO: code using FP
-However, if we run it naively it will take 
-TODO: time or complexity
-A good implementation would instead take
-TODO: time or complexity
-and looks like
-TODO: imperative code
+% % \TODO{find simple yet telling concrete example here}
+% % Consider the following 
+% % TODO: Jacobian, chain rule in maths
+% % It is easy to write a purely functional implementation of it
+% % TODO: code using FP
+% % However, if we run it naively it will take 
+% % TODO: time or complexity
+% % A good implementation would instead take
+% % TODO: time or complexity
+% % and looks like
+% % TODO: imperative code
 
-There is a huge difference between the purely functional yet inefficient implementation and the efficient imperative one.
+% % There is a huge difference between the purely functional yet inefficient implementation and the efficient imperative one.
 The first one is also easy to prove correct, and purely functional code can benefit from a lot of optimisations \cite{}.
@@
 It is often believed \cite{} that a purely functional implementation of reverse-mode differentiation will be unefficient and that there is a need for imperative features such as references or the use of intermediate representations \cite{} to reach good performance.
-\TODO{Add here we want to investigate this for rich HO languages, which offer more possibilities for optimisations and cleaner to write.
+% % \TODO{Add here we want to investigate this for rich HO languages, which offer more possibilities for optimisations and cleaner to write.
 detail more in problem section that this unified way would be good for ||, opti, correctness, etc.}
 
-\subsection{Contribution} % (fold)
-\label{sub:contribution}
+% % \subsection{Contribution} % (fold)
+% % \label{sub:contribution}
 
-\MH{Be clearer and more precise here as to why our contribution solves the problem above.}
+% % \MH{Be clearer and more precise here as to why our contribution solves the problem above.}
 
-This work aims at bridging the gap between proved-correct high-level idealized purely functional differentiation and efficient low-level imperative implementations.
+% % This work aims at bridging the gap between proved-correct high-level idealized purely functional differentiation and efficient low-level imperative implementations.
 We do this by providing a correct, efficient, and purely functional implementation.
@@
 We apply some standard optimisation techniques on the code.
-We finally explore some theoretical guarantees of the resulting code (Sec. \ref{}) and give some experimental results (Sec. \ref{}).
+% % We finally explore some theoretical guarantees of the resulting code (Sec. \ref{}) and give some experimental results (Sec. \ref{}).
 
-% \subsubsection{Brief overview} % (fold)
-% \label{sub:brief_overview}
-%of the problem: code example
+% % % \subsubsection{Brief overview} % (fold)
+% % % \label{sub:brief_overview}
+% % %of the problem: code example
 
-% % diff prog, what state of the art is, what we propose, roughly our contribution
-% Under the slogan of differentiable programming, Automatic Differentiation has seen a renewal in the machine learning community over the past few years.
+% % % % diff prog, what state of the art is, what we propose, roughly our contribution
+% % % Under the slogan of differentiable programming, Automatic Differentiation has seen a renewal in the machine learning community over the past few years.
 The need for fast gradient computations of code implementing functions is ubiquitous and core in deep learning.
@@
 Doing so, we reach state-of-the-art performance and ultimately produce code that is similar to the manually-optimised imperative code that one could get.
-% %maybe add that it's principled, explains more things, is correct, and great for generalisation to other languages/libraries/etc.
+% % % %maybe add that it's principled, explains more things, is correct, and great for generalisation to other languages/libraries/etc.
 
-\subsection{Our approach} % (fold)
-\label{sub:our_approach}
+% % \subsection{Our approach} % (fold)
+% % \label{sub:our_approach}
 
-\MH{start from pure HO functional language with smooth primitives, use an optimised (principled) macro, then further classical optimisations.
+% % \MH{start from pure HO functional language with smooth primitives, use an optimised (principled) macro, then further classical optimisations.
 forward def to things in more detail.
@@
 
-%rename Context?
+% % %rename Context?
 
-\subsection{Differentiable programming} % (fold)
-\label{sub:differentiable_programming}
+% % \subsection{Differentiable programming} % (fold)
+% % \label{sub:differentiable_programming}
 
-%as a "solution" people found? maybe skip this entirely
+% % %as a "solution" people found? maybe skip this entirely
 
-%define AD
-%mostly from Matthijs' POPL paper
-(Source-to-source) Automatic Differentiation (AD) is a technique for transforming code that implements 
-a function $f$ into code that computes $f$'s derivative, essentially by using the chain rule for derivatives. 
-Due to its efficiency and numerical stability, AD is the technique of choice whenever derivatives need to be computed 
-of functions that are implemented as programs, particularly in high dimentional settings. 
-Optimisations and Monte Carlo integration algorithms, such as Gradient Descent and Hamiltonian Monte-Carlo methods, 
-rely crucially on the calculation of derivatives. These algorithms are used in virtually everywhere in machine learning 
-and computational statistics, and the calculation of derivatives is usually the computational bottle-neck. 
-AD, roughly speaking, comes in two modes; forward-mode and reverse-mode. 
-When differentiating a function $\RR^n\to\RR^m$, forward-mode tends to be more efficient if $m \gg n$, while reverse-mode 
-is generally more performant if $n \gg m$. As most applications reduce to optimization of Monte-Carlo 
-integration of an objective function $\RR^n\to\RR$ with $n$ very large ($10^4-10^7$), reverse-mode AD is 
-in many ways the more interesting algorithm.
+% % %define AD
+% % %mostly from Matthijs' POPL paper
+% % (Source-to-source) Automatic Differentiation (AD) is a technique for transforming code that implements 
+% % a function $f$ into code that computes $f$'s derivative, essentially by using the chain rule for derivatives. 
+% % Due to its efficiency and numerical stability, AD is the technique of choice whenever derivatives need to be computed 
+% % of functions that are implemented as programs, particularly in high dimentional settings. 
+% % Optimisations and Monte Carlo integration algorithms, such as Gradient Descent and Hamiltonian Monte-Carlo methods, 
+% % rely crucially on the calculation of derivatives. These algorithms are used in virtually everywhere in machine learning 
+% % and computational statistics, and the calculation of derivatives is usually the computational bottle-neck. 
+% % AD, roughly speaking, comes in two modes; forward-mode and reverse-mode. 
+% % When differentiating a function $\RR^n\to\RR^m$, forward-mode tends to be more efficient if $m \gg n$, while reverse-mode 
+% % is generally more performant if $n \gg m$. As most applications reduce to optimization of Monte-Carlo 
+% % integration of an objective function $\RR^n\to\RR$ with $n$ very large ($10^4-10^7$), reverse-mode AD is 
+% % in many ways the more interesting algorithm.
 
-%define diff prog beyond AD, be specific with references/different define-then-run vs Pytorch/Swift etc.
-%Diff Curry has OK intro
+% % %define diff prog beyond AD, be specific with references/different define-then-run vs Pytorch/Swift etc.
+% % %Diff Curry has OK intro
+
+% \[
+% \begin{tikzcd}
+% 	& \TSSyn \ar[r,"{\Dsynrev[\rho]{-}}"] 
+% 	& \TTSynHO_1 \ar[d,"\UNF{-}^{-1}"] &  \\
+% 	& \SSSyn \ar[u,"\UNF{-}"]  \ar[r,"{\Dsynrev[\rho]{-}^*}"'] 
+% 	& \TTSynHO_2 \ar[d,"Jacob"] \\
+% 	& \SSSyn \ar[u,"ANF"] \ar[r,"{\Dsynrev[\rho]{-}^{**}}"'] 
+% 	&  \TTSynHO_3 \ar[dl,"\partial_2"] \\
+% 	\SSSynHO \ar[ur,"\partial_1"]  \ar[rrr,bend right=20,color=red,"Example\,1"]
+% 	& \TTSyn \ar[r,"OPT"'] 
+% 	& \TTSyn \ar[r,"\partial_3"']
+% 	& \TTSyn 
+% \end{tikzcd}
+% \]
+
+% \begin{itemize}
+% 	\item $\SSSynHO$ is a higher-order language and the language the user uses to write a prgram.
+% 	\item $\partial_1$ denotes partial evaluation of all higher-order variables. We end up with a program whose only higher-order bits are primitive operators.
+% 	\item $\SSSyn$ is a first-order language with a few higher-order primitives like map,fold,etc.
+% 	\item $ANF$ is for (weak) A-normal form. Every operator $\op$ is only given variables as arguments. We do this by introducting additional let bindings. It is only used for the $Jacob$ optimisation.
+% 	\item $\UNF{-}$ is the unary form transformation presented in the next section. Its goal is to deal with the fanout problem before applying the reverse-mode macro.
+% 	\item $\TSSyn$ is the target language for $\UNF{-}$ and also the source for the reverse-mode macro. It has a bit of a special structure.
+% 	\item $\Dsynrev[\rho]{-}$ is a reverse-mode macro. It is efficient on unary operators but does not deal well with fanout. Fortunately, it is only given unary operators at this point.
+% 	\item $\TTSynHO_1$ is the target language of the reverse-mode macro. It is a higher-order language as it needs a continuation for the reverse derivatives.
+% 	\item $\TTSynHO_2,\TTSynHO_3$ are small variations of $\TTSynHO_1$.
+% 	\item $\UNF{-}^{-1}$ is an optimisation which removes the overhead caused by $\UNF{-}$. It reverts the primal part of the derivative to the original term.
+% 	\item $\Dsynrev[\rho]{-}^*$ is an efficient macro for reverse-mode which deals with the fanout problem.
+% 	\item $Jacob$ stands for an efficient representation of the Jacobian.
+It needs a weak form of ANF to compute the Jacobian with a single vector and does not need to use a matrix-vector multiplication.
+In particular, it does not need to store the matrix.
+% 	\item $\Dsynrev[\rho]{-}^{**}$ is the efficient macro for reverse mode that would really be implemented. It will be quicker to run than the composition of the transformations. 
+% 	\item $\partial_2$ stands for partial evaluation. We evaluate (CBN) the higher-order variables binding the lambdas we introduced and inlined with the reverse-mode macro
+% 	\item $\TTSyn$ is the target language of the previous partial evaluation and the final language we use.
+% 	\item $OPT$ is an optimisation. First, we remplace the previously inlined lambdas by let bindings, and then apply forward substitution.
+% 	\item $\partial_3$  is optionnal. We partially evaluate more and do algebraic simplifications, like $0+\trm\to\trm, 1*\trm\to \trm$.
+% 	\item $Example\,1$ is detailed below.
+% \end{itemize}
+
+% \MH{I could add a different name for the target of ANF, of Jacob, and of OPT.}
+
+% \begin{example}
+% 	Let 
+% \begin{align*}
+% 	\var_1,\var_2,\var_3:\RR\vdash \trm:=\big(\lambda f:\RR\to\RR. (f(\var_1))*\var_1\big)(\lambda \var[2]. \var[2]* \var_2)
+% \end{align*}
+% After $\partial_1$ we obtain 
+% \begin{align*}
+% 	(\letin{\var[2]}{\var_1}{\var[2]*\var_2})*\var_1
+% \end{align*}
+% After $ANF$ we obtain
+% \begin{align*}
+% 	&\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
+% \end{align*}
+% After $\Dsynrev[\rho]{-}^*$ we obtain
+% 	\begin{align*}
+% 		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Y\big(J^1(\var_1,\var_2,\var_3)^T(\var[2]_1,...,\var[2]_4)\big)} }{
+% 		\\
+% 		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'\big(J^2(\var_1,\var_2,\var_3,w_1)^T(\var[2]_1,...,\var[2]_5)\big)}}{\\ 
+% 		& w_2'}}
+% 	\end{align*}
+% where $Y$ is a continuation variable of type $\RR^3\to\RR^3$, $J^1,J^2$ are Jacobian matrices for $*$ operators which also return their context, $\inllambda$ is just a normal $\lambda$ but inlined for future optimisation.
+
+% After applying $Jacob$, we obtain
+% 	\begin{align*}
+% 		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Y(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
+% 		\\
+% 		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
+% 		& w_2'}}
+% 	\end{align*}
+% To actually compute the reverse-derivative of the full term, we substitute the identity to $Y$ and apply the whole term to $(0,0,0,0,1)$, the functional equivalent to initialing the sensitivities in the imperative reverse-mode algorithm.
+We obtain
+
+% \begin{align*}
+% 		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
+% 		\\
+% 		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
+% 		& w_2'(0,0,0,0,1)}}
+% 	\end{align*}
+
+% After $\partial_2$, we get
+% 	\begin{align*}
+% 		&\letin{w_1}{\var_1 * \var_2}{\\
+% 		&\letin{w_2}{w_1 * \var_1}{\\
+% 		&\bigg[\inllambda \var[2]_1...\var[2]_5.\Big(\inllambda \var[2]_1'...\var[2]_4'.(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)\Big)(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)\bigg](0,0,0,0,1)
+% 		}
+% 		}
+% 	\end{align*}
+
+% 	Now let's apply the first part of $OPT$:
+% 	\begin{align*}
+% 		&\letin{w_1}{\var_1 * \var_2}{\\
+% 		&\letin{w_2}{w_1 * \var_1}{\\
+% 		&\letin{(\var[2]_1\ldots\var[2]_5)}{(0,0,0,0,1)}{\\
+% 		& \letin{(\var[2]_1'\ldots\var[2]_4')}{(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}{\\
+% 		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2'+\var_1*\var[2]_4',\var[2]_3')
+% 		}}}}
+% 	\end{align*}
+% Let's unsugar the lets:
+% 	\begin{align*}
+% 		&\letin{w_1}{\var_1 * \var_2}{\\
+% 		&\letin{w_2}{w_1 * \var_1}{\\
+% 		&\letin{\var[2]_1}{0}{\\
+% 		&\letin{\var[2]_2}{0}{\\
+% 		&\letin{\var[2]_3}{0}{\\
+% 		&\letin{\var[2]_4}{0}{\\
+% 		&\letin{\var[2]_5}{1}{\\
+% 		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
+% 		&\letin{ {\color{red} \var[2]_2'} }{  {\color{red} \var[2]_2} }{\\
+% 		&\letin{ {\color{red} \var[2]_3'} }{ {\color{red} \var[2]_3}  }{\\
+% 		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
+% 		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2'+\var_1*\var[2]_4',\var[2]_3')
+% 		}}}}}}}}}}}
+% 	\end{align*}
+% 	We apply the second part of $OPT$ to remove the variables in red:
+% 	\begin{align*}
+% 		&\letin{w_1}{\var_1 * \var_2}{\\
+% 		&\letin{w_2}{w_1 * \var_1}{\\
+% 		&\letin{\var[2]_1}{0}{\\
+% 		&\letin{\var[2]_2}{0}{\\
+% 		&\letin{\var[2]_3}{0}{\\
+% 		&\letin{\var[2]_4}{0}{\\
+% 		&\letin{\var[2]_5}{1}{\\
+% 		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
+% 		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
+% 		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2+\var_1*\var[2]_4',\var[2]_3)
+% 		}}}}}}}}}
+% 	\end{align*}
+% 	We apply some algebraic simplifications in $\partial_3$
+% 	\begin{align*}
+% 		&\letin{w_1}{\var_1 * \var_2}{\\
+% 		&\letin{w_2}{w_1 * \var_1}{\\
+% 		& \letin{\var[2]_1'}{w_1}{\\
+% 		&\letin{\var[2]_4'}{\var_1}{\\
+% 		&(\var[2]_1'+\var_2*\var[2]_4',\var_1*\var[2]_4',0)
+% 		}}}}
+% 	\end{align*}
+% 	We can optimise some more with $\partial_3$, and we finally obtain
+% 	\begin{align*}
+% 		&\letin{w_1}{\var_1 * \var_2}{\\
+% 		&\letin{w_2}{w_1 * \var_1}{\\
+% 		&(w_1+\var_2*\var_1,\var_1*\var_1,0)
+% 		}}
+% 	\end{align*}
+% \end{example}
+
+\subsection{Motivation}
+
+\subsection{Problem}
+
+\subsection{Examples}
+
+\subsection{Contribution}
+
+\subsection{Outline of the paper}

### 2021-02-22 rad-icfp-2021/simplediff.tex
change outline

\section{Simple reverse-mode differentiation}

\subsection{Source Language}

\subsection{Source Unary Form}

\subsection{Target Unary Form}

\subsection{Macro for reverse mode}

### 2021-02-22 rad-icfp-2021/figures/unf.tex
adding some structure to the document

\section{Unary Form}

Following the previous insight, we change the interpretation of terms.

Note that every ground type $G$ inherits a pointwise monoid structure from $\reals$ and has zero element $\overline{0}$.
We need a few  more notations.

\begin{notation}
	\begin{itemize}
		\item If $\Gamma\vdash \trm:A$ and $\Delta$ is another context, then $\Gamma,\Delta\vdash \widetilde{\trm}:A$ is the weakening of $\trm$.
		\item If $\Gamma\vdash \trm:A$, $\Delta$ another context and $\Delta'$ is a \underline{ground} context, then  $\Gamma,\Delta\vdash \widehat{\trm}:A\times \Delta'$ where $\widehat{\trm}:= \tPair{\widetilde{\trm}}{\overline{0}}$
	\end{itemize}
\end{notation}

The new idea simply consists in changing each operation so that it takes a single argument (a vector). To do that, each operation returns its arguments in addition to its return value.

We see $\UNF{-}$ as a transformation (not a functor) from $\SSSyn$ to $\TSSyn$, which we describe now. We also denote by $\TSSyn$ the syntactic category of the language. It is a first-order language with $n$-ary products, sums, composition $;$, and the following primitives. For each operator $\op_n:\RR^n\to \RR$ and each type $\Gamma$, we have a primitive operator $\op_{n,\Gamma}:\Gamma\times \RR^n\to \Gamma\times \RR^n\times \RR$. For each type $\Gamma=A_1\times...\times A_n$, we also have a primitive $var_{i,\Gamma}:\Gamma\to \Gamma\times A_i$.

\[
\begin{array}{cc}
	\inferrule{~}{\Gamma=A_1\times\ldots \times A_n\vdash var_{i,\Gamma}: \Gamma\times A_i }(A_i\in G(\reals))
	& \inferrule{~}{\Gamma\times \RR^n \vdash \op_{n,\Gamma}: \Gamma\times \RR^n\times \RR }\\
	\inferrule{A \vdash \trm: B \quad B \vdash \trm[2]: C}{A\vdash \trm;\trm[2]:C} 
	& \inferrule{\Gamma,b:\BB\vdash\trm:\BB\quad \Gamma \vdash \trm_1,\trm_2:A }{\Gamma,b:\BB\vdash \ifelse{\trm}{\trm_1}{\trm_2}:A}\\
	\inferrule{\Gamma\vdash \trm:A \quad \Delta\in G(\reals)}{\Gamma\vdash \widehat{\trm}:A\times \Delta }
	& \inferrule{\Gamma\vdash \trm: A_1\times\ldots A_k}{\Gamma\vdash swap_\sigma(\trm):A_{\sigma(1)}\times\ldots A_{\sigma(k)}}(\sigma\text{ permutation})\\
	\inferrule{~}{\Gamma,b:\BB,\Gamma'\vdash b:\BB}
	&\inferrule{~}{\Gamma,n:\NN,\Gamma'\vdash b:\NN}
\end{array}
\]
The idea is that we are linear in the real-based types and classical on boolean and integer types. This comes from the fact that the derivative of the latter being trivial, we don't need to properly keep track of them to avoid fanout.

Let $G(\reals)$ be a ground type based on reals (not booleans nor integers). It is defined by the following grammar:
\[G(\reals):=\reals \,\gor\, G(\reals)\times\ldots \times G(\reals)\,\gor\,\Array{G(\reals)}\]

We define the transformation $\UNFSymbol$ inductively as follows:
\begin{itemize}
	\item $\UNF{\Gamma\vdash \var_i: G(\reals)}:=var_{i,\Gamma}$
	\item $\UNF{\Gamma\vdash \var_i: \BB\text{ or }\NN}:=\var_i:\BB\text{ or }\NN$
	\item $\UNF{\op_n(\trm_1,\ldots,\trm_n)}:=\UNF{\tTuple{\trm_1,\ldots,\trm_n}}:\Gamma'\times \RR^n;\op_{n,\Gamma'}$
	\item $\UNF{\letin{\var}{\trm}{\trm[2]}}:=\UNF{\trm};\UNF{\widetilde{\trm[2]}}$
	\item $\UNF{\tTuple{\trm_1,\ldots,\trm_n}}:=\UNF{\trm_1};\ldots;\UNF{\widetilde{\trm_n}};swap_\sigma$ where $\UNF{\widetilde{\trm_i}}:\Gamma\times \Delta_1\times A_1 \times \Delta_2 \times A_2\times\ldots \times \Delta_i \times A_i$, $\trm_i:A_i$ and $swap_\sigma:\Gamma\times\Delta_1 \times \ldots \times A_n: \Gamma\times \Delta_1\times\ldots\times\Delta_n\times A_1\times\ldots\times A_n$
	\item $\UNF{\tMatch{\trm}{\var_1,\ldots,\var_n}{\trm[2]}}:=\UNF{\trm};\UNF{\widetilde{\trm[2]}}$
	%\item $\UNF{\ifelse{\trm}{\trm[2]_1}{\trm[2]_2}}:=\ifelse{\UNF{\trm}}{\widehat{\UNF{\widetilde{\trm[2]_1}}};swap_\sigma}{\widehat{\UNF{\widetilde{\trm[2]_2}}};swap_\mu}$ where $swap_\sigma:\Gamma\times \Delta_1\times A \times \Delta_2\to \Gamma\times \Delta_1\times \Delta_2 \times A$ and $swap_\mu:\Gamma\times \Delta_2\times A \times \Delta_1\to \Gamma\times \Delta_1\times \Delta_2 \times A$
\end{itemize}

We denote by $\semu{-}:\TSSyn\to \SSSyn$ the identity-on-objects functor given as follows:
\begin{itemize}
	\item  $\semu{var_{i,\Gamma}}:=\var:\Gamma\vdash \tMatch{\var}{\var_1,\ldots,\var_n}{\tTuple{x_1,\ldots,x_n,x_i}}$
	\item $\semu{\op_{n,\Gamma}}:=\var:\Gamma\times\RR^n\vdash \tMatch{\var}{\var_1,\var_2}{\tTuple{\var_1,\var_2,\op_n(\var_2)}}$
	\item $\semu{\trm;\trm[2]}:=\letin{\var}{\semu{\trm}}{\semu{\trm[2]}}$
	\item $\semu{swap_\sigma}:=\var:A_1\times\ldots\times A_n\vdash\tMatch{\var}{\var_1,\ldots\var_n}{\tTuple{\var_{\sigma(1)},\ldots,\var_{\sigma(n)}}}$
	\item $\semu{b:\BB\text{ or }\NN}:=\var:\Gamma\times (\BB\text{ or }\NN)\vdash\tMatch{\var}{\var[2],b}{b:\BB\text{ or }\NN}$
	\item $\semu{\widehat{\trm}}:=\letin{\var}{\semu{\trm}}{\tPair{x}{\underline{0}}}$
	%\item $\semu{\ifelse{\trm}{\trm_1}{\trm_2}}=\ifelse{\semu{\trm}}{\semu{\trm_1}}{\semu{\trm_2}}$
\end{itemize}

\begin{proposition}
	Let $\Gamma\vdash \trm:\ty$.\\ 
	Then $\Gamma\vdash \UNF{\trm}:\Gamma\times \Delta\times \ty$, where $\Delta$ is the list of bound variables of $\trm$.
	In addition, $\pi_{\Gamma\times \Delta}(\sem{\trm})=id$ and $\pi_{\ty}(\sem{\trm})=\sem{\trm}$. 
\end{proposition}

\begin{proposition}
	$\pi_{last}(\UNF(\trm))=\trm$ where the equality is modulo $\alpha,\beta,\eta$ and some classical algebraic rules.
\end{proposition}

The composition of the 3 transformations from before can be written more compactly and more efficiently into 1 step, by adding a bit of extra information to the reverse-mode transformation.

Semantically, we have the following diagram.\\
\begin{tikzcd}
	& \Dsynrev[\rho]{\reals^n} 
	\ar[rr,"\overleftarrow{\mathcal{D}}_{\rho}(\UNF{\trm})"] 
	\ar[rr, phantom, bend right=10, " "{name=D}, near end]
	\ar[rr, phantom, bend right=10, " "{name=F}, near start]  
	\ar[dd, "\pi", near start] \ar[dl, equal]
	& & \Dsynrev[\rho]{\reals^m\times\reals} 
	\ar[dd,"\pi"] 
	\\
    \Dsynrev[\rho]{\reals^n} 
    \ar[rr, crossing over,"\overleftarrow{\mathcal{D}}_{\rho}^*(\UNF{\trm})"', near end] \ar[dd, "\pi"']
    \ar[rr, phantom, bend left=10, ""{name=E}, near end]
    \ar[rr, phantom, bend right=20, ""{name=G}, near start]  
    & & \Dsynrev[\rho]{\reals}  & \\
    & \reals^n \ar[dl,equal] 
    \ar[rr,"\UNF{\trm}", near start] 
    \ar[rr,bend right=10, phantom, near start," "'{name=A}]
    \ar[rr, phantom, bend left=10, " "{name=C}, near end] 
    & & \reals^m\times\reals \ar[dl, "\pi_2"] \\
    \reals^n 
    \arrow[rr,"\trm"'] 
    \ar[rr, phantom, bend left=10," "{name=B}]
    \ar[rr, phantom, bend left=10," "{name=H}, near start] 
    & & \reals  
    \arrow[from=uu, near start, crossing over, "\pi"] 
    & \arrow[Rightarrow, color=blue, from=B, to=A, "\UNF{-}"']
    \arrow[Rightarrow, color=blue, from=C, to=D, "\overleftarrow{\mathcal{D}}"', bend right=20]
     \arrow[Rightarrow, near start, color=blue, from=F, to=E, "\UNF{-}^{-1}"]
      \arrow[Rightarrow, color=red, from=G, to=H, "\overleftarrow{\mathcal{D}}^*"', bend right=10, crossing over]
    \ar[from=1-4, to=2-3, "\pi_{\reals\times (\reals^m\times \reals\to \rho)}", crossing over]
\end{tikzcd}\\
Of course, the point is that in practice we can run it faster, while keeping correctness.

This macro can also directly perform some of the later optimisations, like the better expression for the Jacobian.

### 2021-02-23 rad-icfp-2021/figures/compilation_scheme.tex
adding some figures

\[
\begin{tikzcd}
    \Source \ar[r] & \UNFSymbol_1 \ar[r,"\Dsynrevsymbol"] & \UNFSymbol_2 \ar[r] & \Target \ar[r,"\partial ev"] & \Target
\end{tikzcd}
\]

### 2021-02-23 rad-icfp-2021/figures/macro_diff.tex
adding more figures

\begin{figure*}[t]
    \begin{tabular}{r c l}
    $\Dsynrevsymbol$(!var!$_{\Gamma,i}$) &=& !<var!$_{\Gamma,i}$!, J!$^T$!var!$_{\Gamma,i}$!>! \\
    $\Dsynrevsymbol$(!op!$_{\Gamma,n}$) &=& !<op!$_{\Gamma,n}$!, J!$^T$!op!$_{\Gamma,n}$!>! \\ 
    $\Dsynrevsymbol$(!e1;e2!) &=& $\Dsynrevsymbol$(!e1!);$\Dsynrevsymbol$(!e2!)\\ 
    $\Dsynrevsymbol$(!map!$_{x.e}$) &=& !<map!$_{x.e}$!, J!$^T$!map!$_{x.e}$!>! \\ 
    $\Dsynrevsymbol$(!map2!$_{x,y.e}$) &=& !<map2!$_{x,y.e}$!, J!$^T$!map2!$_{x,y.e}$!>! \\ 
    $\Dsynrevsymbol$(!foldl!$_{x,y.e}$) &=& !<foldl!$_{x,y.e}$!, J!$^T$!foldl!$_{x,y.e}$!>! \\  
    \end{tabular}
    \caption{Reverse-mode differentiation from Source UNF to Target UNF}
    \label{fig:diff_macro}    \end{figure*}

### 2021-02-23 rad-icfp-2021/simplecase.tex
minor

@@
 \section{Simple case}
-
-\subsection{Language}
-
-We first consider a very simple first-order language. The types and terms are as follows:
-% \input{simpletypesandterms}
-
-The type system is given in Figure \ref{fig:types1}.
-
-% \begin{figure}[b]
-%     \framebox{\scalebox{0.82}{\begin{minipage}{1.2\linewidth}\noindent\input{simpletypesystemsource}\end{minipage}}}
-%     \caption{Typing rules for the simple language.\label{fig:types1}}
-% \end{figure}
-
-Reverse-mode introduces pairs and continuations, so we need to enrich the target language of differentiation with these constructs.
-The target language is given in Figure \ref{fig:types2}.
-
-% \begin{figure}[b]
-%     \framebox{\scalebox{0.82}{\begin{minipage}{1.2\linewidth}\noindent\input{simpletypesystemtarget}\end{minipage}}}
-%     \caption{Typing rules for the simple language.\label{fig:types2}}
-% \end{figure}
-
-\subsection{Naive reverse-mode}
-
-We define the macro $\Dsynrevsymbol[][]{}$ by induction on the syntax of the source language as follows:
-
-% \begin{figure}[b]
-%     \framebox{\scalebox{0.82}{\begin{minipage}{1.2\linewidth}\noindent\input{simplenaivediffmacro}\end{minipage}}}
-%     \caption{Typing rules for the simple language.\label{fig:simplediff}}
-% \end{figure}
-
-We show that our macro $\Dsynrev[]{t}$ is well-typed on the restricted language above.
-Given $x:\RR^n\vdash t:\RR$, we define the type $\TyD{t}\defeq\TyDinter{x:\RR^n}{t}$ where $\TyDinter{L}{t}$ is defined by induction on the structure of $t$ as follows:
-\begin{itemize}
-	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{\letin{y}{s}{t}}\defeq \TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR,y:\RR}{t}$
-	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{\cnst}\defeq \RR\times (\RR^{n+k+1}\to \RR^{n})$
-	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{x_i}\defeq\RR\times (\RR^{n+k+1}\to \RR^{n})$
-	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{y_i}\defeq\RR\times (\RR^{n+k+1}\to \RR^{n})$
-	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{\Op_1(y_i)}\defeq \RR\times (\RR^{n+k+1}\to \RR^{n})$ and the same with $\Op_1(x_i)$
-	\item $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{\Op_2(y_i,y_j)}\defeq \RR\times (\RR^{n+k+1}\to \RR^{n})$
-\end{itemize}
-
-In other words, most items can be summarised as $\TyDinter{x:\RR^n,y_1:\RR,\ldots,y_k:\RR}{W}\defeq \RR\times (\RR^{n+k+1}\to \RR^{n+k})$ where $W$ is an almost immediate expression.
-
-\begin{lemma}
-	The macro $\Dsynrev[]{t}$ is well-typed. More precisely, if $x:\RR^n\vdash t:\RR$, then $(x,x'):\RR^n\times (\RR^n\to\RR^n)\vdash \Dsynrev[x]{t}:\TyD{t}$.
-\end{lemma}
 

### 2021-02-23 rad-icfp-2021/intro.tex
adding some structure to the document
some items in intro

@@
 
-% % \subsection{Problem} % (fold)
-% % \label{sub:problem}
-
-% % \TODO{find simple yet telling concrete example here}
-% % Consider the following 
-% % TODO: Jacobian, chain rule in maths
-% % It is easy to write a purely functional implementation of it
-% % TODO: code using FP
-% % However, if we run it naively it will take 
-% % TODO: time or complexity
-% % A good implementation would instead take
-% % TODO: time or complexity
-% % and looks like
-% % TODO: imperative code
-
-% % There is a huge difference between the purely functional yet inefficient implementation and the efficient imperative one.
-The first one is also easy to prove correct, and purely functional code can benefit from a lot of optimisations \cite{}.
-The imperative version is essentially a manually optimised representation, yet it is error prone.
-In addition, the imperative code will usually be transformed to SSA-form \cite{} for further optimisations by the compiler.
-It is often believed \cite{} that a purely functional implementation of reverse-mode differentiation will be unefficient and that there is a need for imperative features such as references or the use of intermediate representations \cite{} to reach good performance.
-% % \TODO{Add here we want to investigate this for rich HO languages, which offer more possibilities for optimisations and cleaner to write.
-detail more in problem section that this unified way would be good for ||, opti, correctness, etc.}
-
-% % \subsection{Contribution} % (fold)
-% % \label{sub:contribution}
-
-% % \MH{Be clearer and more precise here as to why our contribution solves the problem above.}
-
-% % This work aims at bridging the gap between proved-correct high-level idealized purely functional differentiation and efficient low-level imperative implementations.
-We do this by providing a correct, efficient, and purely functional implementation.
-This allows for more direct optimisations and paves the way for future work.
-In short, after detailing more the problem (Sec.
-\ref{}) and our approach (Sec.
-\ref{}) we provide a source and target language for differentiation (Sec.
-\ref{}) which we see as a source-to-source transformation \cite{}.
-We then propose a macro for differentiation (Sec.
-\ref{}).
-To prove our transformation correct, we give our language a denotational semantics using diffeological spaces (Sec.
-\ref{}) and show correctness via a categorical version of logical relations (Sec.
-\ref{}).
-We apply some standard optimisation techniques on the code.
-% % We finally explore some theoretical guarantees of the resulting code (Sec. \ref{}) and give some experimental results (Sec. \ref{}).
-
-% % % \subsubsection{Brief overview} % (fold)
-% % % \label{sub:brief_overview}
-% % %of the problem: code example
-
-% % % % diff prog, what state of the art is, what we propose, roughly our contribution
-% % % Under the slogan of differentiable programming, Automatic Differentiation has seen a renewal in the machine learning community over the past few years.
-The need for fast gradient computations of code implementing functions is ubiquitous and core in deep learning.
-Recent work has either focused on correct differentiation on high-level functional languages or about fast, usually low-level, imperative and unpure implementations.
-This work aims to bridge the gap between the two, first by providing a high-level source-to-source differentation on a high-level language, which is then optimised in several ways.
-Doing so, we reach state-of-the-art performance and ultimately produce code that is similar to the manually-optimised imperative code that one could get.
-% % % %maybe add that it's principled, explains more things, is correct, and great for generalisation to other languages/libraries/etc.
-
-% % \subsection{Our approach} % (fold)
-% % \label{sub:our_approach}
-
-% % \MH{start from pure HO functional language with smooth primitives, use an optimised (principled) macro, then further classical optimisations.
-forward def to things in more detail.
-list of key insights.}
-
-% % %rename Context?
-
-% % \subsection{Differentiable programming} % (fold)
-% % \label{sub:differentiable_programming}
-
-% % %as a "solution" people found? maybe skip this entirely
-
-% % %define AD
-% % %mostly from Matthijs' POPL paper
 % % (Source-to-source) Automatic Differentiation (AD) is a technique for transforming code that implements 
@@
 % % %Diff Curry has OK intro
-
-% \[
-% \begin{tikzcd}
-% 	& \TSSyn \ar[r,"{\Dsynrev[\rho]{-}}"] 
-% 	& \TTSynHO_1 \ar[d,"\UNF{-}^{-1}"] &  \\
-% 	& \SSSyn \ar[u,"\UNF{-}"]  \ar[r,"{\Dsynrev[\rho]{-}^*}"'] 
-% 	& \TTSynHO_2 \ar[d,"Jacob"] \\
-% 	& \SSSyn \ar[u,"ANF"] \ar[r,"{\Dsynrev[\rho]{-}^{**}}"'] 
-% 	&  \TTSynHO_3 \ar[dl,"\partial_2"] \\
-% 	\SSSynHO \ar[ur,"\partial_1"]  \ar[rrr,bend right=20,color=red,"Example\,1"]
-% 	& \TTSyn \ar[r,"OPT"'] 
-% 	& \TTSyn \ar[r,"\partial_3"']
-% 	& \TTSyn 
-% \end{tikzcd}
-% \]
-
-% \begin{itemize}
-% 	\item $\SSSynHO$ is a higher-order language and the language the user uses to write a prgram.
-% 	\item $\partial_1$ denotes partial evaluation of all higher-order variables. We end up with a program whose only higher-order bits are primitive operators.
-% 	\item $\SSSyn$ is a first-order language with a few higher-order primitives like map,fold,etc.
-% 	\item $ANF$ is for (weak) A-normal form. Every operator $\op$ is only given variables as arguments. We do this by introducting additional let bindings. It is only used for the $Jacob$ optimisation.
-% 	\item $\UNF{-}$ is the unary form transformation presented in the next section. Its goal is to deal with the fanout problem before applying the reverse-mode macro.
-% 	\item $\TSSyn$ is the target language for $\UNF{-}$ and also the source for the reverse-mode macro. It has a bit of a special structure.
-% 	\item $\Dsynrev[\rho]{-}$ is a reverse-mode macro. It is efficient on unary operators but does not deal well with fanout. Fortunately, it is only given unary operators at this point.
-% 	\item $\TTSynHO_1$ is the target language of the reverse-mode macro. It is a higher-order language as it needs a continuation for the reverse derivatives.
-% 	\item $\TTSynHO_2,\TTSynHO_3$ are small variations of $\TTSynHO_1$.
-% 	\item $\UNF{-}^{-1}$ is an optimisation which removes the overhead caused by $\UNF{-}$. It reverts the primal part of the derivative to the original term.
-% 	\item $\Dsynrev[\rho]{-}^*$ is an efficient macro for reverse-mode which deals with the fanout problem.
-% 	\item $Jacob$ stands for an efficient representation of the Jacobian.
-It needs a weak form of ANF to compute the Jacobian with a single vector and does not need to use a matrix-vector multiplication.
-In particular, it does not need to store the matrix.
-% 	\item $\Dsynrev[\rho]{-}^{**}$ is the efficient macro for reverse mode that would really be implemented. It will be quicker to run than the composition of the transformations. 
-% 	\item $\partial_2$ stands for partial evaluation. We evaluate (CBN) the higher-order variables binding the lambdas we introduced and inlined with the reverse-mode macro
-% 	\item $\TTSyn$ is the target language of the previous partial evaluation and the final language we use.
-% 	\item $OPT$ is an optimisation. First, we remplace the previously inlined lambdas by let bindings, and then apply forward substitution.
-% 	\item $\partial_3$  is optionnal. We partially evaluate more and do algebraic simplifications, like $0+\trm\to\trm, 1*\trm\to \trm$.
-% 	\item $Example\,1$ is detailed below.
-% \end{itemize}
-
-% \MH{I could add a different name for the target of ANF, of Jacob, and of OPT.}
 
@@
 
+\begin{itemize}
+    \item rewrite Pearlmutter's paper
+    \item purely functional yet efficient, where P's paper used references
+    \item no purity means harder to optimise, lack of real FP, parallelism, modularity
+    \item show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems
+    \item modularity, claritity and understanding are key.
+\end{itemize}
+
 \subsection{Problem}
+
+How to go from nice theory to efficient practice?
+\begin{itemize}
+    \item tradeoff expressivity versus static knowledge to do Optimisations
+    \item general optimisation versus ad-hoc hand-writen choice
+    \item how to stay provably correct and pure
+    \item how to be probably efficient?
+\end{itemize}
 
@@
 
+\begin{itemize}
+    \item $\Dsynrevsymbol$ of !let! and optimise
+    \item $\Dsynrevsymbol$ of $\prod_i v_i, \sum_i v_i$ and/or dot-product.
+\end{itemize}
+
 \subsection{Contribution}
 
+\begin{itemize}
+    \item expressive source language with arrays but restrictive enough for efficient $\Dsynrevsymbol$
+    \item method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
+    \item new I.R.: UNF
+    \item a very simple $\Dsynrevsymbol$ macro
+    \item expressive purely functional target language
+    \item several generalisation at the end
+    \item correctness and complexity theorems
+    \item recover efficiency via standard optimisation techniques as opposed to ad-hoc ones
+    \item detailed comparison with other work and limitations to our work, some ideas around limitations
+\end{itemize}
+
 \subsection{Outline of the paper}
+
+Our work looks like a compilation scheme, which is the following:
+
+\input{figures/compilation_scheme}

### 2021-02-24 rad-icfp-2021/abstract.tex
redid abstract

--- 2021-02-14 rad-icfp-2021/abstract.tex
@@
 \begin{abstract}
-Our work aims at bridging the gap between high-level functional reverse-mode differentiation and efficient imperative implementations.
-We define a new source-to-source transformation for reverse-mode differentiation on a standard higher-order purely functional language. 
-Using a logical relations argument, we prove our transformation correct using diffeological spaces as a denotational semantics for our language. 
-We then show that combining our macro with some classical optimisations lead to low-level SSA-like code, which is as efficient as low-level imperative implementations. 
-We prove some theoretical complexity guarantees and provide experimental results which showcase the efficiency of our method. 
+    We introduce an efficient source-code transformation for reverse-mode differentiation on an expressive purely functional language that does not introduce mutation.
+    We show that our transformation is correct and verifies the cheap gradient principle. 
+    Our transformation is presented as a compilation scheme through a novel intermediate representation we call unary form.
+    We obtain provably efficient gradients by performing general partial evaluation optimisations after our reverse-mode transformation as opposed to hand-crafted ones.
+    For simple first-order programs, the obtained outputs look like static single assignment (SSA) code. 
+    For arrays, our transformation preserves parallelism as much as possible. 
+    Our work emphasizes modularity and our language can easily be enriched with more optimised primitives as required for some speedups in practice.
 \end{abstract}

### 2021-02-24 rad-icfp-2021/intro.tex
starting to write intro

@@
 
-\begin{itemize}
-    \item rewrite Pearlmutter's paper
-    \item purely functional yet efficient, where P's paper used references
-    \item no purity means harder to optimise, lack of real FP, parallelism, modularity
-    \item show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems
-    \item modularity, claritity and understanding are key.
-\end{itemize}
+In 2008, Pearlmutter and Siskind \cite{pearlmutter2008reverse} introduced AD in a functional setting. 
+They introduced reverse-mode in their setting. 
+Reverse-mode can compute in one pass the whole gradient of a function $\RR^n\to\RR$, and this is very fast.
+By comparison, the easier forward-mode transformation would need $n$ passes. 
+These functions $\RR^n\to\RR$ actually appear a lot in practice, for instance as loss functions in machine learning.
+Reverse-mode is thus important as it computes a whole gradient very quickly for these functions. 
+To be as efficient as the usual imperative version of reverse-mode, the transformation in \cite{pearlmutter2008reverse} introduced references.
+
+The lack of purity of reverse-mode is known to make it significantly harder to optimise and find parrallelism. 
+None of the current implementations of reverse-mode in a functional setting are pure \cite{}, and often complicated heuristics with no guarantees are used \cite{}.
+It means that for every important non trivial operation, a hand-crafted reverse-derivative must be given to be efficient.
+In other words functional implementations seem to stuggle to abstract away from imperative code.
+
+For example, \MH{TODO: show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems}.
+For real world AD systems, modularity is key. They need to be expressive but have a lot of optimised derivatives for operators that are used a lot.
+This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
+An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and potentially reusable.
 
@@
 
-How to go from nice theory to efficient practice?
-\begin{itemize}
-    \item tradeoff expressivity versus static knowledge to do Optimisations
-    \item general optimisation versus ad-hoc hand-writen choice
-    \item how to stay provably correct and pure
-    \item how to be probably efficient?
-\end{itemize}
+Following \cite{pearlmutter2008reverse}, it is relatively easy to define a purely functional reverse-mode on a first-order language using continuations. 
+As remarked by the authors, this is highly inefficient. 
+The problem we tackle in this paper is how to go from a simple and general setting where we can relatively easily prove correctness of reverse-mode, 
+to a representation that matches more efficient implementations.
+
+The problems that this raises and that our paper deals with are the following. First, there is a tradeoff between a general expressive language 
+as opposed to a domain specific language. The latter usually has more static information and a specific representation to perform optimisations.
+Then, a lot of optimisations performed on AD implementations consist of a lot of hand-crafted derivatives for a lot of useful operations like matrix-matrix multiplication, dot-product, etc.
+They don't seem to come with more theoretical justification, are error prone, and can hardly fit more general optimisations.
+\MH{TODO: maybe the example and ref that ML systems ake 90percent of the time outside of numpy.}
+This makes these systems harder to prove correct. The problem is thus to stay provably correct and pure, and yet efficient. What's more is we would like something easily provably efficient. 
+A real implementation on top of our work would of course optimise further, potentially using hand-crafted operations as well, but it would be based on more solid grounds.
 
@@
 
-\begin{itemize}
-    \item $\Dsynrevsymbol$ of !let! and optimise
-    \item $\Dsynrevsymbol$ of $\prod_i v_i, \sum_i v_i$ and/or dot-product.
-\end{itemize}
+We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure. 
+On simple first-order terms, the output code looks like SSA code\cite{}.
+
+\MH{TODO: example !let!.}
+
+On arrays, two simple operations of interest are the dot-product of two vectors and the product of the elements of a vector.
+\MH{TODO: examples $\prod_i v_i, \sum_i v_i$ and/or dot-product.}
 

### 2021-02-24 rad-icfp-2021/intro.tex
derivative fold and map
examples in intro

@@
 % % %Diff Curry has OK intro
-
-% \begin{example}
-% 	Let 
-% \begin{align*}
-% 	\var_1,\var_2,\var_3:\RR\vdash \trm:=\big(\lambda f:\RR\to\RR. (f(\var_1))*\var_1\big)(\lambda \var[2]. \var[2]* \var_2)
-% \end{align*}
-% After $\partial_1$ we obtain 
-% \begin{align*}
-% 	(\letin{\var[2]}{\var_1}{\var[2]*\var_2})*\var_1
-% \end{align*}
-% After $ANF$ we obtain
-% \begin{align*}
-% 	&\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
-% \end{align*}
-% After $\Dsynrev[\rho]{-}^*$ we obtain
-% 	\begin{align*}
-% 		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Y\big(J^1(\var_1,\var_2,\var_3)^T(\var[2]_1,...,\var[2]_4)\big)} }{
-% 		\\
-% 		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'\big(J^2(\var_1,\var_2,\var_3,w_1)^T(\var[2]_1,...,\var[2]_5)\big)}}{\\ 
-% 		& w_2'}}
-% 	\end{align*}
-% where $Y$ is a continuation variable of type $\RR^3\to\RR^3$, $J^1,J^2$ are Jacobian matrices for $*$ operators which also return their context, $\inllambda$ is just a normal $\lambda$ but inlined for future optimisation.
-
-% After applying $Jacob$, we obtain
-% 	\begin{align*}
-% 		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Y(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
-% 		\\
-% 		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
-% 		& w_2'}}
-% 	\end{align*}
-% To actually compute the reverse-derivative of the full term, we substitute the identity to $Y$ and apply the whole term to $(0,0,0,0,1)$, the functional equivalent to initialing the sensitivities in the imperative reverse-mode algorithm.
-We obtain
-
-% \begin{align*}
-% 		&\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
-% 		\\
-% 		&\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
-% 		& w_2'(0,0,0,0,1)}}
-% 	\end{align*}
-
-% After $\partial_2$, we get
-% 	\begin{align*}
-% 		&\letin{w_1}{\var_1 * \var_2}{\\
-% 		&\letin{w_2}{w_1 * \var_1}{\\
-% 		&\bigg[\inllambda \var[2]_1...\var[2]_5.\Big(\inllambda \var[2]_1'...\var[2]_4'.(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)\Big)(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)\bigg](0,0,0,0,1)
-% 		}
-% 		}
-% 	\end{align*}
-
-% 	Now let's apply the first part of $OPT$:
-% 	\begin{align*}
-% 		&\letin{w_1}{\var_1 * \var_2}{\\
-% 		&\letin{w_2}{w_1 * \var_1}{\\
-% 		&\letin{(\var[2]_1\ldots\var[2]_5)}{(0,0,0,0,1)}{\\
-% 		& \letin{(\var[2]_1'\ldots\var[2]_4')}{(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}{\\
-% 		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2'+\var_1*\var[2]_4',\var[2]_3')
-% 		}}}}
-% 	\end{align*}
-% Let's unsugar the lets:
-% 	\begin{align*}
-% 		&\letin{w_1}{\var_1 * \var_2}{\\
-% 		&\letin{w_2}{w_1 * \var_1}{\\
-% 		&\letin{\var[2]_1}{0}{\\
-% 		&\letin{\var[2]_2}{0}{\\
-% 		&\letin{\var[2]_3}{0}{\\
-% 		&\letin{\var[2]_4}{0}{\\
-% 		&\letin{\var[2]_5}{1}{\\
-% 		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
-% 		&\letin{ {\color{red} \var[2]_2'} }{  {\color{red} \var[2]_2} }{\\
-% 		&\letin{ {\color{red} \var[2]_3'} }{ {\color{red} \var[2]_3}  }{\\
-% 		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
-% 		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2'+\var_1*\var[2]_4',\var[2]_3')
-% 		}}}}}}}}}}}
-% 	\end{align*}
-% 	We apply the second part of $OPT$ to remove the variables in red:
-% 	\begin{align*}
-% 		&\letin{w_1}{\var_1 * \var_2}{\\
-% 		&\letin{w_2}{w_1 * \var_1}{\\
-% 		&\letin{\var[2]_1}{0}{\\
-% 		&\letin{\var[2]_2}{0}{\\
-% 		&\letin{\var[2]_3}{0}{\\
-% 		&\letin{\var[2]_4}{0}{\\
-% 		&\letin{\var[2]_5}{1}{\\
-% 		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
-% 		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
-% 		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2+\var_1*\var[2]_4',\var[2]_3)
-% 		}}}}}}}}}
-% 	\end{align*}
-% 	We apply some algebraic simplifications in $\partial_3$
-% 	\begin{align*}
-% 		&\letin{w_1}{\var_1 * \var_2}{\\
-% 		&\letin{w_2}{w_1 * \var_1}{\\
-% 		& \letin{\var[2]_1'}{w_1}{\\
-% 		&\letin{\var[2]_4'}{\var_1}{\\
-% 		&(\var[2]_1'+\var_2*\var[2]_4',\var_1*\var[2]_4',0)
-% 		}}}}
-% 	\end{align*}
-% 	We can optimise some more with $\partial_3$, and we finally obtain
-% 	\begin{align*}
-% 		&\letin{w_1}{\var_1 * \var_2}{\\
-% 		&\letin{w_2}{w_1 * \var_1}{\\
-% 		&(w_1+\var_2*\var_1,\var_1*\var_1,0)
-% 		}}
-% 	\end{align*}
-% \end{example}
 
@@
 They don't seem to come with more theoretical justification, are error prone, and can hardly fit more general optimisations.
-\MH{TODO: maybe the example and ref that ML systems ake 90percent of the time outside of numpy.}
+\MH{TODO: maybe the example and ref that ML systems spend 90percent of the time outside of numpy.}
 This makes these systems harder to prove correct. The problem is thus to stay provably correct and pure, and yet efficient. What's more is we would like something easily provably efficient. 
 A real implementation on top of our work would of course optimise further, potentially using hand-crafted operations as well, but it would be based on more solid grounds.
-
-\subsection{Examples}
-
-We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure. 
-On simple first-order terms, the output code looks like SSA code\cite{}.
-
-\MH{TODO: example !let!.}
-
-On arrays, two simple operations of interest are the dot-product of two vectors and the product of the elements of a vector.
-\MH{TODO: examples $\prod_i v_i, \sum_i v_i$ and/or dot-product.}
 
@@
 
+\subsection{Examples}
+
+\MH{I think it might be better after contribution to show how concrete our work is on some example. Need to say I don't detail UNF here.}
+We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure. 
+Let's have a look at what we obtain on a few examples.
+
+\begin{example}[First-order term]
+On simple first-order terms, the output code looks like SSA code \cite{}.
+Let's consider $\letin{w_1}{\var_1 * \var_2}{\letin{w_2}{w_1 * \var_1}{w_2}}$ in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
+After our inefficient reverse-mode transformation we obtain
+\begin{align*}
+    &\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Id(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
+    \\
+    &\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
+    & w_2'(0,0,0,0,1)}}
+\end{align*}
+where $Id$ is the identity function $\RR^3\to\RR^3$, $\inllambda$ is just a normal $\lambda$ tagged for inlining.
+The part $(0,0,0,0,1)$ comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.
+
+After some general partial evaluation that will be detailed further in the paper, we obtain     
+    	\begin{align*} 
+    		&\letin{w_1}{\var_1 * \var_2}{\\
+    		&\letin{w_2}{w_1 * \var_1}{\\
+    		&\letin{\var[2]_1}{0}{\\
+    		&\letin{\var[2]_2}{0}{\\
+    		&\letin{\var[2]_3}{0}{\\
+    		&\letin{\var[2]_4}{0}{\\
+    		&\letin{\var[2]_5}{1}{\\
+    		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
+    		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
+    		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2+\var_1*\var[2]_4',\var[2]_3)
+    		}}}}}}}}}
+    	\end{align*}  
+This is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.
+Of course this can be further optimised to give
+    	\begin{align*}
+    		&\letin{w_1}{\var_1 * \var_2}{\\
+    		&\letin{w_2}{w_1 * \var_1}{\\
+    		&(w_1+\var_2*\var_1,\var_1*\var_1,0)
+    		}}
+    	\end{align*}
+    \end{example}
+
+ \begin{example}[Simple operations on arrays]
+    On arrays, three simple operations of interest are the dot-product of two vectors, and the product or sum of the elements of a vector.
+    In a functional setting, these can be defined as follows:
+\begin{center}
+    \begin{tabular}{{r c l}}
+        !prod(A)! &:=& !reduce * 1 A! \\
+        !sum(A)! &:=& !reduce + 0 A! \\
+        !dot(A,B)! &:=& !reduce + 0 (map2 * A B)!     
+    \end{tabular}
+\end{center}
+where !reduce! is a known fold left operator for which the function argument is associative and is faster than a fold left.
+
+The gradient of each of these construct w.r.t. !A! will be:
+\begin{center}
+    \begin{tabular}{{r c l}}
+        $\nabla_A$!prod(A)! &:=& !map2 * (scanr x 1 A) (shift1(scanl * 1 A))!\\
+        $\nabla_A$!sum(A)! &:=& !map (x,y -> 1) A!\\
+        $\nabla_A$!dot(A,B)! &:=& !map2 * B (map (x,y -> 1) A! 
+    \end{tabular}
+\end{center}
+where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
+!scanl! and !scanr! are respectively scan left and scan right. These are like fold but store all intermediate results in an array.
+We obtain these gradients, which are efficient and usually derived by hand, as special cases of our work.
+They are obtained automatically in our case.
+\end{example}
+
 \subsection{Outline of the paper}

### 2021-02-24 rad-icfp-2021/intro.tex
outline paper in intro

@@
 On simple first-order terms, the output code looks like SSA code \cite{}.
+\MH{need to rewrite following my notation.}
 Let's consider $\letin{w_1}{\var_1 * \var_2}{\letin{w_2}{w_1 * \var_1}{w_2}}$ in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
@@
 
-Our work looks like a compilation scheme, which is the following:
+Our work looks like the following compilation scheme
+\input{figures/compilation_scheme}
 
-\input{figures/compilation_scheme}
+In Section~\ref{sec:background} we recall reduminents of automatic differentiation, and in particular the idea of reverse-mode and the cheap gradient principle.
+Next, in Section~\ref{sec:simplediff} we introduce the source language we want to differentiate.
+We also introduce our new intermediate representation UNF, and a simple reverse-mode macro on this representation.
+In Section~\ref{sec:efficientrad} we translate back from UNF to a purely functional target language and show several general optimisations to make our reverse-mode transformation efficient.
+In the next Section~\ref{sec:complexity} we show that the previous optimisations are sufficient to ensure our transformation has the right complexity.
+Section~\ref{sec:correctness} shows that our construction is correct and we give standard semantics to our source and target languages. 
+In Section~\ref{sec:generalisation} we show how to generalise the previous sections for a richer source language. 
+Finally, related work and some limitations to our approach are presentend in Section~\ref{sec:conclusion}.

### 2021-02-24 rad-icfp-2021/intro.tex
small fixes
change notation intro example

@@
 
-\MH{I think it might be better after contribution to show how concrete our work is on some example. Need to say I don't detail UNF here.}
 We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure. 
@@
 On simple first-order terms, the output code looks like SSA code \cite{}.
-\MH{need to rewrite following my notation.}
-Let's consider $\letin{w_1}{\var_1 * \var_2}{\letin{w_2}{w_1 * \var_1}{w_2}}$ in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
-After our inefficient reverse-mode transformation we obtain
-\begin{align*}
-    &\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Id(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
-    \\
-    &\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
-    & w_2'(0,0,0,0,1)}}
-\end{align*}
+Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
+After our (inefficient) reverse-mode transformation, we obtain
+\begin{center}
+    \begin{tabular}{l}
+        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> Id(y1+x2*y4, y2+x1*y4, y3)> in!\\
+        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
+        !w_2'(0,0,0,0,1)!
+    \end{tabular}
+\end{center}
+% \begin{align*}
+%     &\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Id(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
+%     \\
+%     &\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
+%     & w_2'(0,0,0,0,1)}}
+% \end{align*}
 where $Id$ is the identity function $\RR^3\to\RR^3$, $\inllambda$ is just a normal $\lambda$ tagged for inlining.
@@
 After some general partial evaluation that will be detailed further in the paper, we obtain     
-    	\begin{align*} 
-    		&\letin{w_1}{\var_1 * \var_2}{\\
-    		&\letin{w_2}{w_1 * \var_1}{\\
-    		&\letin{\var[2]_1}{0}{\\
-    		&\letin{\var[2]_2}{0}{\\
-    		&\letin{\var[2]_3}{0}{\\
-    		&\letin{\var[2]_4}{0}{\\
-    		&\letin{\var[2]_5}{1}{\\
-    		& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
-    		&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
-    		&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2+\var_1*\var[2]_4',\var[2]_3)
-    		}}}}}}}}}
-    	\end{align*}  
+    	% \begin{align*} 
+    	% 	&\letin{w_1}{\var_1 * \var_2}{\\
+    	% 	&\letin{w_2}{w_1 * \var_1}{\\
+    	% 	&\letin{\var[2]_1}{0}{\\
+    	% 	&\letin{\var[2]_2}{0}{\\
+    	% 	&\letin{\var[2]_3}{0}{\\
+    	% 	&\letin{\var[2]_4}{0}{\\
+    	% 	&\letin{\var[2]_5}{1}{\\
+    	% 	& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
+    	% 	&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
+    	% 	&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2+\var_1*\var[2]_4',\var[2]_3)
+    	% 	}}}}}}}}}
+    	% \end{align*}
+        \begin{center}
+            \begin{tabular}{l}
+                !let w1 = x1 * x2 in!\\ 
+                !let w2 = w1 * x1 in!\\
+                !let y1 = 0 in!\\
+                !let y2 = 0 in!\\
+                !let y3 = 0 in!\\
+                !let y4 = 0 in!\\
+                !let y5 = 1 in!\\
+                !let y1'! != y1+w1*y5 in!\\
+                !let y4'! != y4+x1*y5 in!\\
+                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
+            \end{tabular}
+        \end{center}  
 This is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.
 Of course this can be further optimised to give
-    	\begin{align*}
-    		&\letin{w_1}{\var_1 * \var_2}{\\
-    		&\letin{w_2}{w_1 * \var_1}{\\
-    		&(w_1+\var_2*\var_1,\var_1*\var_1,0)
-    		}}
-    	\end{align*}
+    	% \begin{align*}
+    	% 	&\letin{w_1}{\var_1 * \var_2}{\\
+    	% 	&\letin{w_2}{w_1 * \var_1}{\\
+    	% 	&(w_1+\var_2*\var_1,\var_1*\var_1,0)
+    	% 	}}
+    	% \end{align*}
+        \begin{center}
+            \begin{tabular}{{c}}
+                !let w1 = x1 * x2 in!\\ 
+                !let w2 = w1 * x1 in!\\
+                !(w1+x2*x1, x1*x1, 0)!
+            \end{tabular}
+        \end{center}
     \end{example}

### 2021-02-24 rad-icfp-2021/intro.tex
minor

@@
 
-In 2008, Pearlmutter and Siskind \cite{pearlmutter2008reverse} introduced AD in a functional setting. 
+In 2008, Pearlmutter and Siskind \cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting. 
 They introduced reverse-mode in their setting. 
@@
 
+\MH{forward reference to section 2 where reverse-mode is recalled. maybe same for SSA. }
+
 We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure. 
@@
     \begin{tabular}{l}
-        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> Id(y1+x2*y4, y2+x1*y4, y3)> in!\\
+        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
         !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
@@
 \end{center}
-% \begin{align*}
-%     &\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Id(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
-%     \\
-%     &\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
-%     & w_2'(0,0,0,0,1)}}
-% \end{align*}
-where $Id$ is the identity function $\RR^3\to\RR^3$, $\inllambda$ is just a normal $\lambda$ tagged for inlining.
 The part $(0,0,0,0,1)$ comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.
@@
 After some general partial evaluation that will be detailed further in the paper, we obtain     
-    	% \begin{align*} 
-    	% 	&\letin{w_1}{\var_1 * \var_2}{\\
-    	% 	&\letin{w_2}{w_1 * \var_1}{\\
-    	% 	&\letin{\var[2]_1}{0}{\\
-    	% 	&\letin{\var[2]_2}{0}{\\
-    	% 	&\letin{\var[2]_3}{0}{\\
-    	% 	&\letin{\var[2]_4}{0}{\\
-    	% 	&\letin{\var[2]_5}{1}{\\
-    	% 	& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
-    	% 	&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
-    	% 	&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2+\var_1*\var[2]_4',\var[2]_3)
-    	% 	}}}}}}}}}
-    	% \end{align*}
         \begin{center}
@@
 Of course this can be further optimised to give
-    	% \begin{align*}
-    	% 	&\letin{w_1}{\var_1 * \var_2}{\\
-    	% 	&\letin{w_2}{w_1 * \var_1}{\\
-    	% 	&(w_1+\var_2*\var_1,\var_1*\var_1,0)
-    	% 	}}
-    	% \end{align*}
         \begin{center}

### 2021-02-24 rad-icfp-2021/intro.tex
improving intro a bit

@@
 
+Our work provides a sort of compilation scheme from a source functional language to a target functional language for reverse-mode differentiation.
+To emphasize some key ideas and for presentation purposes, we present our work on a simple yet expressive language in Section~\ref{sec:simplediff}, 
+and show some useful generalisation in Section~\ref{sec:generalisation}.
+
+The key ideas of our work can be summarised by saying that reverse-mode differentiation
 \begin{itemize}
-    \item expressive source language with arrays but restrictive enough for efficient $\Dsynrevsymbol$
-    \item method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
-    \item new I.R.: UNF
-    \item a very simple $\Dsynrevsymbol$ macro
-    \item expressive purely functional target language
-    \item several generalisation at the end
-    \item correctness and complexity theorems
-    \item recover efficiency via standard optimisation techniques as opposed to ad-hoc ones
-    \item detailed comparison with other work and limitations to our work, some ideas around limitations
+   \item can be made efficient way more easily on unary operators
+   \item can be done efficiently and purely using second-order array operators for second-order array operators (e.g. !map,foldl!) 
+   \item can be made efficient via standard functional optimisation techniques
+\end{itemize}
+
+Our contribution tries to make these claims precise. Our detailed  contribution consist of
+\begin{itemize}
+    \item introducing an expressive source language with array operations, yet restrictive enough for efficient differentiation
+    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
+    \item a new intermediate representation (I.R.) which we call unary form(UNF), essentially turning every operation as a unary one
+    \item a simple reverse-mode transformation on the I.R.
+    \item an expressive purely functional target language and efficient transformation from the I.R. to this language
+    \item several useful generalisations in Section~\ref{sec:generalisation}. 
+    \item correctness of the differentiation \ref{sec:correctness}, and complexity guarantees \ref{sec:complexity}
+    \item showing how efficiency  is recovered via standard optimisation techniques
+    \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
 \end{itemize}
@@
 
-\MH{forward reference to section 2 where reverse-mode is recalled. maybe same for SSA. }
-
 We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure.
+We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
 Let's have a look at what we obtain on a few examples.
@@
 \end{center}
-The part $(0,0,0,0,1)$ comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.
+The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.
 
@@
         \end{center}  
-This is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.
-Of course this can be further optimised to give
+For the expert viewer, this is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.\\
+Of course, this term can be further optimised to give
         \begin{center}

### 2021-02-25 rad-icfp-2021/intro.tex
some refs

@@
 The lack of purity of reverse-mode is known to make it significantly harder to optimise and find parrallelism. 
-None of the current implementations of reverse-mode in a functional setting are pure \cite{}, and often complicated heuristics with no guarantees are used \cite{}.
+None of the current implementations of reverse-mode in a functional setting \cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.
+\cite{xla}.
 It means that for every important non trivial operation, a hand-crafted reverse-derivative must be given to be efficient.
@@
 \begin{example}[First-order term]
-On simple first-order terms, the output code looks like SSA code \cite{}.
+On simple first-order terms, the output code looks like SSA code \cite{cytron1989efficient}.
 Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
@@
         \end{center}  
-For the expert viewer, this is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.\\
+For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
 Of course, this term can be further optimised to give

### 2021-02-25 rad-icfp-2021/relwork.tex
minor
outline paper in intro
adding a ref

@@
 \section{Conclusion}
+\label{sec:conclusion}
 
@@
 \label{sub:summary}
+
+We have shown how to compute provably efficient gradients via reverse derivatives in a purely functional way 
+on a simple yet expressive language with array functions.
+We introduced a novel intermediate representation, unary form, 
+and used it combined with standard functional optimisations to perform our reverse-mode transformation.
 
@@
 Relatively recent work in AD (2008-2020):
-\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,wang2018demystifying,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing}
+\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,wang2018demystifying,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}
 
@@
     \item more 2nd order functions, though can be derived by hand once and then it's ok
-    \item dynamic arrays
+    \item more dynamic arrays
     \item more dynamic control flow
@@
     \item non smooth and partial functions, although by simply following the ref book it's as usual in AD
-    \item sparsity
+    \item sparsity and data manipulation
     \item other optimisations like implicit function theorem

### 2021-02-25 rad-icfp-2021/background.tex
change outline
outline paper in intro
cleaned up refs.bib

--- 2021-02-16 rad-icfp-2021/background.tex
@@
-\section{Rudiments of AD}
+\section{Reverse-mode in Automatic Differentiation}
+\label{sec:background}
 
-\subsection{Rudiments of differentiation and dual numbers.}
+% \subsection{Rudiments of differentiation and dual numbers.}
+% Recall that the derivative of a function $f:\RR\to \RR$, if it exists, is a function
+% $\nabla f:\RR\to \RR$ such that $\nabla f(x_0)=\frac {\dif f(x)}{\dif x}(x_0)$ is the gradient of $f$ at $x_0$.
+
+% To find $\nabla f$ in a compositional way, two generalizations are reasonable:
+% \begin{itemize}
+% \item We need both $f$ and $\nabla f$ when calculating $\nabla (f;g)$
+% of a composition $f;g$, using the chain rule, so we are really interested in the pair $(f,\nabla f):\RR\to \RR\times \RR$;
+% \item In building $f$ we will need to consider functions of multiple arguments, such as $+:\RR^2\to \RR$, and these functions should propagate derivatives.
+% \end{itemize}
+% Thus we are more generally interested in transforming a function $g:\RR^n\to \RR$ into a function
+% $h:(\RR\times \RR)^n\to \RR\times \RR$ in such a way that for any
+% $f_1\dots f_n:\RR\to\RR$, 
+% \begin{equation}
+%   \label{eqn:dualnumber}
+%   (f_1,\nabla f_1,\dots, f_n,\nabla f_n);h
+%   =
+%   ((f_1,\dots, f_n);g,\nabla ((f_1, \dots, f_n);g))\text.
+% \end{equation}
+
+% An intuition for $h$ is often given in terms of dual numbers.
+% The transformed function operates on pairs of numbers, $(x,x')$, and it is common
+% to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
+% But while this is a helpful intuition, the formalization of infinitesimals can be intricate, 
+% and the development in this paper is focussed on the elementary formulation in~\eqref{eqn:dualnumber}.
+
+% The reader may also notice that $h$ encodes all the partial derivatives of
+% $g$. For example, 
+% if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
+% $h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
+% and similarly 
+% $h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
+% And conversely, if $g$ is differentiable in each argument, then
+% a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
+% combinations of partial derivatives:
+% \[\textstyle h(x_1,x_1',x_2,x_2')=(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))\text.\]
+
+% In summary, the idea of differentiation with dual numbers is 
+% to transform a differentiable function
+% $g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives.
+We packaged this up in~\eqref{eqn:dualnumber} as a sort-of invariant which is useful for building derivatives of compound functions $\RR\to\RR$ in a compositional way.
+% The idea of automatic differentiation is to perform this transformation at the source code level.
+
+% The two main ways in which it is performed in practice is by operator overloading \cite{} or by source code transformation. We focus our work on the latter.
+
+% \subsection{Reverse-mode AD}
+
+% AD comes in two main flavors: forward mode and reverse mode. 
+% Forward mode computes a directional derivative of the Jacobian while reverse-mode computes a directional derivative of the transpose Jaobian. 
+% This has a significant impact in practice as this means reverse mode can compute the whole gradient of a function $f:\RR^n\to \RR$ in one pass, whereas forward mode needs $n$ passes. 
+% Assuming that both have about the same complexity, this can have a huge impact in practice. 
+% Such functions are expremely commonly seen as loss functions in the context of machine learning for instance.
+
+% Reverse-mode computes the derivatives backward, using the symmetry of the chain rule.
+% This means the compute flow is reverted after the first pass to compute the value of the function and its intermediate values. 
+% These values appear in the chain rule so should be stored or recomputed.
+% These things make reverse-mode harder to implement efficiently, and especially harder to analyze and optimize. 
+% Traditional implementations use references.
+% For instance, let 
+%  \begin{align*}
+%      \trm\defeq &\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
+%  \end{align*}
+%  The imperative version of the reverse differentiation of $\trm$ is
+% 	\begin{align*}
+% 		&w_1=\var_1 * \var_2 \\
+% 		&w_2=w_1 * \var_1 \\
+%         &\var_1':=0~;~\var_2':=0 \\
+%         &w_1':=0 \\
+% 		&w_2':=1 \\
+% 		&w_1'+= \var_1 * w_2'; \var_1'+= w_1*w_2' \\
+% 		&\var_1'+= \var_2 * w_1'; \var_2'+=\var_1 * w_1'\\
+% 		&\text{return }(\var_1',\var_2')  
+% 	\end{align*}
+% The tangent part for every variable from the context $\var_i'$ and every intermediate variable $w_i'$ is initialized at $0$, 
+% except for the return variable $w_2$ for which  $w_2':=1$.
+
+% As noted in \cite{pearlmutter2008reverse}, in a functional language, one can have a purely functional reverse-mode using continuations. 
+% This representation is notoriously inefficient, and the existing implementations of reverse mode in a functional setting have followed the lines of the imperative version above \cite{pearlmutter2008reverse, wang2018demystifying, baydin2016diffsharp}.
+
+% In the next section, we first introduce a very simple first-order language. 
+% Then we define reverse-mode on this language as a program transformation. 
+% This first transformation is quite inefficient and we how we can use standard optimizations on functional languages to obtain a provably efficient pure program.
+% The key idea behind that will be generalized and explored in detail in the rest of the paper is that of turning every operator into a unary one.     
+% The remaining of the paper will generalize this idea and explore some consequences of this new approach.
+
+\subsection{Rudiments of forward-mode AD and dual numbers}
+
 Recall that the derivative of a function $f:\RR\to \RR$, if it exists, is a function
@@
 
-The two main ways in which it is performed in practice is by operator overloading \cite{} or by source code transformation. We focus our work on the latter.
+The two main ways in which it is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
+Our work focus on a compiled version so we focus on source code transformation which is more fitted in this case.
 
-\subsection{Reverse-mode AD}
+\subsection{Reverse-mode differentiation}
 
-AD comes in two main flavors: forward mode and reverse mode. 
-Forward mode computes a directional derivative of the Jacobian while reverse-mode computes a directional derivative of the transpose Jaobian. 
-This has a significant impact in practice as this means reverse mode can compute the whole gradient of a function $f:\RR^n\to \RR$ in one pass, whereas forward mode needs $n$ passes. 
-Assuming that both have about the same complexity, this can have a huge impact in practice. 
-Such functions are expremely commonly seen as loss functions in the context of machine learning for instance.
+The problem with the previous approach shows up when one wants to compute the full gradient of a function $\RR^n\to\RR$, for a large $n$. 
+Forward-mode only computes one directional derivative, for instance one partial derivative. This implies $n$ passes must be performed through the forward derivative to compute the whole gradient.
+By using the symmetry in the chain rule, there is a way to compute the whole gradient faster, and this method is reverse-mode differentiation.
+Suppose given a function $f=f_n\circ...\circ f_1 :\RR^n\to\RR$. 
+The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(...(Jf_1v))...)$ for a direction $v\in\RR^n$. 
+Reverse-mode, on the other hand, computes $(Jf)^Tv =Jf_1T^(Jf_{2}^T(...(Jf_n^Tv))...)$ for a vector $v\in\RR$. 
+Because the compute flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
+%%TODO: from Matthijs' paper, need to rewrite.
+Reverse-mode AD is only well-understood as a source-code transformation (also called define-then-run
+style AD) on limited programming languages. Typically, its implementations
+on more expressive languages that have features such as higher-order functions and conditionals
+make use of define-by-run approaches. These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
+first-order program is left, and then they evaluate this new program \cite{carpenter2015stan,paszke2017automatic}. 
+Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. As such, these AD libraries
+need to be implemented using carefully, manually optimized code, that for example does not contain any common subexpressions. This implementation process
+is precarious and labour intensive. Further, some whole-program optimizations
+that a compiler would detect go entirely unused in such systems.
 
-Reverse-mode computes the derivatives backward, using the symmetry of the chain rule.
-This means the compute flow is reverted after the first pass to compute the value of the function and its intermediate values. 
-These values appear in the chain rule so should be stored or recomputed.
-These things make reverse-mode harder to implement efficiently, and especially harder to analyse and optimise. 
-Traditional implementions use references.
-For instance, let 
- \begin{align*}
-     \trm\defeq &\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
- \end{align*}
- The imperative version of the reverse differentiation of $\trm$ is
-	\begin{align*}
-		&w_1=\var_1 * \var_2 \\
-		&w_2=w_1 * \var_1 \\
-        &\var_1':=0~;~\var_2':=0 \\
-        &w_1':=0 \\
-		&w_2':=1 \\
-		&w_1'+= \var_1 * w_2'; \var_1'+= w_1*w_2' \\
-		&\var_1'+= \var_2 * w_1'; \var_2'+=\var_1 * w_1'\\
-		&\text{return }(\var_1',\var_2')  
-	\end{align*}
-The tangent part for every variable from the context $\var_i'$ and every intermediate variable $w_i'$ is initialized at $0$, 
-except for the return variable $w_2$ for which  $w_2':=1$.
+\subsection{Purely functional inefficient reverse-mode}
 
-As noted in \cite{pearlmutter2008reverse}, in a functional language, one can have a purely functional reverse-mode using continuations. 
-This representation is notoriously inefficient and all the implementions of reverse mode in a functional setting have followed the lines of the imperative version above \cite{pearlmutter2008reverse, wang2018demystifying, baydin2016diffsharp}.
+Following \cite{pearlmutter2008reverse}, there is an easy way to define an inefficient yet purely functional reverse-mode transformation for first-order programs.
+\MH{TODO: finish}
 
-In the next section, we show on a very simple language how we can use standard optimisations to turn this inefficient representation into a very efficient one which is still purely functional.
-The remaining of the paper will generalise this idea and explore some consequences of this new approach.
+\subsection{Insights for efficient reverse-mode}
+
+\begin{itemize}
+    \item a
+    \item  
+\end{itemize}

### 2021-02-25 rad-icfp-2021/background.tex
some more background

@@
 The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(...(Jf_1v))...)$ for a direction $v\in\RR^n$. 
-Reverse-mode, on the other hand, computes $(Jf)^Tv =Jf_1T^(Jf_{2}^T(...(Jf_n^Tv))...)$ for a vector $v\in\RR$. 
+Reverse-mode, on the other hand, computes $(Jf)^Tv =J^Tf_1(J^Tf_{2}(...(J^Tf_nv))...)$ for a vector $v\in\RR$.
+In particular, taking $v=1$ computes the gradient of $f$.
+
 Because the compute flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
@@
 
-Following \cite{pearlmutter2008reverse}, there is an easy way to define an inefficient yet purely functional reverse-mode transformation for first-order programs.
-\MH{TODO: finish}
+Following \cite{pearlmutter2008reverse}, there is a simple way to define an inefficient yet purely functional reverse-mode transformation for first-order programs.
+We review a slight modification of the transformation we presented, which is also better explained on an example.
+
+Let the term we want to differentiate be $x1:\RR,...,xn:\RR\vdash exp(cos(xi))$.
+Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
+Instead of considering these as operations from $\RR\to\RR$, we consider then as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
+If we just do that though, we lose compositionality. 
+So $cos$ will also return its context and is now seen as a function $\sem{cos}:\RR^n\to\RR^{n+1}$.
+similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
+We thus obtain $\sem{exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $J\sem{cos} \in Mat_{n,n+1}$, $J\sem{exp} \in Mat_{n+1,n+2}$ also compose nicely.
+The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
+This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ...\circ f_1$. 
+If the original program was of type $\RR$, then the return value of the original program is the last compononent of $f$.
+Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
+\[\nabla f= J^Tf(0,...,0,1)=J^Tf_1(J^Tf_{2}(...(J^Tf_m(0,...,0,1)))...)\]
+
+To actually reverse the order of computation needed for this transpose of jacobians, we use a sort of simple continuation.
+$f_i$ is turned into $\Dsynrevsymbol{f_i}:=<f_i, \lambda Y. Y\circ J^Tf_i>$ where $Y:\RR^{n+i-1}\to \RR^n$. 
+We recover compositionality by noting that $<f_{i+1}(f_i), (\lambda Y. Y\circ J^Tf_{i+1}>)(\lambda Y. Y\circ J^Tf_i)>$ reduces to
+$<f_{i+1}\circ f_i, \lambda Y. Y\circ J^Tf_i \circ J^Tf_{i+1}>$, and thus by induction we can obtain $<f, \lambda Y. YJ^Tf>$.
+By applying the identity continuation $\RR^n\to\RR^n$ on the second component then the result to $(0,...,0,1)$, 
+we have obtained a purely functional to compute $\nabla f$.
+
+This method is quite inefficient though. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
+Second, each $J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.
+
+The imperative version of reverse mode for a binary operator $op(x,y)$ is something like $x'+= \partial_1op(x,y);y'+= \partial_2op(x,y)$. 
+It is for these cases that mutation is usually key. 
+If we see out term as a directed graph, reverse mode needs to backpropagate from the end of the graph to the starting nodes via every path.
+It's hard to keep track of all these information in parallel in a functional way.
+One of the key simple ideas that we used was to transform every operator into a unary one. 
+This sssentially trivialises the compute flow to a line. 
+Even if the starting program was a straight line program, 
+having non unary operators was a source of inefficiency which justified mutation in the first place.
+By returning every variable every time, the problem of using a variable several time does not need to be dealt with via mutation. 
+This simple idea of transforming a program into essentially a straight line is what our new intermediate representation unary-form does.
 
@@
 
-\begin{itemize}
-    \item a
-    \item  
-\end{itemize}
+If we look at $J^Tf_{i+1}$, we notice that this function is almost the identity, except at the last row. 
+Even on the last row, if the original term was a unary or binary operator like $cos, exp, +, *$, 
+the row is zero except for at most 2 indices (1 for unary operators).
+This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,...,y(n+i)).(y1,...,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
+where $[k]$ means that the element is added at the $k$-th index of the tuple.
+
+We know in advance that such jacobian functions are going to be applied one to another, so we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
+Because each function is almost the identity, we obtain a lot of substitutions of the form $[x/y]$ where both $x$ and $y$ are variables. 
+This allows us to drastically reduce the size of $J^Tf$. In fact, for simple programs this is basically enough to get an efficient purely functional reverse derivative transformation.
+We develop this idea further for a richer language.

### 2021-02-25 rad-icfp-2021/generalisation.tex
adding some figures
minor
outline paper in intro
small fixes
some more background

@@
 \section{Generalisations}
+\label{sec:generalisation}
+
+\MH{I think I only present two: conditionals, and general case for array. partial op can be dealt with before. HO function mentioned at the end}
+
+Extended source language:
+
+\input{figures/generalisation_source_grammar}
+
+% \input{figures/generalisation_source_typesystem}
+
+Note that we recover !zip A B! as !map2 (fun x,y -> (x,y)) A B)!. 
+We can lift the restriction on the function of !foldl! by changing a part in $\invUNFSymbol$ of !foldl!.
+\begin{center}
+\begin{tabular}{r c l}
+    !Y((x1,...,xm),...)! & -> &  !Y((x1,...,xm) + map2! \\
+        && !(fun z1 z2 -> let E = map2 (fun x,y -> z2(x,y)) A e3 in! \\
+        &&  !let F = zip B E in!\\
+        && !foldl (fun (x,y), acc  -> x*acc+y) z1 F)! \\
+        && $\nabla_{\Gamma}$!e2! $\nabla_{\Gamma}$!e1,...)!
+\end{tabular}
+\end{center}
+
+Extended target language:
+
+% \input{figures/generalisation_target_grammar}
+
+% \input{figures/generalisation_target_typesystem}
 
@@
 
-\subsubsection{Higher-order functions}
+\subsection{Higher-order functions}

### 2021-02-26 rad-icfp-2021/intro.tex
cleaned up refs.bib
starting section 3

@@
 
-\subsection{Contribution}
+\subsection{Smmary of contributions}
 
@@
 We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
-Let's have a look at what we obtain on a few examples.
+Let us have a look at what we obtain on a few examples.
 
@@
 On simple first-order terms, the output code looks like SSA code \cite{cytron1989efficient}.
-Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
+Let us consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
 After our (inefficient) reverse-mode transformation, we obtain
@@
 
-In Section~\ref{sec:background} we recall reduminents of automatic differentiation, and in particular the idea of reverse-mode and the cheap gradient principle.
+In Section~\ref{sec:background} we recall rudiments of automatic differentiation, and in particular the idea of reverse-mode and the cheap gradient principle.
 Next, in Section~\ref{sec:simplediff} we introduce the source language we want to differentiate.

### 2021-02-26 rad-icfp-2021/simplediff.tex
adding some structure to the document
starting unf figures
outline paper in intro
starting section 3

@@
 \section{Simple reverse-mode differentiation}
+\label{sec:simplediff}
 
@@
 
+We consider a standard call-by-value language. 
+It is a first-order language with array constructs and a few typical second-order array operations. 
+The types !T1,T2! and terms !e1,e2! are in Figure~\ref{fig:source_grammar}.
+
+\input{figures/source_grammar}
+
+The typing rules are in Figure~\ref{fig:source_typesystem}. We have included a minimal set of array operations for the sake of illustration, 
+but it is not difficult to add further operations. For scalar operations, we assume given a set of operations. 
+!op1! and !op2! denote respectively unary and binary smooth operations on reals. 
+We use infix notation for binary operators.
+Typical examples include !cos, exp, +, *!. 
+For the sake of simplicity in the development, the function in the array operations !map, map2, foldl! are quite restricted with and free variables. 
+We also restrict to arrays of reals.
+The general case is also useful and presented in Section~\ref{sec:generalisation}.
+
+\input{figures/source_type_system}
+
 \subsection{Source Unary Form}
+
+Following the intuition higlighted in Section~\ref{subsec:insights}, we present a new language which we call unary form (UNF). 
+It simply consists of a composition of unary operators. That said, we want to compile our source language to this intermediate representation, 
+and we need to remember some information about the initial term.
+
+The grammar of our source UNF is given in Figure~\ref{fig:unf_source_grammar}.
+
+\input{figures/source_unf}
+
+There are a few notable things in this syntax. 
+Every term is indexed by a well-defined contex $\Gamma$ from the source language.
+In addition, a variable has a integer index $1\leq i\leq n$ where $n$ is the size of the context $\Gamma$.
+Every constant, unary, or binary operator is summarised as an $n$-ary operator !op!$_n$.
+Sequential composition is denoted by !;! and !e1;e2! means that !e1! should be performed, and then !e2!.
+Every array operator !map!, !map2!, !foldl! has an extra index which represents a well-formed term in the source language.
+
+The types are the same as in the source language. 
+A context $\Gamma=\{$!x1: T1,...,xn:Tn!$\}$ can always be seen as a type !T1!$\times$...$\times$!Tn!.
+UNF is very syntactic and does not really use variables. 
+In fact it consists only of constants and composition. 
+For this reason, we denote contexts as types in the indices of the terms, and $\times$ is used for context extension.
+But we do emphasize that we mean contexts and the actual variables are remembered and
+this will be key when compiling to the target language after differentiation.
+
+The typing rules are detailed in Figure~\ref{fig:source_unf_typesystem}.
+By $\reals^{\times n}$ we mean the product $\reals\times...\times\reals$ of $n$ factors $\reals$.
+
+\MH{need to explain that return values produce some kind of ANF and introduce new vars!}
+
+\input{figures/source_unf_typesystem}
+
+We give a translation from our source language to source UNF in Figure~\ref{fig:source_to_unf}.
+
+\MH{explain kind of monadic ; need theorem saying this makes sense.}
+
+\input{figures/source_to_unf}
 
@@
 
+\input{figures/target_unf}
+
+\input{figures/target_unf_typesystem}
+
 \subsection{Macro for reverse mode}
+
+\input{figures/macro_diff}

### 2021-02-26 rad-icfp-2021/simplediff.tex
small subsection in Sec3

@@
 
+Now our program in UNF is in an equivalent form to stright line programs. 
+Before presenting our differentiation macro, we present the target language for this macro. 
+We call this intermediate representation target UNF.
+The main idea is similar to source UNF where we only have constants and sequential composition.
+That said, the key difference is that now every constant represents a pair of an operation and what is meant to represent its Jacobian.
+Following the insights from Section~\ref{sec:background}, composition of pairs reverses the order on the second component. T
+his represents the fact that reverse-mode reverts the computation flow, 
+and that pre-composition (using a continuation) is the simplest way to do this in a purely functional setting.
+On pairs, !;! still represents sequential composition for the first component, and reversed composition is represented by $\circ$ on the second component.
+The grammar and types are given in Figure~\ref{fig:unf_target_grammar}.
+For every constant $C$ from the source UNF, we have a new constant $J^TC$ representing its transpose Jacobian.
+
 \input{figures/target_unf}
+
+The typesystem is given in Figure~\ref{fig:target_unf_typesystem}. 
+One thing to note is that one can only form pairs.
+
+\MH{it's currently a bit wrong. need to slightly change to match with UNF-1.}
 

### 2021-02-26 rad-icfp-2021/simplediff.tex
minor

@@
 
+We are now able to present a simple transformation for purely functional reverse-mode on UNF.
+This transformation is somehow straightforward as some of the work was put into the design of UNF.
+Our main takeaway for this paper is that reverse-mode is complicated beause there is a lot happening in the same time. 
+But by decoupling the problems, it becomes muche easier. UNF is somehow dealing with the inherent type-dependency of reverse-mode.
+Then the macro $\Dsynrevsymbol$ adds Jacobians. It allows to compute gradients in a compositional way.
+Finally, we'll see in the next section how to compile back to a normal language and optimise our transformation to get efficiency.
+
+The differentiation macro is given in Figure~\ref{fig:diff_macro}.
+
 \input{figures/macro_diff}

### 2021-02-26 rad-icfp-2021/efficientdiff.tex
adding some structure to the document
part of the operational semantics target
outline paper in intro
writing about target language

@@
 \section{Efficient reverse-mode}
+\label{sec:efficientrad}
 
 \subsection{Target Language}
+
+The last language we present, which we call the target language, is an extension to the source language. 
+The types and terms are presented in Figure~\ref{fig:target_grammar}.
+The language contains lambda abstractions for dealing with continuations. 
+We are not interested here with partial applications and our language is pure, so functions take $n$ arguments.
+Lambda abstractions and applications are going to be removed during partial evaluation and this notation makes reading slightly easier.
+
+\input{figures/target_grammar}
+
+In addition, we require the set of operations to be closed under partial differentiation. 
+In more detail, for every unary scalar operation !op1!, 
+we assumed given an operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
+For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
+representing respectively the first and second partial derivative of !op2!.
+Similarly, we need to add some functions on arrays. 
+!scanl! is a !foldl! that stores all intermediate results. 
+!scanr! is similar, but reads the array from right to left, and also stores from right to left.
+!reduce! is a !foldl! for which the function is associative. 
+This means in practice that it can be computed faster.
+We also include a !map3! operator. 
+Though not necessary as it can be simulated with two !map2!, it is a convenient notation and can somtimes run faster in practice.
+Finally, we add a somewhat strange operator which we call !shift1!. 
+It takes an array of size $n$,  shifts all the element of the array by one, 
+forgetting the last element, and puts a one in the first place. 
+This operator naturally shows up when differentiating !foldl!.
+
+The type system for the extended grammar of target is in Figure~\ref{fig:target_typesystem}.
+
+\input{figures/target_type_system}
+
+A standard semantics of this language is given in Section~\ref{sec:correctness}.
+
+We now present our transformation from target UNF to target. 
+The transformation is presented in Figure~\ref{fig:unf_to_target}.
+
+\input{figures/unf_to_target}
+
+\begin{remark}
+    One could rightfully wonder: where did reverse-mode really happened here?
+    In a way, source UNF does not seem to do much. 
+    Then $\Dsynrevsymbol$ is very simple on UNF.
+    So perhaps everything happens when going from target UNF to target?
+    \MH{finish or remove.}
+\end{remark}
 

### 2021-02-26 rad-icfp-2021/simplediff.tex
writing about target language

@@
 
-Following the intuition higlighted in Section~\ref{subsec:insights}, we present a new language which we call unary form (UNF). 
+Following the intuition highlighted in Section~\ref{subsec:insights}, we present a new language which we call unary form (UNF). 
 It simply consists of a composition of unary operators. That said, we want to compile our source language to this intermediate representation, 
@@
 Sequential composition is denoted by !;! and !e1;e2! means that !e1! should be performed, and then !e2!.
-Every array operator !map!, !map2!, !foldl! has an extra index which represents a well-formed term in the source language.
+Every array operator !map!, !map2! and !foldl! has an extra index which represents a well-formed term in the source language.
 
@@
 This transformation is somehow straightforward as some of the work was put into the design of UNF.
-Our main takeaway for this paper is that reverse-mode is complicated beause there is a lot happening in the same time. 
-But by decoupling the problems, it becomes muche easier. UNF is somehow dealing with the inherent type-dependency of reverse-mode.
+Our main takeaway for this paper is that reverse-mode is complicated because there is a lot happening at the same time. 
+But by decoupling the problems, it becomes much easier. UNF is somehow dealing with the inherent type-dependency of reverse-mode.
 Then the macro $\Dsynrevsymbol$ adds Jacobians. It allows to compute gradients in a compositional way.
-Finally, we'll see in the next section how to compile back to a normal language and optimise our transformation to get efficiency.
+Finally, we'll see in the next section how to compile back to a more standard language and optimize our transformation to get efficiency.
 
 The differentiation macro is given in Figure~\ref{fig:diff_macro}.
+One might be surprised that the macro does not seem to do much, and it is partly correct.
+We tried to decompose reverse-mode into more elementary small steps, and each step is slightly non trivial and dealing with one aspect of reverse-mode.
+The perhaps slightly more involved transformation is the one coming next,  going from target UNF to our target language.
 

### 2021-02-26 rad-icfp-2021/simplediff.tex
some fixes

@@
 We give a translation from our source language to source UNF in Figure~\ref{fig:source_to_unf}.
+UNF acts as a sort of stack. 
+The input represents the current stage of the stack, and we return the new value added to the stack.
+All results are kept, so the stack keeps growing. One might be surprised that the transformation for pairs and let is similar.
+The order of UNF really follows what the trace of doing a sequential call-by-value evaluation would look like if the return value does not matter.
+In other words, when evaluating !let x= e1 in e2! in call by value, we first evaluate !e1! and then !e2!. 
+Similarly in !<e1, e2>!, we evaluate !e1! and then !e2!. 
+Our ultimate goal is to return a gradient, not the original term, so what really matters is the order of evaluation, and what is evaluated.
 
-\MH{explain kind of monadic ; need theorem saying this makes sense.}
+$\widehat{;}$ is similar to $;$, with a sort of context propagation effect. 
+It is a bit reminiscent of monadic composition versus the usual composition.
+Technically though, the transformation $\UNFSymbol$ is not quite compositional.
+\MH{explain in detail how to define it.}
+
+\begin{proposition}
+    Let $\Gamma\vdash$!e!: !T! be a term in the source language.\\
+    Then $\Gamma \vdash \UNF$(!e!):$\Gamma\times\Delta\times$!T! in source UNF for a certain type $\Delta$. 
+\end{proposition}
+
+$\Delta$ corresponds to the list of all the intermediate results from !e!. 
+A different way to say it would be that $\Delta$ is the context made of the bound variables of !e! when !e! is in ANF.
 
@@
 
-Now our program in UNF is in an equivalent form to stright line programs. 
+Now our program in UNF is in an equivalent form to straight-line programs. 
 Before presenting our differentiation macro, we present the target language for this macro. 
@@
 That said, the key difference is that now every constant represents a pair of an operation and what is meant to represent its Jacobian.
-Following the insights from Section~\ref{sec:background}, composition of pairs reverses the order on the second component. T
+Following the insights from Section~\ref{sec:background}, the composition of pairs reverses the order on the second component. T
 his represents the fact that reverse-mode reverts the computation flow, 

### 2021-02-26 rad-icfp-2021/simplediff.tex
fixing target unf

@@
 For every constant $C$ from the source UNF, we have a new constant $J^TC$ representing its transpose Jacobian.
+One thing to note is that one can only form pairs in this language. 
+Similarly to the source UNF, sequential composition is the only operation which is not a constant of the language.
 
@@
 
+Technically, all our terms in target UNF should have an extra index $\rho$ which is a type from Source.
+This will be the return type of the continuation for the jacobian accumulation. 
+In other words, if we want the full gradient of a term $\Gamma\vdash$!e!: $\reals$, $\rho$ should be chosen to be $\Gamma$.
+For every type !T!, let $\Dtype$(!T!)$:=$!T!$\times$(!T!->$\rho$). 
 The typesystem is given in Figure~\ref{fig:target_unf_typesystem}.
-One thing to note is that one can only form pairs.
 

### 2021-02-27 rad-icfp-2021/simplediff.tex
minor fixes

@@
     Let $\Gamma\vdash$!e!: !T! be a term in the source language.\\
-    Then $\Gamma \vdash \UNF$(!e!):$\Gamma\times\Delta\times$!T! in source UNF for a certain type $\Delta$. 
+    Then $\Gamma \vdash \UNFSymbol$(!e!): $\Gamma\times\Delta\times$!T! in source UNF for a certain type $\Delta$. 
 \end{proposition}
@@
 
-\MH{it's currently a bit wrong. need to slightly change to match with UNF-1.}
-
 \input{figures/target_unf_typesystem}
@@
 \input{figures/macro_diff}
+
+Even though the target UNF is designed to achieve this purpose, we have the Following
+
+\begin{proposition}
+    If $\Gamma \vdash$!e!: !T! in source UNF, then
+    $\Dtype(Gamma)\vdash \Dsynrevsymbol$(!e!):  $\Dtype$(!T!) in target UNF.
+\end{proposition}

### 2021-06-09 rad-icfp-2021/appendix1.tex
minor paper

\section{Appendix}

\subsection{gradient from the introduction}

We show that the gradients from the introduction 
can be obtained as special instances of our general construction.

\begin{lemma}
    $\nabla_A$!prod(A)! = !map2 * (scanr x 1 A) (shift1(scanl * 1 A))!
\end{lemma}

\begin{proof}

\end{proof}

\begin{lemma}
    $\nabla_A$!sum(A)! = !map (x -> 1) A!
\end{lemma}

\begin{proof}

\end{proof}

\begin{lemma}
    $\nabla_A$!dot(A,B)! := !map2 * B (map (x -> 1) A)! 
\end{lemma}

\begin{proof}

\end{proof}

### 2021-06-09 rad-icfp-2021/correctness.tex
some comments
change outline
outline paper in intro
writing about target language
minor paper

@@
 \section{Correctness}
+\label{sec:correctness}
 
-%decompose the macro from section 4 into unary form transfo, then principled macro, then opti
-\MH{Question: when is it best to give the idea about our macro? that it's UNF+macro+easy-opti?}
-\MH{probably won't use UNF in this paper}
+We now prove that our transformations are correct. 
+The main result of the section is
 
-\subsection{Unary Form} % (fold)
-\label{sub:unary_form}
+\begin{theorem}
+    For every $\Gamma \vdash$ !e: \reals!, we have 
+    $\sem{\pi_2 \invUNF{\UNF{e}}(Id_\Gamma)(last(1:\reals))}= \nabla_{\Gamma}\sem{e}$.
+\end{theorem}
+where $Id_\Gamma$ is the identity function, 
+$\sem{-}$ an appropriate denotational semantics of smooth functions,
+and $last(a)$ is the hot-vecteur encoding $(0,\ldots,0,a)$.
 
-\subsection{Principled Macro} % (fold)
-\label{sub:principled_macro}
+\subsection{Denotational semantics}
 
-\subsection{Correctness via subsconing} % (fold)
-\label{sub:correctness_via_subsconing}
+\begin{itemize}
+    \item for Syn
+    \item for target
+    \item for UNF
+\end{itemize}
 
-\subsection{Optimizations} % (fold)
-\label{sub:optimizations}
-%the ones in the macro of Sec4 then the further classical ones to ANF
+\MH{find somewhere to put the operational semantics}
+\input{figures/op_semantics_target}
 
-\subsection{Correctness of the macro from section \ref{}} % (fold)
-\label{sub:correctness_of_the_macro_from_section_ref}
+\subsection{Correctness}
+
+In order to prove correctness, we need some key intermediate lemmas.
+
+\begin{lemma}
+    $\sem{\pi_1\invUNF{\UNF{e}}}=\sem{e}$.
+\end{lemma}
+
+\begin{lemma}
+    $\sem{\pi_2\invUNF{e}}=T^*\sem{\pi_1\invUNF{e}}$
+\end{lemma}
+
+From this, we deduce the main theorem
+
+\begin{theorem}
+    \MH{slight generalisation of the corrolary for compositionality reasons}
+\end{theorem}
+
+The main theorem specializes to our case of interest
+
+\begin{corollary}
+    For every $\Gamma \vdash$ !e: \reals!, we have 
+    $\sem{\pi_2 \invUNF{\UNF{e}}(Id_\Gamma)(last(1:\reals))}= \nabla_{\Gamma}\sem{e}$.
+\end{corollary}

### 2021-06-09 rad-icfp-2021/intro.tex
minor paper

@@
 
-For example, \MH{TODO: show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems}.
-For real world AD systems, modularity is key. They need to be expressive but have a lot of optimised derivatives for operators that are used a lot.
-This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
-An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and potentially reusable.
+% For example, \MH{TODO: show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems}.
+% For real world AD systems, modularity is key. They need to be expressive but have a lot of optimised derivatives for operators that are used a lot.
+% This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
+An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and reusable.
 
@@
 
-The problems that this raises and that our paper deals with are the following. First, there is a tradeoff between a general expressive language 
+This raises several problems that are dealt with in our paper. First, there is a tradeoff between a general expressive language 
 as opposed to a domain specific language. The latter usually has more static information and a specific representation to perform optimisations.
@@
     \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
-    \item a new intermediate representation (I.R.) which we call unary form(UNF), essentially turning every operation as a unary one
+    \item a new intermediate representation (I.R.) which we call unary form (UNF), essentially turning every operation to a unary one
     \item a simple reverse-mode transformation on the I.R.
@@
 
-After some general partial evaluation that will be detailed further in the paper, we obtain     
+After some general partial evaluation techniques that will be detailed further in the paper, we obtain     
         \begin{center}
@@
 For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
-Of course, this term can be further optimised to give
+Of course, this term can be further optimised via constant propagation and algebraic simplifications to give
         \begin{center}
@@
     \begin{tabular}{{r c l}}
-        $\nabla_A$!prod(A)! &:=& !map2 * (scanr x 1 A) (shift1(scanl * 1 A))!\\
-        $\nabla_A$!sum(A)! &:=& !map (x,y -> 1) A!\\
-        $\nabla_A$!dot(A,B)! &:=& !map2 * B (map (x,y -> 1) A! 
+        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
+        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
+        $\nabla_A$!dot(A,B)! &:=& !map2 * B (map (x -> 1) A)! 
     \end{tabular}

### 2021-06-09 rad-icfp-2021/notes/main.tex
small note for Sam

\input{preamble}

\usepackage{bussproofs} 
\usepackage{boxedminipage}

\newenvironment{framed}[0]{\begin{boxedminipage}{\linewidth}}{\end{boxedminipage}}

\title{Categorical semantics for UNF}
\author{Mathieu Huot}

\begin{document}
{\large
\texttt{Begun Mai 2021. Draft of \today, \currenttime}
}
\begingroup
\let\newpage\relax
\maketitle
\endgroup

\section{Recall UNF}

\begin{figure}[H]
\setlength{\tabcolsep}{0.3em}
\centering
\begin{tabular}{|l c l|l|}
\hline
\multicolumn{3}{|c|}{\textbf{Core Grammar}} & \multicolumn{1}{c|}{\textbf{Description}}\\\hline
!T! & \mbox{::=} & $\reals$ & \grammarcomment{Real Type} \\
& $\mid$ & !T! $\times$ !T! & \grammarcomment{Product Type}\\
& $\mid$ & $\Array{\reals}$ & \grammarcomment{Real Array Type}\\
\hline
!e! & \mbox{::=} & !x! & \grammarcomment{Variable}\\
& $\mid$ & !c! & \grammarcomment{Real constant}\\
& $\mid$ & !let x = e in e! & \grammarcomment{Variable Binding}\\
& $\mid$ & !< e, e >! $\mid$ $\pi_1$(!e) $\mid$ $\pi_2$(!e) & \grammarcomment{Pair Constructor/Destructor}\\
& $\mid$ & !e op2 e! $\mid$ !op1 e! & \grammarcomment{Binary/Unary operations}\\
& $\mid$ & !map (x.e) e! $\mid$ !map2 (x,y.e) e e! & \grammarcomment{Array map and map2}\\
& $\mid$ & !foldl (x,y.e) e e! & \grammarcomment{Array fold left}\\
\hline
\end{tabular}
\vspace{-0.2cm}
\caption{Grammar of the source language.}
\label{fig:source_grammar}
\end{figure}

\begin{figure}[H]
    \setlength{\tabcolsep}{0.3em}
    \centering
    \begin{tabular}{|l c l|l|}
    \hline
    \multicolumn{3}{|c|}{\textbf{Core Grammar}} & \multicolumn{1}{c|}{\textbf{Description}}\\\hline
    !T! & \mbox{::=} & ... & \grammarcomment{Same as source} \\
    \hline
    !e! & \mbox{::=} & !var!$_{\Gamma,i}$ & \grammarcomment{Variable}\\
    & $\mid$ & !op!$_{\Gamma,n}$ & \grammarcomment{Operations, for $0\leq n\leq 2$}\\
    & $\mid$ & !e;e! & \grammarcomment{Sequential composition}\\
    & $\mid$ & !map!$_{\Gamma,x.e}$ $\mid$ !map2!$_{\Gamma,x,y.e}$ & \grammarcomment{Map and map2}\\
    & $\mid$ & !foldl!$_{\Gamma,x,y.e}$ & \grammarcomment{Fold left}\\
    \hline
    \end{tabular}
    \vspace{-0.2cm}
    \caption{Grammar of the source UNF}
    \label{fig:unf_source_grammar}
    \end{figure}

\begin{figure}[H]
    \begin{tabular}{r c l}
    $\UNFSymbol$($\Gamma\vdash $!c!) &=& !c!$_{\Gamma,0}$ constant seen as a 0-ary operator\\
    $\UNFSymbol$($\Gamma\vdash $!x!) &=& !var!$_{\Gamma,i}$ where !x! is the $i$-th variable in $\Gamma$ \\
    $\UNFSymbol$($\Gamma\vdash $ !let x = e1 in e2!) &=& $\UNFSymbol$(!e1!)$\widehat{;}$ $\UNFSymbol$(!e2!) \\ 
    $\UNFSymbol$($\Gamma\vdash $ !< e1, e2 >!: !T1!$\times$!T2!) &=& $\UNFSymbol$(!e1!)$\widehat{;}$ $\UNFSymbol$(!e2!) $\widehat{;}$ $\pi_{T1\times T2}$\\ 
    $\UNFSymbol$($\Gamma\vdash \pi_i$(!e!)) &=& $\UNFSymbol$(!e!)$\widehat{;}$ $\pi_i$ seen as a unary operator\\
    $\UNFSymbol$($\Gamma\vdash $ !e1 op2 e2!) &=& $\UNFSymbol$(!e1!)$\widehat{;}$ $\UNFSymbol$(!e2!)$\widehat{;}$ !op!$_{\Gamma,2}$\\
    $\UNFSymbol$($\Gamma\vdash $ !op1 e!) &=& $\UNFSymbol$(!e!) $\widehat{;}$ !op!$_{\Gamma,1}$ \\
    $\UNFSymbol$($\Gamma\vdash $ !map (x.e1) e2!) &=& $\UNFSymbol$(!e2!)$\widehat{;}$ !map!$_{\Gamma,x.e}$\\
    $\UNFSymbol$($\Gamma\vdash $ !map2 (x,y.e1) e2 e3!) &=& $\UNFSymbol$(!e2!)$\widehat{;}$ $\UNFSymbol$(!e3!)$\widehat{;}$ !map2!$_{\Gamma, x,y.e}$ \\ 
    $\UNFSymbol$($\Gamma\vdash $ !foldl (x,y.e1) e2 e3!) &=& $\UNFSymbol$(!e2!)$\widehat{;}$ $\UNFSymbol$(!e3!)$\widehat{;}$ !foldl!$_{\Gamma, x,y.e}$ \\  
    \end{tabular}
    \caption{UNF transformation from Source to Source UNF}
    \label{fig:source_to_unf}
    \end{figure}

$\widehat{;}$ is similar to $;$, with a sort of context propagation effect. 
It is a bit reminiscent of monadic composition versus the usual composition.

Formally, if $\Gamma\vdash e1: A$, $\Gamma\vdash \UNFSymbol(e1): \Gamma\times\Delta_1\times A$, $\Gamma\vdash e2: B$, $\Gamma\vdash \UNFSymbol(e2): \Gamma\times\Delta_2\times B$, then $\Gamma\vdash \UNFSymbol(e1) \widehat{;} \UNFSymbol(e2): \Gamma\times \Delta_1\times \Delta_2 \times A \times B$.
This is for most cases above except notably for let binding in which case it is: 
 $\Gamma\vdash \UNFSymbol(e1) \widehat{;} \UNFSymbol(e2): \Gamma\times A \times \Delta_1\times \Delta_2 \times B$.

 These types $\Gamma\times A \times \Delta_1\times \Delta_2 \times B$ ought to be unordered and represent just a set of variables, thus the extra annoyance with permutations in an ordered setting.

 Semantically, $\UNFSymbol(e1) \widehat{;} \UNFSymbol(e2)$ is a macro for $\UNFSymbol(e1) ; \UNFSymbol(\widetilde{e2}); \sigma$ where $\Gamma,\Delta_1,A\vdash \widetilde{e2}:B$ is the weakening of $e2$, and $\sigma: \Gamma\times\Delta_1\times A \times \Delta_2 \times B \to \Gamma\times\Delta_1\times \Delta_2 \times A\times B$ is the obvious permutation with this type.
The weakening and permutation are slightly changed in the obvious way for let binding to make the thing typecheck.

 \MH{I probably need to add permutations $\sigma$ in the language of UNF then.}

\section{Semantics attempts}

Because UNF inspects the code of its agument, it is not invariant under $\beta,\eta$ and so cannot be directly expressed as a structure preserving functor from the syntactic category of the source language modulo $\beta,\eta$ to another category.

\subsection{Syntactic source category}

Let $\Syn$ be the syntactic category of the source language. Its objects are types of source, and a morphism $\tau \to \sigma$ is a well-typed term of source, modulo only alpha-renaming. Composition is by the usual capture-free substitution. It is a free Cartesian category (+ some List-like structure for arrays, map, and fold) where we do not impose any arithmetic equalities, for instance !cos!(\underline{7}) $\neq$ $\underline{cos(7)}$ in the category.

\subsection{The Poly category}

We define the bicategory $\Polysyn$ with objects those of $\Syn$, $\Polysyn(A,B)=\cup_{P\in\Syn}\Syn(A,P\times B)$, and there is a 2-cell from $f:A\to P_1\times B$ to $g:A\to P_2\times B$ if there is an epimorphism $p:P_1\to P_2$ in $\Syn$ such that $g=f;id\times p$.
Composition of $f:A\to P_1\times B$ and $g:B\to P_2\times C$ in $\Polysyn$ is given by $f;P_1\times g;\sigma$ in $\Syn$, where $\sigma:P_1\times(P_2\times C)\to (P_1\times P_2)\times C$ is the obvious isomorphism.

Composition of 1-cells is strictly associoative, but the identity $f;id\cong f\cong id;f$ is not strict as $(1\times P)\times A\cong P\times A$ is usually not the identity.

\subsection{$\UNFSymbol$ categorically}

The nice thing with $\Polysyn$ is that its composition reflects the weakening part of $\widehat{;}$. In addition, the hiding of $P$ in $\Polysyn(A,B)=\cup_{P\in\Syn}\Syn(A,P\times B)$ reflects the type $\Delta$ appearing from $\UNFSymbol$ obtained by collecting bound variables in the term.

The problem is that the hidden state $P$ cannot be accessed anymore. This made me fail defining $\UNFSymbol:\Syn\to\Polysyn$ as a functor.

My attempt was to define $\UNFSymbol:\Syn\to\Polysyn$ as follows. 
\begin{itemize}
	\item $\UNF{u_A} =\langle id_A,u_A\rangle$
	\item $\UNF{id_A} = \langle u_A,id_A \rangle$
	\item $\UNF{\Delta_A} = \langle u_A,\Delta_A \rangle$
	\item $\UNF{\pi_1} = \langle pi_2, \pi_1 \rangle$
	\item $\UNF{op_2:\RR^2\to\RR} = \langle id_{\RR^2},op_2\rangle$ 
\end{itemize}
where $u_A$ is the unique map $A\to 1$.

The problem comes from the fact that asking $\UNFSymbol$ to be a functor, because it means it should be defined on $\Delta$. The semantics of !let e1 in e2! which is $\Delta;(id\times\sem{e1});\sem{e2}$ is sent to something like $\UNF{\Delta};(id\times \UNF{\sem{e1}});\UNF{\sem{e2}}$, which is not of the form I would like, i.e. $\UNF{\sem{e1}};\UNF{\sem{e2}}$. 
With the semantics above, everything is duplicated too much. For instance, the semantics of  !let y=z in cos(y)!  via this UNF would have type $1,\Gamma\to \Gamma\times (\Gamma-A)\times A, A$.

In practice, representing terms with string diagrams, I want to put a box around $e1$ and $\Delta$ and call this $\UNF{e1}$, whereas the categorical semantics forces me to draw a box around each component separately.

\MH{another possible way to look at it is that UNF does some sort of localization, turning every operator to a reversible one, which is simply done by making it also return its context. I wonder if it could simply be seen as the localization of $!$. Assuming $1\times A=A$, we have $\Delta;id\times !=id$, so as $!$ would become invertible, so would be $\Delta$. This is sort of what I need in practice for reverse mode: not seeing any $\Delta$, only unary maps, and an iso can always be seen as such in this context, i.e. by making $\Delta$ an iso, it becomes essentially invisible to reverse mode. One obvious problem is that I don't know if I want all types to be isomorphic to 1.}

\end{document}

### 2021-06-11 rad-icfp-2021/background.tex
starting section 3
minors

@@
 It is for these cases that mutation is usually key. 
-If we see out term as a directed graph, reverse mode needs to backpropagate from the end of the graph to the starting nodes via every path.
+If we see our term as a directed graph, reverse mode needs to backpropagate from the end of the graph to the starting nodes via every path.
 It's hard to keep track of all these information in parallel in a functional way.
 One of the key simple ideas that we used was to transform every operator into a unary one. 
-This sssentially trivialises the compute flow to a line. 
+This essentially trivialises the compute flow to a line. 
 Even if the starting program was a straight line program, 
@@
 
-\subsection{Insights for efficient reverse-mode}
+\subsection{Insights for efficient reverse-mode}\
+\label{subsec:insights}
 

### 2021-06-11 rad-icfp-2021/abstract.tex
some change abstract

@@
 \begin{abstract}
-    We introduce an efficient source-code transformation for reverse-mode differentiation on an expressive purely functional language that does not introduce mutation.
-    We show that our transformation is correct and verifies the cheap gradient principle. 
+    Deep learning is moving towards increasingly sophisticated optimization objectives that employ
+    tensors and operations on tensors.
+    Reverse-mode differentiation is used for optimisation, 
+    but it doesn't preserve purity of the underlying program and makes it notoriously harder to optimise.
+    In a purely functional setting, most array operations can be expressed conveniently using map and fold.
+    We present a differentiable programming language, that is the first to deliver en efficient pure and parallel friendly reverse-mode differentiation on
+    an expressive array language. 
+    We show that our transformation is semantically correct and verifies the cheap gradient principle. 
     Our transformation is presented as a compilation scheme through a novel intermediate representation we call unary form.
-    We obtain provably efficient gradients by performing general partial evaluation optimisations after our reverse-mode transformation as opposed to hand-crafted ones.
-    For simple first-order programs, the obtained outputs look like static single assignment (SSA) code. 
-    For arrays, our transformation preserves parallelism as much as possible. 
-    Our work emphasizes modularity and our language can easily be enriched with more optimised primitives as required for some speedups in practice.
+    We obtain provably efficient gradients by performing general partial evaluation optimisations after our reverse-mode transformation as opposed to manually derived ones.
+    For simple first-order programs, the obtained output programs ressemble static single assignment (SSA) code. 
+    We emphasize the modularity of our approach and show how our language can easily be enriched with more optimised primitives as required for some speedups in practice.
 \end{abstract}

### 2021-06-22 rad-icfp-2021/appendix1.tex
minors
changing makefile and adding todos
writing appendix

@@
 \section{Appendix}
+
+ \subsection{Operational semantics}
+
+In Figure~\ref{fig:op_semantics_target} a small step call-by-value operational semantics for the language. 
+\MH{need to add evaluation contexts}
+
+Some operators are less conventional which we now explain.
+!shift1! takes an array of reals of size $n$ and adds a $1$ as first component and forgets about the last component of the array, returning an array of size !n!.
+!scanr! is similar to !scanl! except that the folding starts from the end of the array. 
+It is equivalent to first reversing the input array, performing !scanl!, and reversing the resulting array.
+
+\MH{need to explain that the !scanl!, !scanr! coming from !reduce! are also parallel ones}
+
+\MH{might need to extend with the operators from the generalized section}
+
+\input{figures/op_semantics_target}
+
+ \subsection{Reverse derivative of array operations}
+
+We now prove that the reverse mode transformation is correct on array operations. 
+The proofs consist in showing that unrolling the computation before differentiation 
+and then differentiating gives the same as differentiation and then unrolling.
+
+ \begin{proposition}
+     The reverse derivative of !map! is correct.
+ \end{proposition}
+
+\begin{proof}
+
+\end{proof}
+
+ \begin{proposition}
+    The reverse derivative of !map2! is correct.
+\end{proposition}
+
+\begin{proof}
+
+\end{proof}
+
+\begin{proposition}
+    The reverse derivative of !foldl! is correct.
+\end{proposition}
+
+\begin{proof}
+
+\end{proof}
 
@@
 
-We show that the gradients from the introduction 
-can be obtained as special instances of our general construction.
+We show that the gradients from Section \ref{sec:intro} are obtained as instances of our general construction. 
+The proofs consist in instantiating the general derivatives to these cases 
+and showing that each rewrite step is a simple known optimization.
+
+Similarly to numpy, we use the notation OnesLike(A) to mean map (x -> 1) A 
+and ZerosLike(A) to mean map (x -> 0) A. 
+Finding these constant arrays is key to a lot of optimisations that leverage the ring algebraic structure of the reals to arrays.
+
+\MH{change to make it nice. be more explicit about optimisations. 0 should be ZerosLike. 
+Also, I am freely referring to the variables from the context or bound in theoriginal term, should be clarified.}
 
@@
  \begin{proof}
-
+\begin{align*}
+    \nabla_A \text{prod(A)} 
+    &= \nabla_A(\text{reduce} * 1 A) \\
+    &\defeq  \Big(\lambda A',z.
+    \begin{aligned}
+        & let A_1 = scanl * 1 A in \\
+        & let A_2 = map2 (x,y -> y) A_1 A in\\
+        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in\\
+        & let A_4 = scanr * 1 A_2 in \\
+        & map3 (a,b,c -> a+b*c*z) A' A_4 A_3
+    \end{aligned}
+    \Big)(0,1)\\
+    &\stackrel{\beta-reduction}{=}  \begin{aligned}
+        & let A_1 = scanl * 1 A in &\\
+        & let A_2 = map2 (x,y -> y) A_1 A in & A_2=  A\\
+        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in & A_3 = shift1 A1\\
+        & let A_4 = scanr * 1 A_2 in & \\
+        & map3 (a,b,c -> a+b*c*1) 0 A_4 A_3 & map2 * A_4 A_3
+    \end{aligned}\\
+    &=  \begin{aligned}
+        & let A_1 = scanl * 1 A in & \\
+        & let A_3 = shift1 A1 in &\\
+        & let A_4 = scanr * 1 A in &\text{forward substitution} A_2\\
+        & map2 * A_4 A_3
+    \end{aligned}\\
+    &\stackrel{\eta-reduction}{=} map2 * (scanr x 1 A) (shift1(scanl * 1 A))
+\end{align*}
  \end{proof}
@@
 
+\begin{align*}
+    \nabla_A sum(A) 
+    &= \nabla_A reduce + 0 A \\
+    & \defeq \Big(\lambda A',z.
+    \begin{aligned}
+        & let A_1 = scanl * 1 A in \\
+        & let A_2 = map2 (x,y -> 1) A_1 A in \\
+        & let A_3 = shift1 (map2 (x,y -> 1) A_1 A) in \\
+        & let A_4 = scanr * 1 A_2 in \\
+        & map3 (a,b,c -> a+b*c*z) A' A_4 A_3
+    \end{aligned}
+    \Big)(0,1)\\
+    &{\beta-reduction}{=}
+    \begin{aligned}
+        & let A_1 = scanl * 1 A in &\\
+        & let A_2 = map2 (x,y -> y) A_1 A in & A_2= OnesLike(A) \\
+        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in & A_3 = shift1 OnesLike(A)\\
+        & let A_4 = scanr * 1 A_2 in & \\
+        & map3 (a,b,c -> a+b*c*1) 0 A_4 A_3 & map2 * A_4 A_3
+    \end{aligned}\\
+    &= \begin{aligned}
+        & let A_1 = scanl * 1 A in & \\
+        & let A_2= OnesLike(A) in &  \\
+        & let A_3 = shift1 OnesLike(A) in & A_3 = OnesLike(A) \\
+        & let A_4 = scanr * 1 A_2 in & A_4 = OnesLike(A_2) \\
+        & map2 * A_4 A_3 &
+    \end{aligned}\\
+    &= map2 * OnesLike(A) OnesLike(A) \\
+    &= OnesLike(A)
+\end{align*}
  \end{proof}
@@
  \begin{lemma}
-    $\nabla_A$!dot(A,B)! := !map2 * B (map (x -> 1) A)! 
+     $\nabla_A$!dot(A,B)! = !B! 
  \end{lemma}
@@
  \begin{proof}
+The reverse derivative of !map2 * A B! is given by 
+\begin{align*}
+    & \lambda A',B',C'. \\
+    & let C_1 =map2 (a,b -> b) A B in \\
+    & let C_2 =map2 (a,b -> a) A B in \\
+    & (map3 (a,b,c -> a+b*c) A' C_1 C', map3 (a,b,c -> a+b*c) A' C_2 C')
+\end{align*}
 
+Let's call this term Y.
+For convenience, let us also rewrite !map2 * B (map (x -> 1) A)! 
+as !let C = map (x -> 1) A in map2 * B C!.
+
+Then the gradient of !dot(A,B)! is given by
+\begin{align*}
+    \nabla dot(A,B) 
+    &= \nabla let C = map (x -> 1) A in map2 * B C\\
+    &\defeq \lambda A',B',C',z. 
+    \left.\begin{aligned}
+        & let A_1 = scanl * 1 A in \\
+        & let A_2 = map2 (x,y -> 1) A_1 A in \\
+        & let A_3 = shift1 (map2 (x,y -> 1) A_1 A) in \\
+        & let A_4 = scanr * 1 A_2 in \\
+        & Y(A', B', map3 (a,b,c -> a+b*c*z) A' A_4 A_3)
+    \end{aligned}\right\}same reduction as previously
+    \\
+    &\quad(0,0,0,1) \\
+    &= Y(0,0,OnesLike(C)) \\
+    &\defeq 
+    \begin{aligned}
+        & let C_1 =map2 (a,b -> b) A B in & C_1 = B\\
+        & let C_2 =map2 (a,b -> a) A B in & C_2 = A\\
+        & (map3 (a,b,c -> a+b*c) 0 C_1 OnesLike(C), & map2 & C_1 OnesLike(C)
+        & map3 (a,b,c -> a+b*c) 0 C_2 OnesLike(C)) & map2 & C_2 OnesLike(C)
+    \end{aligned}\\
+    &= \begin{aligned}
+        & let C_1 = B & \text{forward substitution}\\
+        & let C_2 = A \text{forward substitution}\\
+        & map2 & C_1 OnesLike(C) & C_1
+        & map2 & C_2 OnesLike(C) & C_2
+    \end{aligned}\\
+    &= (B, A)
+\end{align*}
+So if we are only interested in the gradient w.r.t. A, this indeed gives us B.
  \end{proof}

### 2021-06-22 rad-icfp-2021/complexity.tex
outline paper in intro
writing appendix

@@
 \section{Complexity}
+\label{sec:complexity}
 
@@
 
+We follow a simple model similar to the one in \cite{griewank2008evaluating}.
+We assume the cost is divided into 4 elementary measures, being the number of MOVES, ADDS, MULTS, and NLOPS.
+MOVES assumes a flat memory and represents moving a fixed size information  (e.g. 64 bits). 
+ADDS represents the number of additions, 
+MULTS the number of multiplications, 
+and NLOPS the number of elementary non linear operations like !cos!, !exp!.
+
+The gives a complexity function $\cost$ valued in $\RR^4$. 
+For instance, we have
+
+\begin{itemize}
+    \item $\cost(*)=(3,0,1,0)$
+    \item $\cost(+)=(3,1,0,0)$
+    \item $\cost(c)=(1,0,0,0)$
+    \item $\cost($!sin!$)=(2,0,0,1)$
+\end{itemize}
+
+\MH{need cost array operations}
+\MH{need to extend the cost function in a compositional way}
+
 \subsection{Cheap gradient principle}

### 2021-06-28 rad-icfp-2021/purerad.tex
changing outline

--- 2021-02-26 rad-icfp-2021/efficientdiff.tex
@@
-\section{Efficient reverse-mode}
-\label{sec:efficientrad}
+\section{Simple pure reverse-mode differentiation}
+\label{sec:simplediff}
+
+\subsection{Source Language}
+
+We consider a standard call-by-value language. 
+It is a first-order language with array constructs and a few typical second-order array operations. 
+The types !T1,T2! and terms !e1,e2! are in Figure~\ref{fig:source_grammar}.
+
+\input{figures/source_grammar}
+
+The typing rules are in Figure~\ref{fig:source_typesystem}. We have included a minimal set of array operations for the sake of illustration, 
+but it is not difficult to add further operations. For scalar operations, we assume given a set of operations. 
+!op1! and !op2! denote respectively unary and binary smooth operations on reals. 
+We use infix notation for binary operators.
+Typical examples include !cos, exp, +, *!. 
+For the sake of simplicity in the development, the function in the array operations !map, map2, foldl! are quite restricted with and free variables. 
+We also restrict to arrays of reals.
+The general case is also useful and presented in Section~\ref{sec:generalisation}.
+
+\input{figures/source_type_system}
 
@@
 
-\begin{remark}
-    One could rightfully wonder: where did reverse-mode really happened here?
-    In a way, source UNF does not seem to do much. 
-    Then $\Dsynrevsymbol$ is very simple on UNF.
-    So perhaps everything happens when going from target UNF to target?
-    \MH{finish or remove.}
-\end{remark}
+\subsection{Macro for pure reverse mode transformation} % (fold)
+\label{sub:Macro for pure reverse mode transformation}
 
-\subsection{Efficient representation}
+\subsection{Partial evaluation and optimisation} % (fold)
+\label{sub:Partial evaluation and optimisation}
 
-\subsection{Optimisations via partial evaluation}
+% subsectionPartial evaluation and optimisation (end)

### 2021-06-28 rad-icfp-2021/unf.tex
changing outline

--- 2021-02-27 rad-icfp-2021/simplediff.tex
@@
-\section{Simple reverse-mode differentiation}
-\label{sec:simplediff}
+\section{UNF}
 
-\subsection{Source Language}
-
-We consider a standard call-by-value language. 
-It is a first-order language with array constructs and a few typical second-order array operations. 
-The types !T1,T2! and terms !e1,e2! are in Figure~\ref{fig:source_grammar}.
-
-\input{figures/source_grammar}
-
-The typing rules are in Figure~\ref{fig:source_typesystem}. We have included a minimal set of array operations for the sake of illustration, 
-but it is not difficult to add further operations. For scalar operations, we assume given a set of operations. 
-!op1! and !op2! denote respectively unary and binary smooth operations on reals. 
-We use infix notation for binary operators.
-Typical examples include !cos, exp, +, *!. 
-For the sake of simplicity in the development, the function in the array operations !map, map2, foldl! are quite restricted with and free variables. 
-We also restrict to arrays of reals.
-The general case is also useful and presented in Section~\ref{sec:generalisation}.
-
-\input{figures/source_type_system}
-
-\subsection{Source Unary Form}
+\subsection{Source UNF} % (fold)
+\label{sub:Source UNF}
 
@@
 
-\subsection{Target Unary Form}
+% subsubsectionSource UNF (end)
+
+\subsection{Target UNF} % (fold)
+\label{sub:Target UNF}
 
@@
 
-\subsection{Macro for reverse mode}
+\subsection{Simple reverse mode transformation} % (fold)
+\label{sub:Simple reverse mode transformation}
 
@@
 \end{proposition}
+
+\subsection{transformations to and from UNF} % (fold)
+\label{sub:transformations to and from UNF}
+
+% subsection transformations to and from UNF (end)

### 2021-06-29 rad-icfp-2021/correctness.tex
minors
writing appendix
changing outline
simple writing phase

@@
  \begin{theorem}
+    TODO: rephrase following Section~\ref{sec:simplediff}.
      For every $\Gamma \vdash$ !e: \reals!, we have 
@@
 
-\subsection{Denotational semantics}
+\subsection{Denotational semantics for the source and target languages}
 
 \begin{itemize}
-    \item for Syn
-    \item for target
-    \item for UNF
+    \item syntactic (multi?)category category
+    \item semantics in Manifolds or Eucl: TODO: need be clearer arrays of fixed size in syntax, should reflect the semantics
+    \item semantics for the array part
 \end{itemize}
 
-\MH{find somewhere to put the operational semantics}
-\input{figures/op_semantics_target}
+ \subsection{UNF semantics} % (fold)
+ \label{sub:UNF semantics}
 
-\subsection{Correctness}
+\begin{itemize}
+    \item TODO: what do I do here?
+\end{itemize}
 
-In order to prove correctness, we need some key intermediate lemmas.
+ \subsection{Correctness transformations} % (fold)
+ \label{sub:Correctness transformations}
 
-\begin{lemma}
-    $\sem{\pi_1\invUNF{\UNF{e}}}=\sem{e}$.
-\end{lemma}
-
-\begin{lemma}
-    $\sem{\pi_2\invUNF{e}}=T^*\sem{\pi_1\invUNF{e}}$
-\end{lemma}
-
-From this, we deduce the main theorem
-
-\begin{theorem}
-    \MH{slight generalisation of the corrolary for compositionality reasons}
-\end{theorem}
-
-The main theorem specializes to our case of interest
-
-\begin{corollary}
-    For every $\Gamma \vdash$ !e: \reals!, we have 
-    $\sem{\pi_2 \invUNF{\UNF{e}}(Id_\Gamma)(last(1:\reals))}= \nabla_{\Gamma}\sem{e}$.
-\end{corollary}
+ \subsection{Correctness optimisations} % (fold)
+ \label{sub:Correctness optimisations}

### 2021-06-29 rad-icfp-2021/purerad.tex
simple writing phase

@@
 We consider a standard call-by-value language. 
-It is a first-order language with array constructs and a few typical second-order array operations. 
-The types !T1,T2! and terms !e1,e2! are in Figure~\ref{fig:source_grammar}.
+It consist in a first-order functional language enriched with arrays and a few typical second-order array operations. 
+The types !T1,T2! and terms !e1,e2! are given in Figure~\ref{fig:source_grammar}.
 
@@
 The typing rules are in Figure~\ref{fig:source_typesystem}. We have included a minimal set of array operations for the sake of illustration, 
-but it is not difficult to add further operations. For scalar operations, we assume given a set of operations. 
-!op1! and !op2! denote respectively unary and binary smooth operations on reals. 
+but it is not difficult to add further operations. 
+For scalar operations, we assume given a set of operations, including $+$ and $*$. 
+!op1! and !op2! denote respectively a unary and a binary smooth operation on reals. 
+Typical examples include !cos, exp, +, *!. 
 We use infix notation for binary operators.
-Typical examples include !cos, exp, +, *!. 
-For the sake of simplicity in the development, the function in the array operations !map, map2, foldl! are quite restricted with and free variables. 
-We also restrict to arrays of reals.
-The general case is also useful and presented in Section~\ref{sec:generalisation}.
+For the sake of simplicity in the presentation, a bound function in an array operation !map, map2, foldl! is restricted to having no free variables.
+For the same reason, we restrict arrays of reals.
+The general cases are presented in Section~\ref{sec:generalisation}.
 
@@
 
-The last language we present, which we call the target language, is an extension to the source language. 
+The target language for differentiation is an extension to the source language, 
+which is higher-order as needed to carry the continuation used to have a pure reverse-mode transformation, 
+and contains more array primitives. In addition, we require the set of operations to be closed under partial differentiation.
+
 The types and terms are presented in Figure~\ref{fig:target_grammar}.
-The language contains lambda abstractions for dealing with continuations. 
 We are not interested here with partial applications and our language is pure, so functions take $n$ arguments.
-Lambda abstractions and applications are going to be removed during partial evaluation and this notation makes reading slightly easier.
+Lambda abstractions and applications will be removed during partial evaluation and this notation makes reading slightly easier.
 
@@
 
-In addition, we require the set of operations to be closed under partial differentiation. 
 In more detail, for every unary scalar operation !op1!, 
@@
 representing respectively the first and second partial derivative of !op2!.
-Similarly, we need to add some functions on arrays. 
+Similarly, some additional operations on arrays are needed. 
 !scanl! is a !foldl! that stores all intermediate results. 
@@
 We also include a !map3! operator. 
-Though not necessary as it can be simulated with two !map2!, it is a convenient notation and can somtimes run faster in practice.
+Though not necessary as it can be simulated with two !map2!, it is a convenient notation and can sometimes run faster in practice.
 Finally, we add a somewhat strange operator which we call !shift1!. 
@@
 
-The type system for the extended grammar of target is in Figure~\ref{fig:target_typesystem}.
+The type system for the extended grammar of target is presented in Figure~\ref{fig:target_typesystem}.
 
@@
 
-We now present our transformation from target UNF to target. 
-The transformation is presented in Figure~\ref{fig:unf_to_target}.
-
-\input{figures/unf_to_target}
-
 \subsection{Macro for pure reverse mode transformation} % (fold)
 \label{sub:Macro for pure reverse mode transformation}
+
+In Figure~\ref{fig:direct_diff_macro} we present the direct transformation for pure reverse mode differentiation.
+We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimise it further and compute the gradient.
+This transformation may be hard to read and prove correct. 
+In Section~\ref{sec:unf} we decompose this transformation into several easier 
+and more intuitive steps via a new intermediate representation which we call UNF.
+
+TODO: on the examples
+
+\input{figures/direct_macro_diff}
 
@@
 
-% subsectionPartial evaluation and optimisation (end)
+TODO: how to get gradient or subpart of the gradient from the previosu term.
+
+TODO: maybe intuition around here
+
+TODO: on the examples
+
+TODO: partial evaluation to get a tuple (i.e no lambdas)
+
+TODO: use the optimisations to be efficient, just on example?
+
+The full list of optimisations can be found on Figure~\ref{fig:optim}.

### 2021-06-29 rad-icfp-2021/generalisation.tex
writing optim table

@@
 
-\MH{I think I only present two: conditionals, and general case for array. partial op can be dealt with before. HO function mentioned at the end}
+% Extended source language:
 
-Extended source language:
-
-\input{figures/generalisation_source_grammar}
+% \input{figures/generalisation_source_grammar}
 
 % \input{figures/generalisation_source_typesystem}
+
+Extended target language:
+
+% \input{figures/generalisation_target_grammar}
+
+% \input{figures/generalisation_target_typesystem}
+
+\subsection{Arrays}
+
+\subsubsection{Lifting the restriction on map and fold}
+\label{ssub:Lifting the restriction on map and fold}
 
@@
 
-Extended target language:
-
-% \input{figures/generalisation_target_grammar}
-
-% \input{figures/generalisation_target_typesystem}
-
-\subsection{Arrays}
+\subsubsection{Lifting the restriction on the type of arrays}
+\label{ssub:Lifting the restriction on the type of arrays}
 
@@
 \subsection{Partial operations}
-
-\subsection{Higher-order functions}

### 2021-06-29 rad-icfp-2021/figures/notation_D.tex
writing direct D macro

Ground types are defined inductively by 
(TODO: chose whether we want n-ary, binary, or both kind of products)
$$G::= \RR \mid G\times \ldots \times G \mid A[G]$$

$(\RR,+,\underline{0})$ forms a monoid and this monoid structure extends canonically 
to a monoid structure $(G,\widehat{+},0_G)$ for every ground type $G$. 
It is defined inductively on $G$ as follows
\begin{itemize}
    \item $0_\RR \defeq \underline{0}$
    \item $0_{G_1\times G_2}\defeq < O_{G_1}, 0_{G_2} >$
    \item $0_{A[\RR]_n}\defeq$ ZerosLike($n$) TODO: general array? what about index n?
    \item $a\widehat{+}_\RR b\defeq a+b$
    \item $(a_1,\ldots,a_n)\widehat{+}_{G_1\times\ldots\times G_n}(b_1,\ldots,b_n)\defeq (a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n)$
    \item $A\widehat{+}_{A[\RR]}B \defeq$ !map! + $A$ $B$
\end{itemize}

A ground context is a context only containing variables of ground type.
The previous monoid structure again extends canonically on ground contexts $\Gamma$ by defining
$0_{x1:G1,\ldots,x_n:G_n}\defeq 0_{G_1},\ldots,0_{G_n}$ and 
$a_1,\ldots,a_n\widehat{+}_{x1:G_1,\ldots, x_n:G_n}b_1,\ldots,b_n\defeq a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n$.

\begin{notation}
    \begin{itemize}
        \item Given a context $\Gamma=x_1:A_1,\ldots,x_n:A_n$, $Id_\Gamma\defeq$ !fun! $(y_1,\ldots,y_n)$! -> !$(y_1,\ldots,y_n)$
        \item 
        \item If $\Gamma\vdash e:\RR$, $\grad_\Gamma e 
                    \defeq \directD{\Gamma}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,\underline{1})$
        \item If $\Gamma=\Gamma_1,\Gamma_2$ and $|\Gamma_1|=k$ and $\grad_{\Gamma_1,\Gamma_2}e=(d_1,\ldots,d_n)$, 
                then $\grad_{\Gamma_1}(e)\defeq(e_1,\ldots,e_k)$ and $\grad_{\Gamma_2}(e)\defeq(e_{k+1},\ldots,e_n)$
        \item Given a context $\Gamma=x_1:A_1,\ldots,x_n:A_n$ and a variable !x!, we define !pos x! to be its position $i$ in $\Gamma$, assuming all variables are distinct and that $x$ is indeed in $\Gamma$.
        \item Given $G_1,\ldots,G_n$ ground types, $1\leq i\leq n$ and a term !e! of type $G_i$, 
                $[i]e \defeq (0_{G_1},\ldots,0_{G_{i-1}},e,0_{G_{i+1}},\ldots,0_{G_n})$
    \end{itemize}
\end{notation}

### 2021-06-30 rad-icfp-2021/intro.tex
minors
writing appendix
simple writing phase
minor

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 
-The lack of purity of reverse-mode is known to make it significantly harder to optimise and find parrallelism. 
+The lack of purity of reverse-mode is known to make it significantly harder to optimise and find parallelism. 
 None of the current implementations of reverse-mode in a functional setting \cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.
@@
 It means that for every important non trivial operation, a hand-crafted reverse-derivative must be given to be efficient.
-In other words functional implementations seem to stuggle to abstract away from imperative code.
+In other words functional implementations seem to struggle to abstract away from imperative code.
 
@@
 % This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
+% An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and potentially reusable.
  An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and reusable.
@@
 
-\subsection{Smmary of contributions}
+\subsection{Summary of contributions}
 
@@
     \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
-    \item a new intermediate representation (I.R.) which we call unary form (UNF), essentially turning every operation to a unary one
+    \item a new intermediate representation (I.R.) which we call unary form(UNF), essentially turning every operation to a unary one
     \item a simple reverse-mode transformation on the I.R.
@@
 \end{center}
-where !reduce! is a known fold left operator for which the function argument is associative and is faster than a fold left.
+where !reduce! is a known fold left operator for which the function argument is associative. 
+It is notably faster to execute than a fold left as it parallel friendly.
 
@@
         $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
-        $\nabla_A$!dot(A,B)! &:=& !map2 * B (map (x -> 1) A)! 
+        $\nabla_A$!dot(A,B)! &:=& !B! 
     \end{tabular}

### 2021-07-01 rad-icfp-2021/background.tex
minor corrections intro

@@
 
-The two main ways in which it is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
-Our work focus on a compiled version so we focus on source code transformation which is more fitted in this case.
+The main two ways in which it is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
+Our approach focuses on source code transformation which is better fitted for compilation and optimizations.
 
@@
 
-Because the compute flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
-%%TODO: from Matthijs' paper, need to rewrite.
-Reverse-mode AD is only well-understood as a source-code transformation (also called define-then-run
-style AD) on limited programming languages. Typically, its implementations
-on more expressive languages that have features such as higher-order functions and conditionals
-make use of define-by-run approaches. These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
+Because the computation flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
+(TODO: paragraph below from Matthijs' paper, need to rewrite.)
+Reverse-mode AD is only well-understood as a source-code transformation 
+(also called define-then-run style AD) on limited programming languages. 
+Typically, its implementations on more expressive languages that have features such as higher-order functions and conditionals
+make use of define-by-run approaches. 
+These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
 first-order program is left, and then they evaluate this new program \cite{carpenter2015stan,paszke2017automatic}. 
-Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. As such, these AD libraries
-need to be implemented using carefully, manually optimized code, that for example does not contain any common subexpressions. This implementation process
-is precarious and labour intensive. Further, some whole-program optimizations
-that a compiler would detect go entirely unused in such systems.
+Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. 
+As such, these AD libraries need to be implemented using carefully, manually optimized code, that for example does not contain any common subexpressions. 
+This implementation process is precarious and labour intensive. 
+Further, some whole-program optimizations that a compiler would detect go entirely unused in such systems.
 
@@
 Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
-Instead of considering these as operations from $\RR\to\RR$, we consider then as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
-If we just do that though, we lose compositionality. 
+Instead of considering these as operations from $\RR\to\RR$, we consider them as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
+However, by simply doing this we lose compositionality. 
 So $cos$ will also return its context and is now seen as a function $\sem{cos}:\RR^n\to\RR^{n+1}$.
-similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
+Similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
 We thus obtain $\sem{exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $J\sem{cos} \in Mat_{n,n+1}$, $J\sem{exp} \in Mat_{n+1,n+2}$ also compose nicely.
@@
 This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ...\circ f_1$. 
-If the original program was of type $\RR$, then the return value of the original program is the last compononent of $f$.
+If the original program was of type $\RR$, then the return value of the original program is the last component of $f$.
 Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
@@
 
-This method is quite inefficient though. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
+This method is quite inefficient still. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
 Second, each $J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.
@@
 One of the key simple ideas that we used was to transform every operator into a unary one. 
-This essentially trivialises the compute flow to a line. 
+This essentially trivialises the computation flow to a line. 
 Even if the starting program was a straight line program, 
-having non unary operators was a source of inefficiency which justified mutation in the first place.
-By returning every variable every time, the problem of using a variable several time does not need to be dealt with via mutation. 
+having non-unary operators was a source of inefficiency which justified mutation in the first place.
+By returning every variable every time, the problem of using a variable several times does not need to be dealt with via mutation. 
 This simple idea of transforming a program into essentially a straight line is what our new intermediate representation unary-form does.
@@
 Even on the last row, if the original term was a unary or binary operator like $cos, exp, +, *$, 
-the row is zero except for at most 2 indices (1 for unary operators).
+the row is zero except for at most two indices (one for unary operators).
 This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,...,y(n+i)).(y1,...,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
@@
 
-We know in advance that such jacobian functions are going to be applied one to another, so we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
+We know in advance that such jacobian functions are going to be applied one to another and we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
 Because each function is almost the identity, we obtain a lot of substitutions of the form $[x/y]$ where both $x$ and $y$ are variables. 
-This allows us to drastically reduce the size of $J^Tf$. In fact, for simple programs this is basically enough to get an efficient purely functional reverse derivative transformation.
+This allows us to drastically reduce the size of $J^Tf$. In fact, for simple programs, this is basically enough to get an efficient purely functional reverse derivative transformation.
 We develop this idea further for a richer language.

### 2021-07-01 rad-icfp-2021/figures/compilation_scheme.tex
minor corrections intro

@@
 \begin{tikzcd}
-    \Source \ar[r] & \UNFSymbol_1 \ar[r,"\Dsynrevsymbol"] & \UNFSymbol_2 \ar[r] & \Target \ar[r,"\partial ev"] & \Target
+    \Source \ar[rrr,"\text{efficient }\Dsynrevsymbol (Fig.~\ref{fig:direct_diff_macro})"] \ar[dr,"(Fig.~\ref{fig:source_to_unf})"'] &&& \Target \arrow[r,loop right,"\text{optim }(Fig.~\ref{fig:optim})"] \\
+    & Source\UNFSymbol \ar[r,"\Dsynrevsymbol (Fig.~\ref{fig:diff_macro})"] & Target\UNFSymbol \ar[ur,"(Fig.~\ref{fig:unf_to_target})"'] & 
 \end{tikzcd}

### 2021-07-01 rad-icfp-2021/intro.tex
minor corrections intro

@@
 After some general partial evaluation techniques that will be detailed further in the paper, we obtain     
+        % \begin{center}
+        %     \begin{tabular}{l}
+        %         !let w1 = x1 * x2 in!\\ 
+        %         !let w2 = w1 * x1 in!\\
+        %         !let y1 = 0 in!\\
+        %         !let y2 = 0 in!\\
+        %         !let y3 = 0 in!\\
+        %         !let y4 = 0 in!\\
+        %         !let y5 = 1 in!\\
+        %         !let y1'! != y1+w1*y5 in!\\
+        %         !let y4'! != y4+x1*y5 in!\\
+        %         !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
+        %     \end{tabular}
+        % \end{center}
     \begin{center}
@@
                 !let w2 = w1 * x1 in!\\
-                !let y1 = 0 in!\\
-                !let y2 = 0 in!\\
-                !let y3 = 0 in!\\
-                !let y4 = 0 in!\\
-                !let y5 = 1 in!\\
+                !let y1,y2,y3,y4,y5 = 0,0,0,0,1 in!\\
                 !let y1'! != y1+w1*y5 in!\\
@@
 For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
-Of course, this term can be further optimised via constant propagation and algebraic simplifications to give
+This term can be further optimised via constant propagation and algebraic simplifications to give
         \begin{center}
@@
 where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
-!scanl! and !scanr! are respectively scan left and scan right. These are like fold but store all intermediate results in an array.
-We obtain these gradients, which are efficient and usually derived by hand, as special cases of our work.
-They are obtained automatically in our case.
+!scanl! and !scanr! are respectively scan left and scan right. These return all the intermediate result of a fold in an array.
+These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
 \end{example}
@@
 
-Our work looks like the following compilation scheme
+Our main transformation is present on the top of the diagram and presented in Section~\ref{sec:simplediff}. 
+It it decomposed 3 steps which are presented in Section~\ref{sec:unf}.
 \input{figures/compilation_scheme}
 
-In Section~\ref{sec:background} we recall rudiments of automatic differentiation, and in particular the idea of reverse-mode and the cheap gradient principle.
-Next, in Section~\ref{sec:simplediff} we introduce the source language we want to differentiate.
-We also introduce our new intermediate representation UNF, and a simple reverse-mode macro on this representation.
-In Section~\ref{sec:efficientrad} we translate back from UNF to a purely functional target language and show several general optimisations to make our reverse-mode transformation efficient.
-In the next Section~\ref{sec:complexity} we show that the previous optimisations are sufficient to ensure our transformation has the right complexity.
-Section~\ref{sec:correctness} shows that our construction is correct and we give standard semantics to our source and target languages. 
-In Section~\ref{sec:generalisation} we show how to generalise the previous sections for a richer source language. 
-Finally, related work and some limitations to our approach are presentend in Section~\ref{sec:conclusion}.
+In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
+Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
+In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
+We introduce transformations from the source language to UNF and from UNF to the target language.
+Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
+In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimisations.
+In Section~\ref{sec:generalisation} we show how to adapt our work for a richer source language. 
+Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

### 2021-07-01 rad-icfp-2021/relwork.tex
minor corrections intro

@@
 
-We have shown how to compute provably efficient gradients via reverse derivatives in a purely functional way 
-on a simple yet expressive language with array functions.
-We introduced a novel intermediate representation, unary form, 
-and used it combined with standard functional optimisations to perform our reverse-mode transformation.
+We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
+on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), combined with standard functional optimisations (\ref{fig:optim}).  
+We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
+to decompose our transformation into simpler and more insightful transformations.
+We showed (\ref{sec:generalisation}) how to lift some of the restrictions that
+we introduced on arrays and how to extend our approach.
 

### 2021-07-02 rad-icfp-2021/purerad.tex
writing direct D macro
redoing unf figures
minor corrections intro
rewrote notation

@@
 \subsection{Source Language}
+
+\label{sub:sourcelang}
 
@@
 
+We introduce several notations which are useful when defining the transformation for reverse-mode.
+
+\input{figures/notation_D}
+
 \input{figures/direct_macro_diff}
@@
 
-TODO: how to get gradient or subpart of the gradient from the previosu term.
-
-TODO: maybe intuition around here
+Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
+$\directD{\rho}{\Gamma}{Y}(e)$. First, $\rho, Y$ specifies if we want to compute the whole gradient regarding the variables from $\Gamma$ or a subset of it.
+For a subset $\rho\subset \Gamma$, one chooses $Y$ to be the projection function sending a variable 
+$x_i:G$ of $\Gamma$ to $x_i$ if it belongs to $\rho$ and to $0_G$ otherwise.
+In particular, we take $Y=Id_\Gamma$ to compute the whole gradient.
+Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
+and we need to initialise the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
+which we initialise at $1$ to run the backpropagation. 
+All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.
 
@@
 
-TODO: use the optimisations to be efficient, just on example?
+Though this transformation already has the right complexity, 
+its purity allows us to take the most out of generic optimisations.  
+A list of common optimisations that are useful for this language can be found in Figure~\ref{fig:optim}.
 
-The full list of optimisations can be found on Figure~\ref{fig:optim}.
+TODO: use the optimisations to be efficient on example

### 2021-07-02 rad-icfp-2021/figures/compilation_scheme.tex
minor

--- 2021-02-23 rad-icfp-2021/figures/compilation_scheme.tex
@@
 \begin{tikzcd}
-    \Source \ar[r] & \UNFSymbol_1 \ar[r,"\Dsynrevsymbol"] & \UNFSymbol_2 \ar[r] & \Target \ar[r,"\partial ev"] & \Target
+    \Source \ar[rrrr,"\text{efficient }\Dsynrevsymbol (Fig.~\ref{fig:direct_diff_macro})"] \ar[dr,"(Fig.~\ref{fig:source_to_unf})"'] &&&& \Target \arrow[r,loop right,"\text{optim }(Fig.~\ref{fig:optim})"] \\
+    & Source\UNFSymbol \ar[rr,"\Dsynrevsymbol (Fig.~\ref{fig:diff_macro})"] && Target\UNFSymbol \ar[ur,"(Fig.~\ref{fig:unf_to_target})"'] & 
 \end{tikzcd}

### 2021-07-02 rad-icfp-2021/figures/notation_D.tex
rewrote notation
minor

@@
 Ground types are defined inductively by 
-(TODO: chose whether we want n-ary, binary, or both kind of products)
 $$G::= \RR \mid G\times \ldots \times G \mid A[G]$$
@@
     \item $0_\RR \defeq \underline{0}$
-    \item $0_{G_1\times G_2}\defeq < O_{G_1}, 0_{G_2} >$
+    \item $0_{G_1\times \ldots \times G_n}\defeq < O_{G_1},\ldots, 0_{G_n} >$
     \item $0_{A[\RR]_n}\defeq$ ZerosLike($n$) TODO: general array? what about index n?
@@
 
-\begin{notation}
-    \begin{itemize}
-        \item Given a context $\Gamma=x_1:A_1,\ldots,x_n:A_n$, $Id_\Gamma\defeq$ !fun! $(y_1,\ldots,y_n)$! -> !$(y_1,\ldots,y_n)$
-        \item 
-        \item If $\Gamma\vdash e:\RR$, $\grad_\Gamma e 
-                    \defeq \directD{\Gamma}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,\underline{1})$
-        \item If $\Gamma=\Gamma_1,\Gamma_2$ and $|\Gamma_1|=k$ and $\grad_{\Gamma_1,\Gamma_2}e=(d_1,\ldots,d_n)$, 
-                then $\grad_{\Gamma_1}(e)\defeq(e_1,\ldots,e_k)$ and $\grad_{\Gamma_2}(e)\defeq(e_{k+1},\ldots,e_n)$
-        \item Given a context $\Gamma=x_1:A_1,\ldots,x_n:A_n$ and a variable !x!, we define !pos x!
-to be its position $i$ in $\Gamma$, assuming all variables are distinct and that $x$ is indeed in $\Gamma$.
-        \item Given $G_1,\ldots,G_n$ ground types, $1\leq i\leq n$ and a term !e! of type $G_i$, 
-                $[i]e \defeq (0_{G_1},\ldots,0_{G_{i-1}},e,0_{G_{i+1}},\ldots,0_{G_n})$
-    \end{itemize}
-\end{notation}
+\input{figures/notation}

### 2021-07-02 rad-icfp-2021/purerad.tex
minor

@@
 
-TODO: partial evaluation to get a tuple (i.e no lambdas)
+Following the insight from Section~\ref{subsec:insights}, 
+we don't want to keep all the lambda abstractions as this is costly. 
+Fortunately, the transformation is designed in such a way that we only have applications 
+and we can use partial evaluation to do beta redudction.
 
-Though this transformation already has the right complexity, 
+By inspection in Figure~\ref{fig:direct_diff_macro}, 
+we see that there is a linear usage of  each continuation $Y$ 
+and that it is always applied to almost the identity. 
+More precisely, we have applications of the form !fun (x1,...,xn) -> Y(e1,...,en)! 
+where the !ei! are !xi! except for at most $k$ (independant of $n$) terms.
+In fact we have $k=2$, except for the !map2! case which requires a more detailed analysis (see Section~\ref{sec:complexity}).
+
+Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
+!fun (x1,...,xn) -> (fun (y1,...,yn) -> (f1,...,fn))(e1,...,en)! 
+$\transto$ 
+!fun (x1,...,xn) -> let y1,...,yn = e1,...,en in (f1,...,fn)!
+$\transto$
+!fun (x1,...,xn) -> (f1[e1/y1],...,fn[en/yn])!
+Now, importantly, every !yi! is used at most once (except for !map2! again and !reduce! which require a more detailed analysis)
+which garantees that this will not blow up the complexity.
+
+TODO: finish explanation, not sure how to conclude at the moment
+
+Though this transformation after the inlining above already has the right complexity, 
 its purity allows us to take the most out of generic optimisations.  
@@
 
-TODO: use the optimisations to be efficient on example
+TODO: use the optimisations to be efficient on examples

### 2021-07-02 rad-icfp-2021/appendix1.tex
writing figures
simple lemma appendix

@@
 In Figure~\ref{fig:op_semantics_target} a small step call-by-value operational semantics for the language. 
-\MH{need to add evaluation contexts}
+The evaluation contexts aregiven in Figure~\ref{fig:ev_contexts}.
+
+\input{figures/ev_contexts}
 
@@
 
+\begin{lemma}
+    if !op2! is an associative binary operation with unit $\Gamma \vdash$ !v:! $\reals$, then 
+    $\frac{\partial op2(v,e)}{\partial y_1}\times\frac{\partial v}{\partial x_i}=0$ 
+    and $\frac{\partial op2(e,v)}{\partial y_2}\times\frac{\partial v}{\partial x_i}=0$ for all $x_i$.
+\end{lemma}
+
+\begin{proof}
+    For any $\Gamma \vdash$ !e:! $\reals$, we have !v op2 e!=!e!.
+    Differentiating and using the chain rule we get 
+    $$\frac{\partial op2(v,e)}{\partial y_1}\times\frac{\partial v}{\partial x_i}
+    +\frac{\partial op2(v,e)}{\partial y_2}\times\frac{\partial e}{\partial x_i}
+    = \frac{\partial e}{\partial x_i}$$
+As $\frac{\partial e}{\partial x_i}$ is arbitrary, 
+this shows that $\frac{\partial op2(v,e)}{\partial y_2}=1$ and $\frac{\partial op2(v,e)}{\partial y_1}\times\frac{\partial v}{\partial x_i}=0$.
+Similarly for the other case.
+\end{proof}
+
 \begin{proposition}
-    The reverse derivative of !foldl! is correct.
+    The reverse derivative of !reduce! is correct.
 \end{proposition}

### 2021-07-02 rad-icfp-2021/figures/macro_diff.tex
part of the operational semantics target
some corrections from Sam,Amir comments
rewrote notation
changing UNF to make it correct
still fixing UNF

@@
     \begin{tabular}{r c l}
-    $\Dsynrevsymbol$(!var!$_{\Gamma,i}$) &=& !<var!$_{\Gamma,i}$!, J!$^T$!var!$_{\Gamma,i}$!>! \\
-    $\Dsynrevsymbol$(!op!$_{\Gamma,n}$) &=& !<op!$_{\Gamma,n}$!, J!$^T$!op!$_{\Gamma,n}$!>! \\ 
-    $\Dsynrevsymbol$(!e1;e2!) &=& $\Dsynrevsymbol$(!e1!);$\Dsynrevsymbol$(!e2!)\\ 
-    $\Dsynrevsymbol$(!map!$_{x.e}$) &=& !<map!$_{x.e}$!, J!$^T$!map!$_{x.e}$!>! \\ 
-    $\Dsynrevsymbol$(!map2!$_{x,y.e}$) &=& !<map2!$_{x,y.e}$!, J!$^T$!map2!$_{x,y.e}$!>! \\ 
-    $\Dsynrevsymbol$(!foldl!$_{x,y.e}$) &=& !<foldl!$_{x,y.e}$!, J!$^T$!foldl!$_{x,y.e}$!>! \\  
+        $\Dsynrevsymbol_{\rho}$(!A1,...,An!) &=& !A1,...,An,(A1x...xAn)->!$\rho$
+    \end{tabular}
+
+    \medskip
+    \begin{tabular}{r c l}
+    $\Dsynrevsymbol_{\rho}$(!var!$_{T;i}$) &=& !<proj!$_{T;\underline{T}->\rho;[]} \comp$ !var!$_{T;i}$!, proj!$_{[];T;\underline{T}->\rho} \icomp$ (!proj!$_{T;\underline{T}->\rho;[]}$ $\comp$ !J!$^T$!var!$_{T;i}$)!>!
+\\
+    $\Dsynrevsymbol_{\rho}$(!op!$_{\Gamma;n}$) &=& !<proj!$_{T;\underline{T}->\rho;[]} \comp$ !op!$_{T;n}$!, proj!$_{[];T;\underline{T}->\rho} \icomp$ (!proj!$_{T;\underline{T}->\rho;[]}$ $\comp$ !J!$^T$!op!$_{T;,n}$)!>!
+\\
+    $\Dsynrevsymbol_{\rho}$(!pair!$_{T;A\times B}$) &=& !<proj!$_{T;\underline{T}->\rho;[]} \comp$ !pair!$_{T;A\times B}$!, proj!$_{[];T;\underline{T}->\rho} \icomp$ (!proj!$_{T;\underline{T}->\rho;[]}$ $\comp$ !J!$^T$!pair!$_{T;A\times B}$)!>!
+\\
+    $\Dsynrevsymbol_{\rho}$(!proj!$_{T1;T2;T3}$) &=& !<proj!$_{T;\underline{T}->\rho;[]} \comp$ !proj!$_{T1;T2;T3}$!, proj!$_{[];T;\underline{T}->\rho} \icomp$ (!proj!$_{T;\underline{T}->\rho;[]}$ $\comp$ !J!$^T$!proj!$_{T1;T2;T3}$)!>!
+\\
+    $\Dsynrevsymbol_{\rho}$(!e1!$\comp$!e2!) &=& $\Dsynrevsymbol_{\rho}$(!e1!); $\Dsynrevsymbol_{\rho}$(!e2!)\\ 
+    $\Dsynrevsymbol_{\rho}$(!e1!$\pcomp$!e2!) &=& TODO \\
+    $\Dsynrevsymbol_{\rho}$(!map2!$_{T;x,y.e}$) &=& !<proj!$_{T;\underline{T}->\rho;[]} \comp$ !map2!$_{T;x,y.e}$, !proj!$_{[];T;\underline{T}->\rho} \icomp$ (!proj!$_{T;\underline{T}->\rho;[]}$ $\comp$ !J!$^T$!map2!$_{T;x,y.e}$)!>!
+\\
+    $\Dsynrevsymbol_{\rho}$(!reduce!$_{T;x,y.e1;e2}$) &=& !<proj!$_{T;\underline{T}->\rho;[]} \comp$ !reduce!$_{T;x,y.e1;e2}$, !proj!$_{[];T;\underline{T}->\rho} \icomp$ (!proj!$_{T;\underline{T}->\rho;[]}$ $\comp$ !J!$^T$!reduce!$_{T;x,y.e1;e2}$)!>!
+\\
     \end{tabular}
     \caption{Reverse-mode differentiation from Source UNF to Target UNF}
-    \label{fig:diff_macro}    \end{figure*}
+    \label{fig:diff_macro}    
+\end{figure*}

### 2021-07-03 rad-icfp-2021/generalisation.tex
writing figures
writing a bit generalisation section

--- 2021-02-25 rad-icfp-2021/generalisation.tex
@@
 
-\MH{I think I only present two: conditionals, and general case for array. partial op can be dealt with before. HO function mentioned at the end}
+\subsection{Lifting the restriction on map and fold}
+\label{sub:Lifting the restriction on map and fold}
 
-Extended source language:
-
-\input{figures/generalisation_source_grammar}
-
-% \input{figures/generalisation_source_typesystem}
+TODO: not sure about this, and my derivative for reduce/foldl is currently a bit wrong
 
@@
 
-Extended target language:
+\subsection{Adding more array operators}
+\label{sub:Adding more array operators}
 
-% \input{figures/generalisation_target_grammar}
+\subsection{Adding more scalar operators} % (fold)
+\label{sub:Adding more scalar operators}
 
-% \input{figures/generalisation_target_typesystem}
+We assumed the unary and binary operators were denoted by smooth functions $\RR^n\to\RR$. 
+There is no additional difficulty in considering operators which are partial functions 
+like division or operators which are not smooth at a point like square root.
 
-\subsection{Arrays}
+These functions are then given intentional derivatives which provide valid derivatives 
+on the domain of definition and differentiability of the operator. 
+These functions are well known to be the bete noire of AD \cite{griewank2008evaluating} 
+and we do not provide novel solutions to these.  
+Several recent work have shown how to give semantics to such operators in the context of AD \cite{vakar2020denotational,mazza2021automatic,sherman2021,lee2020correctness}.
 
-\subsection{Conditionals}
+\subsection{Adding conditionals} % (fold)
+\label{sub:Adding conditionals}
 
-\subsection{Partial operations}
+Adding conditionals to the source language can be done easily, for instance via defining
 
-\subsection{Higher-order functions}
+\begin{tabular}{r c l}
+$\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) &=& !if e1 then !$\directD{\rho}{\Gamma}{Y}$!(e2) else !$\directD{\rho}{\Gamma}{Y}$!(e3)! 
+\end{tabular}
+
+The problem is that this could break the complexity of reverse mode because of the non linear usage of $Y$, and makes everything harder to optimise.
+
+A slighly better option would be to define
+
+\begin{tabular}{r c l}
+    $\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) 
+    &=& !let b=e1 in!   \\
+    && !<if b then e2 else e3, fun (x1,...,xn,z) ->! \\
+    && !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e2)(x1,...,xn,z)+!\\
+    && \quad!(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e3)(x1,...,xn,z))>!
+\end{tabular}
+
+It is slighly better because both derivatives of !e2! and !e3! are put together which might unlock some optimisations, but there is still a non linear usage of the variables.
+One solution, if we know we want to compute the whole gradient of the expression, is to define instead
+
+\begin{tabular}{r c l}
+    $\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) 
+    &=& !let b=e1 in!   \\
+    && !<if b then e2 else e3, fun (x1,...,xn,z) ->! \\
+    && !Y((x1,...,xn)!$\widehat{+}$($\grad_\Gamma$!(e2)*b!$\widehat{+}\grad_\Gamma$!(e3)*(1-b))*z)>!\\
+\end{tabular}
+
+This time, there is linear usage of !Y,z! and the !xi!.
+
+TODO: explain why it's correct?\\
+TODO: adding conditionals does not break differentiability as long as no non smooth function like max(0,-):R->B\\
+TODO: re explain how it's usually done by partial eval of the conditional first.
+
+\subsection{Adding general arrays} % (fold)
+\label{sub:Adding general arrays}
+
+We now show how to consider arrays over any ground type $G$.
+
+TODO: introduce sum and tensor of semantics of ground type\\
+TODO: then * becomes a dot product\\
+TODO: minor changes from there\\
+TODO: except I'm not sure with reduce and shift1...
+
+Though this has the right complexity, it is an interesting avenue for future research to find
+better representations to allow for more optimisations. 
+In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
+and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.
+
+TODO: more generally about tensors \cite{liao2019differentiable,bernstein2020differentiating}

### 2021-07-03 rad-icfp-2021/correctness.tex
outline correctness section

--- 2021-02-14 rad-icfp-2021/correctness.tex
@@
 \section{Correctness}
+\label{sec:correctness}
 
-%decompose the macro from section 4 into unary form transfo, then principled macro, then opti
-\MH{Question: when is it best to give the idea about our macro? that it's UNF+macro+easy-opti?}
-\MH{probably won't use UNF in this paper}
+This section provides a categorical analysis of our language and transformations.
+The end goal is to prove the correctness of the reverse mode transformation from Section~\ref{sub:Macro for pure reverse mode transformation}.
 
-\subsection{Unary Form} % (fold)
-\label{sub:unary_form}
+ \begin{theorem}
+     For every $\Gamma \vdash$ !e:!$\reals$, we have 
+     $\sem{\grad_\Gamma e}= \grad_\Gamma\sem{e}$.
+ \end{theorem}
+where $\sem{-}$ is ab appropriate denotational semantics of smooth functions.
 
-\subsection{Principled Macro} % (fold)
-\label{sub:principled_macro}
+\subsection{Denotational semantics source and Target} % (fold)
+\label{sub:Denotational semantics source and Target}
 
-\subsection{Correctness via subsconing} % (fold)
-\label{sub:correctness_via_subsconing}
+\begin{definition}{multicategory}
+    TODO
+\end{definition}
 
-\subsection{Optimizations} % (fold)
-\label{sub:optimizations}
-%the ones in the macro of Sec4 then the further classical ones to ANF
+\begin{definition}{free multicategory}
+    TODO
+\end{definition}
 
-\subsection{Correctness of the macro from section \ref{}} % (fold)
-\label{sub:correctness_of_the_macro_from_section_ref}
+\begin{example}{Source and Target as syntactic multicategories}
+    TODO
+\end{example}
+
+\begin{definition}{CartSp}
+    TODO
+\end{definition}
+
+TODO: semantics of Source in CartSp, what about semantics Target?
+
+\subsection{Semantics for UNF using concategories} % (fold)
+\label{sub:Semantics for UNF using concategories}
+
+\begin{definition}{concategory}
+    TODO
+\end{definition}
+
+\begin{definition}{free concategory}
+    TODO
+\end{definition}
+
+\begin{example}{Concat1, Concat2}
+    TODO
+\end{example}
+
+TODO: semantics of UN1, UNF2 inside Concat1, Concat2
+
+\subsection{Semantics for UNF transformations} % (fold)
+\label{sub:Semantics for UNF transformations}
+
+\begin{definition}{Mutlicat from concat}
+    TODO
+\end{definition}
+
+\begin{proposition}{construction above gives UNF}
+    TODO
+\end{proposition}
+
+\begin{definition}{lax functor between concat}
+    TODO
+\end{definition}
+
+\begin{example}{D as a lax functor}
+    TODO
+\end{example}
+
+\begin{proposition}{semantic of D syntactic matches D lax functor}
+    TODO
+\end{proposition}
+
+\begin{definition}{Concat from multicat}
+    TODO
+\end{definition}
+
+\begin{proposition}{Initiality of concat2 gives $UNF^-1$}
+    TODO
+\end{proposition}
+
+\subsection{Correctness theorem} % (fold)
+\label{sub:Correctness theorem}
+
+\begin{proposition}{The transpose Jacobian of !map2! and !reduce! are correct.}
+    TODO
+\end{proposition}
+
+\begin{theorem}{UNF;D;$UNF^-1$ correct}
+    TODO
+\end{theorem}
+
+\begin{proposition}{UNF;D;$UNF^-1$=D}
+    TODO
+\end{proposition}
+
+\begin{theorem}{D is correct}
+    TODO
+\end{theorem}

### 2021-07-03 rad-icfp-2021/relwork.tex
some refs and related work

@@
 
-We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
-on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), combined with standard functional optimisations (\ref{fig:optim}).  
+We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) 
+gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
+on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), 
+combined with standard functional optimisations (\ref{fig:optim}).  
 We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
-to decompose our transformation into simpler and more insightful transformations.
+to decompose our transformation into simpler transformations.
+We gave denotational semantics to our languages and our transformations (\ref{sec:correctness}) 
+and showed correctness of the reverse-mode transformation (\ref{sub:Correctness theorem}).
 We showed (\ref{sec:generalisation}) how to lift some of the restrictions that
-we introduced on arrays and how to extend our approach.
+we introduced on arrays and how to extend our approach to a richer language with more primitive operations and conditionals.
 
@@
 \noindent \textbf{Correctness of AD in functional languages.}
+Several recent work \cite{huot2020correctness,vakar2020reverse,vakar2020denotational,brunel2019backpropagation,barthe2020versatility,mazza2021automatic} have focused on correctness of AD in a purely functional setting,
+often leaving efficiency on the side, especially for reverse-mode differentiation. 
+We see our work as a complement and a first bridge between these works 
+and more practical considerations of efficiency and implementations, 
+which often require a lot more care than is acknowledegd in more theoretical works.
+
+An exception is \cite{abadi-plotkin2020} which gives a reverse-mode transformation that is proved correct and with the right complexity.
+This work is purely first-order, does not have arrays and uses a non standard semantics. 
+Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.
+They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
+Instead, they follow the principle of unrolling for recursive calls and evaluating the conditional before differentiating, 
+which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
+This is a well-known limit in TensorFlow which TeaserFlow EagerMode tries to bypass.
 
@@
 \noindent \textbf{More dynamic language}
-%TF is losing ground to Pytorch, Swift tries to renew with Pytorch-like dynamic features.
+JAX \cite{bradbury2020jax,frostig2018compiling} is a modern framework that provides state-of-the-art automatic differentiation. 
+One feature that makes it popular is that it uses JIT compilation, 
+which seems to be the best avenue to help best ML practionners as it allows them 
+to quickly experiment but also benefit from the full power of static optimisations.
+We are interested in seeing how our work could sit in a JIT compiler.
 
 \noindent \textbf{More optimisations.}
-%maybe?
+As pointed in \ref{sub:Adding general arrays} it would be interesting to have a more principled representation 
+of tensors which could lead to more optimisations. 
+There is always a tension between dynamicity and the possibility for static optimisations 
+and we believe our language offers a good tradeoff, but it will be interesting to tailor it 
+further for specific applications.
 
-\noindent \textbf{Limitations.}
+\noindent \textbf{Limitations and future work}
 \begin{itemize}

### 2021-07-03 rad-icfp-2021/relwork.tex
more related work

@@
 
-Some work in ML where some kind of iteration/recursion is used.
-Dynamic neural networks: \cite{jin2017manipulability,wu2016deep}
-Recursive NN: \cite{socher2011parsing,biancofiore2017recursive}
-Reccurent NN: \cite{bahdanau2014neural,luong2015effective}
-Tree LSTM: \cite{tai2015improved,chen2016enhanced}
-Dynamic Recursive NN: \cite{guo2019dynamic}. It's recent with very few citations, but their point is close to ours. 
-Top-down Tree LSTM: \cite{zhang2015top}
-Recursion in DNN: \cite{jeong2018improving}
-
-\noindent \textbf{Automatic Differentiation.}
-
-\noindent \textbf{Array Languages and Fusion.}
-
-\noindent \textbf{Numerical DSLs.}
+\noindent \textbf{Usage of iteration mechanics}
+A immense effort in machine learning for the past decade has been in finding
+good architectures, to limit computational costs, 
+avoid vanishing and exploding gradients, 
+and have better building blocks of large complicated systems than traditional layers of a neural network.
+Different approaches 
+(Dynamic neural networks \cite{jin2017manipulability,wu2016deep}, 
+Recursive NN \cite{socher2011parsing,biancofiore2017recursive}, 
+Reccurent NN \cite{bahdanau2014neural,luong2015effective}, 
+Tree LSTM \cite{tai2015improved,chen2016enhanced}, 
+Dynamic Recursive NN \cite{guo2019dynamic}, 
+Top-down Tree LSTM \cite{zhang2015top}, 
+Recursion in DNN \cite{jeong2018improving}) 
+have found that recursive data structures such as trees are good candidates.
+We have emphasized here on differentiating fold left on arrays for efficiency, 
+but one should be able to adapt this on any algebraic data type. 
+It will be interesting to see if and how we recover efficient backpropagation on the proposed architectures 
+which is usually derived by hand and one main goal of these papers.
 
 \noindent \textbf{Correctness of AD in functional languages.}
-Several recent work \cite{huot2020correctness,vakar2020reverse,vakar2020denotational,brunel2019backpropagation,barthe2020versatility,mazza2021automatic} have focused on correctness of AD in a purely functional setting,
+Several recent work \cite{huot2020correctness,vakar2020reverse,vakar2020denotational,brunel2019backpropagation,barthe2020versatility,mazza2021automatic,ee2020correctness} have focused on correctness of AD in a purely functional setting,
 often leaving efficiency on the side, especially for reverse-mode differentiation. 
@@
 which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
-This is a well-known limit in TensorFlow which TeaserFlow EagerMode tries to bypass.
+This is a well-known limit in TensorFlow \cite{abadi2016tensorflow} which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.
+
+\noindent \textbf{Array Languages and AD}
+
+Given the enormous computation needs for state-of-the-art large scale machine learning applications, 
+which require extremely efficient tensor computations and automatic differentiation for backpropagation, 
+combining array languages and automatic differentiation (tensor calculus) in the best 
+fitted intermediate representation for optimisations is of key interest, and active research.
+\cite{bernstein2020differentiating} 
+\cite{laue2018computing,laue2020simple}
 
 \noindent \textbf{Comparison to other recent papers.}
+
+\cite{ee2020correctness}: non diff function
+\cite{sherman2021}: non diff HO functions
+\cite{pearlmutter2008reverse}: based on this
+\cite{elliott2018simple}: bit similar because using CT as useful language, but we take impl and opti more seriously.
+\cite{shaikhha2019efficient}: similar approach but here more complicated for reversemode, would be interesting to investigate more loop fusions here.
 

### 2021-07-03 rad-icfp-2021/relwork.tex
bit more related work

@@
 \label{sub:related_work}
-
-Relatively recent work in AD (2008-2020):
-\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,wang2018demystifying,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}
 
@@
 
+Relatively recent work in AD (2008-2020):
+\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,wang2018demystifying,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}
+
+Good recent surveys: \cite{van2018automatic,baydin2017automatic}
+
 \subsection{Discussion and future work} % (fold)
@@
 
-\noindent \textbf{Limitations and future work}
+\noindent \textbf{Cost function and memory accesses}
+Our cost model \ref{sub:costModel} is simple but has limits. 
+In particular, it does not capture well cost in term of
+parallel computing, and memory management (cache optimisation is one of best way to optimise such linear algebra intensive computations).
+For instance, in the earlier days of automatic differentiation, 
+advanced checkpointing techniques were used to avoid storing all the intermediate values just to save memory 
+and try avoiding saturation of RAM. 
+These problems are harder to analyse and eventually it seems to come down to a lot of experimenting and benchmarking.
+
+\noindent \textbf{Parallelism and GPU}
+We tried to locally preserve parallelism as much as possible, 
+and having a pure transformation should help with found parallelism. 
+It would be interesting to implement on transformations and see how they perform on GPU.
+It is well-kwown that the addition of references provoked by the standard reverse-mode transformation
+restricts a lot of optimisations and it is not hard to find examples in C where the compiler does not detect simple optimisations 
+because of the references introduced by reverse-mode.
+
+\noindent \textbf{implementation, BLaS, benchmark}
+
+\noindent \textbf{Higher-order functions}
+Matthijs'work, use of defunctionalization, question efficiency and usefulness, 
+interested in functions like root of function or minimal of a function. Implicit function theorem
+
+\noindent \textbf{More array primitives}
+filter, flatten, gather, scanl, ...
+can be derived by hand once and then it's ok
+
+\noindent \textbf{Recursion}
+
+\noindent \textbf{Higher derivatives}
+
+\noindent \textbf{Sparse data and matrix representation}
+
 \begin{itemize}
-    \item memory access for cost function 
-    \item parallelism
-    \item dedicated tensors for ML, and BLaS considerations
-    \item implementation and benchmark
-    \item conditionals
-    \item recursion
-    \item general HO function
-    \item more 2nd order functions, though can be derived by hand once and then it's ok
-    \item more dynamic arrays
-    \item more dynamic control flow
-    \item all store vs all recompute reverse-mode, and advanced checkpointing
-    \item numerical accuracy
     \item fancier types and datatypes (eg dependent, adt) 
-    \item iterated ad and higher derivatives
     \item use in Probabilistic programming
-    \item non smooth and partial functions, although by simply following the ref book it's as usual in AD
-    \item sparsity and data manipulation
     \item other optimisations like implicit function theorem
@@
 
-\MH{useful links (Swift is close to our project, and the doc is very readable):}
-\begin{verbatim}https://docs.google.com/document/d/1_BirmTqdotglwNTOcYAW-ib6mx_jl-gH9Dbg4WmHZh0/edit# \end{verbatim}
-\begin{verbatim}
-	https://github.com/apple/swift/blob/master/docs/DifferentiableProgramming.md#approaches-to-automatic-differentiation
-\end{verbatim}
+% \MH{useful links (Swift is close to our project, and the doc is very readable):}
+% \begin{verbatim}https://docs.google.com/document/d/1_BirmTqdotglwNTOcYAW-ib6mx_jl-gH9Dbg4WmHZh0/edit# \end{verbatim}
+% \begin{verbatim}
+% 	https://github.com/apple/swift/blob/master/docs/DifferentiableProgramming.md#approaches-to-automatic-differentiation
+% \end{verbatim}
 
-\MH{understand related work sentence from Diff.
-Curry paper: The idea of using closures as back-propagators is receiving recent attention.
-For example Julia Zy- gote [15] and Swift AD adopts this design.
+\MH{understand related work sentence from Diff. Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
+    For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
     Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.}

### 2021-07-03 rad-icfp-2021/correctness.tex

@@
 This section provides a categorical analysis of our language and transformations.
-The end goal is to prove the correctness of the reverse mode transformation from Section~\ref{sub:Macro for pure reverse mode transformation}.
+The end result is the correctness of the reverse mode transformation from Section~\ref{sub:Macro for pure reverse mode transformation}:
 
@@
 
-\begin{definition}{multicategory}
-    TODO
+It is standard to give a denotational semantics of a first-order language in a Cartesian category.
+An alternative approach consists in giving a semantics in a multicategory.
+A multicategory generalises a category by allowing multimorphims, that is morphisms from a list of objects to an object.
+Most categorical structures from category theory can be phrases similarly in multicategories.
+
+In terms of semantics, a term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$ in a multicategory.
+Substitution is still interpreted as composition.
+
+One can consider a syntactic model for a language, which consists in a free multicategory.
+Our source and target languages induce syntactic multicategories as follows.
+
+\begin{definition}{Syntactic multicategory for Source}
+    Let $\SynSource$ be the multicategory whose objects are types of Source, and where a morphism 
+    $[A_1,\ldots,A_n]\to A$ is a term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ of Source modulo the $\eta\beta$-laws.
+    Composition is by substitution.
 \end{definition}
 
-\begin{definition}{free multicategory}
-    TODO
+We similarly define $\SynTarget$, the syntactic multicategory for Target.
+
+As is standard, $\SynSource$ satisfies the following universal property: 
+for every Cartesian multicategory $\catC$ with arrays map and reduce,
+and every object $F(\reals)\in\catC$ and morphisms $F(\underline{c})\in\catC(1;F(\reals))$, 
+$F(op1)\in\catC(F(\reals);F(\reals))$, $F(op2)\in\catC(F(\reals),F(\reals);F(\reals))$, there is a unique
+multifunctor $F:\SynSource\to\catC$ respecting the interpretation and preserving all the structure.
+
+We can for instance use this to give a simple semantics of Source 
+in the multicategory of Cartesian spaces and smooth maps between them.
+
+\begin{definition}{CartSp}
+    TODO, its definition and semantics of Source in it.
 \end{definition}
 
-\begin{example}{Source and Target as syntactic multicategories}
-    TODO
-\end{example}
-
-\begin{definition}{CartSp}
-    TODO
-\end{definition}
-
-TODO: semantics of Source in CartSp, what about semantics Target?
+TODO: what about semantics Target?
 

### 2021-07-03 rad-icfp-2021/abstract.tex

@@
     tensors and operations on tensors.
-    Reverse-mode differentiation is used for optimisation, 
-    but it doesn't preserve purity of the underlying program and makes it notoriously harder to optimise.
+    Reverse-mode differentiation is used for optimization, 
+    but it does not preserve purity of the underlying program and makes it notoriously harder to optimize.
     In a purely functional setting, most array operations can be expressed conveniently using map and fold.
-    We present a differentiable programming language, that is the first to deliver en efficient pure and parallel friendly reverse-mode differentiation on
+    We present a differentiable programming language which is the first to deliver on efficient, pure and parallel-friendly reverse-mode differentiation on
     an expressive array language. 
     We show that our transformation is semantically correct and verifies the cheap gradient principle. 
-    Our transformation is presented as a compilation scheme through a novel intermediate representation we call unary form.
-    We obtain provably efficient gradients by performing general partial evaluation optimisations after our reverse-mode transformation as opposed to manually derived ones.
-    For simple first-order programs, the obtained output programs ressemble static single assignment (SSA) code. 
-    We emphasize the modularity of our approach and show how our language can easily be enriched with more optimised primitives as required for some speedups in practice.
+    Our transformation is presented as a compilation scheme through a novel intermediate representation that we call `unary form`.
+    We obtain provably efficient gradients by performing general partial evaluation optimizations after our reverse-mode transformation, as opposed to manually derived ones.
+    For simple first-order programs, the obtained output programs resemble static single assignment (SSA) code. 
+    We emphasize the modularity of our approach and show how our language can easily be enriched with more optimized primitives, as required for some speedups in practice.
 \end{abstract}

### 2021-07-03 rad-icfp-2021/correctness.tex

@@
 
-\begin{definition}{concategory}
-    TODO
-\end{definition}
+One main reason for introducing UNF is to have a better control 
+of the computation flow of the term, similarly to what ANF or CPS do. 
+A convenient categorical setting for this is to use string diagrams.
+To better fit the standard denotational semantics of languages, 
+we can instead use concategories.
 
-\begin{definition}{free concategory}
-    TODO
-\end{definition}
+Concategories are a generalisation of categories and multicategories in that 
+they allow morphisms both from and to a list of objects. 
+Similarly to categories and multicategories, 
+most categorical abstractions used for the semantics of languages are easily adapted to them.
 
-\begin{example}{Concat1, Concat2}
+In particular, one can form a syntactic concategory, 
+which satisfies a similar universal property as syntactic multicategories.
+Two such examples are of particular interest for us, 
+because they will allow us to interpret Source and Target UNF, 
+but also explain the UNF and $UNF^{-1}$ transformations.
+
+\begin{example}{Concat1}
     TODO
@@
 
-TODO: semantics of UN1, UNF2 inside Concat1, Concat2
+We can interpret Source UNF in $\ConcatS$ as follows.
+
+TODO
+
+Similarly, we introduce a second concategory for the Target part.
+
+\begin{example}{Concat2}
+    TODO
+\end{example}
+
+We can interpret Target UNF in $\ConcatT$ as follows.
+
+TODO
 
@@
 \label{sub:Semantics for UNF transformations}
+
+We interpret Source in a new multicategory, 
+whose morphisms are particular morphisms of $\ConcatS$.
+As Source UNF is itself interpreted in $\ConcatS$, 
+this gives us a way to compare terms in Source with terms in Source UNF.
+This comparison precisely give the UNF transformation.
 
@@
 \begin{proposition}{construction above gives UNF}
+    TODO
+\end{proposition}
+
+\begin{definition}{Concat from multicat}
+    TODO
+\end{definition}
+
+\begin{proposition}{Initiality of concat2 gives $UNF^-1$}
     TODO
@@
 \begin{proposition}{semantic of D syntactic matches D lax functor}
-    TODO
-\end{proposition}
-
-\begin{definition}{Concat from multicat}
-    TODO
-\end{definition}
-
-\begin{proposition}{Initiality of concat2 gives $UNF^-1$}
     TODO

### 2021-07-03 rad-icfp-2021/correctness.tex
writing bit more in correctness section

@@
 
+In a similar vein, we form a concategory from the syntactic multicategory for Target.
+We then use the universal property of $\ConcatT$ to construct a functor from $\ConcatT$ to this concategory.
+This allows us to compare terms of Target UNF and terms of Target. 
+$UNF^-1$ arises in this way.
+
 \begin{definition}{Concat from multicat}
@@
 
-\begin{definition}{lax functor between concat}
+What remains to explain now is the reverse mode transformation between Source UNF and Target UNF.
+We construct a functor $D:\ConcatS\to\ConcatT$. 
+This functor will easily be shown to compute reverse mode derivatives.
+Because terms of Source UNF in $\ConcatS$, we observe the effect of $D$ on them
+and show that it matches the syntactic $\Dsynrevsymbol_{\rho}$ 
+from Section~\ref{sub:Simple reverse mode transformation}.
+
+\begin{definition}{D as a lax functor}
     TODO
@@
 
-\begin{example}{D as a lax functor}
-    TODO
-\end{example}
-
-\begin{proposition}{semantic of D syntactic matches D lax functor}
+\begin{proposition}{semantic of syntactic D matches D lax functor}
     TODO
@@
 
+First, we start from the correctness of $D$ which is easy to establish, 
+and then somehow propagate this information to Source and Target.
+
+\begin{proposition}{Correctness $D$}
+    TODO
+\end{proposition}
+
+Recall that the intuition from Source UNF is that it consist of terms of Source that also return their context.
+From there, the intuition for the Jacobian of a primitive in Target UNF is that it should be the Jacobian of
+the corresponding term in Target.
+
+\MH{I don't want to talk too much about the cotangent bundle functor, but to phrase correctness properties clearly I might have to.}
+
 \begin{proposition}{The transpose Jacobian of !map2! and !reduce! are correct.}
@@
 \end{proposition}
+
+From this, we can now deduce that the composite transformation doing UNF then $D$ and then $UNF^-1$ is correct
+in the sense that it produces a term which computes the gradient of the original term.
 
@@
 
+By inspection of what that composition actually give on terms from Source, 
+we show that the previous composition of transformation indeed gives the transformation 
+$\directD{\rho}{\Gamma}{Y}$ from Section~\ref{sub:Macro for pure reverse mode transformation}.
+
 \begin{proposition}{UNF;D;$UNF^-1$=D}
@@
 
+Without any more work, the previous claims have shown that 
+$\directD{\rho}{\Gamma}{Y}$ is correct.
+
 \begin{theorem}{D is correct}

### 2021-07-03 rad-icfp-2021/correctness.tex
bit more progress correctness

@@
 It is standard to give a denotational semantics of a first-order language in a Cartesian category.
-An alternative approach consists in giving a semantics in a multicategory.
-A multicategory generalises a category by allowing multimorphims, that is morphisms from a list of objects to an object.
+An alternative approach consists of giving semantics in a multicategory.
+A multicategory generalizes a category by allowing multimorphisms, that is, morphisms from a list of objects to an object.
 Most categorical structures from category theory can be phrases similarly in multicategories.
@@
 
-\begin{definition}{Syntactic multicategory for Source}
+\begin{definition}[Syntactic multicategory for Source]
     Let $\SynSource$ be the multicategory whose objects are types of Source, and where a morphism 
@@
 
-We can for instance use this to give a simple semantics of Source 
+We can, for instance, use this to give a simple semantics of Source 
 in the multicategory of Cartesian spaces and smooth maps between them.
 
-\begin{definition}{CartSp}
+\begin{definition}[CartSp]
     TODO, its definition and semantics of Source in it.
@@
 
-One main reason for introducing UNF is to have a better control 
-of the computation flow of the term, similarly to what ANF or CPS do. 
+One main reason for introducing UNF is to have a better handle over
+the computation flow of the term, similar to what ANF or CPS provide. 
 A convenient categorical setting for this is to use string diagrams.
@@
 
-Concategories are a generalisation of categories and multicategories in that 
+Concategories are a generalization of categories and multicategories in that 
 they allow morphisms both from and to a list of objects. 
@@
 
-\begin{example}{Concat1}
-    TODO
-\end{example}
+\begin{definition}[$\ConcatS$]
+Let $\ConcatS$ the syntactic Cartesian concategory whose types are those of Source and 
+with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$ $map2_{x,y.e}$
+and $reduce_{x,y.e1;e2}$.
+\MH{actually I just need projection and a representing object for pairs, it might be weaker than Cartesian?}
+\end{definition}
 
+One may notice that the syntax of $\ConcatS$ is somewhere in between the syntax of Source and of Source UNF.
 We can interpret Source UNF in $\ConcatS$ as follows.
 
-TODO
+\begin{tabular}{l c r}
+   $\seml$!var!$_{T;i} \semr$ &=& \\
+   $\seml$!op!$_{T;n} \semr$ &=& \\
+   $\seml$!pair!$_{T;A\times B} \semr$ &=& \\
+   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& \\
+   $\seml$!e!$\comp$!e!$\semr$  &=& \\
+   $\seml$!e!$\pcomp$!e!$\semr$ &=& \\
+   $\seml$!map2!$_{T;x,y.e}\semr$  &=& \\
+   $\seml$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
+\end{tabular}
 
-Similarly, we introduce a second concategory for the Target part.
+Similarly, we introduce a second concategory for the Target part of our transformations.
 
-\begin{example}{Concat2}
-    TODO
-\end{example}
+\begin{definition}[$\ConcatT$]
+Let $\ConcatT$ be the syntactic Cartesian concategory whose types are those of Target and 
+with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$ $map2_{x,y.e}$, 
+$reduce_{x,y.e1;e2}$, TODO
+\MH{not too sure, should I say what a Cartesian concat is?}
+\end{definition}
 
@@
 
-TODO
+\begin{tabular}{l c r}
+    $\seml$!var!$_{T;i} \semr$ &=& \\
+    $\seml$!op!$_{T;n} \semr$ &=& \\
+    $\seml$!pair!$_{T;A\times B} \semr$ &=& \\
+    $\seml$!proj!$_{T1;T2;T3} \semr$ &=& \\
+    $\seml$!e!$\comp$!e!$\semr$  &=& \\
+    $\seml$!e!$\pcomp$!e!$\semr$ &=& \\
+    $\seml$!map2!$_{T;x,y.e}\semr$  &=& \\
+    $\seml$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
+ \end{tabular}
+\MH{Do I need any kind of proposition saying that this makes sense?
+Might need to define on types as well to be clearer: It's an identity on type functor...}
 
@@
 this gives us a way to compare terms in Source with terms in Source UNF.
-This comparison precisely give the UNF transformation.
+This comparison precisely gives the UNF transformation.
 
-\begin{definition}{Mutlicat from concat}
+\begin{definition}[Mutlicat from concat]
     TODO
@@
 
-\begin{proposition}{construction above gives UNF}
+\begin{proposition}[construction above gives UNF]
     TODO
@@
 
-\begin{definition}{Concat from multicat}
+\begin{definition}[Concat from multicat]
     TODO
@@
 
-\begin{proposition}{Initiality of concat2 gives $UNF^-1$}
+\begin{proposition}[Initiality of concat2 gives $UNF^-1$]
     TODO
@@
 We construct a functor $D:\ConcatS\to\ConcatT$. 
-This functor will easily be shown to compute reverse mode derivatives.
+This functor will easily be shown to compute reverse-mode derivatives.
 Because terms of Source UNF in $\ConcatS$, we observe the effect of $D$ on them
@@
 
-\begin{definition}{D as a lax functor}
+\begin{definition}[D as a lax functor]
     TODO
@@
 
-\begin{proposition}{semantic of syntactic D matches D lax functor}
+\begin{proposition}[semantic of syntactic D matches D lax functor]
     TODO
@@
 
-First, we start from the correctness of $D$ which is easy to establish, 
-and then somehow propagate this information to Source and Target.
+First, we start from the correctness of $D$, which is easy to establish, 
+and then propagate this information to Source and Target.
 
-\begin{proposition}{Correctness $D$}
+\begin{proposition}[Correctness $D$]
     TODO
@@
 
-Recall that the intuition from Source UNF is that it consist of terms of Source that also return their context.
+Recall that the intuition from Source UNF is that it consists of terms of Source that also return their context.
 From there, the intuition for the Jacobian of a primitive in Target UNF is that it should be the Jacobian of
@@
 
-\begin{proposition}{The transpose Jacobian of !map2! and !reduce! are correct.}
+\begin{proposition}[The transpose Jacobian of !map2! and !reduce! are correct.]
     TODO
@@
 From this, we can now deduce that the composite transformation doing UNF then $D$ and then $UNF^-1$ is correct
-in the sense that it produces a term which computes the gradient of the original term.
+in the sense that it produces a term that computes the gradient of the original term.
 
-\begin{theorem}{UNF;D;$UNF^-1$ correct}
+\begin{theorem}[UNF;D;$UNF^-1$ correct]
     TODO
@@
 
-By inspection of what that composition actually give on terms from Source, 
+By inspection of what that composition actually gives on terms of Source, 
 we show that the previous composition of transformation indeed gives the transformation 
@@
 
-\begin{proposition}{UNF;D;$UNF^-1$=D}
+\begin{proposition}[UNF;D;$UNF^-1$=D]
     TODO
@@
 
-\begin{theorem}{D is correct}
+\begin{theorem}[D is correct]
     TODO

### 2021-07-03 rad-icfp-2021/relwork.tex
bit more progress correctness

@@
 combining array languages and automatic differentiation (tensor calculus) in the best 
-fitted intermediate representation for optimisations is of key interest, and active research.
+fitted intermediate representation for optimizations is of key interest and active research.
 \cite{bernstein2020differentiating} 
@@
 \cite{elliott2018simple}: bit similar because using CT as useful language, but we take impl and opti more seriously.
-\cite{shaikhha2019efficient}: similar approach but here more complicated for reversemode, would be interesting to investigate more loop fusions here.
+\cite{shaikhha2019efficient}: similar approach but here more complicated for reverse mode, would be interesting to investigate more loop fusions here.
 
@@
 
+\MH{should talk a bit about very recent reverse AD in Dex \cite{paszke2021getting}}
 \clearpage

### 2021-07-03 rad-icfp-2021/intro.tex

@@
 
-In 2008, Pearlmutter and Siskind \cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting. 
-They introduced reverse-mode in their setting. 
-Reverse-mode can compute in one pass the whole gradient of a function $\RR^n\to\RR$, and this is very fast.
+In 2008, Pearlmutter and Siskind~\cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting, where they introduced reverse-mode. 
+Reverse-mode computes in a single fast pass the whole gradient of a function $\RR^n\to\RR$.
 By comparison, the easier forward-mode transformation would need $n$ passes. 
-These functions $\RR^n\to\RR$ actually appear a lot in practice, for instance as loss functions in machine learning.
-Reverse-mode is thus important as it computes a whole gradient very quickly for these functions. 
-To be as efficient as the usual imperative version of reverse-mode, the transformation in \cite{pearlmutter2008reverse} introduced references.
+These functions $\RR^n\to\RR$ appear a lot in practice: for instance, as loss functions in machine learning.
+Reverse-mode is thus important, as it computes the whole gradient of such functions very quickly. 
+However, in order to be as efficient as the usual imperative version of reverse-mode, the transformation in~\cite{pearlmutter2008reverse} resorted to references.
 
-The lack of purity of reverse-mode is known to make it significantly harder to optimise and find parallelism. 
-None of the current implementations of reverse-mode in a functional setting \cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.
-\cite{xla}.
-It means that for every important non trivial operation, a hand-crafted reverse-derivative must be given to be efficient.
-In other words functional implementations seem to struggle to abstract away from imperative code.
+The lack of purity of reverse-mode is known to make it significantly harder to optimize and parallelize. 
+None of the current implementations of reverse-mode in a functional setting~\cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.~\cite{xla}.
+
+This means that for every important non trivial operation, a hand-crafted reverse-derivative must be given, to optimize for efficiency.
+In other words, abstracting away from imperative code is still a hurdle that functional implementations need to overcome.
 
@@
 % An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and potentially reusable.
- An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and reusable.
+An efficient purely functional reverse-mode transformation would be more easily explainable, optimizable, and reusable.
 
@@
 Following \cite{pearlmutter2008reverse}, it is relatively easy to define a purely functional reverse-mode on a first-order language using continuations. 
-As remarked by the authors, this is highly inefficient. 
-The problem we tackle in this paper is how to go from a simple and general setting where we can relatively easily prove correctness of reverse-mode, 
-to a representation that matches more efficient implementations.
+Yet, as remarked by the authors, this is highly inefficient. 
+The problem we tackle in this paper is how to go from a simple and general setting, where we can relatively easily prove correctness of reverse-mode, 
+to a representation that leads to more efficient implementations.
 
-This raises several problems that are dealt with in our paper. First, there is a tradeoff between a general expressive language 
-as opposed to a domain specific language. The latter usually has more static information and a specific representation to perform optimisations.
-Then, a lot of optimisations performed on AD implementations consist of a lot of hand-crafted derivatives for a lot of useful operations like matrix-matrix multiplication, dot-product, etc.
-They don't seem to come with more theoretical justification, are error prone, and can hardly fit more general optimisations.
+This raises several issues that are dealt with throughout the paper.
+First, there is a tradeoff to reach between a general expressive language and a domain specific one.
+The latter usually has more static information and a specific representation that lends itself to better optimizations.
+Then, a lot of optimizations performed on AD implementations consist of a lot of hand-crafted derivatives for a lot of useful operations like matrix-matrix multiplication, dot-product, etc.
+They don't seem to come with more theoretical justification, are error prone, and can hardly fit more general optimizations.
 \MH{TODO: maybe the example and ref that ML systems spend 90percent of the time outside of numpy.}
 This makes these systems harder to prove correct. The problem is thus to stay provably correct and pure, and yet efficient. What's more is we would like something easily provably efficient. 
-A real implementation on top of our work would of course optimise further, potentially using hand-crafted operations as well, but it would be based on more solid grounds.
+A real implementation on top of our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on more solid grounds.
 

### 2021-07-03 rad-icfp-2021/correctness.tex
minor

@@
 
-\begin{tabular}{l c r}
-   $\seml$!var!$_{T;i} \semr$ &=& \\
+\begin{tabular}{l c l}
+   $\seml$!var!$_{T;i} \semr$ &=& TODO \\
    $\seml$!op!$_{T;n} \semr$ &=& \\
@@
 
-\begin{tabular}{l c r}
-    $\seml$!var!$_{T;i} \semr$ &=& \\
+\begin{tabular}{l c l}
+    $\seml$!var!$_{T;i} \semr$ &=& TODO \\
     $\seml$!op!$_{T;n} \semr$ &=& \\

### 2021-07-03 rad-icfp-2021/relwork.tex
minor

@@
 In particular, it does not capture well cost in term of
-parallel computing, and memory management (cache optimisation is one of best way to optimise such linear algebra intensive computations).
+parallel computing, and memory management (cache optimization is one of best way to optimize such linear algebra intensive computations).
 For instance, in the earlier days of automatic differentiation, 
@@
 and try avoiding saturation of RAM. 
-These problems are harder to analyse and eventually it seems to come down to a lot of experimenting and benchmarking.
+These problems are harder to analyze and, eventually, it seems to come down to a lot of experimenting and benchmarking.
 
 \noindent \textbf{Parallelism and GPU}
-We tried to locally preserve parallelism as much as possible, 
-and having a pure transformation should help with found parallelism. 
+We tried to locally preserve parallelism. Having a pure transformation should help with found parallelism. 
 It would be interesting to implement on transformations and see how they perform on GPU.
-It is well-kwown that the addition of references provoked by the standard reverse-mode transformation
-restricts a lot of optimisations and it is not hard to find examples in C where the compiler does not detect simple optimisations 
+It is well-known that the addition of references provoked by the standard reverse-mode transformation
+restricts a lot of optimizations and it is not hard to find examples in C where the compiler does not detect simple optimisations 
 because of the references introduced by reverse-mode.
@@
 
-\begin{itemize}
-    \item fancier types and datatypes (eg dependent, adt) 
-    \item use in Probabilistic programming
-    \item other optimisations like implicit function theorem
-\end{itemize}
+\noindent \textbf{Further applications}
+Reverse-mode differentiation is notoriously difficult to grasp, implement, and optimize.
+We hope this work could help develop a new perspective on pure reverse-mode automatic differentiation.
+Other applications include the use of fancier datatypes e.g. dependent types, 
+in probabilistic programming, e.g. for HMC or Variational Inference. 
+It is also of interest to see how to extend this work with other techniques which 
+have had renewed interest such as implicit differentiation \cite{blondel2021efficient,lorraine2020optimizing}.
 
-% \MH{useful links (Swift is close to our project, and the doc is very readable):}
-% \begin{verbatim}https://docs.google.com/document/d/1_BirmTqdotglwNTOcYAW-ib6mx_jl-gH9Dbg4WmHZh0/edit# \end{verbatim}
-% \begin{verbatim}
 % 	https://github.com/apple/swift/blob/master/docs/DifferentiableProgramming.md#approaches-to-automatic-differentiation
-% \end{verbatim}
 
@@
 
-\MH{should talk a bit about very recent reverse AD in Dex \cite{paszke2021getting}}
+\MH{should talk a bit about the recent paper: reverse AD in Dex \cite{paszke2021getting}}
 \clearpage

### 2021-07-04 rad-icfp-2021/intro.tex
rewrote contribution

@@
 
+TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
+all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g.
+in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress.
+For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.
+
 \subsection{Problem}
@@
 
+TODO: same as above, maybe rewrite for more theory audience.
+
 \subsection{Summary of contributions}
 
-Our work provides a sort of compilation scheme from a source functional language to a target functional language for reverse-mode differentiation.
-To emphasize some key ideas and for presentation purposes, we present our work on a simple yet expressive language in Section~\ref{sec:simplediff}, 
-and show some useful generalisation in Section~\ref{sec:generalisation}.
+We propose a source-code transformation on simple purely functional language for a purely functional reverse-mode differentiation.
+Our transformation is decomposed in a sort of compilation scheme through a novel intermediate representation, unary form (UNF).
+To emphasize some key ideas and for presentation purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
+and show how to extend our work to a richer language in Section~\ref{sec:generalisation}.
 
-The key ideas of our work can be summarised by saying that reverse-mode differentiation
+We summarize the key ideas of our work. Reverse-mode can be made efficient
 \begin{itemize}
-   \item can be made efficient way more easily on unary operators
-   \item can be done efficiently and purely using second-order array operators for second-order array operators (e.g. !map,foldl!) 
-   \item can be made efficient via standard functional optimisation techniques
+   \item easily if there are only unary operators
+   \item in a purely functional way
+   \item on expressive constructs such as second-order array operators (e.g. !map2, reduce!) 
+   \item via standard functional optimization techniques
 \end{itemize}
 
-Our contribution tries to make these claims precise. Our detailed  contribution consist of
+Our contribution makes these claims precise. In detail we introduce
 \begin{itemize}
-    \item introducing an expressive source language with array operations, yet restrictive enough for efficient differentiation
-    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
-    \item a new intermediate representation (I.R.) which we call unary form(UNF), essentially turning every operation to a unary one
-    \item a simple reverse-mode transformation on the I.R.
-    \item an expressive purely functional target language and efficient transformation from the I.R. to this language
-    \item several useful generalisations in Section~\ref{sec:generalisation}. 
-    \item correctness of the differentiation \ref{sec:correctness}, and complexity guarantees \ref{sec:complexity}
-    \item showing how efficiency  is recovered via standard optimisation techniques
+    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\ref{sub:sourcelang})
+    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice (\ref{sec:generalisation})
+    \item a new intermediate representation (I.R), UNF, essentially turning every operation to a unary one which inputs and outputs a tuple (\ref{sec:unf})
+    \item a transformation to and from the I.R. from the main language \ref{sub:transformations to and from UNF}
+    \item a simple reverse-mode transformation on the I.R. (\ref{sub:Simple reverse mode transformation})
+    \item the direct reverse-mode transformation obtained from the source to the target language (\ref{sub:Macro for pure reverse mode transformation})
+    \item denotational semantics of our languages and transformations using multicategories and concategories \ref{sec:correctness}
+    \item correctness of the transformations \ref{sec:correctness}
+    \item complexity guarantees \ref{sec:complexity}
+    \item how efficiency is recovered via standard optimization techniques \ref{sub:Optimisations}
+    \item several additions to the source language (\ref{sec:generalisation}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
     \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
@@
 After some general partial evaluation techniques that will be detailed further in the paper, we obtain
-        % \begin{center}
-        %     \begin{tabular}{l}
-        %         !let w1 = x1 * x2 in!\\ 
-        %         !let w2 = w1 * x1 in!\\
-        %         !let y1 = 0 in!\\
-        %         !let y2 = 0 in!\\
-        %         !let y3 = 0 in!\\
-        %         !let y4 = 0 in!\\
-        %         !let y5 = 1 in!\\
-        %         !let y1'! != y1+w1*y5 in!\\
-        %         !let y4'! != y4+x1*y5 in!\\
-        %         !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
-        %     \end{tabular}
-        % \end{center}
+
     \begin{center}

### 2021-07-04 rad-icfp-2021/intro.tex
minor

@@
 \begin{itemize}
-    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\ref{sub:sourcelang})
-    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice (\ref{sec:generalisation})
-    \item a new intermediate representation (I.R), UNF, essentially turning every operation to a unary one which inputs and outputs a tuple (\ref{sec:unf})
-    \item a transformation to and from the I.R. from the main language \ref{sub:transformations to and from UNF}
-    \item a simple reverse-mode transformation on the I.R. (\ref{sub:Simple reverse mode transformation})
-    \item the direct reverse-mode transformation obtained from the source to the target language (\ref{sub:Macro for pure reverse mode transformation})
-    \item denotational semantics of our languages and transformations using multicategories and concategories \ref{sec:correctness}
-    \item correctness of the transformations \ref{sec:correctness}
-    \item complexity guarantees \ref{sec:complexity}
-    \item how efficiency is recovered via standard optimization techniques \ref{sub:Optimisations}
-    \item several additions to the source language (\ref{sec:generalisation}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
-    \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
+    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
+    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice (\S\ref{sec:generalisation})
+    \item a new intermediate representation (I.R), UNF, essentially turning every operation to a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
+    \item a transformation to and from the I.R. from the main language (\S\ref{sub:transformations to and from UNF})
+    \item a simple reverse-mode transformation on the I.R. (\S\ref{sub:Simple reverse mode transformation})
+    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
+    \item denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
+    \item correctness of the transformations (\S\ref{sec:correctness})
+    \item complexity guarantees (\S\ref{sec:complexity})
+    \item how efficiency is recovered via standard optimization techniques (\S\ref{sub:Optimisations})
+    \item several additions to the source language (\S\ref{sec:generalisation}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
+    \item a detailed comparison with other work and some limitations to our work (\S\ref{sec:conclusion})
 \end{itemize}
@@
 
-We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure.
+We introduce the general idea of how we do efficient reverse-mode in a functional setting on some examples.
 We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
-Let us have a look at what we obtain on a few examples.
 

### 2021-07-04 rad-icfp-2021/purerad.tex
rewrote contribution
rewriting section 3

@@
 \subsection{Source Language}
+\label{sub:sourcelang}
 
@@
 We consider a standard call-by-value language. 
-It consist in a first-order functional language enriched with arrays and a few typical second-order array operations. 
+It consists of a first-order functional language with arrays and a few typical second-order array operations. 
 The types !T1,T2! and terms !e1,e2! are given in Figure~\ref{fig:source_grammar}.
+We have included a minimal set of array operations for the sake of illustration,  it is not hard to add more.
+See Section~\ref{sec:generalisation}.
 
@@
 
-The typing rules are in Figure~\ref{fig:source_typesystem}. We have included a minimal set of array operations for the sake of illustration, 
-but it is not difficult to add further operations. 
+The typing rules are given in Figure~\ref{fig:source_typesystem}. 
 For scalar operations, we assume given a set of operations, including $+$ and $*$. 
-!op1! and !op2! denote respectively a unary and a binary smooth operation on reals. 
+!op1! and !op2! denote respectively a unary and a binary operation on reals. 
+These operations represent smooth total functions, 
+but again this can be easily generalised (\S\ref{sec:generalisation}).  
 Typical examples include !cos, exp, +, *!. 
 We use infix notation for binary operators.
-For the sake of simplicity in the presentation, a bound function in an array operation !map, map2, foldl! is restricted to having no free variables.
-For the same reason, we restrict arrays of reals.
-The general cases are presented in Section~\ref{sec:generalisation}.
+
+For the sake of simplicity in the presentation, the bound function in !reduce! 
+is restricted to having no free variables. For the same reason, we currently restrict to arrays of reals.
+We show how to lift these restrictions in Section~\ref{sec:generalisation}.
 
@@
 
-The target language for differentiation is an extension to the source language, 
-which is higher-order as needed to carry the continuation used to have a pure reverse-mode transformation, 
-and contains more array primitives. In addition, we require the set of operations to be closed under partial differentiation.
+The target language of our source-code transformation is an extension to the source language.
+It is a higher-order language, as our purely functional reverse-mode introduces a continuation.
+The set of scalar operations should also be closed under partial differentiation. 
+In more detail, for every unary scalar operation !op1!, 
+we assumed given an operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
+For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
+representing respectively the first and second partial derivative of !op2!.
+
+TODO:redo
+Similarly, the target language contains more array primitives which are used to define the reverse derivatives of array operations. 
+!scanl! is a !foldl! that stores all intermediate results. 
+!scanr! is similar, but reads the array from right to left, and also stores from right to left.
+!reduce! is a !foldl! for which the function is associative. 
+This means in practice that it can be computed faster.
+Finally, we add a somewhat strange operator which we call !shift1!. 
+It takes an array of size $n$,  shifts all the element of the array by one, 
+forgetting the last element, and puts a one in the first place. 
+This operator naturally shows up when differentiating !reduce!.
 
@@
 
-In more detail, for every unary scalar operation !op1!, 
-we assumed given an operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
-For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
-representing respectively the first and second partial derivative of !op2!.
-Similarly, some additional operations on arrays are needed. 
-!scanl! is a !foldl! that stores all intermediate results. 
-!scanr! is similar, but reads the array from right to left, and also stores from right to left.
-!reduce! is a !foldl! for which the function is associative. 
-This means in practice that it can be computed faster.
-We also include a !map3! operator. 
-Though not necessary as it can be simulated with two !map2!, it is a convenient notation and can sometimes run faster in practice.
-Finally, we add a somewhat strange operator which we call !shift1!. 
-It takes an array of size $n$,  shifts all the element of the array by one, 
-forgetting the last element, and puts a one in the first place. 
-This operator naturally shows up when differentiating !foldl!.
-
-The type system for the extended grammar of target is presented in Figure~\ref{fig:target_typesystem}.
+The type system for the extended grammar of the target language is presented in Figure~\ref{fig:target_typesystem}.
 
 \input{figures/target_type_system}
-
-A standard semantics of this language is given in Section~\ref{sec:correctness}.
 
@@
 
+TODO:redo
 In Figure~\ref{fig:direct_diff_macro} we present the direct transformation for pure reverse mode differentiation.

### 2021-07-04 rad-icfp-2021/appendix1.tex
redoing format appendix proof

@@
 
-\MH{change to make it nice. be more explicit about optimisations. 0 should be ZerosLike. 
-Also, I am freely referring to the variables from the context or bound in theoriginal term, should be clarified.}
+\MH{be more explicit about optimisations. 0 should be ZerosLike. A bit outdated compared to current D transformation.
+Also, I am freely referring to the variables from the context or bound in the original term, should be clarified.}
 
@@
  \begin{proof}
-\begin{align*}
-    \nabla_A \text{prod(A)} 
-    &= \nabla_A(\text{reduce} * 1 A) \\
-    &\defeq  \Big(\lambda A',z.
-    \begin{aligned}
-        & let A_1 = scanl * 1 A in \\
-        & let A_2 = map2 (x,y -> y) A_1 A in\\
-        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in\\
-        & let A_4 = scanr * 1 A_2 in \\
-        & map3 (a,b,c -> a+b*c*z) A' A_4 A_3
-    \end{aligned}
-    \Big)(0,1)\\
-    &\stackrel{\beta-reduction}{=}  \begin{aligned}
-        & let A_1 = scanl * 1 A in &\\
-        & let A_2 = map2 (x,y -> y) A_1 A in & A_2=  A\\
-        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in & A_3 = shift1 A1\\
-        & let A_4 = scanr * 1 A_2 in & \\
-        & map3 (a,b,c -> a+b*c*1) 0 A_4 A_3 & map2 * A_4 A_3
-    \end{aligned}\\
-    &=  \begin{aligned}
-        & let A_1 = scanl * 1 A in & \\
-        & let A_3 = shift1 A1 in &\\
-        & let A_4 = scanr * 1 A in &\text{forward substitution} A_2\\
-        & map2 * A_4 A_3
-    \end{aligned}\\
-    &\stackrel{\eta-reduction}{=} map2 * (scanr x 1 A) (shift1(scanl * 1 A))
-\end{align*}
+The gradient of  derivative $\nabla_A$!prod(A)! is given by
+
+\begin{tabular}{c l r}
+    & $\nabla_A$ !prod(A)! & \\
+    =&  $\nabla_A$(!reduce * 1 A!) & \\
+    $\defeq$ & \Big($\lambda$ !A',z. let A1 = scanl * 1 A in! & \\
+    & !let A2 = map2 (x,y -> y) A1 A in! & \\ 
+    & !let A3 = shift1 (map2 (x,y -> x) A1 A) in! & \\
+    & !let A4 = scanr * 1 A2 in! &\\
+    & !map3 (a,b,c -> a+b*c*z) A' A4 A3! & \\
+    & \Big)(0,1)\\
+     $\stackrel{\beta-reduction}{=}$  & !let A1 = scanl * 1 A in! & \\
+    & !let A2 = map2 (x,y -> y) A1 A in! & !A2=A! \\
+    & !let A3 = shift1 (map2 (x,y -> x) A1 A) in! & !A3=shift1 A1!\\
+    & !let A4 = scanr * 1 A2 in! & \\
+    & !map3 (a,b,c -> a+b*c*1) 0 A4 A3! & !map2 * A4 A3! \\
+    = & !let A1 = scanl * 1 A in! & \\
+    & !let A3 = shift1 A1 in! &\\
+    & !let A4 = scanr * 1 A in! & forward substitution !A2!\\
+    & !map2 * A4 A3! & \\
+    $\stackrel{\eta-reduction}{=}$ & !map2 * (scanr x 1 A) (shift1(scanl * 1 A))! &
+\end{tabular}
  \end{proof}
@@
  \begin{proof}
+The gradient of  derivative $\nabla_A$!sum(A)! is given by
 
-\begin{align*}
-    \nabla_A sum(A) 
-    &= \nabla_A reduce + 0 A \\
-    & \defeq \Big(\lambda A',z.
-    \begin{aligned}
-        & let A_1 = scanl * 1 A in \\
-        & let A_2 = map2 (x,y -> 1) A_1 A in \\
-        & let A_3 = shift1 (map2 (x,y -> 1) A_1 A) in \\
-        & let A_4 = scanr * 1 A_2 in \\
-        & map3 (a,b,c -> a+b*c*z) A' A_4 A_3
-    \end{aligned}
-    \Big)(0,1)\\
-    &{\beta-reduction}{=}
-    \begin{aligned}
-        & let A_1 = scanl * 1 A in &\\
-        & let A_2 = map2 (x,y -> y) A_1 A in & A_2= OnesLike(A) \\
-        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in & A_3 = shift1 OnesLike(A)\\
-        & let A_4 = scanr * 1 A_2 in & \\
-        & map3 (a,b,c -> a+b*c*1) 0 A_4 A_3 & map2 * A_4 A_3
-    \end{aligned}\\
-    &= \begin{aligned}
-        & let A_1 = scanl * 1 A in & \\
-        & let A_2= OnesLike(A) in &  \\
-        & let A_3 = shift1 OnesLike(A) in & A_3 = OnesLike(A) \\
-        & let A_4 = scanr * 1 A_2 in & A_4 = OnesLike(A_2) \\
-        & map2 * A_4 A_3 &
-    \end{aligned}\\
-    &= map2 * OnesLike(A) OnesLike(A) \\
-    &= OnesLike(A)
+    \begin{tabular}{c l r}
+    & $\nabla_A$ !sum(A)! & \\
+    = & $\nabla_A$ !reduce + 0 A! & \\
+    $\defeq$ & \Big($\lambda$ !A',z. let A1 = scanl * 1 A in! \\
+    & !let A2 = map2 (x,y -> 1) A1 A in! & \\
+    & !let A3 = shift1 (map2 (x,y -> 1) A1 A) in! & \\
+    & !let A4 = scanr * 1 A2 in! & \\
+    & !map3 (a,b,c -> a+b*c*z) A' A4 A3! &
+    &  \Big)(0,1) & \\
+    $\stackrel{\beta-reduction}{=}$  & !let A1 = scanl * 1 A in! & \\
+    & !let A2 = map2 (x,y -> y) A1 A in! & !A2=OnesLike(A)! \\
+    & !let A3 = shift1 (map2 (x,y -> x) A1 A) in! & !A3=shift1 OnesLike(A)!\\
+    & !let A4 = scanr * 1 A2 in! & \\
+    & !map3 (a,b,c -> a+b*c*1) 0 A4 A3! & !map2 * A4 A3! & \\
+    = & !let A1 = scanl * 1 A in! & \\
+    & !let A2= OnesLike(A) in! &  \\
+    & !let A3 = shift1 OnesLike(A) in! & !A3=OnesLike(A)! \\
+    & !let A4 = scanr * 1 A2 in! & !A4=OnesLike(A2)! \\
+    & !map2 * A4 A3! & \\
+    = & !map2 * OnesLike(A) OnesLike(A)! & \\
+    = & !OnesLike(A)!
+    \end{tabular}
 \end{align*}
@@
 The reverse derivative of !map2 * A B! is given by
-\begin{align*}
-    & \lambda A',B',C'. \\
-    & let C_1 =map2 (a,b -> b) A B in \\
-    & let C_2 =map2 (a,b -> a) A B in \\
-    & (map3 (a,b,c -> a+b*c) A' C_1 C', map3 (a,b,c -> a+b*c) A' C_2 C')
-\end{align*}
+
+\begin{tabular}{c l}
+    & $\lambda$ !A',B',C'.! \\
+    & !let C1 =map2 (a,b -> b) A B in! \\
+    & !let C2 =map2 (a,b -> a) A B in! \\
+    & !(map3 (a,b,c -> a+b*c) A' C1 C', map3 (a,b,c -> a+b*c) A' C2 C')!
+\end{tabular}
 
@@
 Then the gradient of !dot(A,B)! is given by
-\begin{align*}
-    \nabla dot(A,B) 
-    &= \nabla let C = map (x -> 1) A in map2 * B C\\
-    &\defeq \lambda A',B',C',z. 
-    \left.\begin{aligned}
-        & let A_1 = scanl * 1 A in \\
-        & let A_2 = map2 (x,y -> 1) A_1 A in \\
-        & let A_3 = shift1 (map2 (x,y -> 1) A_1 A) in \\
-        & let A_4 = scanr * 1 A_2 in \\
-        & Y(A', B', map3 (a,b,c -> a+b*c*z) A' A_4 A_3)
-    \end{aligned}\right\}same reduction as previously
-    \\
-    &\quad(0,0,0,1) \\
-    &= Y(0,0,OnesLike(C)) \\
-    &\defeq 
-    \begin{aligned}
-        & let C_1 =map2 (a,b -> b) A B in & C_1 = B\\
-        & let C_2 =map2 (a,b -> a) A B in & C_2 = A\\
-        & (map3 (a,b,c -> a+b*c) 0 C_1 OnesLike(C), & map2 & C_1 OnesLike(C)
-        & map3 (a,b,c -> a+b*c) 0 C_2 OnesLike(C)) & map2 & C_2 OnesLike(C)
-    \end{aligned}\\
-    &= \begin{aligned}
-        & let C_1 = B & \text{forward substitution}\\
-        & let C_2 = A \text{forward substitution}\\
-        & map2 & C_1 OnesLike(C) & C_1
-        & map2 & C_2 OnesLike(C) & C_2
-    \end{aligned}\\
-    &= (B, A)
-\end{align*}
+
+\begin{tabular}{c l r}
+    & $\nabla$ !dot(A,B)! & \\
+    = & $\nabla$ !let C = map (x -> 1) A in map2 * B C! & \\
+    $\defeq$ & $\Big(\lambda$ A',B',C',z. & \\
+    & !let A1 = scanl * 1 A in!  \hspace{4cm}\rdelim\}{5}{3mm}[\parbox{40mm}{same reduction\\as previously}]\\
+    & !let A1 = scanl * 1 A in! & \\
+    & !let A2 = map2 (x,y -> 1) A1 A in! & \\
+    & !let A3 = shift1 (map2 (x,y -> 1) A1 A) in! &\\
+    & !let A4 = scanr * 1 A2 in! &\\
+    & !Y(A', B', map3 (a,b,c -> a+b*c*z) A' A4 A3)!\Big) &  \\
+    & \quad !(0,0,0,1)! & \\
+    = & !Y(0,0,OnesLike(C))! & \\
+    $\defeq$ & !let C1=map2 (a,b -> b) A B in! & !C1 = B! \\
+    & !let C2=map2 (a,b -> a) A B in! & !C2 = A! \\
+    & !(map3 (a,b,c -> a+b*c) 0 C1 OnesLike(C)!, & !map2 C1 OnesLike(C)! \\
+    & !map3 (a,b,c -> a+b*c) 0 C2 OnesLike(C))!  & !map2 C2 OnesLike(C)! \\
+    = & !let C1=B! & forward substitution \\
+    & !let C2=A! & forward substitution \\
+    & !map2 C1 OnesLike(C)! & !C1! \\
+    & !map2 C2 OnesLike(C)! & !C2! \\
+    = & !(B, A)!
+\end{tabular}
+
 So if we are only interested in the gradient w.r.t. A, this indeed gives us B.

### 2021-07-04 rad-icfp-2021/purerad.tex
redoing format appendix proof

@@
 
-For the sake of simplicity in the presentation, the bound function in !reduce! 
+!reduce! is a fold left operator for which the function is assumed to be associative, 
+and the provided initial value should be a unit of the binary operation.
+It is a well-known parallel friendly construct. For the sake of simplicity in the presentation, the bound function in !reduce! 
 is restricted to having no free variables. For the same reason, we currently restrict to arrays of reals.
@@
 
-TODO:redo
 Similarly, the target language contains more array primitives which are used to define the reverse derivatives of array operations. 
-!scanl! is a !foldl! that stores all intermediate results. 
-!scanr! is similar, but reads the array from right to left, and also stores from right to left.
-!reduce! is a !foldl! for which the function is associative. 
-This means in practice that it can be computed faster.
-Finally, we add a somewhat strange operator which we call !shift1!. 
+Scan left !scanl! is similar to fold left but also stores all the intermediate results in a array, which it returns.
+In the same vein, scan right !scanr! perform a fold left by reading the array from right to left and also storing 
+the intermediate results in a array from right to left.
+
+Finally, we add a new shift operator !shift1!. 
 It takes an array of size $n$,  shifts all the element of the array by one, 
 forgetting the last element, and puts a one in the first place. 
-This operator naturally shows up when differentiating !reduce!.
+This somewhat ad-hoc operator naturally shows up when differentiating fold-like operators.
 
 The types and terms are presented in Figure~\ref{fig:target_grammar}.
-We are not interested here with partial applications and our language is pure, so functions take $n$ arguments.
-Lambda abstractions and applications will be removed during partial evaluation and this notation makes reading slightly easier.
+Our lambda abstractions take $n$ arguments as we are not interested here with partial applications. 
+In fact, the lambda abstractions introduced by reverse-mode will be removed during partial evaluation 
+and the notation with lambda abstractions having $n$ bound variables makes reading slightly easier.
+We note that we don't actually need the full power of higher-order because we only use lambda abstractions over variables of ground types
+and let expressions binding such lambda abstractions. So we only require the target language to be second-order.
+The type system for the extended grammar of the target language is presented in Figure~\ref{fig:target_typesystem}.
 
 \input{figures/target_grammar}
-
-The type system for the extended grammar of the target language is presented in Figure~\ref{fig:target_typesystem}.
 
@@
 TODO:redo
-In Figure~\ref{fig:direct_diff_macro} we present the direct transformation for pure reverse mode differentiation.
+In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
 We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimise it further and compute the gradient.
-This transformation may be hard to read and prove correct. 
-In Section~\ref{sec:unf} we decompose this transformation into several easier 
-and more intuitive steps via a new intermediate representation which we call UNF.
+Admittedly, this transformation may be hard to read, and prove correct. 
+In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).
 
-TODO: on the examples
+\begin{example}
+    TODO
+\end{example}
 
@@
 
-TODO: on the examples
+\begin{example}
+    TODO
+\end{example}
 
@@
 !fun (x1,...,xn) -> (f1[e1/y1],...,fn[en/yn])!
+
+TODO: redo
 Now, importantly, every !yi! is used at most once (except for !map2! again and !reduce! which require a more detailed analysis)

### 2021-07-04 rad-icfp-2021/purerad.tex
minor

@@
 
-TODO:redo
 In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
-We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimise it further and compute the gradient.
-Admittedly, this transformation may be hard to read, and prove correct. 
+We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimize it further and compute the gradient.
+Admittedly, this transformation may be hard to read, and it is not straightforward to show its correctness directly. 
 In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).
@@
 \begin{example}
-    TODO
+    The reverse-mode transformation of the terms from the introduction is
+
+    \begin{tabular}{c r}
+        &$\directD{\rho}{\Gamma}{Y}$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
+        =& TODO \\
+        \\
+        $\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
+        =& \\
+        % !dot(A,B)!
+    \end{tabular}
 \end{example}
 
-We introduce several notations which are useful when defining the transformation for reverse-mode.
+The idea is that $\rho$ represents the return type of the derivative part, which should be !A1x...,xAn! 
+if we want the whole gradient of a term !e! in context $\Gamma$=!x1:A1,...,xn:An!.
+The subscript $\Gamma$ represents the current context, 
+which is locally augmented for instance when differentiating a !let! rule.
+For non-unary operations, we differentiate the arguents from left to right and add their derivatives to the current stack, 
+which is modeled by the continuation $Y$. 
+Importantly for performance, each continuation variable $Y$ is only used once.
+
+We now introduce several notations which are useful when defining the transformation for reverse-mode.
 
 \input{figures/notation_D}
+
+We have the following typing lemma for $\directD{\rho}{\Gamma}{Y}$.
+\begin{lemma}[Typing $\directD{\rho}{\Gamma}{Y}$]
+    If $\Gamma \vdash$ !e: A!, then $\Gamma$!,Y: !$\Gamma\to \rho \vdash \directD{\rho}{\Gamma}{Y}$!(e): !$\directD{\rho}{\Gamma}{Y}$!(A)!.
+\end{lemma}
 
@@
 
+TODO: redo
 Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
@@
 Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
-and we need to initialise the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
-which we initialise at $1$ to run the backpropagation. 
+and we need to initialize the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
+which we initialize at $1$ to run the backpropagation. 
 All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.
@@
 \begin{example}
-    TODO
+    The gradient of the terms from the introduction is
+
+    \begin{tabular}{c r}
+        & $\grad_\Gamma$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
+        =& TODO \\
+        \\
+        $\grad_\Gamma$(!prod(A)!) \\
+        =& \\
+        % !dot(A,B)!
+    \end{tabular}
 \end{example}
@@
 Fortunately, the transformation is designed in such a way that we only have applications 
-and we can use partial evaluation to do beta redudction.
+and we can use partial evaluation to do beta-reduction.
 
@@
 
-TODO: redo
 Now, importantly, every !yi! is used at most once (except for !map2! again and !reduce! which require a more detailed analysis)
@@
 Though this transformation after the inlining above already has the right complexity, 
-its purity allows us to take the most out of generic optimisations.  
-A list of common optimisations that are useful for this language can be found in Figure~\ref{fig:optim}.
+its purity allows us to make the most out of generic optimizations.  
+A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.
 
-TODO: use the optimisations to be efficient on examples
+TODO: use the optimizations to be efficient on examples

### 2021-07-04 rad-icfp-2021/purerad.tex
adding example

@@
 \begin{example}
-    The reverse-mode transformation of the terms from the introduction is
+    The reverse-mode transformation of the terms from the introduction are given by
 
-    \begin{tabular}{c r}
+    \begin{tabular}{c l}
         &$\directD{\rho}{\Gamma}{Y}$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
-        =& TODO \\
-        \\
-        $\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
-        =& \\
-        % !dot(A,B)!
+        =& !let w1,Y1=!\\
+        & \quad\quad !let y11,Y11= <x1, fun (y1,y2,y3,z) -> Y(y1+z,y2,y3)> in! \\
+        & \quad\quad !let y12,Y12= <x2, fun (y1,y2,y3,y4,z) -> Y11(y1,y2+z,y3,y4)> in! \\
+        & \quad\quad !<y11 * y12, fun (y1,y2,y3,z) -> Y12(y1,y2,y3,y12*z,y11*z) > in! \\
+        & !let w2,Y2=!\\
+        & \quad\quad !let y21,Y21= <w1, fun (y1,y2,y3,y4,z) -> Y(y1,y2,y3,y4+z)> in! \\
+        & \quad\quad !let y22,Y22= <x1, fun (y1,y2,y3,y4,y5,z) -> Y11(y1+z,y2,y3,y4,y5)> in! \\
+        & \quad\quad !<y21 * y22, fun (y1,y2,y3,y4,z) -> Y22(y1,y2,y3,y4,y22*z,y21*z) > in! \\
+        & !let y,Y3= <w, fun (y1,y2,y3,y4,z) -> Y2(y1,y2,y3,y4+z)> in! \\
+        & !<y, fun (y1,y2,y3,z) -> Y3(y1,y2,y3,0,z) >!
+    \end{tabular}
+    \medskip
+
+    \begin{tabular}{c l}
+        &$\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
+        =& !let x,Y1= <0, fun (X,z) -> Y(X)> in! \\
+        & !let B,Y2= <A, fun (X,x,Z) -> Y(X+Z,x)> in!\\
+        & !let A0= scanl * x B in!\\
+        & !let A1= map2 (a,b.b) A0 B in!\\
+        & !let A2= shift1 (map2 (a,b.a) A0 B) in!\\
+        & !let A3= scanr * 1 A1 in!\\
+        & !<prod(B), fun (X,z) -> Y2(X,0,map2 (a,b. a*z*b) A2 A3)>! 
     \end{tabular}
@@
         =& \\
-        % !dot(A,B)!
     \end{tabular}

### 2021-07-04 rad-icfp-2021/purerad.tex
adding more examples

@@
 
-\begin{example}
-    The gradient of the terms from the introduction is
-
-    \begin{tabular}{c r}
-        & $\grad_\Gamma$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
-        =& TODO \\
-        \\
-        $\grad_\Gamma$(!prod(A)!) \\
-        =& \\
-    \end{tabular}
-\end{example}
-
 Following the insight from Section~\ref{subsec:insights}, 
@@
 
-Now, importantly, every !yi! is used at most once (except for !map2! again and !reduce! which require a more detailed analysis)
-which garantees that this will not blow up the complexity.
+TODO: finish explanation
 
-TODO: finish explanation, not sure how to conclude at the moment
+\begin{example}
+    The gradient of the terms from the introduction reduces to
+
+    \begin{tabular}{c l}
+        & $\grad_\Gamma$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
+        =& TODO \\
+    \end{tabular}
+    \medskip
+
+    \begin{tabular}{c l}
+        & $\grad_\Gamma$(!prod(A)!) \\
+        =& TODO \\
+    \end{tabular}
+\end{example}
 
@@
 
-TODO: use the optimizations to be efficient on examples
+\begin{example}
+As shown in appendix \ref{sub:gradintro}, the optimizations from Figure~\ref{fig:optim} 
+are sufficient to show that the gradients of the terms of the introduction reduce to the following.
+
+\begin{tabular}{{r c l}}
+    $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
+    $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
+    $\nabla_A$!dot(A,B)! &:=& !B! 
+\end{tabular}
+\end{example}

### 2021-07-04 rad-icfp-2021/correctness.tex
writing a bit in correctness

--- 2021-07-03 rad-icfp-2021/correctness.tex
@@
 and $reduce_{x,y.e1;e2}$.
-\MH{actually I just need projection and a representing object for pairs, it might be weaker than Cartesian?}
+\MH{actually I just need projection and a representing object for pairs, it might be weaker than Cartesian? edit: I think I need Cartesian.}
 \end{definition}
@@
 One may notice that the syntax of $\ConcatS$ is somewhere in between the syntax of Source and of Source UNF.
+
+\begin{notation}[Morphisms in a Cartesian concategory]
+    TODO: projection, unique map, $\concatcomp$, swap, pair, id, ...
+\end{notation}
+
 We can interpret Source UNF in $\ConcatS$ as follows.
 
-\begin{tabular}{l c r}
-   $\seml$!var!$_{T;i} \semr$ &=& \\
-   $\seml$!op!$_{T;n} \semr$ &=& \\
-   $\seml$!pair!$_{T;A\times B} \semr$ &=& \\
-   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& \\
-   $\seml$!e!$\comp$!e!$\semr$  &=& \\
-   $\seml$!e!$\pcomp$!e!$\semr$ &=& \\
-   $\seml$!map2!$_{T;x,y.e}\semr$  &=& \\
-   $\seml$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
+\begin{tabular}{l c l}
+   $\seml$!var!$_{T;i} \semr$ &=& $<id_\Gamma,\pi_i>$ \\
+   $\seml$!op!$_{T;n} \semr$ &=& $id_\Gamma\concatcomp op_n$\\
+   $\seml$!pair!$_{T;A\times B} \semr$ &=& $pair_{T;A\times B}$ \\
+   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T1}\concatcomp 1_{T2}\concatcomp id_{T3}$\\
+   $\seml$!e1!$\comp$!e2!$\semr$  &=& $\seml$!e1!$\semr \comp \seml$!e2!$\semr$ \\
+   $\seml$!e1!$\pcomp$!e2!$\semr$ &=& $\seml$!e1!$\semr \comp (\seml$!e2!$\semr\concatcomp id )\comp swap$ \\
+   $\seml$!map2!$_{T;x,y.e}\semr$  &=& $id_\Gamma\concatcomp map2_{x,y.e}$ \\
+   $\seml$!reduce!$_{T;x,y.e;e}\semr$ &=& $id_\Gamma\concatcomp reduce_{x,y.e1;e2}$ \\
 \end{tabular}
@@
 Let $\ConcatT$ be the syntactic Cartesian concategory whose types are those of Target and 
-with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$ $map2_{x,y.e}$, 
-$reduce_{x,y.e1;e2}$, TODO
-\MH{not too sure, should I say what a Cartesian concat is?}
+with primitives given by 
+$op1: \reals \to \reals$, 
+$op2: \reals,\reals \to \reals$, 
+$map2_{x,y.e}: \Array{\reals},\Array{\reals}\to \Array{\reals}$, 
+$reduce_{x,y.e1;e2}: \Array{\reals} \to \Array{\reals}$, 
+$J^Top1: \reals \to \reals$, 
+$J^Top2: \reals \to \reals, \reals$, 
+$J^Tmap2_{x,y.e}: \Array{\reals} \to \Array{\reals}, \Array{\reals}$, 
+$J^Treduce_{x,y.e1;e2}: \Array{\reals} \to \Array{\reals}$, 
+\MH{not too sure, should I say what a Cartesian concat is? do I need $J^Tproj$?}
 \end{definition}
@@
 We can interpret Target UNF in $\ConcatT$ as follows. 
+The part common with Source UNF is interpreted in the same way as for Source UNF.
 
-\begin{tabular}{l c r}
-    $\seml$!var!$_{T;i} \semr$ &=& \\
-    $\seml$!op!$_{T;n} \semr$ &=& \\
-    $\seml$!pair!$_{T;A\times B} \semr$ &=& \\
-    $\seml$!proj!$_{T1;T2;T3} \semr$ &=& \\
-    $\seml$!e!$\comp$!e!$\semr$  &=& \\
-    $\seml$!e!$\pcomp$!e!$\semr$ &=& \\
-    $\seml$!map2!$_{T;x,y.e}\semr$  &=& \\
-    $\seml$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
+\MH{I think need to explain here the (co)monoid structure induced by $D$}
+
+\begin{tabular}{l c l}
+    $\seml$!J!$^T$!var!$_{T;i} \semr$ &=& TODO \\
+    $\seml$!J!$^T$!op!$_{T;n} \semr$ &=&  \\
+    $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& \\
+    $\seml$!J!$^T$!proj!$_{T1;T2;T3} \semr$ &=& \\
+    $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& \\
+    $\seml$!J!$^T$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
+    $\seml$!<e1, e2>!$\semr$  &=& !<!$\seml$!e1!$\semr$, $\seml$!e2!$\semr$!>! \\
+    $\seml$!e1!$\icomp$!e2!$\semr$ &=&  \\
  \end{tabular}
 \MH{Do I need any kind of proposition saying that this makes sense?
-Might need to define on types as well to be clearer: It's an identity on type functor...}
+Might need to define on types as well to be clearer: It's an identity on types functor...}
 
@@
 \begin{definition}[Mutlicat from concat]
-    TODO
+    A multicategory $\catC$ naturally defines a concategory with the same objects as $\catC$ and
+    with morphisms $A\to [B1,...,Bn]$ being $n$ morphisms $A\to Bi$ of $\catC$. 
 \end{definition}

### 2021-07-04 rad-icfp-2021/appendix1.tex
adding more examples
minor

@@
 \subsection{gradient from the introduction}
+\label{sub:gradintro}
 
@@
 and ZerosLike(A) to mean map (x -> 0) A. 
-Finding these constant arrays is key to a lot of optimisations that leverage the ring algebraic structure of the reals to arrays.
+Finding these constant arrays is key to a lot of optimizations that leverage the ring algebraic structure of the reals to arrays.
 
@@
  \begin{proof}
-The gradient of  derivative $\nabla_A$!sum(A)! is given by
+The gradient of !sum(A)! is given by
 
@@
     & !let A4 = scanr * 1 A2 in! & \\
-    & !map3 (a,b,c -> a+b*c*z) A' A4 A3! &
-    &  \Big)(0,1) & \\
+    & !map3 (a,b,c -> a+b*c*z) A' A4 A3!\Big)(0,1) & \\
     $\stackrel{\beta-reduction}{=}$  & !let A1 = scanl * 1 A in! & \\
@@
     & !let A4 = scanr * 1 A2 in! & \\
-    & !map3 (a,b,c -> a+b*c*1) 0 A4 A3! & !map2 * A4 A3! & \\
+    & !map3 (a,b,c -> a+b*c*1) 0 A4 A3! & !map2 * A4 A3! \\
     = & !let A1 = scanl * 1 A in! & \\

### 2021-07-04 rad-icfp-2021/correctness.tex
minor fixes
minor

@@
  \end{theorem}
-where $\sem{-}$ is ab appropriate denotational semantics of smooth functions.
+where $\sem{-}$ is an appropriate denotational semantics of smooth functions.
 
@@
 We then use the universal property of $\ConcatT$ to construct a functor from $\ConcatT$ to this concategory.
-This allows us to compare terms of Target UNF and terms of Target. 
+This allows us to compare the terms of Target UNF and the terms of Target. 
 $UNF^-1$ arises in this way.

### 2021-07-05 rad-icfp-2021/generalisation.tex
writing generalisation section

@@
 
-\subsection{Lifting the restriction on map and fold}
-\label{sub:Lifting the restriction on map and fold}
+\subsection{Lifting the restriction on reduce}
+\label{sub:Lifting the restriction on reduce}
 
-TODO: not sure about this, and my derivative for reduce/foldl is currently a bit wrong
+Assume we allow $\Gamma$, !x!: $\reals$, !y!: $\reals$ $\vdash$ !e!: $\reals$ to be the function argument in
+!reduce (x,y.e) v A!. We need to add a term depending on $\grad_{\{xi\}}e$ to !yi! in the continuation.
+Writing the derivative becomes quite cumbersome, the notation is more compact using arrays of tuples. 
+Similarly to he monoid $\widehat{+},0_\Gamma$ defined in Section~\ref{sub:Macro for pure reverse mode transformation},
+we use $\widehat{\times},1_\Gamma$ for obvious extension of the monoid $(\reals,\times,1)$.
 
-Note that we recover !zip A B! as !map2 (fun x,y -> (x,y)) A B)!. 
-We can lift the restriction on the function of !foldl! by changing a part in $\invUNFSymbol$ of !foldl!.
+We can now compute the reverse-mode transformation as follows.
 \begin{center}
 \begin{tabular}{r c l}
-    !Y((x1,...,xm),...)! & -> &  !Y((x1,...,xm) + map2! \\
-        && !(fun z1 z2 -> let E = map2 (fun x,y -> z2(x,y)) A e3 in! \\
-        &&  !let F = zip B E in!\\
-        && !foldl (fun (x,y), acc  -> x*acc+y) z1 F)! \\
-        && $\nabla_{\Gamma}$!e2! $\nabla_{\Gamma}$!e1,...)!
+$\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
+&=& ... as before up until !A3! \\
+&& !let B0 = map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A0 A) in!\\
+&& !let B1 = scanr !$1_\Gamma$ $\widehat{\times}$ !B0 in! \\
+&& !let B2 = reduce !$\widehat{+}$ $0_\Gamma$! B1 in! \\
+&& !<reduce (x,y.e1) y1 A, fun (x1,...,xn,z) ->! \\
+&& !Y((x1,...,xn)!$\widehat{+}$!B2,0,map2 (a,b. a*b*z) A2 A3)!
 \end{tabular}
@@
 \label{sub:Adding more array operators}
+
+There are two main differences with fold left !foldl! copmpared to !reduce!. 
+First, in !foldl (x,y.e) v A! the starting accumulation element !v! is not a unit for !(x,y.e)!,
+so it will have non trivial derivatives in general and we need to account for that.
+Second, we will need a more general !scanl! which allows !(x,y.e)! as a function argument. 
+In other words, we need the general scan left computing the intermediate values of a fold left.
+This should be a different primitive but we will still call it !scanl! in this section.
+
+Finally, the former point implies we need a few more array manipulations. 
+We write !c::[v1,...,vn]! to be the array ![c,v1,...,vn]!. 
+Given an array ![v1,...,vn]! of size $n>0$, we write !let x,A=[v1,...,vn]! to mean 
+that !x=v1! and !A=[v2,...,vn]!. The reverse derivative of fold left is then
+\begin{center}
+\begin{tabular}{r c l}
+    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !foldl (x,y.e1) e2 e3!) &=&
+            !let v,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
+            && !let A,Y2 = !$\directD{\rho}{\Gamma,v}{Y1}$!(e3) in! \\
+            && !let A0 = (scanl (x,y.e1) v A)[1..n] in! \\
+            && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
+            && !let A2 = map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A! \\
+            && !let A3 = scanr * 1 A1! \\
+            && !<reduce (x,y.e1) v A, fun (x1,...,xn,z) ->! \\
+            && !let y1,B = map2 (x,y. x*y*z) (1::A2) A3 in! \\
+            && !Y2(x1,...,xn,y1,B)>! \\
+\end{tabular}
+\end{center}
+
+TODO: other examples/recipe
 
@@
 
-It is slighly better because both derivatives of !e2! and !e3! are put together which might unlock some optimisations, but there is still a non linear usage of the variables.
+It is slightly better because both derivatives of !e2! and !e3! are put together, and this might unlock some optimizations, 
+but there is still a non linear usage of the continuation variable !Y!.
 One solution, if we know we want to compute the whole gradient of the expression, is to define instead
@@
 
-This time, there is linear usage of !Y,z! and the !xi!.
+This time, there is a linear usage of the continuation variable !Y! (and of the variables !x1,...xn! too).
 
-TODO: explain why it's correct?\\
-TODO: adding conditionals does not break differentiability as long as no non smooth function like max(0,-):R->B\\
-TODO: re explain how it's usually done by partial eval of the conditional first.
+We note that adding conditionals does not break differentiability as long as there are no non-trivial primitives $\reals\to\BB$.
+Non smooth functions such as the Rectified Linear Unit (ReLU) in machine learning which can be defined as $reLU(x)\defeq max(0,x)$ 
+requires a non smooth primitive such as $>0: \reals\to\BB$. In several AD systems such as TensorFlow, 
+the condition is evaluated before differentiation is applied, and so conditionals are never directly differentiated.
 
@@
 
-We now show how to consider arrays over any ground type $G$.
+We now show how to generalise our reverse-mode transformation to be defined on arrays over any ground type $G$.
+That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.
 

### 2021-07-05 rad-icfp-2021/abstract.tex
rewrote contribution
Minor fixes

@@
     but it does not preserve purity of the underlying program and makes it notoriously harder to optimize.
-    In a purely functional setting, most array operations can be expressed conveniently using map and fold.
-    We present a differentiable programming language which is the first to deliver on efficient, pure and parallel-friendly reverse-mode differentiation on
+    In a purely functional setting, most array operations can be expressed conveniently using \texttt{map} and \texttt{fold}.
+    We present a differentiable programming language which is the first one to deliver on efficient, pure and parallel-friendly reverse-mode differentiation on
     an expressive array language. 
     We show that our transformation is semantically correct and verifies the cheap gradient principle. 
-    Our transformation is presented as a compilation scheme through a novel intermediate representation that we call `unary form`.
+    Our transformation is presented as a compilation scheme through a novel intermediate representation that we call `unary form'.
     We obtain provably efficient gradients by performing general partial evaluation optimizations after our reverse-mode transformation, as opposed to manually derived ones.
@@
     We emphasize the modularity of our approach and show how our language can easily be enriched with more optimized primitives, as required for some speedups in practice.
+    TODO: rewrite for more theory people
 \end{abstract}

### 2021-07-05 rad-icfp-2021/intro.tex
minor
Intro until section 1.4 + US spelling

@@
 % % Due to its efficiency and numerical stability, AD is the technique of choice whenever derivatives need to be computed 
-% % of functions that are implemented as programs, particularly in high dimentional settings. 
-% % Optimisations and Monte Carlo integration algorithms, such as Gradient Descent and Hamiltonian Monte-Carlo methods, 
+% % of functions that are implemented as programs, particularly in high dimensional settings. 
+% % Optimizations and Monte Carlo integration algorithms, such as Gradient Descent and Hamiltonian Monte-Carlo methods, 
 % % rely crucially on the calculation of derivatives. These algorithms are used in virtually everywhere in machine learning 
@@
 
-In 2008, Pearlmutter and Siskind~\cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting, where they introduced reverse-mode. 
+In 2008, Pearlmutter and Siskind~\cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting, where they defined reverse-mode. 
 Reverse-mode computes in a single fast pass the whole gradient of a function $\RR^n\to\RR$.
-By comparison, the easier forward-mode transformation would need $n$ passes. 
+By comparison, the easier forward-mode transformation would require $n$ passes. 
 These functions $\RR^n\to\RR$ appear a lot in practice: for instance, as loss functions in machine learning.
 Reverse-mode is thus important, as it computes the whole gradient of such functions very quickly. 
-However, in order to be as efficient as the usual imperative version of reverse-mode, the transformation in~\cite{pearlmutter2008reverse} resorted to references.
+However, in order to reach the efficiency of the usual imperative version of reverse-mode, the transformation in~\cite{pearlmutter2008reverse} resorts to references.
 
-The lack of purity of reverse-mode is known to make it significantly harder to optimize and parallelize. 
+The lack of purity in reverse-mode is known to make it significantly harder to optimize and parallelize. 
 None of the current implementations of reverse-mode in a functional setting~\cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.~\cite{xla}.
 
-This means that for every important non trivial operation, a hand-crafted reverse-derivative must be given, to optimize for efficiency.
+As a result, to optimize for efficiency, a hand-crafted reverse-derivative must be given, for every important non trivial operation.
 In other words, abstracting away from imperative code is still a hurdle that functional implementations need to overcome.
 
-% For example, \MH{TODO: show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems}.
-% For real world AD systems, modularity is key. They need to be expressive but have a lot of optimised derivatives for operators that are used a lot.
-% This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
-% An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and potentially reusable.
 An efficient purely functional reverse-mode transformation would be more easily explainable, optimizable, and reusable.
 
-TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
+\MH{TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
 all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g.
 in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress.
-For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.
+For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.}
 
@@
 
-This raises several issues that are dealt with throughout the paper.
+This raises several issues that we deal with throughout the paper.
 First, there is a tradeoff to reach between a general expressive language and a domain specific one.
 The latter usually has more static information and a specific representation that lends itself to better optimizations.
-Then, a lot of optimizations performed on AD implementations consist of a lot of hand-crafted derivatives for a lot of useful operations like matrix-matrix multiplication, dot-product, etc.
-They don't seem to come with more theoretical justification, are error prone, and can hardly fit more general optimizations.
+Then, many optimizations performed on AD implementations consist in hand-crafted derivatives for useful operations like matrix-matrix multiplication, dot-product, etc.
+They don't seem to arise from theoretical justifications, are error prone, and can hardly fit with more general optimizations.
 \MH{TODO: maybe the example and ref that ML systems spend 90percent of the time outside of numpy.}
-This makes these systems harder to prove correct. The problem is thus to stay provably correct and pure, and yet efficient. What's more is we would like something easily provably efficient. 
-A real implementation on top of our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on more solid grounds.
+This makes these systems harder to prove correct.
+The problem is thus to ensure provable correctness and pureness, while not compromising on efficiency.
+In addition, we would like something easily provably efficient.
+A real-world implementation based on our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on solid grounds.
 
-TODO: same as above, maybe rewrite for more theory audience.
+\MH{TODO: same as above, maybe rewrite for more theory audience.}
 
@@
 
-We propose a source-code transformation on simple purely functional language for a purely functional reverse-mode differentiation.
-Our transformation is decomposed in a sort of compilation scheme through a novel intermediate representation, unary form (UNF).
-To emphasize some key ideas and for presentation purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
+We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
+Our transformation is comprised of compilation scheme through a novel intermediate representation: unary form (UNF).
+To emphasize the key ideas and for expository purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
 and show how to extend our work to a richer language in Section~\ref{sec:generalisation}.
@@
 
-Our contribution makes these claims precise. In detail we introduce
+Our contribution makes these claims precise. In more details, we introduce
 \begin{itemize}
     \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
-    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice (\S\ref{sec:generalisation})
-    \item a new intermediate representation (I.R), UNF, essentially turning every operation to a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
-    \item a transformation to and from the I.R. from the main language (\S\ref{sub:transformations to and from UNF})
-    \item a simple reverse-mode transformation on the I.R. (\S\ref{sub:Simple reverse mode transformation})
+    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalisation})
+    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
+    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
+    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
     \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
@@
     \item complexity guarantees (\S\ref{sec:complexity})
-    \item how efficiency is recovered via standard optimization techniques (\S\ref{sub:Optimisations})
+    \item how efficiency is recovered via standard optimization techniques (\S\ref{sub:Optimizations})
     \item several additions to the source language (\S\ref{sec:generalisation}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
@@
 Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
-In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimisations.
+In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimizations.
 In Section~\ref{sec:generalisation} we show how to adapt our work for a richer source language. 

### 2021-07-05 rad-icfp-2021/relwork.tex
Intro until section 1.4 + US spelling

@@
 on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), 
-combined with standard functional optimisations (\ref{fig:optim}).  
+combined with standard functional optimizations (\ref{fig:optim}).  
 We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
@@
 One feature that makes it popular is that it uses JIT compilation, 
-which seems to be the best avenue to help best ML practionners as it allows them 
-to quickly experiment but also benefit from the full power of static optimisations.
+which seems to be the best avenue to help best ML practitioners as it allows them 
+to quickly experiment but also benefit from the full power of static optimizations.
 We are interested in seeing how our work could sit in a JIT compiler.
 
-\noindent \textbf{More optimisations.}
+\noindent \textbf{More optimizations.}
 As pointed in \ref{sub:Adding general arrays} it would be interesting to have a more principled representation 
-of tensors which could lead to more optimisations. 
-There is always a tension between dynamicity and the possibility for static optimisations 
+of tensors which could lead to more optimizations. 
+There is always a tension between dynamicity and the possibility for static optimizations 
 and we believe our language offers a good tradeoff, but it will be interesting to tailor it 
@@
 It is well-known that the addition of references provoked by the standard reverse-mode transformation
-restricts a lot of optimizations and it is not hard to find examples in C where the compiler does not detect simple optimisations 
+restricts a lot of optimizations and it is not hard to find examples in C where the compiler does not detect simple optimizations 
 because of the references introduced by reverse-mode.

### 2021-07-05 rad-icfp-2021/intro.tex
US spelling "Generalization"

@@
 We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
-Our transformation is comprised of compilation scheme through a novel intermediate representation: unary form (UNF).
+Our transformation is comprised of a compilation scheme through a novel intermediate representation (IR): unary form (UNF).
 To emphasize the key ideas and for expository purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
-and show how to extend our work to a richer language in Section~\ref{sec:generalisation}.
+and show how to extend our work to a richer language in Section~\ref{sec:generalization}.
 
@@
     \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
-    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalisation})
+    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
     \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
@@
     \item how efficiency is recovered via standard optimization techniques (\S\ref{sub:Optimizations})
-    \item several additions to the source language (\S\ref{sec:generalisation}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
+    \item several additions to the source language (\S\ref{sec:generalization}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
     \item a detailed comparison with other work and some limitations to our work (\S\ref{sec:conclusion})
@@
 In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimizations.
-In Section~\ref{sec:generalisation} we show how to adapt our work for a richer source language. 
+In Section~\ref{sec:generalization} we show how to adapt our work for a richer source language. 
 Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

### 2021-07-05 rad-icfp-2021/purerad.tex
writing a bit in correctness
Intro until section 1.4 + US spelling
US spelling "Generalization"

@@
 We have included a minimal set of array operations for the sake of illustration,  it is not hard to add more.
-See Section~\ref{sec:generalisation}.
+See Section~\ref{sec:generalization}.
 
@@
 These operations represent smooth total functions, 
-but again this can be easily generalised (\S\ref{sec:generalisation}).  
+but again this can be easily generalized (\S\ref{sec:generalization}).  
 Typical examples include !cos, exp, +, *!. 
@@
 is restricted to having no free variables. For the same reason, we currently restrict to arrays of reals.
-We show how to lift these restrictions in Section~\ref{sec:generalisation}.
+We show how to lift these restrictions in Section~\ref{sec:generalization}.
 
@@
 In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
-We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimize it further and compute the gradient.
+We explain in the next Section (\ref{sub:Partial evaluation and optimization}) how to optimize it further and compute the gradient.
 Admittedly, this transformation may be hard to read, and it is not straightforward to show its correctness directly. 
@@
 
-\subsection{Partial evaluation and optimisation} % (fold)
-\label{sub:Partial evaluation and optimisation}
+\subsection{Partial evaluation and optimization} % (fold)
+\label{sub:Partial evaluation and optimization}
 
@@
 \begin{tabular}{{r c l}}
-    $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
-    $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
-    $\nabla_A$!dot(A,B)! &:=& !B! 
+    $\nabla_A$!prod(A)! &=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
+    $\nabla_A$!sum(A)! &=& !map (x -> 1) A!\\
+    $\nabla_A$!dot(A,B)! &=& !B! 
 \end{tabular}

### 2021-07-05 rad-icfp-2021/generalisation.tex
Intro until section 1.4 + US spelling
more in gen section

@@
 
-There are two main differences with fold left !foldl! copmpared to !reduce!. 
+There are two main differences with fold left !foldl! compared to !reduce!. 
 First, in !foldl (x,y.e) v A! the starting accumulation element !v! is not a unit for !(x,y.e)!,
-so it will have non trivial derivatives in general and we need to account for that.
+so it will have non-trivial derivatives in general and we need to account for that.
 Second, we will need a more general !scanl! which allows !(x,y.e)! as a function argument. 
@@
 
-The problem is that this could break the complexity of reverse mode because of the non linear usage of $Y$, and makes everything harder to optimise.
+The problem is that this could break the complexity of reverse mode because of the non-linear usage of $Y$, and makes everything harder to optimize.
 
-A slighly better option would be to define
+A slightly better option would be to define
 
@@
 It is slightly better because both derivatives of !e2! and !e3! are put together, and this might unlock some optimizations, 
-but there is still a non linear usage of the continuation variable !Y!.
+but there is still a non-linear usage of the continuation variable !Y!.
 One solution, if we know we want to compute the whole gradient of the expression, is to define instead
@@
 
-TODO: introduce sum and tensor of semantics of ground type\\
-TODO: then * becomes a dot product\\
-TODO: minor changes from there\\
-TODO: except I'm not sure with reduce and shift1...
+A ground type $G$ is interpreted as an Euclidean space $A$. 
+It is in particular a real vector space.
+Similarly, a ground context $\Gamma=x1:G1,...,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
+The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $()\bigoplus_{1\leq i\leq n}Ai)\otimes A$
+where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.
 
-Though this has the right complexity, it is an interesting avenue for future research to find
-better representations to allow for more optimisations. 
+We can define $\otimes$ on the types of our language inductively by
+
+\begin{tabular}{r c l}
+    $\reals \otimes A$ & $\defeq$ & $A$ \\
+    $(A1 \times ... \times An)\otimes A$ & $\defeq$ & $(A1\otimes A) \times ... \times (An \otimes A)$ \\
+    $\Array{A1} \otimes A$ & $\defeq$ & $\Array{A1\otimes A}$
+\end{tabular}
+
+With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times...\times An$ as expected.
+TODO: define *, finish D(reduce)
+
+We will need a new primitive $\widehat{*}$ whose denotation is this map.
+
+Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!.
+
+\begin{center}
+\begin{tabular}{r c l}
+    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !map2 (x,y.e1: G) e2 e3!) &=&  
+    !let A,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
+    && !let B,Y2 = !$\directD{\rho}{\Gamma,A}{Y1}$!(e3) in! \\
+    && !let C=!$\grad_{\Gamma}$!e1 in!\\
+    && !<map2 (x,y.e1) A B, fun (x1,...,xn,Z) -> !\\
+    && !Y2( (x1,...,xn)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
+    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x]!$\widehat{*}$!b) A Z,!\\
+    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e1)[a/x]!$\widehat{*}$!b) B Z )>!\\
+    \end{tabular}
+    \end{center}
+
+\begin{center}
+\begin{tabular}{r c l}
+        $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
+        &=&  !let y1,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
+        && !let A,Y2 = !$\directD{\rho}{\Gamma,y1}{Y1}$!(e3) in! \\
+        && !let A0 = scanl (x,y.e1) y1 A in! \\
+        && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
+        && !let A2 = shift1 (map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A)! \\
+        && !let A3 = scanr * 1 A1! \\
+        && !<reduce (x,y.e1) y1 A, fun (x1,...,xn,z) ->! \\
+        && !Y((x1,...,xn),0,map2 (a,b. a*b*z) A2 A3)!
+\end{tabular}
+\end{center}
+
+Evidently, one can combine all the generalizations from the previous subsections.
+Even though this transformation has the correct complexity, it is open for future research to find
+even better representations to allow for more optimizations. 
 In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
 and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.
-
-TODO: more generally about tensors \cite{liao2019differentiable,bernstein2020differentiating}
+More generally, there is growing interest in tensor calculus \cite{liao2019differentiable,bernstein2020differentiating}.

### 2021-07-05 rad-icfp-2021/generalization.tex
US spelling "Generalization"

--- 2021-07-05 rad-icfp-2021/generalisation.tex
@@
-\section{Generalisations}
-\label{sec:generalisation}
+\section{Generalizations}
+\label{sec:generalization}
 

### 2021-07-05 rad-icfp-2021/relwork.tex

@@
 and showed correctness of the reverse-mode transformation (\ref{sub:Correctness theorem}).
-We showed (\ref{sec:generalisation}) how to lift some of the restrictions that
+We showed (\ref{sec:generalization}) how to lift some of the restrictions that
 we introduced on arrays and how to extend our approach to a richer language with more primitive operations and conditionals.

### 2021-07-05 rad-icfp-2021/intro.tex
Listings improvements + Fixed acmart error

@@
     \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
-    \item denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
+    \item a denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
     \item correctness of the transformations (\S\ref{sec:correctness})
     \item complexity guarantees (\S\ref{sec:complexity})
-    \item how efficiency is recovered via standard optimization techniques (\S\ref{sub:Optimizations})
+    \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
     \item several additions to the source language (\S\ref{sec:generalization}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
-    \item a detailed comparison with other work and some limitations to our work (\S\ref{sec:conclusion})
+    \item a detailed comparison with other work and some limitations of our work (\S\ref{sec:conclusion}) \YK{Is it really necessary to specify this point in the contributions?
+(as it is supposed to be part of every paper)}
 \end{itemize}
@@
 
-We introduce the general idea of how we do efficient reverse-mode in a functional setting on some examples.
+We introduce the general idea of efficient reverse-mode in a functional setting on some examples.
 We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
@@
 \begin{example}[First-order term]
-On simple first-order terms, the output code looks like SSA code \cite{cytron1989efficient}.
-Let us consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
+On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
+Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
 After our (inefficient) reverse-mode transformation, we obtain

### 2021-07-05 rad-icfp-2021/relwork.tex
minor

@@
-\section{Conclusion}
-\label{sec:conclusion}
+\section{Discussion and future work} % (fold)
+\label{sec:discussion_and_future_work}
 
-\subsection{Summary} % (fold)
-\label{sub:summary}
+\noindent \textbf{More dynamic language}
+JAX \cite{bradbury2020jax,frostig2018compiling} is a modern framework that provides state-of-the-art automatic differentiation. 
+One feature that makes it popular is that it uses JIT compilation, 
+which seems to be the best avenue to help best ML practitioners as it allows them 
+to quickly experiment but also benefit from the full power of static optimizations.
+We are interested in seeing how our work could sit in a JIT compiler.
 
-We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) 
-gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
-on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), 
-combined with standard functional optimizations (\ref{fig:optim}).  
-We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
-to decompose our transformation into simpler transformations.
-We gave denotational semantics to our languages and our transformations (\ref{sec:correctness}) 
-and showed correctness of the reverse-mode transformation (\ref{sub:Correctness theorem}).
-We showed (\ref{sec:generalization}) how to lift some of the restrictions that
-we introduced on arrays and how to extend our approach to a richer language with more primitive operations and conditionals.
+\noindent \textbf{More optimizations.}
+As pointed in \ref{sub:Adding general arrays} it would be interesting to have a more principled representation 
+of tensors which could lead to more optimizations. 
+There is always a tension between dynamicity and the possibility for static optimizations 
+and we believe our language offers a good tradeoff, but it will be interesting to tailor it 
+further for specific applications.
 
-\subsection{Related Work} % (fold)
-\label{sub:related_work}
+\noindent \textbf{Cost function and memory accesses}
+Our cost model \ref{sub:costModel} is simple but has limits. 
+In particular, it does not capture well cost in term of
+parallel computing, and memory management (cache optimization is one of best way to optimize such linear algebra intensive computations).
+For instance, in the earlier days of automatic differentiation, 
+advanced checkpointing techniques were used to avoid storing all the intermediate values just to save memory 
+and try avoiding saturation of RAM. 
+These problems are harder to analyze and, eventually, it seems to come down to a lot of experimenting and benchmarking.
+
+\noindent \textbf{Parallelism and GPU}
+We tried to locally preserve parallelism. Having a pure transformation should help with found parallelism. 
+It would be interesting to implement on transformations and see how they perform on GPU.
+It is well-known that the addition of references provoked by the standard reverse-mode transformation
+restricts a lot of optimizations and it is not hard to find examples in C where the compiler does not detect simple optimizations 
+because of the references introduced by reverse-mode.
+
+% \noindent \textbf{implementation, BLaS, benchmark}
+
+\noindent \textbf{Higher-order functions}
+Matthijs'work, use of defunctionalization, question efficiency and usefulness, 
+interested in functions like root of function or minimal of a function. Implicit function theorem
+Some recent work \cite{vakar2020reverse,sherman2021} present reverse-mode in a higher-order language.
+\cite{vakar2020reverse} uses categorical semantics to show correctness of the reverse-mode transformation 
+and \cite{sherman2021} uses sophisticated higher-order primitives such as root finding, max, argmax or integral. 
+Their work focuses on computable reals which is hard to compare in terms of efficiency 
+with our more standard approach of AD.
+
+\noindent \textbf{More array primitives}
+filter, flatten, gather, scanl, ...
+can be derived by hand once and then it's ok
+
+\noindent \textbf{Recursion}
+
+\noindent \textbf{Higher derivatives}
+
+\noindent \textbf{Sparse data and matrix representation}
+
+\noindent \textbf{Further applications}
+Reverse-mode differentiation is notoriously difficult to grasp, implement, and optimize.
+We hope this work could help develop a new perspective on pure reverse-mode automatic differentiation.
+Other applications include the use of fancier datatypes e.g. dependent types, 
+in probabilistic programming, e.g. for HMC or Variational Inference. 
+It is also of interest to see how to extend this work with other techniques which 
+have had renewed interest such as implicit differentiation \cite{blondel2021efficient,lorraine2020optimizing}.
+
+% 	https://github.com/apple/swift/blob/master/docs/DifferentiableProgramming.md#approaches-to-automatic-differentiation
+
+\MH{understand related work sentence from Diff. Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
+    For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
+    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.}
+
+\MH{should talk a bit about the recent paper: reverse AD in Dex \cite{paszke2021getting}}
+
+\section{Related Work} % (fold)
+\label{sec:related_work}
 
@@
 
-\subsection{Discussion and future work} % (fold)
-\label{sub:discussion_and_future_work}
+\section{Conclusion}
+\label{sec:conclusion}
 
-\noindent \textbf{More dynamic language}
-JAX \cite{bradbury2020jax,frostig2018compiling} is a modern framework that provides state-of-the-art automatic differentiation. 
-One feature that makes it popular is that it uses JIT compilation, 
-which seems to be the best avenue to help best ML practitioners as it allows them 
-to quickly experiment but also benefit from the full power of static optimizations.
-We are interested in seeing how our work could sit in a JIT compiler.
-
-\noindent \textbf{More optimizations.}
-As pointed in \ref{sub:Adding general arrays} it would be interesting to have a more principled representation 
-of tensors which could lead to more optimizations. 
-There is always a tension between dynamicity and the possibility for static optimizations 
-and we believe our language offers a good tradeoff, but it will be interesting to tailor it 
-further for specific applications.
-
-\noindent \textbf{Cost function and memory accesses}
-Our cost model \ref{sub:costModel} is simple but has limits. 
-In particular, it does not capture well cost in term of
-parallel computing, and memory management (cache optimization is one of best way to optimize such linear algebra intensive computations).
-For instance, in the earlier days of automatic differentiation, 
-advanced checkpointing techniques were used to avoid storing all the intermediate values just to save memory 
-and try avoiding saturation of RAM. 
-These problems are harder to analyze and, eventually, it seems to come down to a lot of experimenting and benchmarking.
-
-\noindent \textbf{Parallelism and GPU}
-We tried to locally preserve parallelism. Having a pure transformation should help with found parallelism. 
-It would be interesting to implement on transformations and see how they perform on GPU.
-It is well-known that the addition of references provoked by the standard reverse-mode transformation
-restricts a lot of optimizations and it is not hard to find examples in C where the compiler does not detect simple optimizations 
-because of the references introduced by reverse-mode.
-
-\noindent \textbf{implementation, BLaS, benchmark}
-
-\noindent \textbf{Higher-order functions}
-Matthijs'work, use of defunctionalization, question efficiency and usefulness, 
-interested in functions like root of function or minimal of a function. Implicit function theorem
-
-\noindent \textbf{More array primitives}
-filter, flatten, gather, scanl, ...
-can be derived by hand once and then it's ok
-
-\noindent \textbf{Recursion}
-
-\noindent \textbf{Higher derivatives}
-
-\noindent \textbf{Sparse data and matrix representation}
-
-\noindent \textbf{Further applications}
-Reverse-mode differentiation is notoriously difficult to grasp, implement, and optimize.
-We hope this work could help develop a new perspective on pure reverse-mode automatic differentiation.
-Other applications include the use of fancier datatypes e.g. dependent types, 
-in probabilistic programming, e.g. for HMC or Variational Inference. 
-It is also of interest to see how to extend this work with other techniques which 
-have had renewed interest such as implicit differentiation \cite{blondel2021efficient,lorraine2020optimizing}.
-
-% 	https://github.com/apple/swift/blob/master/docs/DifferentiableProgramming.md#approaches-to-automatic-differentiation
-
-\MH{understand related work sentence from Diff. Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
-    For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
-    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.}
-
-\MH{should talk a bit about the recent paper: reverse AD in Dex \cite{paszke2021getting}}
+We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) 
+gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
+on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), 
+combined with standard functional optimizations (\ref{fig:optim}).  
+We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
+to decompose our transformation into simpler transformations.
+We gave denotational semantics to our languages and our transformations (\ref{sec:correctness}) 
+and showed correctness of the reverse-mode transformation (\ref{sub:Correctness theorem}).
+We showed (\ref{sec:generalization}) how to lift some of the restrictions that
+we introduced on arrays and how to extend our approach to a richer language with more primitive operations and conditionals.
 \clearpage

### 2021-07-05 rad-icfp-2021/intro.tex
Part1: Subscripts and math environments in listings

@@
 On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
-Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
+Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
 After our (inefficient) reverse-mode transformation, we obtain
@@
     \begin{tabular}{l}
-        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
-        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
-        !w_2'(0,0,0,0,1)!
+        !let w$_1$,w$_1'$! = !< x$_1$ * x$_2$, fun (y$_1$, ..., y$_4$) -> (y$_1$+x$_2$*y$_4$, y$_2$+x$_1$*y$_4$, y$_3$)> in!\\
+        !let w$_2$,w$_2'$! = !< w$_1$*x$_1$, fun (y$_1$, ..., y$_5$) -> w$_1'$!!(y$_1$+w$_1$*y$_5$, y$_2$, y$_3$, y$_4$+x$_1$*y$_5$)> in!\\
+        !w$_2'$(0,0,0,0,1)!
     \end{tabular}
@@
             \begin{tabular}{l}
-                !let w1 = x1 * x2 in!\\ 
-                !let w2 = w1 * x1 in!\\
-                !let y1,y2,y3,y4,y5 = 0,0,0,0,1 in!\\
-                !let y1'! != y1+w1*y5 in!\\
-                !let y4'! != y4+x1*y5 in!\\
-                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
+                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
+                !let w$_2$ = w$_1$ * x$_1$ in!\\
+                !let y$_1$,y$_2$,y$_3$,y$_4$,y$_5$ = 0,0,0,0,1 in!\\
+                !let y$_1'$! != y$_1$+w$_1$*y$_5$ in!\\
+                !let y$_4'$! != y$_4$+x$_1$*y$_5$ in!\\
+                !(y$_1'$+x$_2$*y$_4'$!!, y$_2$+x$_1$*y$_4'$!!, y$_3$)!
             \end{tabular}
@@
 For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
-This term can be further optimised via constant propagation and algebraic simplifications to give
+This term can be further optimized via constant propagation and algebraic simplifications to give
         \begin{center}
             \begin{tabular}{{c}}
-                !let w1 = x1 * x2 in!\\ 
-                !let w2 = w1 * x1 in!\\
-                !(w1+x2*x1, x1*x1, 0)!
+                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
+                !let w$_2$ = w$_1$ * x$_1$ in!\\
+                !(w$_1$+x$_2$*x$_1$, x$_1$*x$_1$, 0)!
             \end{tabular}
@@
 \end{center}
-where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
+where !shift1 [v$_1$,...,v$_n$] = [1,v$_1$,...,v$_{n-1}$]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
 !scanl! and !scanr! are respectively scan left and scan right. These return all the intermediate result of a fold in an array.

### 2021-07-05 rad-icfp-2021/purerad.tex
Part1: Subscripts and math environments in listings

@@
     \begin{tabular}{c l}
-        &$\directD{\rho}{\Gamma}{Y}$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
-        =& !let w1,Y1=!\\
-        & \quad\quad !let y11,Y11= <x1, fun (y1,y2,y3,z) -> Y(y1+z,y2,y3)> in! \\
-        & \quad\quad !let y12,Y12= <x2, fun (y1,y2,y3,y4,z) -> Y11(y1,y2+z,y3,y4)> in! \\
-        & \quad\quad !<y11 * y12, fun (y1,y2,y3,z) -> Y12(y1,y2,y3,y12*z,y11*z) > in! \\
-        & !let w2,Y2=!\\
-        & \quad\quad !let y21,Y21= <w1, fun (y1,y2,y3,y4,z) -> Y(y1,y2,y3,y4+z)> in! \\
-        & \quad\quad !let y22,Y22= <x1, fun (y1,y2,y3,y4,y5,z) -> Y11(y1+z,y2,y3,y4,y5)> in! \\
-        & \quad\quad !<y21 * y22, fun (y1,y2,y3,y4,z) -> Y22(y1,y2,y3,y4,y22*z,y21*z) > in! \\
-        & !let y,Y3= <w, fun (y1,y2,y3,y4,z) -> Y2(y1,y2,y3,y4+z)> in! \\
-        & !<y, fun (y1,y2,y3,z) -> Y3(y1,y2,y3,0,z) >!
+        &$\directD{\rho}{\Gamma}{Y}$(!let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$!) \\
+        =& !let w$_1$,Y$_1$=!\\
+        & \quad\quad !let y$_1$1,Y$_1$1= <x$_1$, fun (y$_1$,y$_2$,y$_3$,z) -> Y(y$_1$+z,y$_2$,y$_3$)> in! \\
+        & \quad\quad !let y$_{12}$,Y$_{12}$= <x$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{11}$(y$_{1}$,y$_{2}$+z,y$_{3}$,y$_{4}$)> in! \\
+        & \quad\quad !<y$_{11}$ * y$_{12}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{12}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{12}$*z,y$_{11}$*z) > in! \\
+        & !let w$_{2}$,Y$_{2}$=!\\
+        & \quad\quad !let y$_{21}$,Y$_{21}$= <w$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
+        & \quad\quad !let y$_{22}$,Y$_{22}$= <x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$,z) -> Y$_{11}$(y$_{1}$+z,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$)> in! \\
+        & \quad\quad !<y$_{21}$ * y$_{22}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{22}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{22}$*z,y$_{21}$*z) > in! \\
+        & !let y,Y$_{3}$= <w, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{2}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
+        & !<y, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{3}$(y$_{1}$,y$_{2}$,y$_{3}$,0,z) >!
     \end{tabular}
@@
         &$\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
-        =& !let x,Y1= <0, fun (X,z) -> Y(X)> in! \\
-        & !let B,Y2= <A, fun (X,x,Z) -> Y(X+Z,x)> in!\\
-        & !let A0= scanl * x B in!\\
-        & !let A1= map2 (a,b.b) A0 B in!\\
-        & !let A2= shift1 (map2 (a,b.a) A0 B) in!\\
-        & !let A3= scanr * 1 A1 in!\\
-        & !<prod(B), fun (X,z) -> Y2(X,0,map2 (a,b. a*z*b) A2 A3)>! 
+        =& !let x,Y$_{1}$= <0, fun (X,z) -> Y(X)> in! \\
+        & !let B,Y$_{2}$= <A, fun (X,x,Z) -> Y(X+Z,x)> in!\\
+        & !let A$_{0}$= scanl * x B in!\\
+        & !let A$_{1}$= map2 (a,b.b) A$_{0}$ B in!\\
+        & !let A$_{2}$= shift1 (map2 (a,b.a) A$_{0}$ B) in!\\
+        & !let A$_{3}$= scanr * 1 A$_{1}$ in!\\
+        & !<prod(B), fun (X,z) -> Y$_{2}$(X,0,map2 (a,b. a*z*b) A$_{2}$ A$_{3}$)>! 
     \end{tabular}
@@
 
-The idea is that $\rho$ represents the return type of the derivative part, which should be !A1x...,xAn! 
-if we want the whole gradient of a term !e! in context $\Gamma$=!x1:A1,...,xn:An!.
-The subscript $\Gamma$ represents the current context, 
+The idea is that $\rho$ represents the return type of the derivative part, which should be !A$_{1} \times \ldots \times$ A$_n$! 
+if we want the whole gradient of a term !e! in context !$\Gamma = \;$x$_{1}$:A$_{1}$,...,x$_n$:A$_n$!.
+The subscript $\Gamma$ denotes the current context, 
 which is locally augmented for instance when differentiating a !let! rule.
-For non-unary operations, we differentiate the arguents from left to right and add their derivatives to the current stack, 
+For non-unary operations, we differentiate the arguments from left to right and add their derivatives to the current stack, 
 which is modeled by the continuation $Y$. 
@@
 Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
-!fun (x1,...,xn) -> (fun (y1,...,yn) -> (f1,...,fn))(e1,...,en)! 
+!fun (x$_{1}$,...,x$_n$) -> (fun (y$_{1}$,...,yn) -> (f$_{1}$,...,f$_n$))(e$_{1}$,...,e$_n$)! 
 $\transto$ 
-!fun (x1,...,xn) -> let y1,...,yn = e1,...,en in (f1,...,fn)!
+!fun (x$_{1}$,...,x$_n$) -> let y$_{1}$,...,y$_n$ = e$_1$,...,e$_n$ in (f$_{1}$,...,f$_n$)!
 $\transto$
-!fun (x1,...,xn) -> (f1[e1/y1],...,fn[en/yn])!
+!fun (x$_{1}$,...,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],...,f$_n$[e$_n$/y$_n$])!
 
@@
     \begin{tabular}{c l}
-        & $\grad_\Gamma$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
+        & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
         =& TODO \\

### 2021-07-05 rad-icfp-2021/correctness.tex
writing generalisation section
minor

@@
 
-\begin{definition}[CartSp]
-    TODO, its definition and semantics of Source in it.
+\begin{definition}[$\CartSp$]
+Let $\CartSp$ be the Cartesian category whose objects are Euclidean spaces
+and whose morphisms are smooth functions between Euclidean spaces.
+\MH{it's not a multicategory yet}
+
+We can interpret Source in $\CartSp$ as follows.
+    \begin{tabular}{r c l}
+    $\seml \reals \semr$ &\defeq& $\RR$ \\
+    $\seml$!T1xT2!$\semr$ &\defeq& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
+    $\seml \Array{\reals}{n}\semr$ &\defeq& $\Prod_{1\leq i \leq n} \RR$ 
+    \end{tabular}
+A context $\Gamma=\{x_1:A_1,\ldots,x_n:A_n\}$ is interpreted as the product $\Prod_{1\leq i \leq n}\sem{A_i}$.
+Well typed terms $\Gamma\vdash$!e: !$A$ are interpreted as functions $\sem{\Gamma}\to\sem{A}$. This is
+routinely defined by induction on the structure of typing derivations. Constants
+!c! $\reals$ are interpreted as constant functions; and the primitives operations !op1,op2!
+are interpreted by composing with the corresponding functions, which are smooth. 
+Variables are interpreted as $\sem{x_i}(\rho)\defeq \rho_i$. 
+The remaining constructions are interpreted as follows.
+\begin{tabular}{r c l}
+    $\seml$!map2 (x,y.e1) e2 e3!$\semr$ &=& TODO\\
+    $\seml$!reduce (x,y.e1) e2 e3!$\semr$ &=& TODO
+\end{tabular}
 \end{definition}
@@
 
-\begin{proposition}[The transpose Jacobian of !map2! and !reduce! are correct.]
+\begin{proposition}[The transpose Jacobian of map2 and reduce are correct.]
     TODO

### 2021-07-05 rad-icfp-2021/correctness.tex
fixing array notation

@@
 $op2: \reals,\reals \to \reals$, 
-$map2_{x,y.e}: \Array{\reals},\Array{\reals}\to \Array{\reals}$, 
-$reduce_{x,y.e1;e2}: \Array{\reals} \to \Array{\reals}$, 
+$map2_{x,y.e}: \Array{\reals}{n},\Array{\reals}{n}\to \Array{\reals}{n}$, 
+$reduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
 $J^Top1: \reals \to \reals$, 
 $J^Top2: \reals \to \reals, \reals$, 
-$J^Tmap2_{x,y.e}: \Array{\reals} \to \Array{\reals}, \Array{\reals}$, 
-$J^Treduce_{x,y.e1;e2}: \Array{\reals} \to \Array{\reals}$, 
+$J^Tmap2_{x,y.e}: \Array{\reals}{n} \to \Array{\reals}{n}, \Array{\reals}{n}$, 
+$J^Treduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
 \MH{not too sure, should I say what a Cartesian concat is? do I need $J^Tproj$?}

### 2021-07-05 rad-icfp-2021/unf.tex
simple writing phase
writing a bit in correctness
Few fixes (+ Jacobian defined as a math operator)

@@
 \section{UNF}
+\label{sec:unf}
 
@@
 There are a few notable things in this syntax. 
-Every term is indexed by a well-defined contex $\Gamma$ from the source language.
+Every term is indexed by a well-defined context $\Gamma$ from the source language.
 In addition, a variable has a integer index $1\leq i\leq n$ where $n$ is the size of the context $\Gamma$.
-Every constant, unary, or binary operator is summarised as an $n$-ary operator !op!$_n$.
-Sequential composition is denoted by !;! and !e1;e2! means that !e1! should be performed, and then !e2!.
+Every constant, unary, or binary operator is summarized as an $n$-ary operator !op$_n$!.
+Sequential composition is denoted by !;! and !e$_{1}$;e$_{2}$! means that !e$_{1}$! should be performed, and then !e$_{2}$!.
 Every array operator !map!, !map2! and !foldl! has an extra index which represents a well-formed term in the source language.
@@
 The types are the same as in the source language. 
-A context $\Gamma=\{$!x1: T1,...,xn:Tn!$\}$ can always be seen as a type !T1!$\times$...$\times$!Tn!.
+A context $\Gamma=\{$!x$_{1}$: T$_{1}$,...,x$_n$:T$_n$!$\}$ can always be seen as a type !T$_{1}$$\times$...$\times$ T$_n$!.
 UNF is very syntactic and does not really use variables. 
@@
 
-We give a translation from our source language to source UNF in Figure~\ref{fig:source_to_unf}.
-UNF acts as a sort of stack. 
-The input represents the current stage of the stack, and we return the new value added to the stack.
-All results are kept, so the stack keeps growing. One might be surprised that the transformation for pairs and let is similar.
-The order of UNF really follows what the trace of doing a sequential call-by-value evaluation would look like if the return value does not matter.
-In other words, when evaluating !let x= e1 in e2! in call by value, we first evaluate !e1! and then !e2!. 
-Similarly in !<e1, e2>!, we evaluate !e1! and then !e2!. 
-Our ultimate goal is to return a gradient, not the original term, so what really matters is the order of evaluation, and what is evaluated.
-
-$\widehat{;}$ is similar to $;$, with a sort of context propagation effect. 
-It is a bit reminiscent of monadic composition versus the usual composition.
-Technically though, the transformation $\UNFSymbol$ is not quite compositional.
-\MH{explain in detail how to define it.}
-
-\begin{proposition}
-    Let $\Gamma\vdash$!e!: !T! be a term in the source language.\\
-    Then $\Gamma \vdash \UNFSymbol$(!e!): $\Gamma\times\Delta\times$!T! in source UNF for a certain type $\Delta$. 
-\end{proposition}
-
-$\Delta$ corresponds to the list of all the intermediate results from !e!. 
-A different way to say it would be that $\Delta$ is the context made of the bound variables of !e! when !e! is in ANF.
-
-\input{figures/source_to_unf}
+\begin{example}
+    TODO: simple term in the language, explain intuition.
+\end{example}
 
@@
 The grammar and types are given in Figure~\ref{fig:unf_target_grammar}. 
-For every constant $C$ from the source UNF, we have a new constant $J^TC$ representing its transpose Jacobian.
+For every constant $C$ from the source UNF, we have a new constant $\J^TC$ representing its transpose Jacobian.
 One thing to note is that one can only form pairs in this language. 
@@
 Technically, all our terms in target UNF should have an extra index $\rho$ which is a type from Source.
-This will be the return type of the continuation for the jacobian accumulation. 
-In other words, if we want the full gradient of a term $\Gamma\vdash$!e!: $\reals$, $\rho$ should be chosen to be $\Gamma$.
-For every type !T!, let $\Dtype$(!T!)$:=$!T!$\times$(!T!->$\rho$). 
-The typesystem is given in Figure~\ref{fig:target_unf_typesystem}.
+This will be the return type of the continuation for the Jacobian accumulation. 
+In other words, if we want the full gradient of a term $\Gamma\vdash$!e:$\reals$!, $\rho$ should be chosen to be $\Gamma$.
+For every type !T!, let !$\Dtype($!T!$):= \,$T$\times$(T->$\rho$)!. 
+The type system is given in Figure~\ref{fig:target_unf_typesystem}.
 
@@
 
-% subsection transformations to and from UNF (end)
+We give a translation from our source language to source UNF in Figure~\ref{fig:source_to_unf}.
+UNF acts as a sort of stack. 
+The input represents the current stage of the stack, and we return the new value added to the stack.
+All results are kept, so the stack keeps growing. One might be surprised that the transformation for pairs and let is similar.
+The order of UNF really follows what the trace of doing a sequential call-by-value evaluation would look like if the return value does not matter.
+In other words, when evaluating !let x= e1 in e2! in call by value, we first evaluate !e1! and then !e2!. 
+Similarly in !<e1, e2>!, we evaluate !e1! and then !e2!. 
+Our ultimate goal is to return a gradient, not the original term, so what really matters is the order of evaluation, and what is evaluated.
+
+$\widehat{;}$ is similar to $;$, with a sort of context propagation effect. 
+It is a bit reminiscent of monadic composition versus the usual composition.
+Technically though, the transformation $\UNFSymbol$ is not quite compositional.
+\MH{explain in detail how to define it.}
+
+\begin{proposition}
+    Let $\Gamma\vdash$!e!: !T! be a term in the source language.\\
+    Then $\Gamma \vdash \UNFSymbol$(!e!): $\Gamma\times\Delta\times$!T! in source UNF for a certain type $\Delta$. 
+\end{proposition}
+
+$\Delta$ corresponds to the list of all the intermediate results from !e!. 
+A different way to say it would be that $\Delta$ is the context made of the bound variables of !e! when !e! is in ANF.
+
+\input{figures/source_to_unf}
+
+The transformation from the target UNF to the target language is presented in Figure~\ref{fig:unf_to_target}.
+
+\input{figures/unf_to_target}

### 2021-07-05 rad-icfp-2021/purerad.tex

@@
 
-Finally, we add a new shift operator !shift1!. 
-It takes an array of size $n$,  shifts all the element of the array by one, 
-forgetting the last element, and puts a one in the first place. 
-This somewhat ad-hoc operator naturally shows up when differentiating fold-like operators.
+Finally, we add two new shift operator !shift1L, shift1R!. 
+They take an array of size $n$, and shift respectively forget the first and the last element of the array.
+These somewhat ad-hoc operators naturally show up when differentiating fold-like operators.
 

### 2021-07-05 rad-icfp-2021/background.tex
minor syntax

@@
 combinations of partial derivatives:
-\[\textstyle h(x_1,x_1',x_2,x_2')=(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))\text.\]
+
+\begin{center}
+    \begin{tabular}{r c l}
+        $h(x_1,x_1',x_2,x_2')$ &= & $(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))$
+    \end{tabular}
+\end{center}
 
@@
 By using the symmetry in the chain rule, there is a way to compute the whole gradient faster, and this method is reverse-mode differentiation.
-Suppose given a function $f=f_n\circ...\circ f_1 :\RR^n\to\RR$. 
-The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(...(Jf_1v))...)$ for a direction $v\in\RR^n$. 
-Reverse-mode, on the other hand, computes $(Jf)^Tv =J^Tf_1(J^Tf_{2}(...(J^Tf_nv))...)$ for a vector $v\in\RR$.
+Suppose given a function $f=f_n\circ\ldots\circ f_1 :\RR^n\to\RR$. 
+The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(\ldots(Jf_1v))\ldots)$ for a direction $v\in\RR^n$. 
+Reverse-mode, on the other hand, computes $(Jf)^Tv =J^Tf_1(J^Tf_{2}(\ldots(J^Tf_nv))\ldots)$ for a vector $v\in\RR$.
 In particular, taking $v=1$ computes the gradient of $f$.
@@
 
-Let the term we want to differentiate be $x1:\RR,...,xn:\RR\vdash exp(cos(xi))$.
+Let the term we want to differentiate be $x1:\RR,\ldots,xn:\RR\vdash exp(cos(xi))$.
 Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
@@
 The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
-This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ...\circ f_1$. 
+This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ\ldots\circ f_1$. 
 If the original program was of type $\RR$, then the return value of the original program is the last component of $f$.
 Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
-\[\nabla f= J^Tf(0,...,0,1)=J^Tf_1(J^Tf_{2}(...(J^Tf_m(0,...,0,1)))...)\]
+\begin{center}
+    \begin{tabular}{r c l}
+        $\nabla f$ &=& $J^Tf(0,\ldots,0,1)=J^Tf_1(J^Tf_{2}(\ldots(J^Tf_m(0,\ldots,0,1))\ldots)$
+    \end{tabular}
+\end{center}
 
@@
 $<f_{i+1}\circ f_i, \lambda Y. Y\circ J^Tf_i \circ J^Tf_{i+1}>$, and thus by induction we can obtain $<f, \lambda Y. YJ^Tf>$.
-By applying the identity continuation $\RR^n\to\RR^n$ on the second component then the result to $(0,...,0,1)$, 
+By applying the identity continuation $\RR^n\to\RR^n$ on the second component then the result to $(0,\ldots,0,1)$, 
 we have obtained a purely functional to compute $\nabla f$.
@@
 the row is zero except for at most two indices (one for unary operators).
-This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,...,y(n+i)).(y1,...,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
+This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,\ldots,y(n+i)).(y1,\ldots,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
 where $[k]$ means that the element is added at the $k$-th index of the tuple.

### 2021-07-05 rad-icfp-2021/correctness.tex
minor syntax

@@
     A multicategory $\catC$ naturally defines a concategory with the same objects as $\catC$ and
-    with morphisms $A\to [B1,...,Bn]$ being $n$ morphisms $A\to Bi$ of $\catC$. 
+    with morphisms $A\to [B_1,\ldots,B_n]$ being $n$ morphisms $A\to Bi$ of $\catC$. 
 \end{definition}

### 2021-07-05 rad-icfp-2021/generalization.tex
fixing array notation
minor
minor syntax

@@
 $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
-&=& ... as before up until !A3! \\
+&=& $\ldots$ as before up until !A3! \\
 && !let B0 = map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A0 A) in!\\
@@
 && !let B2 = reduce !$\widehat{+}$ $0_\Gamma$! B1 in! \\
-&& !<reduce (x,y.e1) y1 A, fun (x1,...,xn,z) ->! \\
-&& !Y((x1,...,xn)!$\widehat{+}$!B2,0,map2 (a,b. a*b*z) A2 A3)!
+&& !<reduce (x,y.e1) y1 A, fun (x1,$\ldots$,xn,z) ->! \\
+&& !Y((x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!B2,0,map2 (a,b. a*b*z) A2 A3)!
 \end{tabular}
@@
 Finally, the former point implies we need a few more array manipulations. 
-We write !c::[v1,...,vn]! to be the array ![c,v1,...,vn]!. 
-Given an array ![v1,...,vn]! of size $n>0$, we write !let x,A=[v1,...,vn]! to mean 
-that !x=v1! and !A=[v2,...,vn]!. The reverse derivative of fold left is then
+We write !c::[v1,$\ldots$,vn]! to be the array ![c,v1,$\ldots$,vn]!. 
+Given an array ![v1,$\ldots$,vn]! of size $n>0$, we write !let x,A=[v1,$\ldots$,vn]! to mean 
+that !x=v1! and !A=[v2,$\ldots$,vn]!. The reverse derivative of fold left is then
+
+TODO: reduce recomputed after scanl can be fixed here, explain let v,B notation somewhere
 \begin{center}
@@
             && !let A3 = scanr * 1 A1! \\
-            && !<reduce (x,y.e1) v A, fun (x1,...,xn,z) ->! \\
+            && !<reduce (x,y.e1) v A, fun (x1,$\ldots$,xn,z) ->! \\
             && !let y1,B = map2 (x,y. x*y*z) (1::A2) A3 in! \\
-            && !Y2(x1,...,xn,y1,B)>! \\
+            && !Y2(x1,$\ldots$,xn,y1,B)>! \\
 \end{tabular}
@@
     &=& !let b=e1 in!   \\
-    && !<if b then e2 else e3, fun (x1,...,xn,z) ->! \\
-    && !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e2)(x1,...,xn,z)+!\\
-    && \quad!(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e3)(x1,...,xn,z))>!
+    && !<if b then e2 else e3, fun (x1,$\ldots$,xn,z) ->! \\
+    && !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e2)(x1,$\ldots$,xn,z)+!\\
+    && \quad!(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e3)(x1,$\ldots$,xn,z))>!
 \end{tabular}
@@
     &=& !let b=e1 in!   \\
-    && !<if b then e2 else e3, fun (x1,...,xn,z) ->! \\
-    && !Y((x1,...,xn)!$\widehat{+}$($\grad_\Gamma$!(e2)*b!$\widehat{+}\grad_\Gamma$!(e3)*(1-b))*z)>!\\
+    && !<if b then e2 else e3, fun (x1,$\ldots$,xn,z) ->! \\
+    && !Y((x1,$\ldots$,xn)!$\widehat{+}$($\grad_\Gamma$!(e2)*b!$\widehat{+}\grad_\Gamma$!(e3)*(1-b))*z)>!\\
 \end{tabular}
 
-This time, there is a linear usage of the continuation variable !Y! (and of the variables !x1,...xn! too).
+This time, there is a linear usage of the continuation variable !Y! (and of the variables !x1,$\ldots$xn! too).
 
@@
 It is in particular a real vector space.
-Similarly, a ground context $\Gamma=x1:G1,...,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
+Similarly, a ground context $\Gamma=x1:G1,\ldots,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
 The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $()\bigoplus_{1\leq i\leq n}Ai)\otimes A$
@@
     $\reals \otimes A$ & $\defeq$ & $A$ \\
-    $(A1 \times ... \times An)\otimes A$ & $\defeq$ & $(A1\otimes A) \times ... \times (An \otimes A)$ \\
-    $\Array{A1} \otimes A$ & $\defeq$ & $\Array{A1\otimes A}$
+    $(A1 \times \ldots \times An)\otimes A$ & $\defeq$ & $(A1\otimes A) \times \ldots \times (An \otimes A)$ \\
+    $\Array{A1}{n} \otimes A$ & $\defeq$ & $\Array{A1\otimes A}{n}$
 \end{tabular}
 
-With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times...\times An$ as expected.
+With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldotstimes An$ as expected.
 TODO: define *, finish D(reduce)
@@
     && !let C=!$\grad_{\Gamma}$!e1 in!\\
-    && !<map2 (x,y.e1) A B, fun (x1,...,xn,Z) -> !\\
-    && !Y2( (x1,...,xn)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
+    && !<map2 (x,y.e1) A B, fun (x1,$\ldots$,xn,Z) -> !\\
+    && !Y2( (x1,$\ldots$,xn)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
     && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x]!$\widehat{*}$!b) A Z,!\\
@@
         && !let A3 = scanr * 1 A1! \\
-        && !<reduce (x,y.e1) y1 A, fun (x1,...,xn,z) ->! \\
-        && !Y((x1,...,xn),0,map2 (a,b. a*b*z) A2 A3)!
+        && !<reduce (x,y.e1) y1 A, fun (x1,$\ldots$,xn,z) ->! \\
+        && !Y((x1,$\ldots$,xn),0,map2 (a,b. a*b*z) A2 A3)!
 \end{tabular}

### 2021-07-05 rad-icfp-2021/purerad.tex
minor syntax

@@
 The idea is that $\rho$ represents the return type of the derivative part, which should be !A$_{1} \times \ldots \times$ A$_n$! 
-if we want the whole gradient of a term !e! in context !$\Gamma = \;$x$_{1}$:A$_{1}$,...,x$_n$:A$_n$!.
+if we want the whole gradient of a term !e! in context !$\Gamma = \;$x$_{1}$:A$_{1}$,$\ldots$,x$_n$:A$_n$!.
 The subscript $\Gamma$ denotes the current context, 
@@
 and that it is always applied to almost the identity. 
-More precisely, we have applications of the form !fun (x1,...,xn) -> Y(e1,...,en)! 
+More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
 where the !ei! are !xi! except for at most $k$ (independant of $n$) terms.
@@
 Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
-!fun (x$_{1}$,...,x$_n$) -> (fun (y$_{1}$,...,yn) -> (f$_{1}$,...,f$_n$))(e$_{1}$,...,e$_n$)! 
+!fun (x$_{1}$,$\ldots$,x$_n$) -> (fun (y$_{1}$,$\ldots$,y$_n$) -> (f$_{1}$,$\ldots$,f$_n$))(e$_{1}$,$\ldots$,e$_n$)! 
 $\transto$ 
-!fun (x$_{1}$,...,x$_n$) -> let y$_{1}$,...,y$_n$ = e$_1$,...,e$_n$ in (f$_{1}$,...,f$_n$)!
+!fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)!
 $\transto$
-!fun (x$_{1}$,...,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],...,f$_n$[e$_n$/y$_n$])!
+!fun (x$_{1}$,$\ldots$,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],$\ldots$,f$_n$[e$_n$/y$_n$])!
 

### 2021-07-05 rad-icfp-2021/unf.tex
minor syntax

@@
 The types are the same as in the source language. 
-A context $\Gamma=\{$!x$_{1}$: T$_{1}$,...,x$_n$:T$_n$!$\}$ can always be seen as a type !T$_{1}$$\times$...$\times$ T$_n$!.
+A context $\Gamma=\{$!x$_{1}$: T$_{1}$,$\ldots$,x$_n$:T$_n$!$\}$ can always be seen as a type !T$_{1}$$\times\ldots\times$ T$_n$!.
 UNF is very syntactic and does not really use variables. 
@@
 The typing rules are detailed in Figure~\ref{fig:source_unf_typesystem}.
-By $\reals^{\times n}$ we mean the product $\reals\times...\times\reals$ of $n$ factors $\reals$.
+By $\reals^{\times n}$ we mean the product $\reals\times\ldotstimes\reals$ of $n$ factors $\reals$.
 

### 2021-07-05 rad-icfp-2021/appendix1.tex
Minor fixes
fixing shift1

@@
 \input{figures/ev_contexts}
-
-Some operators are less conventional which we now explain.
-!shift1! takes an array of reals of size $n$ and adds a $1$ as first component and forgets about the last component of the array, returning an array of size !n!.
-!scanr! is similar to !scanl! except that the folding starts from the end of the array. 
-It is equivalent to first reversing the input array, performing !scanl!, and reversing the resulting array.
-
-\MH{need to explain that the !scanl!, !scanr! coming from !reduce! are also parallel ones}
-
-\MH{might need to extend with the operators from the generalized section}
 
@@
  \begin{proposition}
-     The reverse derivative of !map! is correct.
- \end{proposition}
-
-\begin{proof}
-
-\end{proof}
-
- \begin{proposition}
     The reverse derivative of !map2! is correct.
@@
 \begin{proof}
-
+    TODO
 \end{proof}
@@
 
-\MH{be more explicit about optimisations. 0 should be ZerosLike. A bit outdated compared to current D transformation.
-Also, I am freely referring to the variables from the context or bound in the original term, should be clarified.}
+TODO: add ZerosLike ONesLike notation officially somewhere.
 
  \begin{lemma}
-     $\nabla_A$!prod(A)! = !map2 * (scanr x 1 A) (shift1(scanl * 1 A))!
+    !$\nabla_A$prod(A)= map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))!
  \end{lemma}
@@
     =&  $\nabla_A$(!reduce * 1 A!) & \\
-    $\defeq$ & \Big($\lambda$ !A',z. let A1 = scanl * 1 A in! & \\
-    & !let A2 = map2 (x,y -> y) A1 A in! & \\ 
-    & !let A3 = shift1 (map2 (x,y -> x) A1 A) in! & \\
-    & !let A4 = scanr * 1 A2 in! &\\
-    & !map3 (a,b,c -> a+b*c*z) A' A4 A3! & \\
-    & \Big)(0,1)\\
-     $\stackrel{\beta-reduction}{=}$  & !let A1 = scanl * 1 A in! & \\
-    & !let A2 = map2 (x,y -> y) A1 A in! & !A2=A! \\
-    & !let A3 = shift1 (map2 (x,y -> x) A1 A) in! & !A3=shift1 A1!\\
-    & !let A4 = scanr * 1 A2 in! & \\
-    & !map3 (a,b,c -> a+b*c*1) 0 A4 A3! & !map2 * A4 A3! \\
-    = & !let A1 = scanl * 1 A in! & \\
-    & !let A3 = shift1 A1 in! &\\
-    & !let A4 = scanr * 1 A in! & forward substitution !A2!\\
-    & !map2 * A4 A3! & \\
-    $\stackrel{\eta-reduction}{=}$ & !map2 * (scanr x 1 A) (shift1(scanl * 1 A))! &
+    $\defeq$ & \Big(!fun z ->! & \\ 
+    & !let A$_0$ = shift1R (scanl * 1 A) in! & \\
+    & !let A$_1$ = shift1L (map2 (x,y -> y) A$_0$ A) in! & \\ 
+    & !let A$_2$ = map2 (x,y -> x) A$_0$ A in! & \\
+    & !let A$_3$ = scanr * 1 A$_1$ in! &\\
+    & !map2 (a,b -> a*b*z) A$_3$ A$_2$! & \\
+    & \Big)(1)\\
+     $\stackrel{\beta-reduction}{=}$  & !let A$_0$ = shift1R (scanl * 1 A) in! & \\
+    & !let A$_1$ = shift1L (map2 (x,y -> y) A$_0$ A) in! & !A$_1$=shift1L A! \\
+    & !let A$_2$ = map2 (x,y -> x) A$_0$ A in! & !A$_2$= A$_0$!\\
+    & !let A$_3$ = scanr * 1 A$_1$ in! & \\
+    & !map2 (a,b -> a*b*1) A$_3$ A$_2$! & !map2 * A$_3$ A$_2$! \\
+    = & !let A$_0$ = shift1R (scanl * 1 A) in! & \\
+    & !let A$_1$=shift1L A! &\\
+    & !let A$_3$ = scanr * 1 A$_1$! & \\
+    & !map2 * A$_3$ A$_0$! & forward substitution !A$_2$! \\
+    $\stackrel{\eta-reduction}{=}$ & !map2 * (scanr * 1 shift1L A)! &\\
+    & \quad\quad\quad\quad !(shift1R (scanl * 1 A))! &
 \end{tabular}
@@
     = & $\nabla_A$ !reduce + 0 A! & \\
-    $\defeq$ & \Big($\lambda$ !A',z. let A1 = scanl * 1 A in! \\
-    & !let A2 = map2 (x,y -> 1) A1 A in! & \\
-    & !let A3 = shift1 (map2 (x,y -> 1) A1 A) in! & \\
-    & !let A4 = scanr * 1 A2 in! & \\
-    & !map3 (a,b,c -> a+b*c*z) A' A4 A3!\Big)(0,1) & \\
-    $\stackrel{\beta-reduction}{=}$  & !let A1 = scanl * 1 A in! & \\
-    & !let A2 = map2 (x,y -> y) A1 A in! & !A2=OnesLike(A)! \\
-    & !let A3 = shift1 (map2 (x,y -> x) A1 A) in! & !A3=shift1 OnesLike(A)!\\
-    & !let A4 = scanr * 1 A2 in! & \\
-    & !map3 (a,b,c -> a+b*c*1) 0 A4 A3! & !map2 * A4 A3! \\
-    = & !let A1 = scanl * 1 A in! & \\
-    & !let A2= OnesLike(A) in! &  \\
-    & !let A3 = shift1 OnesLike(A) in! & !A3=OnesLike(A)! \\
-    & !let A4 = scanr * 1 A2 in! & !A4=OnesLike(A2)! \\
-    & !map2 * A4 A3! & \\
+    $\defeq$ & \Big(!fun z ->! & \\ 
+    & !let A$_0$ = shift1R (scanl + 0 A) in! & \\
+    & !let A$_1$ = shift1L (map2 (x,y -> 1) A$_0$ A) in! & \\ 
+    & !let A$_2$ = map2 (x,y -> 1) A$_0$ A in! & \\
+    & !let A$_3$ = scanr * 1 A$_1$ in! &\\
+    & !map2 (a,b -> a*b*z) A$_3$ A$_2$! & \\
+    & \Big)(1)\\
+    $\stackrel{\beta-reduction}{=}$  & !let A$_0$ = shift1R (scanl + 0 A) in! & \\
+    & !let A$_1$ = shift1L (map2 (x,y -> 1) A$_0$ A) in! & !A$_1$=OnesLike(shift1L(A))! \\
+    & !let A$_2$ = map2 (x,y -> 1) A$_0$ A in! & !A$_2$=OnesLike(A)!\\
+    & !let A$_3$ = scanr * 1 A$_1$ in! & \\
+    & !map2 (a,b -> a*b*1) A$_3$ A$_2$! & !map2 * A$_3$ A$_2$! \\
+    = & !let A$_0$ = shift1R (scanl + 0 A) in! & \\
+    & !let A$_1$ = OnesLike(shift1L(A))! & forward substitution \\
+    & !let A$_2$ = OnesLike(A)! & forward substitution \\
+    & !let A$_3$ = scanr * 1 A$_1$ in! & \\
+    & !map2 * A$_3$ A$_2$! & \\
+    = & !let A$_3$ = scanr * 1 OnesLike(shift1L(A)) in! & !A$_3$=OnesLike(A)!\\
+    & !map2 * A$_3$ OnesLike(A)! & forward substitution \\
     = & !map2 * OnesLike(A) OnesLike(A)! & \\
@@
     \end{tabular}
-\end{align*}
 \end{proof}
@@
 \begin{tabular}{c l}
-    & $\lambda$ !A',B',C'.! \\
-    & !let C1 =map2 (a,b -> b) A B in! \\
-    & !let C2 =map2 (a,b -> a) A B in! \\
-    & !(map3 (a,b,c -> a+b*c) A' C1 C', map3 (a,b,c -> a+b*c) A' C2 C')!
+    & !fun (Z) ->! \\
+    & !let C$_1$ = map2 (a,b -> b) A B in! \\
+    & !let C$_2$ = map2 (a,b -> a) A B in! \\
+    & !(map2 * C$_1$ Z, map2 * C$_2$ Z)!
 \end{tabular}
 
-Let's call this term Y.
+Let's call this term !Y!.
 For convenience, let us also rewrite !map2 * B (map (x -> 1) A)! 
@@
     = & $\nabla$ !let C = map (x -> 1) A in map2 * B C! & \\
-    $\defeq$ & $\Big(\lambda$ A',B',C',z. & \\
-    & !let A1 = scanl * 1 A in!  \hspace{4cm}\rdelim\}{5}{3mm}[\parbox{40mm}{same reduction\\as previously}]\\
-    & !let A1 = scanl * 1 A in! & \\
-    & !let A2 = map2 (x,y -> 1) A1 A in! & \\
-    & !let A3 = shift1 (map2 (x,y -> 1) A1 A) in! &\\
-    & !let A4 = scanr * 1 A2 in! &\\
-    & !Y(A', B', map3 (a,b,c -> a+b*c*z) A' A4 A3)!\Big) &  \\
-    & \quad !(0,0,0,1)! & \\
-    = & !Y(0,0,OnesLike(C))! & \\
-    $\defeq$ & !let C1=map2 (a,b -> b) A B in! & !C1 = B! \\
-    & !let C2=map2 (a,b -> a) A B in! & !C2 = A! \\
-    & !(map3 (a,b,c -> a+b*c) 0 C1 OnesLike(C)!, & !map2 C1 OnesLike(C)! \\
-    & !map3 (a,b,c -> a+b*c) 0 C2 OnesLike(C))!  & !map2 C2 OnesLike(C)! \\
-    = & !let C1=B! & forward substitution \\
-    & !let C2=A! & forward substitution \\
-    & !map2 C1 OnesLike(C)! & !C1! \\
-    & !map2 C2 OnesLike(C)! & !C2! \\
+    $\defeq$ & \Big(!fun z ->! & \\ 
+    & !let A$_0$ = shift1R (scanl + 0 A) in!  \hspace{4cm}\rdelim\}{5}{3mm}[\parbox{40mm}{same reduction\\as previously}]\\
+    & !let A$_1$ = shift1L (map2 (x,y -> 1) A$_0$ A) in! & \\ 
+    & !let A$_2$ = map2 (x,y -> 1) A$_0$ A in! & \\
+    & !let A$_3$ = scanr * 1 A$_1$ in! &\\
+    & !Y(map2 (a,b -> a*b*z) A$_3$ A$_2$)! & \\
+    & \Big)(ZerosLike(A),1)\\
+    = & !Y(OnesLike(A))! & \\
+    = & !let C$_1$ = map2 (a,b -> b) A B in! & !C$_1$ = B! \\
+    & !let C$_2$ = map2 (a,b -> a) A B in! & !C$_2$ = A! \\
+    & !(map2 * C$_1$ OnesLike(A), map2 * C$_2$ OnesLike(A))!  &\\
+    = & !let C$_1$=B in! & forward substitution \\
+    & !let C$_2$=A in! & forward substitution \\
+    & !(map2 * C$_1$ OnesLike(A),! & !C$_1$! \\
+    & !map2 * C$_2$ OnesLike(A))! & !C$_2$! \\
+    = & !(map2 * B OnesLike(A), map2 * A OnesLike(A))! \\
     = & !(B, A)!
@@
 
-So if we are only interested in the gradient w.r.t. A, this indeed gives us B.
+If we are only interested in the gradient w.r.t. !A!, this indeed gives !B!.
 \end{proof}

### 2021-07-05 rad-icfp-2021/purerad.tex
finishing to erase shift1

@@
         &$\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
-        =& !let x,Y$_{1}$= <0, fun (X,z) -> Y(X)> in! \\
+        =& !let y,Y$_{1}$= <0, fun (X,z) -> Y(X)> in! \\
         & !let B,Y$_{2}$= <A, fun (X,x,Z) -> Y(X+Z,x)> in!\\
-        & !let A$_{0}$= scanl * x B in!\\
-        & !let A$_{1}$= map2 (a,b.b) A$_{0}$ B in!\\
-        & !let A$_{2}$= shift1 (map2 (a,b.a) A$_{0}$ B) in!\\
+        & !let A$_{0}$= shift1R (scanl * y B) in!\\
+        & !let A$_{1}$= shift1L (map2 (a,b.b) A$_{0}$ B) in!\\
+        & !let A$_{2}$= map2 (a,b.a) A$_{0}$ B in!\\
         & !let A$_{3}$= scanr * 1 A$_{1}$ in!\\
@@
 \begin{tabular}{{r c l}}
-    $\nabla_A$!prod(A)! &=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
+    $\nabla_A$!prod(A)! &=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))!\\
     $\nabla_A$!sum(A)! &=& !map (x -> 1) A!\\

### 2021-07-05 rad-icfp-2021/generalization.tex
fixing shift1
finishing some examples

@@
 Similarly, a ground context $\Gamma=x1:G1,\ldots,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
-The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $()\bigoplus_{1\leq i\leq n}Ai)\otimes A$
+The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}Ai)\otimes A$
 where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.
@@
         && !let A,Y2 = !$\directD{\rho}{\Gamma,y1}{Y1}$!(e3) in! \\
-        && !let A0 = scanl (x,y.e1) y1 A in! \\
-        && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
-        && !let A2 = shift1 (map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A)! \\
-        && !let A3 = scanr * 1 A1! \\
+        && !let A0 = shift1R (scanl (x,y.e1) y1 A) in! \\
+        && !let A1 = shiftL (map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A) in! \\
+        && !let A2 = map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A in! \\
+        && !let A3 = scanr * 1 A1 in! \\
         && !<reduce (x,y.e1) y1 A, fun (x1,$\ldots$,xn,z) ->! \\

### 2021-07-05 rad-icfp-2021/intro.tex
minor syntax
small fixes
finishing to erase shift1
finishing some examples

@@
 They don't seem to arise from theoretical justifications, are error prone, and can hardly fit with more general optimizations.
-\MH{TODO: maybe the example and ref that ML systems spend 90percent of the time outside of numpy.}
 This makes these systems harder to prove correct.
@@
     \item several additions to the source language (\S\ref{sec:generalization}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
-    \item a detailed comparison with other work and some limitations of our work (\S\ref{sec:conclusion}) \YK{Is it really necessary to specify this point in the contributions?
-(as it is supposed to be part of every paper)}
 \end{itemize}
@@
     \begin{tabular}{l}
-        !let w$_1$,w$_1'$! = !< x$_1$ * x$_2$, fun (y$_1$, ..., y$_4$) -> (y$_1$+x$_2$*y$_4$, y$_2$+x$_1$*y$_4$, y$_3$)> in!\\
-        !let w$_2$,w$_2'$! = !< w$_1$*x$_1$, fun (y$_1$, ..., y$_5$) -> w$_1'$!!(y$_1$+w$_1$*y$_5$, y$_2$, y$_3$, y$_4$+x$_1$*y$_5$)> in!\\
+        !let w$_1$,w$_1'$! = !< x$_1$ * x$_2$, fun (y$_1$,$\ldots$, y$_4$) -> (y$_1$+x$_2$*y$_4$, y$_2$+x$_1$*y$_4$, y$_3$)> in!\\
+        !let w$_2$,w$_2'$! = !< w$_1$*x$_1$, fun (y$_1$,$\ldots$, y$_5$) -> w$_1'$!!(y$_1$+w$_1$*y$_5$, y$_2$, y$_3$, y$_4$+x$_1$*y$_5$)> in!\\
         !w$_2'$(0,0,0,0,1)!
@@
     \begin{tabular}{{r c l}}
-        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
+        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))! \\
         $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
@@
 \end{center}
-where !shift1 [v$_1$,...,v$_n$] = [1,v$_1$,...,v$_{n-1}$]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
-!scanl! and !scanr! are respectively scan left and scan right. These return all the intermediate result of a fold in an array.
+where !scanl! and !scanr! are respectively scan left and scan right. 
+These return all the intermediate result of a fold in an array.
+!shift1L [v$_1$,$\ldots$,v$_n$] \defeq [v$_2$,$\ldots$,v$_{n}$]! and 
+!shift1R [v$_1$,$\ldots$,v$_n$] \defeq [v$_1$,$\ldots$,v$_{n-1}$]! are shift operators.
 These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.

### 2021-07-05 rad-icfp-2021/purerad.tex
finishing some examples

@@
         =& !let w$_1$,Y$_1$=!\\
-        & \quad\quad !let y$_1$1,Y$_1$1= <x$_1$, fun (y$_1$,y$_2$,y$_3$,z) -> Y(y$_1$+z,y$_2$,y$_3$)> in! \\
+        & \quad\quad !let y$_{11}$,Y$_{11}$= <x$_1$, fun (y$_1$,y$_2$,y$_3$,z) -> Y(y$_1$+z,y$_2$,y$_3$)> in! \\
         & \quad\quad !let y$_{12}$,Y$_{12}$= <x$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{11}$(y$_{1}$,y$_{2}$+z,y$_{3}$,y$_{4}$)> in! \\
@@
         & !let w$_{2}$,Y$_{2}$=!\\
-        & \quad\quad !let y$_{21}$,Y$_{21}$= <w$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
-        & \quad\quad !let y$_{22}$,Y$_{22}$= <x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$,z) -> Y$_{11}$(y$_{1}$+z,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$)> in! \\
+        & \quad\quad !let y$_{21}$,Y$_{21}$= <w$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_1$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
+        & \quad\quad !let y$_{22}$,Y$_{22}$= <x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$,z) -> Y$_{21}$(y$_{1}$+z,y$_{2}$,y$_{3}$,y$_{4}$,y$_{5}$)> in! \\
         & \quad\quad !<y$_{21}$ * y$_{22}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{22}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,y$_{22}$*z,y$_{21}$*z) > in! \\
@@
         &$\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
-        =& !let y,Y$_{1}$= <0, fun (X,z) -> Y(X)> in! \\
-        & !let B,Y$_{2}$= <A, fun (X,x,Z) -> Y(X+Z,x)> in!\\
+        =& !let y,Y$_{1}$= <1, fun (X,z) -> Y(X)> in! \\
+        & !let B,Y$_{2}$= <A, fun (X,x,Z) -> Y$_{1}$(X+Z,x)> in!\\
         & !let A$_{0}$= shift1R (scanl * y B) in!\\
@@
 
-TODO: redo
 Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
@@
 Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
+
+\begin{tabular}{l}
 !fun (x$_{1}$,$\ldots$,x$_n$) -> (fun (y$_{1}$,$\ldots$,y$_n$) -> (f$_{1}$,$\ldots$,f$_n$))(e$_{1}$,$\ldots$,e$_n$)! 
-$\transto$ 
+$\transto$ \\
 !fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)!
-$\transto$
+$\transto$ \\
 !fun (x$_{1}$,$\ldots$,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],$\ldots$,f$_n$[e$_n$/y$_n$])!
+\end{tabular}
 
@@
 \begin{example}
-    The gradient of the terms from the introduction reduces to
+    After forward substitution and inlining the inner !Y!$_{i}$, the gradient of the terms from the introduction reduces to
 
@@
         & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
-        =& TODO \\
+        =& !let w$_1$,Y$_1$= <x$_1$ * x$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y(y$_{1}$+x$_2$*z,y$_{2}$+x$_1$*z,y$_{3}$) > in! \\
+        & !let w$_{2}$,Y$_{2}$= <w$_{1}$ * x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{1}$(y$_{1}$+w$_1$*z,y$_{2}$,y$_{3}$,y$_{4}$+x$_{1}$*z) > in! \\
+        & !let y,Y$_{3}$= <w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{2}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
+        & !<y, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{3}$(y$_{1}$,y$_{2}$,y$_{3}$,0,z) >!
     \end{tabular}
-    \medskip
+
+After another simplification step, we obtain
+
+\begin{tabular}{c l}
+    & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
+    =& !let w$_1$= x$_1$ * x$_{2}$ in! \\
+        & !let w$_{2}$= w$_{1}$ * x$_{1}$ in! \\
+        & !<w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> !\\
+        & \quad\quad\quad!Y(y$_{1}$+w$_1$*(y$_4$+z)+x$_2$*(y$_4$+x$_1$*(y$_4$+z)),y$_{2}$+x$_1$*(y$_4$+x$_1$*(y$_4$+z)),y$_{3}$) >!
+\end{tabular}
+
+Similarly, for the gradient of !prod(A)! we obtain
 
@@
         & $\grad_\Gamma$(!prod(A)!) \\
-        =& TODO \\
+        & !let A$_{0}$= shift1R (scanl * 1 A) in!\\
+        & !let A$_{1}$= shift1L (map2 (a,b.b) A$_{0}$ A) in!\\
+        & !let A$_{2}$= map2 (a,b.a) A$_{0}$ A in!\\
+        & !let A$_{3}$= scanr * 1 A$_{1}$ in!\\
+        & !<prod(A), fun (X,z) -> Y(X+map2 (a,b. a*z*b) A$_{2}$ A$_{3}$)>! 
     \end{tabular}

### 2021-07-06 rad-icfp-2021/purerad.tex
minor

@@
 \begin{example}
-    After forward substitution and inlining the inner !Y!$_{i}$, the gradient of the terms from the introduction reduces to
+    After forward-substitution and inlining the inner !Y!$_{i}$, the gradient of the terms from the introduction reduces to
 
+    TODO: bit wrong, should introduce more let binding
     \begin{tabular}{c l}

### 2021-07-06 rad-icfp-2021/complexity.tex
changing outline
simple writing phase
bit more related work
minor

@@
 \subsection{Cost model}
+\label{sub:costModel}
 
@@
 
-\MH{need cost array operations}
-\MH{need to extend the cost function in a compositional way}
+TODO: extend cost array
+TODO: extend cost compositional way
 
 \subsection{Cheap gradient principle}
+
+The cheap gradient principle (see e.g. \cite{griewank2008evaluating}) for reverse-mode
+asserts that evaluating the gradient of a function $e:\RR^n\to\RR$ 
+should be the same order of cost as evaluating $e$. 
+More precisely, there should be a constant $K$ such that for each program !$\Gamma \vdash$ e: $\reals$! in the context $\Gamma=\{x_1:\reals,\ldots,x_n:\reals\}$,
+ !$\cost$($\grad$e) $\leq$ $K$*$\cost$(e)!.
+
+As $\directD{\rho}{\Gamma}{Y}$ is defined by induction on programs, it suffices to show locally 
+that $\directD{\rho}{\Gamma}{Y}$ verifies the cheap gradient principle. 
+However, this is not quite the case with $\directD{\rho}{\Gamma}{Y}$ 
+before the optimization (inlining and forward substitution) step from Section~\ref{sub:Partial evaluation and optimization}.
+
+\begin{proposition}
+    TODO: cost $\directD{\rho}{\Gamma}{Y}$. 
+\end{proposition}
+
+As we can see, the only problem is that too many variables are being moved, 
+due to the continuation carrying and copying $n$ variables at each step.
+This is exactly the overhead that our inlining and forward substitution removes.
+
+\begin{proposition}
+    TODO: complexity gain using our optim.
+\end{proposition}
+
+Combining the two previous points, we conclude that our reverse mode transformation satisfies the cheap gradient principle.
+
+\begin{theorem}
+    TODO: The reverse mode transformation satisfies the cheap gradient principle.
+\end{theorem}
+
+\subsection{Optimizations} % (fold)
+\label{sub:Optimizations}
+
+In Figure~\ref{fig:optim} we present a of optimizations. 
+As our language is purely functional and effect free, 
+we can use these optimizations aggressively. 
+A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
+After these optimizations, the output program is essentially a sequence of let bindings 
+and resembles SSA-code.
+
+\input{figures/optims}

### 2021-07-06 rad-icfp-2021/correctness.tex
small fixes
minor

@@
 and whose morphisms are smooth functions between Euclidean spaces.
-\MH{it's not a multicategory yet}
+TODO:it's not a multicategory yet
 
@@
 and $reduce_{x,y.e1;e2}$.
-\MH{actually I just need projection and a representing object for pairs, it might be weaker than Cartesian? edit: I think I need Cartesian.}
+TODO: actually I just need projection and a representing object for pairs, it might be weaker than Cartesian? edit: I think I need Cartesian.
 \end{definition}
@@
 $J^Treduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
-\MH{not too sure, should I say what a Cartesian concat is? do I need $J^Tproj$?}
+TODO: not too sure, should I say what a Cartesian concat is? do I need $J^Tproj$?
 \end{definition}
@@
 
-\MH{I think need to explain here the (co)monoid structure induced by $D$}
+TODO: I think need to explain here the (co)monoid structure induced by $D$
 
@@
  \end{tabular}
-\MH{Do I need any kind of proposition saying that this makes sense?
-Might need to define on types as well to be clearer: It's an identity on types functor...}
+TODO: Do I need any kind of proposition saying that this makes sense?
+Might need to define on types as well to be clearer: It's an identity on types functor...
 
@@
 
-\MH{I don't want to talk too much about the cotangent bundle functor, but to phrase correctness properties clearly I might have to.}
+TODO: I don't want to talk too much about the cotangent bundle functor, but to phrase correctness properties clearly I might have to.
 

### 2021-07-06 rad-icfp-2021/intro.tex
minor

@@
 
-\MH{TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
+TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
 all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g.
 in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress.
-For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.}
+For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.
 
@@
 
-\MH{TODO: same as above, maybe rewrite for more theory audience.}
+TODO: same as above, maybe rewrite for more theory audience.
 
@@
     \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
-    \item several additions to the source language (\S\ref{sec:generalization}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
+    \item several additions to the source language (\S\ref{sec:generalization}) TODO:should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.
 \end{itemize}

### 2021-07-06 rad-icfp-2021/notes/main.tex
minor syntax
minor

@@
     \multicolumn{3}{|c|}{\textbf{Core Grammar}} & \multicolumn{1}{c|}{\textbf{Description}}\\\hline
-    !T! & \mbox{::=} & ... & \grammarcomment{Same as source} \\
+    !T! & \mbox{::=} & $\ldots$ & \grammarcomment{Same as source} \\
     \hline
@@
 
- \MH{I probably need to add permutations $\sigma$ in the language of UNF then.}
+ TODO: I probably need to add permutations $\sigma$ in the language of UNF then.
 
@@
 
-\MH{another possible way to look at it is that UNF does some sort of localization, turning every operator to a reversible one, which is simply done by making it also return its context.
+TODO: another possible way to look at it is that UNF does some sort of localization, turning every operator to a reversible one, which is simply done by making it also return its context.
 I wonder if it could simply be seen as the localization of $!$.
@@
 by making $\Delta$ an iso, it becomes essentially invisible to reverse mode.
-One obvious problem is that I don't know if I want all types to be isomorphic to 1.}
+One obvious problem is that I don't know if I want all types to be isomorphic to 1.
 

### 2021-07-06 rad-icfp-2021/relwork.tex
minor

@@
 \noindent \textbf{Recursion}
+TODO
 
 \noindent \textbf{Higher derivatives}
+TODO
 
 \noindent \textbf{Sparse data and matrix representation}
+TODO
 
@@
 
-\MH{understand related work sentence from Diff. Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
+TODO: understand related work sentence from Diff. Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
     For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
-    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.}
+    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.
 
-\MH{should talk a bit about the recent paper: reverse AD in Dex \cite{paszke2021getting}}
+TODO: should talk a bit about the recent paper: reverse AD in Dex \cite{paszke2021getting}
 
@@
 \cite{bernstein2020differentiating} 
+TODO
 \cite{laue2018computing,laue2020simple}
@@
 
+TODO
 \cite{ee2020correctness}: non diff function

### 2021-07-06 rad-icfp-2021/purerad.tex
typo

@@
 
-Finally, we add two new shift operator !shift1L, shift1R!. 
-They take an array of size $n$, and shift respectively forget the first and the last element of the array.
+Finally, we add two new shift operator !shift1L!, !shift1R!. 
+They take an array of size $n$, and respectively forget the first and the last element of the array.
 These somewhat ad-hoc operators naturally show up when differentiating fold-like operators.
@@
 We note that we don't actually need the full power of higher-order because we only use lambda abstractions over variables of ground types
-and let expressions binding such lambda abstractions. So we only require the target language to be second-order.
+and let expressions binding such lambda abstractions. We only need the target language to be second-order.
 The type system for the extended grammar of the target language is presented in Figure~\ref{fig:target_typesystem}.

### 2021-07-06 rad-icfp-2021/purerad.tex
typos

@@
 For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
-representing respectively the first and second partial derivative of !op2!.
+respectively representing the first and second partial derivative of !op2!.
 
 Similarly, the target language contains more array primitives which are used to define the reverse derivatives of array operations. 
-Scan left !scanl! is similar to fold left but also stores all the intermediate results in a array, which it returns.
-In the same vein, scan right !scanr! perform a fold left by reading the array from right to left and also storing 
-the intermediate results in a array from right to left.
+Scan left !scanl! is similar to fold left but also stores all the intermediate results in an array, which it returns.
+In the same vein, scan right !scanr! performs a fold left by reading the array from right to left and stores 
+the intermediate results in an array from right to left.
 
-Finally, we add two new shift operator !shift1L!, !shift1R!. 
+Finally, we add two new shift operators !shift1L! and !shift1R!. 
 They take an array of size $n$, and respectively forget the first and the last element of the array.
@@
 Our lambda abstractions take $n$ arguments as we are not interested here with partial applications. 
-In fact, the lambda abstractions introduced by reverse-mode will be removed during partial evaluation 
+In fact, the lambda abstractions introduced by reverse-mode will be removed during partial evaluation, 
 and the notation with lambda abstractions having $n$ bound variables makes reading slightly easier.
@@
 The subscript $\Gamma$ denotes the current context, 
-which is locally augmented for instance when differentiating a !let! rule.
-For non-unary operations, we differentiate the arguments from left to right and add their derivatives to the current stack, 
+which is locally augmented, for instance when differentiating a !let! rule.
+For non-unary operations, we differentiate the arguments from the left to the right and add their derivatives to the current stack, 
 which is modeled by the continuation $Y$. 
@@
 we don't want to keep all the lambda abstractions as this is costly. 
-Fortunately, the transformation is designed in such a way that we only have applications 
+Fortunately, the transformation is designed in such a way that we only have applications, 
 and we can use partial evaluation to do beta-reduction.

### 2021-07-06 rad-icfp-2021/purerad.tex
finishing section 3

@@
 we don't want to keep all the lambda abstractions as this is costly. 
-Fortunately, the transformation is designed in such a way that we only have applications, 
-and we can use partial evaluation to do beta-reduction.
+The transformation is designed in such a way that all the lambda abstractions are given arguments,  
+and we can use partial evaluation to do beta-reduce all these lambda abstractions.
 
@@
 
-Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
+Using inlining as given in Figure~\ref{fig:optim}, this allows us to rewrite
 
@@
 !fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)!
-$\transto$ \\
-!fun (x$_{1}$,$\ldots$,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],$\ldots$,f$_n$[e$_n$/y$_n$])!
 \end{tabular}
 
-TODO: finish explanation
+Using the invariant above, most of the !e!$_{i}$ are variables. 
+Without loss of generality, assume the onlt terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
+we can use forward subsitution on the other !e!$_{i}$ and we obtain
+
+\begin{tabular}{l}
+    !fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)!  $\transto$ \\
+    !fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{n-1}$,y$_n$ =e$_{n-1},$e$_1$ in (f$_{1}$[x$_{1}$/y$_{1}$],$\ldots$,f$_{n-2}[$x$_{n-2}$/y$_{n-2}]$,f$_{n-1}$,f$_{n}$)!
+\end{tabular}
+It means that it does not change the evaluation cost of most of the !f!$_{i}$ for $1\leq i \leq n-2$.
+The new evaluation cost is reduced to the sum of the cost of evaluating 
+the !f!$_{i}$ in addition to the cost of evaluating !e!$_{n-1}$ and !e!$_{n}$, O($n$) movement of variables.
 
@@
 
-    TODO: bit wrong, should introduce more let binding
     \begin{tabular}{c l}
@@
     & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
-    =& !let w$_1$= x$_1$ * x$_{2}$ in! \\
-        & !let w$_{2}$= w$_{1}$ * x$_{1}$ in! \\
-        & !<w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> !\\
-        & \quad\quad\quad!Y(y$_{1}$+w$_1$*(y$_4$+z)+x$_2$*(y$_4$+x$_1$*(y$_4$+z)),y$_{2}$+x$_1$*(y$_4$+x$_1$*(y$_4$+z)),y$_{3}$) >!
+    =& !let w$_1$= x$_1$ * x$_{2}$ in let w$_{2}$= w$_{1}$ * x$_{1}$ in! \\
+        & !<w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> let y'$_{1}$=y$_{1}$+w$_1$*z in !\\
+        & \quad\quad\quad!let z'=y$_4$+x$_1$*z in Y(y'+x$_2$*z',y$_{2}$+x$_1$*z',y$_{3}$) >!
 \end{tabular}

### 2021-07-06 rad-icfp-2021/unf.tex
minor
rewriting UNF section

--- 2021-07-05 rad-icfp-2021/unf.tex
@@
 \label{sec:unf}
+
+Following the intuition highlighted in Section~\ref{subsec:insights}, 
+we present a new language which we call unary form (UNF). 
+There is a Source UNF for the Source language and a Target UNF for the Target language.
+Just as Target is an extension of Source, Target UNF is an extension of Source UNF.
+
+One takeaway from this paper is that efficient pure reverse-mode is complicated because of the several things it does.
+It goes through the term and keeps a store for the gradients to be updated. 
+It keeps track of the new bound variables found while going through the term.
+The updates to the gradient are done via pre-composition and not post-composition.
+Due to the linearity requirement in the usage of the continuation variable, 
+it needs to process like a call-by-value evaluation and evaluate the arguments of operators in a certain order before 
+dealing with the operator itself.
+
+By introducing UNF, we are trying to decouple some of the problems.
+UNF introduces the call-by-value evaluation of the arguments before evaluating the arguments and a good management of the environment.
+Differentiation on UNF deals with the purely functional store and gradients update via pre-composition.
+Finally, going back from UNF to Target uses the stored information in UNF to make sure the Jacobians are computed efficiently.
 
@@
 
-Following the intuition highlighted in Section~\ref{subsec:insights}, we present a new language which we call unary form (UNF). 
-It simply consists of a composition of unary operators. That said, we want to compile our source language to this intermediate representation, 
-and we need to remember some information about the initial term.
-
-The grammar of our source UNF is given in Figure~\ref{fig:unf_source_grammar}.
+Intuitively, a term consists of a composition of unary operators. 
+To compile our source language to this intermediate representation, 
+we need to remember some information about the context of the initial term. 
+The grammar of Source UNF is given in Figure~\ref{fig:unf_source_grammar}.
 
@@
 There are a few notable things in this syntax. 
-Every term is indexed by a well-defined context $\Gamma$ from the source language.
-In addition, a variable has a integer index $1\leq i\leq n$ where $n$ is the size of the context $\Gamma$.
-Every constant, unary, or binary operator is summarized as an $n$-ary operator !op$_n$!.
+Every primitive is indexed by a list of types from the source language, 
+which corresponds to a context being carried. 
+Every constant, unary, or binary operator from Source has a corresponding $n$-ary operator !op$_{T,n}$! in UNF.
 Sequential composition is denoted by !;! and !e$_{1}$;e$_{2}$! means that !e$_{1}$! should be performed, and then !e$_{2}$!.
-Every array operator !map!, !map2! and !foldl! has an extra index which represents a well-formed term in the source language.
-
-The types are the same as in the source language. 
-A context $\Gamma=\{$!x$_{1}$: T$_{1}$,...,x$_n$:T$_n$!$\}$ can always be seen as a type !T$_{1}$$\times$...$\times$ T$_n$!.
-UNF is very syntactic and does not really use variables. 
-In fact it consists only of constants and composition. 
-For this reason, we denote contexts as types in the indices of the terms, and $\times$ is used for context extension.
-But we do emphasize that we mean contexts and the actual variables are remembered and
-this will be key when compiling to the target language after differentiation.
+The array operators !map2! and !reduce! have extra indices that represent well-formed terms in Source.
+The language has a second kind of sequential composition, which we call passing sequential compsition $\pcomp$.
+In !e$_{1}$ $\pcomp$ e$_{2}$! the idea is first that !e$_{1}$! is computed, followed by !e$_{2}$!. 
+However, !e$_{1}$! outputs an argument which is not used by !e$_{2}$! and is just passed by !e$_{2}$! 
+along and returned with the output of !e$_{2}$!.
+The language also contains a projection !proj! which forgets some of the elements, 
+and a pairing operator !pair! which takes the last pair of elements and returns the pairing of these elements.
 
 The typing rules are detailed in Figure~\ref{fig:source_unf_typesystem}.
-By $\reals^{\times n}$ we mean the product $\reals\times...\times\reals$ of $n$ factors $\reals$.
-
-\MH{need to explain that return values produce some kind of ANF and introduce new vars!}
+$\reals^{\times n}$ is a notation for $\reals,\ldots,\reals$ with $n$ factors $\reals$.
 
@@
 \begin{example}
-    TODO: simple term in the language, explain intuition.
+    !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$! 
+    represents !x$_1$:$\reals$, x$_2$:$\reals$, x$_3$:$\reals$ $\vdash$ let (x$_1$,x$_2$,x$_3$,x$_4$)=(x$_1$,x$_2$,x$_3$,cos(x$_3$)) in (x$_1$,x$_2$,<x$_3$,x$_4$>)!.
+    $\comp$ can more generally be understood as a let binding of a tuple, and primitives only act on the last parts of the tuple, always returning the context.
 \end{example}
-
-% subsubsectionSource UNF (end)
 
@@
 
-Now our program in UNF is in an equivalent form to straight-line programs. 
-Before presenting our differentiation macro, we present the target language for this macro. 
-We call this intermediate representation target UNF.
-The main idea is similar to source UNF where we only have constants and sequential composition.
-That said, the key difference is that now every constant represents a pair of an operation and what is meant to represent its Jacobian.
-Following the insights from Section~\ref{sec:background}, the composition of pairs reverses the order on the second component. T
-his represents the fact that reverse-mode reverts the computation flow, 
-and that pre-composition (using a continuation) is the simplest way to do this in a purely functional setting.
-On pairs, !;! still represents sequential composition for the first component, and reversed composition is represented by $\circ$ on the second component.
-The grammar and types are given in Figure~\ref{fig:unf_target_grammar}. 
-For every constant $C$ from the source UNF, we have a new constant $\J^TC$ representing its transpose Jacobian.
-One thing to note is that one can only form pairs in this language. 
-Similarly to the source UNF, sequential composition is the only operation which is not a constant of the language.
+The Target UNF is an extension of Source UNF. 
+Its types are lists of types of Target. 
+Higher-order types are needed for the continuation introduced by reverse-mode. 
+In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
+Lastly, Target UNF has an internal composition $\icomp$. 
+!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!
+It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
+The grammar of Target UNF is given in Figure~\ref{fig:unf_source_grammar} and its type system in Figure~\ref{fig:target_unf_typesystem}.
 
 \input{figures/target_unf}
-
-Technically, all our terms in target UNF should have an extra index $\rho$ which is a type from Source.
-This will be the return type of the continuation for the Jacobian accumulation. 
-In other words, if we want the full gradient of a term $\Gamma\vdash$!e:$\reals$!, $\rho$ should be chosen to be $\Gamma$.
-For every type !T!, let !$\Dtype($!T!$):= \,$T$\times$(T->$\rho$)!. 
-The type system is given in Figure~\ref{fig:target_unf_typesystem}.
 
@@
 
-We are now able to present a simple transformation for purely functional reverse-mode on UNF.
-This transformation is somehow straightforward as some of the work was put into the design of UNF.
-Our main takeaway for this paper is that reverse-mode is complicated because there is a lot happening at the same time. 
-But by decoupling the problems, it becomes much easier. UNF is somehow dealing with the inherent type-dependency of reverse-mode.
-Then the macro $\Dsynrevsymbol$ adds Jacobians. It allows to compute gradients in a compositional way.
-Finally, we'll see in the next section how to compile back to a more standard language and optimize our transformation to get efficiency.
-
-The differentiation macro is given in Figure~\ref{fig:diff_macro}.
-One might be surprised that the macro does not seem to do much, and it is partly correct.
-We tried to decompose reverse-mode into more elementary small steps, and each step is slightly non trivial and dealing with one aspect of reverse-mode.
-The perhaps slightly more involved transformation is the one coming next,  going from target UNF to our target language.
+We are now able to present a simpler transformation for purely functional reverse-mode from Source UNF to Target UNF.
+The differentiation transformation is given in Figure~\ref{fig:diff_macro}.
+To a list of types, the transformation adds a higher-order type. 
+It is to be understood as the continuation for the pure storage of the gradient.
+On primitive constants of Source UNF, it returns a pair. 
+The first part is the initial term (pre-composed by a projection because it is not using the continuation variable).
+The second part returns the continuation term pre-composed by the transpose Jacobian of the primitive.
 
 \input{figures/macro_diff}
-
-Even though the target UNF is designed to achieve this purpose, we have the Following
-
-\begin{proposition}
-    If $\Gamma \vdash$!e!: !T! in source UNF, then
-    $\Dtype(Gamma)\vdash \Dsynrevsymbol$(!e!):  $\Dtype$(!T!) in target UNF.
-\end{proposition}
 
@@
 
-We give a translation from our source language to source UNF in Figure~\ref{fig:source_to_unf}.
+We give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}.
+
+TODO:redo
+
 UNF acts as a sort of stack. 
@@
 
-$\widehat{;}$ is similar to $;$, with a sort of context propagation effect. 
-It is a bit reminiscent of monadic composition versus the usual composition.
-Technically though, the transformation $\UNFSymbol$ is not quite compositional.
-\MH{explain in detail how to define it.}
-
-\begin{proposition}
-    Let $\Gamma\vdash$!e!: !T! be a term in the source language.\\
-    Then $\Gamma \vdash \UNFSymbol$(!e!): $\Gamma\times\Delta\times$!T! in source UNF for a certain type $\Delta$. 
-\end{proposition}
-
-$\Delta$ corresponds to the list of all the intermediate results from !e!. 
-A different way to say it would be that $\Delta$ is the context made of the bound variables of !e! when !e! is in ANF.
+TODO: need to explain that return values produce some kind of ANF and introduce new vars.
 

### 2021-07-06 rad-icfp-2021/correctness.tex
rewriting UNF section
writing section 5.4

@@
     \begin{tabular}{r c l}
-    $\seml \reals \semr$ &\defeq& $\RR$ \\
-    $\seml$!T1xT2!$\semr$ &\defeq& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
-    $\seml \Array{\reals}{n}\semr$ &\defeq& $\Prod_{1\leq i \leq n} \RR$ 
+    $\seml \reals \semr$ & $\defeq$& $\RR$ \\
+    $\seml$!T1xT2!$\semr$ & $\defeq$& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
+    $\seml \Array{\reals}{n}\semr$ & $\defeq$ & $\prod_{1\leq i \leq n} \RR$ 
     \end{tabular}
-A context $\Gamma=\{x_1:A_1,\ldots,x_n:A_n\}$ is interpreted as the product $\Prod_{1\leq i \leq n}\sem{A_i}$.
+A context $\Gamma=\{x_1:A_1,\ldots,x_n:A_n\}$ is interpreted as the product $\prod_{1\leq i \leq n}\sem{A_i}$.
 Well typed terms $\Gamma\vdash$!e: !$A$ are interpreted as functions $\sem{\Gamma}\to\sem{A}$. This is
@@
 
-First, we start from the correctness of $D$, which is easy to establish, 
-and then propagate this information to Source and Target.
+First, we start from the correctness of $\Dsynrevsymbol_{\rho}$:Source UNF $\to$ Target UNF, which is easy to establish, 
+and then propagate this information to Source and Target via the $\UNFSymbol, \UNFSymbol^{-1}$ transformations.
 
-\begin{proposition}[Correctness $D$]
-    TODO
+\begin{proposition}[Correctness $\Dsynrevsymbol_{\rho}$]
+    For every term !$\reals^{\times n} \vdash$ e: $\reals^{\times n+1}$! in Source UNF,\\ 
+    $\pi_2 \seml \Dsynrevsymbol_{\rho}$!e!$\semr(x_1,\ldots x_n,Id_{\RR^n})$=$J^T_{(x_1,\ldots x_n)}\seml$!e!$\semr$.
 \end{proposition}
+
+This is routinely proved by induction as the language is first-order. 
+This uses the fact that for every primitive constant $A$, $\sem{J^TA}=J^T\sem{A}$.
 
@@
 From there, the intuition for the Jacobian of a primitive in Target UNF is that it should be the Jacobian of
-the corresponding term in Target.
+the corresponding term in Target. This is easily checked for scalar operations.
 
-TODO: I don't want to talk too much about the cotangent bundle functor, but to phrase correctness properties clearly I might have to.
-
-\begin{proposition}[The transpose Jacobian of map2 and reduce are correct.]
-    TODO
+\begin{proposition}
+    \begin{center}
+\begin{tabular}{r c l}
+    !$\seml\UNFSymbol^{-1}(J^T$map2$_{x,y.e})\semr$! &=& !$J^T\seml\UNFSymbol^{-1}($map2$_{x,y.e})\semr$!\\
+    !$\seml\UNFSymbol^{-1}(J^T$reduce$_{x,y.e_1;e_2})\semr$! &=& !$J^T\seml\UNFSymbol^{-1}($reduce$_{x,y.e_1;e_2})\semr$!
+\end{tabular}
+\end{center}
 \end{proposition}
 
-From this, we can now deduce that the composite transformation doing UNF then $D$ and then $UNF^-1$ is correct
+This is proved in Appendix \ref{sub:Reverse derivative of array operations}.
+
+From this, we now deduce that the composite transformation $\UNFSymbol$, $\Dsynrevsymbol_{\rho}$, $\UNFSymbol^{-1}$ is correct
 in the sense that it produces a term that computes the gradient of the original term.
 
-\begin{theorem}[UNF;D;$UNF^-1$ correct]
+\begin{proposition}
+    If !x$_1:\reals$,$\ldots$,x$_n:\reals$ $\vdash$ e: $\reals$! then \\
+    $\seml \pi_2 \UNFSymbol^{-1}(\Dsynrevsymbol_{\rho}(\UNFSymbol($!e!$)))\semr(x_1,\ldots x_n,Id_{\RR^n})$=$J^T_{(x_1,\ldots x_n)}$!<!$Id_{\RR^n}$,$\seml$!e!$\semr$!>!
+\end{proposition}
+
+\begin{proof}
     TODO
-\end{theorem}
+\end{proof}
 
-By inspection of what that composition actually gives on terms of Source, 
-we show that the previous composition of transformation indeed gives the transformation 
+By inspecting what that composition of transformations does on terms of Source, 
+we show that this indeed gives computes the same as the transformation 
 $\directD{\rho}{\Gamma}{Y}$ from Section~\ref{sub:Macro for pure reverse mode transformation}.
 
-\begin{proposition}[UNF;D;$UNF^-1$=D]
-    TODO
+\begin{proposition}
+    $\seml\UNFSymbol^{-1}(\Dsynrevsymbol_{\rho}(\UNFSymbol($!e!$)))\semr$ = $\seml\directD{\rho}{\Gamma}{Y}($!e!$)\semr$ 
 \end{proposition}
 
-Without any more work, the previous claims have shown that 
-$\directD{\rho}{\Gamma}{Y}$ is correct.
+\begin{proof}
+    TODO
+\end{proof}
 
-\begin{theorem}[D is correct]
-    TODO
+Combining the the previous propositions, we have shown that $\directD{\rho}{\Gamma}{Y}$ is correct.
+
+\begin{theorem}
+    For every $\Gamma \vdash$ !e:!$\reals$, we have 
+    $\sem{\grad_\Gamma e}= \grad_\Gamma\sem{e}$.
 \end{theorem}

### 2021-07-06 rad-icfp-2021/complexity.tex
expanding complexity section

@@
 We follow a simple model similar to the one in \cite{griewank2008evaluating}.
-We assume the cost is divided into 4 elementary measures, being the number of MOVES, ADDS, MULTS, and NLOPS.
+We assume the cost is divided into $4$ elementary measures, being the number of MOVES, ADDS, MULTS, and NLOPS.
 MOVES assumes a flat memory and represents moving a fixed size information  (e.g. 64 bits). 
@@
 MULTS the number of multiplications, 
-and NLOPS the number of elementary non linear operations like !cos!, !exp!.
+and NLOPS the number of elementary non-linear operations like !cos! or !exp!.
 
@@
 
-\begin{itemize}
-    \item $\cost(*)=(3,0,1,0)$
-    \item $\cost(+)=(3,1,0,0)$
-    \item $\cost(c)=(1,0,0,0)$
-    \item $\cost($!sin!$)=(2,0,0,1)$
-\end{itemize}
+\begin{tabular}{ll}
+    $\cost(*)=(3,0,1,0)$ & $\cost(+)=(3,1,0,0)$\\
+    $\cost(c)=(1,0,0,0)$ & $\cost($!sin!$)=(2,0,0,1)$
+\end{tabular}
 
-TODO: extend cost array
-TODO: extend cost compositional way
+For simplicity, we will not address any parallelism in our cost model. 
+So the cost of !map (x.e)! on an array of size $n$ will be $n*\cost$(!e!).
+The cost function extends compositionally to the whole language in the obvious way.
+For instance we have !$\cost$(let x=e$_1$ in e$_2$)!= !$\cost$(e$_1$)+$\cost$(e$_2$)!.
+TODO: for let binding and same in map, do I need to add some MOVES?
 
@@
 \begin{proposition}
-    TODO: cost $\directD{\rho}{\Gamma}{Y}$. 
+    $\cost(\directD{\rho}{\Gamma}{Y}($!e!$))\leq$ $K\cost$(!e!) + O((0,0,0,n*|!e!|)) 
 \end{proposition}
 
-As we can see, the only problem is that too many variables are being moved, 
+\begin{proof}
+    TODO: routinely by induction and inspection. be specific with K? 
+    One problem still is for map and reduce because they call twice the gradient of !e$_1$!.
+\end{proof}
+
+As we can see, the only problem is that too many variables are being moved, and this is
 due to the continuation carrying and copying $n$ variables at each step.
-This is exactly the overhead that our inlining and forward substitution removes.
-
-\begin{proposition}
-    TODO: complexity gain using our optim.
-\end{proposition}
+This overhead is precisely what that our inlining and forward substitution removes, 
+as was examplified in Section~\ref{sub:Partial evaluation and optimization}.
 

### 2021-07-06 rad-icfp-2021/generalization.tex
rewriting UNF section
fixing todo generalisation

@@
 Finally, the former point implies we need a few more array manipulations.
-We write !c::[v1,$\ldots$,vn]! to be the array ![c,v1,$\ldots$,vn]!. 
-Given an array ![v1,$\ldots$,vn]! of size $n>0$, we write !let x,A=[v1,$\ldots$,vn]! to mean 
-that !x=v1! and !A=[v2,$\ldots$,vn]!. The reverse derivative of fold left is then
+These can be elegantly dealt with by changing the semantics of !scanl! and !scanr!. 
+Let's assume that they now return a pair of an array of size $n$ 
+of the intermediate computations and the final result.
+The reverse derivative of fold left is then
 
-TODO: reduce recomputed after scanl can be fixed here, explain let v,B notation somewhere
 \begin{center}
 \begin{tabular}{r c l}
-    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !foldl (x,y.e1) e2 e3!) &=&
-            !let v,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
-            && !let A,Y2 = !$\directD{\rho}{\Gamma,v}{Y1}$!(e3) in! \\
-            && !let A0 = (scanl (x,y.e1) v A)[1..n] in! \\
-            && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
-            && !let A2 = map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A! \\
-            && !let A3 = scanr * 1 A1! \\
-            && !<reduce (x,y.e1) v A, fun (x1,$\ldots$,xn,z) ->! \\
-            && !let y1,B = map2 (x,y. x*y*z) (1::A2) A3 in! \\
-            && !Y2(x1,$\ldots$,xn,y1,B)>! \\
+    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !foldl (x,y.e$_1$) e$_2$ e$_3$!) &=&
+            !let v,Y$_1$ = $\directD{\rho}{\Gamma}{Y}$(e$_2$) in! \\
+            && !let A,Y$_2$ = $\directD{\rho}{\Gamma,v}{Y_1}$(e$_3$) in! \\
+            && !let A$_0$,r$_1$ = (scanl (x,y.e$_1$) v A) in! \\
+            && !let A$_1$ = map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x,b/y]) A$_0$ A! \\
+            && !let A$_2$ = map2 (a,b.(!$\grad_{\{y\}}$!)e$_1$[a/x,b/y]) A$_0$ A! \\
+            && !let r$_2$, A$_3$ = scanr * 1 A$_1$! \\
+            && !<r$_1$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
+            && !let y$_1$,B = r$_2$*z ,map2 (x,y. x*y*z) A$_2$ A$_3$ in! \\
+            && !Y$_2$(x$_1$,$\ldots$,x$_n$,y$_1$,B)>! \\
 \end{tabular}
@@
 
-TODO: other examples/recipe
+Another useful example is !map!, which is very similar to !map2!.
+
+\begin{center}
+    \begin{tabular}{r c l}
+        $\directD{\rho}{\Gamma}{Y}$(!map (x.e$_{1}$) e$_{2}$!) &=&  
+            !let A,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
+            && !let G=!$\grad_{\Gamma}$!e$_{1}$ in!\\
+            && !<map (x.e$_{1}$) A, fun (x$_{1}$,$\ldots$,x$_n$,Z) -> !\\
+            && !Y$_{1}$( (x$_{1}$,$\ldots$,x$_n$)!$\widehat{+}$!(G*(reduce + 0 Z)),!\\
+            && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x]*b) A Z )>!\\
+    \end{tabular}
+\end{center}
 
@@
 
-With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldotstimes An$ as expected.
+With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldots\times An$ as expected.
 TODO: define *, finish D(reduce)

### 2021-07-06 rad-icfp-2021/generalization.tex
finishing gen section

@@
         $\directD{\rho}{\Gamma}{Y}$(!map (x.e$_{1}$) e$_{2}$!) &=&  
-            !let A,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
-            && !let G=!$\grad_{\Gamma}$!e$_{1}$ in!\\
+            !let A,Y$_{1}$ = $\directD{\rho}{\Gamma}{Y}$(e$_{2}$) in! \\
+            && !let G=$\grad_{\Gamma}$ e$_{1}$ in!\\
             && !<map (x.e$_{1}$) A, fun (x$_{1}$,$\ldots$,x$_n$,Z) -> !\\
-            && !Y$_{1}$( (x$_{1}$,$\ldots$,x$_n$)!$\widehat{+}$!(G*(reduce + 0 Z)),!\\
-            && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x]*b) A Z )>!\\
+            && !Y$_{1}$( (x$_{1}$,$\ldots$,x$_n$)$\widehat{+}$(G*(reduce + 0 Z)),!\\
+            && \quad\quad !map2 (a,b.($\grad_{\{x\}}$e$_{1}$)[a/x]*b) A Z )>!\\
     \end{tabular}
@@
 
-We now show how to generalise our reverse-mode transformation to be defined on arrays over any ground type $G$.
+We now show how to generalize our reverse-mode transformation to be defined on arrays over any ground type $G$.
 That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.
@@
 With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldots\times An$ as expected.
-TODO: define *, finish D(reduce)
-
-We will need a new primitive $\widehat{*}$ whose denotation is this map.
-
+For !map2!, we need a generalization of $*:\reals\to\reals$. 
+If !e1: A!, then $\grad_{\{x\}}$!e1: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
+If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.
+Similarly, for !A$_{3}$! in !reduce! we need a new primitive $\widetilde{*}:(A\otimes A)\times(A\otimes A) \to (A\otimes A)$.
+$\widetilde{*}$ corresponds to matrix-matrix multiplication.
 Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!.
@@
 \begin{tabular}{r c l}
-        $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
-        &=&  !let y1,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
-        && !let A,Y2 = !$\directD{\rho}{\Gamma,y1}{Y1}$!(e3) in! \\
-        && !let A0 = shift1R (scanl (x,y.e1) y1 A) in! \\
-        && !let A1 = shiftL (map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A) in! \\
-        && !let A2 = map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A in! \\
-        && !let A3 = scanr * 1 A1 in! \\
-        && !<reduce (x,y.e1) y1 A, fun (x1,$\ldots$,xn,z) ->! \\
-        && !Y((x1,$\ldots$,xn),0,map2 (a,b. a*b*z) A2 A3)!
+    $\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e$_{1}$) e$_{2}$ e$_{3}$!) &=&
+    !let y$_{1}$,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
+    && !let A,Y$_{2}$ = !$\directD{\rho}{\Gamma,y_1}{Y_1}$!(e$_{3}$) in! \\
+    && !let A$_{0}$ = shift1R (scanl (x,y.e$_{1}$) y$_{1}$ A) in! \\
+    && !let A$_{1}$ = shift1L (map2 (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A$_{0}$ A) in! \\
+    && !let A$_{2}$ = map2 (a,b.(!$\grad_{\{y\}}$!)e$_{1}$[a/x,b/y]) A$_{0}$ A in! \\
+    && !let A$_{3}$ = scanr $\widetilde{*}$ 1 A$_{1}$ in! \\
+    && !<reduce (x,y.e$_{1}$) y$_{1}$ A, fun (x$_{1}$,$\ldots$,x$_n$,z) ->! \\
+    && !Y$_{2}$(x$_{1}$,$\ldots$,x$_n$, map2 (x,y. x$\widehat{*}$(y$\widehat{*}$z)) A$_{2}$ A$_{3}$>! \\
 \end{tabular}

### 2021-07-06 rad-icfp-2021/complexity.tex
improving complexity section

@@
 that $\directD{\rho}{\Gamma}{Y}$ verifies the cheap gradient principle.
-However, this is not quite the case with $\directD{\rho}{\Gamma}{Y}$ 
-before the optimization (inlining and forward substitution) step from Section~\ref{sub:Partial evaluation and optimization}.
 
-\begin{proposition}
-    $\cost(\directD{\rho}{\Gamma}{Y}($!e!$))\leq$ $K\cost$(!e!) + O((0,0,0,n*|!e!|)) 
-\end{proposition}
+There is a first restriction that prevents our transformation from satisfying the cheap gradient principle.
+If we try to show that the cheap gradient principle holds by induction on the term, it fails on the !map2!.
+When differentiating !map2!, there are two series of calls to (sub)gradients of !e$_1$!. 
+The induction hypothesis is then too weak to conclude. 
+The problem comes from the fact that !e$_1$! could itself use !map2!. 
+So the constant $K$ can be independant of $n$ and of 
+the size of the term if we allow it to be dependent on the level of nestings of !map2!.
+A similar phenomenon happens with !reduce!.
 
-\begin{proof}
-    TODO: routinely by induction and inspection. be specific with K? 
-    One problem still is for map and reduce because they call twice the gradient of !e$_1$!.
-\end{proof}
+Another problem is that the continuation part of $\directD{\rho}{\Gamma}{Y}$ adds $O(n)$ MOVES at each step.
+This overhead is precisely what that our inlining and forward substitution removes, 
+as was exemplified in Section~\ref{sub:Partial evaluation and optimization}.
 
-As we can see, the only problem is that too many variables are being moved, and this is
-due to the continuation carrying and copying $n$ variables at each step.
-This overhead is precisely what that our inlining and forward substitution removes, 
-as was examplified in Section~\ref{sub:Partial evaluation and optimization}.
-
-Combining the two previous points, we conclude that our reverse mode transformation satisfies the cheap gradient principle.
+Combining the two previous points, we conclude that our reverse mode transformation satisfies a version of the cheap gradient principle.
 
 \begin{theorem}
-    TODO: The reverse mode transformation satisfies the cheap gradient principle.
+    Given a term !$\Gamma \vdash$ e: $\reals$! in Source with at most $p$ nested !map2! or !reduce!. 
+    Denote by $G$ the term $\grad_\Gamma$!e! after inlining and forward substitution.
+    Then !$\cost$(G) $\leq$ $4*2^{p}*\cost$(e)!.
 \end{theorem}
+
+\begin{proof}[notes]
+    The proof is by routine induction on !e!. 
+    First, one computes the cost of $\directD{\rho}{\Gamma}{Y}$(!e!).
+    It has $N$ too many MOVES which are removed by inlining and forward substitution using the invariant that at each step, 
+    at most 2 !x!$_i$ in the continuation are not variables.
+\end{proof}
 

### 2021-07-06 rad-icfp-2021/intro.tex
.

@@
 
-\subsection{Motivation}
+% \subsection{Motivation}
+Deep learning is moving towards increasingly sophisticated optimization objectives that employ tensors and operations on tensors.
+Reverse-mode Automatic Differentiation (AD) is a technique to automatically compute the gradient of objective functions of form $\RR^n\to\RR$.
+Such functions appear a lot in practice: for instance, as loss functions in machine learning.
 
-In 2008, Pearlmutter and Siskind~\cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting, where they defined reverse-mode. 
-Reverse-mode computes in a single fast pass the whole gradient of a function $\RR^n\to\RR$.
-By comparison, the easier forward-mode transformation would require $n$ passes. 
-These functions $\RR^n\to\RR$ appear a lot in practice: for instance, as loss functions in machine learning.
-Reverse-mode is thus important, as it computes the whole gradient of such functions very quickly. 
-However, in order to reach the efficiency of the usual imperative version of reverse-mode, the transformation in~\cite{pearlmutter2008reverse} resorts to references.
-
+In order to reach the efficiency of the usual imperative version of reverse-mode, the transformations even in functional languages~\cite{pearlmutter2008reverse} resort to references.
 The lack of purity in reverse-mode is known to make it significantly harder to optimize and parallelize. 
 None of the current implementations of reverse-mode in a functional setting~\cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.~\cite{xla}.
-
-As a result, to optimize for efficiency, a hand-crafted reverse-derivative must be given, for every important non trivial operation.
+As a result, to optimize for efficiency, a hand-crafted reverse-derivative must be given, for every important non-trivial operation.
 In other words, abstracting away from imperative code is still a hurdle that functional implementations need to overcome.
 
-An efficient purely functional reverse-mode transformation would be more easily explainable, optimizable, and reusable.
+% An efficient purely functional reverse-mode transformation would be more easily explainable, optimizable, and reusable.
 
-TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
-all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g.
+% TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
+% all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g.
 in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress.
-For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.
+% For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.
 
-\subsection{Problem}
+% \subsection{Problem}
 
-Following \cite{pearlmutter2008reverse}, it is relatively easy to define a purely functional reverse-mode on a first-order language using continuations. 
-Yet, as remarked by the authors, this is highly inefficient. 
-The problem we tackle in this paper is how to go from a simple and general setting, where we can relatively easily prove correctness of reverse-mode, 
-to a representation that leads to more efficient implementations.
+% Following \cite{pearlmutter2008reverse}, it is relatively easy to define a purely functional reverse-mode on a first-order language using continuations. 
+% Yet, as remarked by the authors, this is highly inefficient. 
+% The problem we tackle in this paper is how to go from a simple and general setting, where we can relatively easily prove correctness of reverse-mode, 
+% to a representation that leads to more efficient implementations.
+
+In this paper, we define a purely functional language with efficient and provably correct reverse-mode AD.
+To do so, we go start from a simple and general representation, where we can relatively easily prove correctness of reverse-mode.
+Then, we define a representation that leads to more efficient implementations.
 
@@
 \end{itemize}
+
+% \begin{table}
+% \begin{tabular}{|l|c|c|c|c|c|c|c|c|c|}
+% \hline
+%  & \rot{Reverse Mode}  & \rot{Complexity Proofs} & \rot{Pure} & \rot{Correctness Proofs} & \rot{Tensor Support}  & \rot{Recursion} & \rot{Conditional} \\
+% \hline
+% This Paper &
+% \supfull & \supfull & \supfull  & \supfull  & \supfull  & \suphalf & \supfull \\ \hline
+% Lantern~\cite{lantern_icfp} & 
+% \supfull & \suphalf & \supnone  & \supnone & \supfull & ? & \supfull \\ \hline
+% \dfsmooth{}~\cite{shaikhha2019efficient} & 
+% \supnone & \supnone & \supfull  & \supnone & \supfull  & \suphalf  & \supnone \\ \hline
+% \cite{huot2020correctness} &
+% \supfull & \supnone & \supfull   & \supfull & \supnone & \suphalf & \supfull \\ \hline
+% \cite{brunel2019backpropagation} &
+% \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & ? \\ \hline
+% \cite{abadi2019simple} &
+% \supfull & ? & \supfull & \supfull & \supnone & \supfull & \supfull \\ \hline
+% \cite{barthe2020versatility} &
+% \supnone & \supnone & \supfull & \supfull & \supnone & \supnone & \supfull \\ \hline
+% \cite{pearlmutter2008reverse} &
+% \supfull & \suphalf & \supnone & \supnone & ? & ? & ? \\ \hline
+% \cite{Elliott:2018:SEA:3243631.3236765} &
+% \supfull & \suphalf & \supfull & \supnone & ? & \supnone & ? \\ \hline
+% \end{tabular}
+% \caption{Comparison of different functional differentiable programming frameworks.}
+% \end{table}
+
+\begin{table}
+ \label{fig:comparison-table}
+ \begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+ \hline
+  & \rot{Reverse Mode}  & \rot{Complexity Proofs} & \rot{Pure} & \rot{Correctness Proofs} & \rot{Tensor Support}  & \rot{Recursion} & \rot{Conditional} \\
+  % & Reverse  & Complexity & Pure & Correct & HO Fun. & Tensors & Rec. & Cond. \\
+ \hline
+ \system{} (This Paper) &
+ \supfull & \supfull & \supfull & \supfull  & \supfull  & \supfull & \supfull \\ 
+ \hline
+ Lantern~\cite{lantern_icfp} & 
+ \supfull & \suphalf & \supnone & \supnone & \supfull & \supfull & \supfull \\ 
+ \hline
+ \dfsmooth{}~\cite{shaikhha2019efficient} 
+ & 
+ \supnone & \supnone & \supfull & \supnone & \supfull  & \supnone  & \supnone \\ 
+ \hline
+ \cite{huot2020correctness} &
+ \supfull & \supnone & \supfull & \supfull & \supnone & \suphalf & \supfull \\ 
+ \hline
+ \cite{brunel2019backpropagation} &
+ \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
+ \hline
+ \cite{abadi2019simple} &
+ \supfull & \supfull? & \supfull & \supfull & \supnone? & \supfull & \supfull \\ 
+ \hline
+ \cite{barthe2020versatility} &
+ \supnone & \supnone & \supfull & \supfull  & \supnone & \supnone & \supfull \\ 
+ \hline
+ \cite{pearlmutter2008reverse} &
+ \supfull & \supnone  & \supnone & \supnone & \supnone? & \supfull & \supfull \\ 
+ \hline
+ \cite{Elliott:2018:SEA:3243631.3236765} &
+ \supfull & \suphalf & \supfull & \supnone & \supnone? & \supnone & \supnone? \\ 
+ \hline
+ \cite{sherman2021} & 
+ \suphalf or \supnone? & \supnone & \supfull & \supfull? & \supnone? & \supnone & \supnone \\ 
+ \hline
+ \cite{vytiniotis2019differentiable} &
+ \supfull & \suphalf & \supfull & \supnone & \supnone? & \supnone & \supnone? \\ 
+ \hline
+ \cite{mak2020differential} & 
+ \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
+ \hline
+ \cite{vakar2020reverse} & 
+ \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
+ \hline
+ \cite{Manzyuk2012} & 
+ \supnone & \supnone & \supfull & \supfull? & \supnone & \supnone & \supnone \\ 
+ \hline 
+ \cite{gallagher-sdg} & 
+ \supnone & \supnone & \supfull & \supfull ? & \supnone & \supfull & \supfull  \\ 
+ \hline
+ \end{tabular}
+ \caption{Comparison of different differentiable programming frameworks.
+ $\supfull$ means the property is verified, and $\supnone$ means its absent from the work.}
+ \end{table}
 

### 2021-07-06 rad-icfp-2021/intro.tex
.

@@
 
-In this paper, we define a purely functional language with efficient and provably correct reverse-mode AD.
+In this paper, we define a compilation scheme for a purely functional, provably correct, and efficient reverse-mode AD. 
 To do so, we go start from a simple and general representation, where we can relatively easily prove correctness of reverse-mode.
-Then, we define a representation that leads to more efficient implementations.
+Then, we define the UNF representation inspired by SSA~\cite{cytron1989efficient} and ANF, that leads to more efficient implementations.
 
-This raises several issues that we deal with throughout the paper.
-First, there is a tradeoff to reach between a general expressive language and a domain specific one.
-The latter usually has more static information and a specific representation that lends itself to better optimizations.
-Then, many optimizations performed on AD implementations consist in hand-crafted derivatives for useful operations like matrix-matrix multiplication, dot-product, etc.
-They don't seem to arise from theoretical justifications, are error prone, and can hardly fit with more general optimizations.
-This makes these systems harder to prove correct.
-The problem is thus to ensure provable correctness and pureness, while not compromising on efficiency.
-In addition, we would like something easily provably efficient.
-A real-world implementation based on our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on solid grounds.
-
-TODO: same as above, maybe rewrite for more theory audience.
-
-\subsection{Summary of contributions}
-
-We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
-Our transformation is comprised of a compilation scheme through a novel intermediate representation (IR): unary form (UNF).
-To emphasize the key ideas and for expository purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
-and show how to extend our work to a richer language in Section~\ref{sec:generalization}.
-
-We summarize the key ideas of our work. Reverse-mode can be made efficient
-\begin{itemize}
-   \item easily if there are only unary operators
-   \item in a purely functional way
-   \item on expressive constructs such as second-order array operators (e.g. !map2, reduce!) 
-   \item via standard functional optimization techniques
-\end{itemize}
-
-Our contribution makes these claims precise. In more details, we introduce
-\begin{itemize}
-    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
-    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
-    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
-    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
-    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
-    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
-    \item a denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
-    \item correctness of the transformations (\S\ref{sec:correctness})
-    \item complexity guarantees (\S\ref{sec:complexity})
-    \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
-    \item several additions to the source language (\S\ref{sec:generalization}) TODO:should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.
-\end{itemize}
-
-% \begin{table}
-% \begin{tabular}{|l|c|c|c|c|c|c|c|c|c|}
-% \hline
-%  & \rot{Reverse Mode}  & \rot{Complexity Proofs} & \rot{Pure} & \rot{Correctness Proofs} & \rot{Tensor Support}  & \rot{Recursion} & \rot{Conditional} \\
-% \hline
-% This Paper &
-% \supfull & \supfull & \supfull  & \supfull  & \supfull  & \suphalf & \supfull \\ \hline
-% Lantern~\cite{lantern_icfp} & 
-% \supfull & \suphalf & \supnone  & \supnone & \supfull & ? & \supfull \\ \hline
-% \dfsmooth{}~\cite{shaikhha2019efficient} & 
-% \supnone & \supnone & \supfull  & \supnone & \supfull  & \suphalf  & \supnone \\ \hline
-% \cite{huot2020correctness} &
-% \supfull & \supnone & \supfull   & \supfull & \supnone & \suphalf & \supfull \\ \hline
-% \cite{brunel2019backpropagation} &
-% \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & ? \\ \hline
-% \cite{abadi2019simple} &
-% \supfull & ? & \supfull & \supfull & \supnone & \supfull & \supfull \\ \hline
-% \cite{barthe2020versatility} &
-% \supnone & \supnone & \supfull & \supfull & \supnone & \supnone & \supfull \\ \hline
-% \cite{pearlmutter2008reverse} &
-% \supfull & \suphalf & \supnone & \supnone & ? & ? & ? \\ \hline
-% \cite{Elliott:2018:SEA:3243631.3236765} &
-% \supfull & \suphalf & \supfull & \supnone & ? & \supnone & ? \\ \hline
-% \end{tabular}
-% \caption{Comparison of different functional differentiable programming frameworks.}
-% \end{table}
-
-\begin{table}
- \label{fig:comparison-table}
- \begin{tabular}{|l|c|c|c|c|c|c|c|c|}
- \hline
-  & \rot{Reverse Mode}  & \rot{Complexity Proofs} & \rot{Pure} & \rot{Correctness Proofs} & \rot{Tensor Support}  & \rot{Recursion} & \rot{Conditional} \\
-  % & Reverse  & Complexity & Pure & Correct & HO Fun. & Tensors & Rec. & Cond. \\
- \hline
- \system{} (This Paper) &
- \supfull & \supfull & \supfull & \supfull  & \supfull  & \supfull & \supfull \\ 
- \hline
- Lantern~\cite{lantern_icfp} & 
- \supfull & \suphalf & \supnone & \supnone & \supfull & \supfull & \supfull \\ 
- \hline
- \dfsmooth{}~\cite{shaikhha2019efficient} 
- & 
- \supnone & \supnone & \supfull & \supnone & \supfull  & \supnone  & \supnone \\ 
- \hline
- \cite{huot2020correctness} &
- \supfull & \supnone & \supfull & \supfull & \supnone & \suphalf & \supfull \\ 
- \hline
- \cite{brunel2019backpropagation} &
- \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
- \hline
- \cite{abadi2019simple} &
- \supfull & \supfull? & \supfull & \supfull & \supnone? & \supfull & \supfull \\ 
- \hline
- \cite{barthe2020versatility} &
- \supnone & \supnone & \supfull & \supfull  & \supnone & \supnone & \supfull \\ 
- \hline
- \cite{pearlmutter2008reverse} &
- \supfull & \supnone  & \supnone & \supnone & \supnone? & \supfull & \supfull \\ 
- \hline
- \cite{Elliott:2018:SEA:3243631.3236765} &
- \supfull & \suphalf & \supfull & \supnone & \supnone? & \supnone & \supnone? \\ 
- \hline
- \cite{sherman2021} & 
- \suphalf or \supnone? & \supnone & \supfull & \supfull? & \supnone? & \supnone & \supnone \\ 
- \hline
- \cite{vytiniotis2019differentiable} &
- \supfull & \suphalf & \supfull & \supnone & \supnone? & \supnone & \supnone? \\ 
- \hline
- \cite{mak2020differential} & 
- \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
- \hline
- \cite{vakar2020reverse} & 
- \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
- \hline
- \cite{Manzyuk2012} & 
- \supnone & \supnone & \supfull & \supfull? & \supnone & \supnone & \supnone \\ 
- \hline 
- \cite{gallagher-sdg} & 
- \supnone & \supnone & \supfull & \supfull ? & \supnone & \supfull & \supfull  \\ 
- \hline
- \end{tabular}
- \caption{Comparison of different differentiable programming frameworks.
- $\supfull$ means the property is verified, and $\supnone$ means its absent from the work.}
- \end{table}
+\input{figures/related_work_table}
 
@@
 
-We introduce the general idea of efficient reverse-mode in a functional setting on some examples.
-We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
+We introduce the general idea of efficient reverse-mode in a functional setting the next examples.
+% We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
 
 \begin{example}[First-order term]
-On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
+% On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
 Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
-After our (inefficient) reverse-mode transformation, we obtain
+After an (inefficient) reverse-mode transformation, we obtain:
 \begin{center}
@@
 The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm. 
-
-After some general partial evaluation techniques that will be detailed further in the paper, we obtain
+After some general partial evaluation techniques that will be detailed further in the paper, we obtain:
 
@@
     \end{center}   
-For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
+For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.
 This term can be further optimized via constant propagation and algebraic simplifications to give
@@
 \end{center}
-where !reduce! is a known fold left operator for which the function argument is associative. 
-It is notably faster to execute than a fold left as it parallel friendly.
+where !reduce! is a known fold-left operator for which the function argument is associative. 
+It is notably faster to execute than a fold-left as it is parallel friendly.
 
-The gradient of each of these construct w.r.t. !A! will be:
+The gradient of each of these expressions w.r.t. !A! will be:
 \begin{center}
@@
 \end{center}
-where !scanl! and !scanr! are respectively scan left and scan right. 
-These return all the intermediate result of a fold in an array.
-!shift1L [v$_1$,$\ldots$,v$_n$] \defeq [v$_2$,$\ldots$,v$_{n}$]! and 
-!shift1R [v$_1$,$\ldots$,v$_n$] \defeq [v$_1$,$\ldots$,v$_{n-1}$]! are shift operators.
+where:
+\begin{itemize}
+\item !scanl! is the scan-left operator that returns all the intermediate results of fold-left,
+\item !scanr! is the scan-right operator that returns all the intermediate results of fold-right,
+\item !shift1L [v$_1$,$\ldots$,v$_n$]! is the shift-left operator and returns ![v$_2$,$\ldots$,v$_{n}$]!, and 
+\item !shift1R [v$_1$,$\ldots$,v$_n$]! is the shift-right operator and returns ![v$_1$,$\ldots$,v$_{n-1}$]!.
+\end{itemize}
 These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
@@
 
-\subsection{Outline of the paper}
-
-Our main transformation is present on the top of the diagram and presented in Section~\ref{sec:simplediff}. 
-It it decomposed 3 steps which are presented in Section~\ref{sec:unf}.
 \input{figures/compilation_scheme}
 
-In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
-Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
-In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
-We introduce transformations from the source language to UNF and from UNF to the target language.
-Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
-In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimizations.
-In Section~\ref{sec:generalization} we show how to adapt our work for a richer source language. 
-Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.
+\subsection{Contributions}
+We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
+Our transformation is comprised of a compilation scheme that is outlined in Figure~\ref{fig:outline}.
+We make the following contributions:
+
+\begin{itemize}[leftmargin=*]
+\item We present our work with a simple yet expressive array-based language (with constructs such as !map2! and !reduce!) in Section~\ref{sec:simplediff}. 
+We show how to directly compute an efficient reverse-mode AD for the expressions of this program (top of Figure~\ref{fig:outline}).  
+Furthermore, we show how to extend our work to a richer language in Section~\ref{sec:generalization}.
+\item One of the key insights behind efficient reverse-mode AD is to only consider unary operators. 
+Inspired by this insight and following IRs such as SSA and ANF, we introduce a novel IR, UNF (Section~\ref{sec:unf})).
+We introduce an alternative and easier to follow compilation pipeline for efficient reverse-mode AD (bottom of Figure~\ref{fig:outline}).
+\item We proof the correctness of our transformations (top/bottom parts of Figure~\ref{fig:outline}) by defining the denotational semantics of our languages using multicategories and concategories (Section~\ref{sec:correctness}).
+\item We prove the complexity guarantees for the reverse ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
+\end{itemize}
+
+Next, we recall rudiments of automatic differentiation, forward and reverse-mode differentiation
+
+% \subsection{Summary of contributions}
+
+% We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
+% Our transformation is comprised of a compilation scheme through a novel intermediate representation (IR): unary form (UNF).
+% To emphasize the key ideas and for expository purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
+% and show how to extend our work to a richer language in Section~\ref{sec:generalization}.
+
+% We summarize the key ideas of our work. Reverse-mode can be made efficient
+% \begin{itemize}
+%    \item easily if there are only unary operators
+%    \item in a purely functional way
+%    \item on expressive constructs such as second-order array operators (e.g. !map2, reduce!) 
+%    \item via standard functional optimization techniques
+% \end{itemize}
+
+% Our contribution makes these claims precise. In more details, we introduce
+% \begin{itemize}
+%     \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
+%     \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
+%     \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
+%     \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
+%     \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
+%     \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
+%     \item a denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
+%     \item correctness of the transformations (\S\ref{sec:correctness})
+%     \item complexity guarantees (\S\ref{sec:complexity})
+%     \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
+%     \item several additions to the source language (\S\ref{sec:generalization}) TODO:should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.
+% \end{itemize}
+
+% \subsection{Outline of the paper}
+
+% Our main transformation is present on the top of the diagram and presented in Section~\ref{sec:simplediff}. 
+% It it decomposed 3 steps which are presented in Section~\ref{sec:unf}.
+
+% \input{figures/compilation_scheme}
+
+% In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
+% Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
+% In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
+% We introduce transformations from the source language to UNF and from UNF to the target language.
+% Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
+% In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimizations.
+% In Section~\ref{sec:generalization} we show how to adapt our work for a richer source language. 
+% Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

### 2021-07-06 rad-icfp-2021/relwork.tex
.

@@
 \label{sec:discussion_and_future_work}
+
+\noindent \textbf{Design space}
+First, there is a tradeoff to reach between a general expressive language and a domain specific one.
+The latter usually has more static information and a specific representation that lends itself to better optimizations.
+Then, many optimizations performed on AD implementations consist in hand-crafted derivatives for useful operations like matrix-matrix multiplication, dot-product, etc.
+They don't seem to arise from theoretical justifications, are error prone, and can hardly fit with more general optimizations.
+This makes these systems harder to prove correct.
+The problem is thus to ensure provable correctness and pureness, while not compromising on efficiency.
+In addition, we would like something easily provably efficient.
+A real-world implementation based on our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on solid grounds.
 

### 2021-07-06 rad-icfp-2021/unf.tex

@@
 
-We give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}.
+We first give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}, which we call $\UNFSymbol$.
+As discussed above, $\UNFSymbol$ sends a constant or an operator to a primitive in Source UNF. 
+This term in Source UNF carries the context of the original term in Source.
+$\UNFSymbol$ sequentializes a term, mimicking a left-to-right call-by-value evaluation.
+Because of this, we we're trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
+!e!$_{1} \pcomp$ !e!$_{2}$  is used to pass the result !e!$_{1}$ which is at the last position in the tuple through !e!$_{2}$.
+It required when the coming operator represents a non-unary primitive such as $+$ or !pair!.
+We want to preserve the invariant that $\UNFSymbol$(!e!) represents the term !e! which is also returning its context. 
+Because !x! is not free in $\Gamma \vdash$ !let x=e$_1$in e$_2$!, we need to hide !x! after $\UNFSymbol$(!e$_2$!).
+This explain the projection !proj! in $\UNFSymbol$ of a !let!. 
+More generally, in !e!$_{1} \comp$ !e!$_{2}$ the return value of !e!$_{1}$ acts as a new variable for !e!$_{2}$, 
+is not in the context of !e!$_{1}$, but will be returned by !e!$_{2}$.
+This means we need to hide all these intermediate values to preserve the invariant,
+and this explains the projection !proj! in the $\UNFSymbol$ of other terms.
 
-TODO:redo
-
-UNF acts as a sort of stack. 
-The input represents the current stage of the stack, and we return the new value added to the stack.
-All results are kept, so the stack keeps growing. One might be surprised that the transformation for pairs and let is similar.
-The order of UNF really follows what the trace of doing a sequential call-by-value evaluation would look like if the return value does not matter.
-In other words, when evaluating !let x= e1 in e2! in call by value, we first evaluate !e1! and then !e2!. 
-Similarly in !<e1, e2>!, we evaluate !e1! and then !e2!. 
-Our ultimate goal is to return a gradient, not the original term, so what really matters is the order of evaluation, and what is evaluated.
-
-TODO: need to explain that return values produce some kind of ANF and introduce new vars.
+TODO: we could merge everything of the form op;proj into just op by changing the type of op.
 
@@
 
-The transformation from the target UNF to the target language is presented in Figure~\ref{fig:unf_to_target}.
+Next, the transformation from Target UNF to Target is presented in Figure~\ref{fig:unf_to_target}. 
+We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
+Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some kind of ANF. 
+TODO: finish
 

### 2021-07-07 rad-icfp-2021/unf.tex
fixing unf

@@
 The array operators !map2! and !reduce! have extra indices that represent well-formed terms in Source.
-The language has a second kind of sequential composition, which we call passing sequential compsition $\pcomp$.
-In !e$_{1}$ $\pcomp$ e$_{2}$! the idea is first that !e$_{1}$! is computed, followed by !e$_{2}$!. 
-However, !e$_{1}$! outputs an argument which is not used by !e$_{2}$! and is just passed by !e$_{2}$! 
-along and returned with the output of !e$_{2}$!.
+
+% The language has a second kind of sequential composition, which we call passing sequential compsition $\pcomp$.
+% In !e$_{1}$ $\pcomp$ e$_{2}$! the idea is first that !e$_{1}$! is computed, followed by !e$_{2}$!. 
+% However, !e$_{1}$! outputs an argument which is not used by !e$_{2}$! and is just passed by !e$_{2}$! 
+% along and returned with the output of !e$_{2}$!.
+TODO: redo
+
 The language also contains a projection !proj! which forgets some of the elements, 
@@
 Because of this, we we're trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
-!e!$_{1} \pcomp$ !e!$_{2}$  is used to pass the result !e!$_{1}$ which is at the last position in the tuple through !e!$_{2}$.
-It required when the coming operator represents a non-unary primitive such as $+$ or !pair!.
+
+% !e!$_{1} \pcomp$ !e!$_{2}$  is used to pass the result !e!$_{1}$ which is at the last position in the tuple through !e!$_{2}$.
+% It required when the coming operator represents a non-unary primitive such as $+$ or !pair!.
+TODO:redo
+
 We want to preserve the invariant that $\UNFSymbol$(!e!) represents the term !e! which is also returning its context. 
@@
 Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some kind of ANF. 
-TODO: finish
+TODO: finish. explain what is happening. explain carefully the context story. then alpha rename the Figure to make it correct.
 

### 2021-07-07 rad-icfp-2021/unf.tex
fixing unf

--- 2021-07-06 rad-icfp-2021/unf.tex
@@
 The array operators !map2! and !reduce! have extra indices that represent well-formed terms in Source.
-The language has a second kind of sequential composition, which we call passing sequential compsition $\pcomp$.
-In !e$_{1}$ $\pcomp$ e$_{2}$! the idea is first that !e$_{1}$! is computed, followed by !e$_{2}$!. 
-However, !e$_{1}$! outputs an argument which is not used by !e$_{2}$! and is just passed by !e$_{2}$! 
-along and returned with the output of !e$_{2}$!.
 The language also contains a projection !proj! which forgets some of the elements, 
@@
 In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
-Lastly, Target UNF has an internal composition $\icomp$. 
-!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!
+Given a type !T$\defeq$ [A$_1$,$\ldots$,A$_n$]!, we write \underline{T} for the type !A$_1$ $\times\ldots\times$ A$_n$! of Target. 
+Target UNF has an internal composition $\icomp$. 
+!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!.
 It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
@@
 Because of this, we we're trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
-!e!$_{1} \pcomp$ !e!$_{2}$  is used to pass the result !e!$_{1}$ which is at the last position in the tuple through !e!$_{2}$.
-It required when the coming operator represents a non-unary primitive such as $+$ or !pair!.
+When we need to pass the result of type !A! of !e$_{1}$! through !e$_{2}$! which does not take it as an input, we can transform !e$_{2}$! to accept and pass it.
+It's the equivalent in UNF of a weakening.
+This weakening is required when we transform a non-unary operator such as $+$ or !pair!, as we transform the arguments in order and we need to keep all of their results.
+We write this new term $(\widetilde{e_{2}})_{A}$ and it defined by induction on !e$_{2}$! as follows. 
+We assume given a type !A! that needs to be passed, and drop the index !A!.
+
+\begin{tabular}{r c l}
+    $\widetilde{var_{T;i}}$  &=& $var_{T,A;i}$ \\
+    $\widetilde{op_{T;n}}$  &=& $op_{T,A;n}$ \\§
+    $\widetilde{pair_{T;B\times C}}$  &=& $pair_{T,A;B\times C}$ \\
+    $\widetilde{proj_{T_1;T_2;T_3}}$  &=& $proj_{T_1;T_2;T_3,A}$ \\
+    $\widetilde{e_1 \comp e_2}$  &=& $\widetilde{e_1}\comp\widetilde{e_2}$ \\
+    $\widetilde{map2_{T;x,y.e}}$  &=& $map2_{T,A;x,y.e}$ \\
+    $\widetilde{reduce_{T;x,y.e_1;e_2}}$ &=& $reduce_{T,A;x,y.e_1;e_2}$ \\
+\end{tabular}
+
 We want to preserve the invariant that $\UNFSymbol$(!e!) represents the term !e! which is also returning its context. 
@@
 
-TODO: we could merge everything of the form op;proj into just op by changing the type of op.
+\begin{lemma}[Well typedness $\UNFSymbol$]
+    If !$\Gamma$=x$_1$:A$_{1}$,$\ldots$,x$_n$:A$_{n}$ $\vdash$ e:B! then
+   !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ $\UNFSymbol$(e): A$_{1}$,$\ldots$,A$_{n}$,B!.
+\end{lemma}
+
+% TODO: we could merge everything of the form op;proj into just op by changing the type of op.
 
@@
 We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
-Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some kind of ANF. 
-TODO: finish
+Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of ANF transformation. 
+$\UNFSymbol^{-1}$ does not preserve a typing property as nicely as $\UNFSymbol$, 
+because it treats primitives with and without a $J$ differently.
+
+Target UNF does not have variables. 
+A context type !T=[A$_1$,$\ldots$,A$_n$]! is transformed to a context !$\Gamma$=x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$!.
+We use this convention when definiting $\UNFSymbol^{-1}$. 
+For operators like !op$_{\Gamma;m}$!, we extend the notation above by saying that the canonical variables are 
+!x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$,x$_{n+1}$:$\reals$,$\ldots$,x$_{n+m}$:$\reals$!.
 

### 2021-07-07 rad-icfp-2021/purerad.tex
rewriting UNF section
.

@@
 \subsection{Source Language}
-\label{sub:sourcelang}
-
 \label{sub:sourcelang}
@@
 The typing rules are given in Figure~\ref{fig:source_typesystem}. 
-For scalar operations, we assume given a set of operations, including $+$ and $*$. 
+For scalar operations, we assume a given set of operations, including $+$ and $*$. 
 !op1! and !op2! denote respectively a unary and a binary operation on reals. 
@@
 
-!reduce! is a fold left operator for which the function is assumed to be associative, 
+The !reduce! operator is a fold-left operator for which the function is assumed to be associative, 
 and the provided initial value should be a unit of the binary operation.
-It is a well-known parallel friendly construct. For the sake of simplicity in the presentation, the bound function in !reduce! 
-is restricted to having no free variables. For the same reason, we currently restrict to arrays of reals.
+It is a well-known parallel friendly construct. 
+For the sake of simplicity in the presentation, the bound function in !reduce! 
+is restricted to having no free variables. Furthermore, as our main focus is on AD and not array processing,
+we currently restrict to arrays of reals.
 We show how to lift these restrictions in Section~\ref{sec:generalization}.
@@
 In more detail, for every unary scalar operation !op1!, 
-we assumed given an operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
+we assumed a given operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
 For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
@@
 In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).
+
+\input{figures/notation}
+
+\begin{notation}
+We now introduce several notations which are useful when defining the transformation for reverse-mode in Figure~\ref{tbl:notation:one}.
+Ground types are defined inductively by 
+$$G::= \RR \mid G\times \ldots \times G \mid A[G]$$
+
+$(\RR,+,\underline{0})$ forms a monoid and this monoid structure extends canonically 
+to a monoid structure $(G,\widehat{+},0_G)$ for every ground type $G$. 
+It is defined inductively on $G$ as follows
+
+\begin{tabular}{l c l}
+    $0_\RR$  & $\defeq$ & $\underline{0}$ \\
+    $0_{G_1\times \ldots \times G_n}$ & $\defeq$ &  $< O_{G_1},\ldots, 0_{G_n} >$ \\
+    $0_{\Array{\reals}{n}}$& $\defeq$ & !ZerosLike(n)! \\
+    $a\widehat{+}_\RR b$ & $\defeq$ & $ a+b$ \\
+    $(a_1,\ldots,a_n)\widehat{+}_{G_1\times\ldots\times G_n}(b_1,\ldots,b_n)$ & $\defeq$ & $(a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n)$ \\
+    $A\widehat{+}_{A[\RR]}B $ & $\defeq$ & !map2! + $A$ $B$ 
+\end{tabular}
+
+A ground context is a context only containing variables of ground type.
+The previous monoid structure again extends canonically on ground contexts $\Gamma$ by defining
+$0_{x1:G1,\ldots,x_n:G_n}\defeq 0_{G_1},\ldots,0_{G_n}$ and 
+$a_1,\ldots,a_n\widehat{+}_{x1:G_1,\ldots, x_n:G_n}b_1,\ldots,b_n\defeq a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n$.
+
+\end{notation}
 
@@
 Importantly for performance, each continuation variable $Y$ is only used once.
-
-We now introduce several notations which are useful when defining the transformation for reverse-mode.
-
-\input{figures/notation_D}
 

### 2021-07-07 rad-icfp-2021/correctness.tex
fixing unf

@@
 This section provides a categorical analysis of our language and transformations.
-The end result is the correctness of the reverse mode transformation from Section~\ref{sub:Macro for pure reverse mode transformation}:
+The main result at the end of section is the correctness of the reverse mode transformation from Section~\ref{sub:Macro for pure reverse mode transformation}:
 
@@
 
-\subsection{Denotational semantics source and Target} % (fold)
+\subsection{Denotational semantics} % (fold)
 \label{sub:Denotational semantics source and Target}
@@
 
-\subsection{Semantics for UNF using concategories} % (fold)
+\subsection{Semantics for UNF languages with concategories} % (fold)
 \label{sub:Semantics for UNF using concategories}
@@
 
-\subsection{Semantics for UNF transformations} % (fold)
+\subsection{Semantics of $\UNFSymbol$ transformations} % (fold)
 \label{sub:Semantics for UNF transformations}

### 2021-07-07 rad-icfp-2021/figures/macro_diff.tex
minor syntax

@@
     \begin{tabular}{r c l}
-        $\Dsynrevsymbol_{\rho}$(!A1,...,An!) &=& !A1,...,An,(A1x...xAn)->!$\rho$
+        $\Dsynrevsymbol_{\rho}$(!A1,$\ldots$,An!) &=& !A1,$\ldots$,An,(A1x$\ldots$xAn)->!$\rho$
     \end{tabular}
@@
     $\Dsynrevsymbol_{\rho}$(!e1!$\comp$!e2!) &=& $\Dsynrevsymbol_{\rho}$(!e1!); $\Dsynrevsymbol_{\rho}$(!e2!)\\ 
-    $\Dsynrevsymbol_{\rho}$(!e1!$\pcomp$!e2!) &=& TODO \\
     $\Dsynrevsymbol_{\rho}$(!map2!$_{T;x,y.e}$) &=& !<proj!$_{T;\underline{T}->\rho;[]} \comp$ !map2!$_{T;x,y.e}$, !proj!$_{[];T;\underline{T}->\rho} \icomp$ (!proj!$_{T;\underline{T}->\rho;[]}$ $\comp$ !J!$^T$!map2!$_{T;x,y.e}$)!>!

### 2021-07-07 rad-icfp-2021/purerad.tex
.

@@
 In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
-We explain in the next Section (\ref{sub:Partial evaluation and optimization}) how to optimize it further and compute the gradient.
-Admittedly, this transformation may be hard to read, and it is not straightforward to show its correctness directly. 
-In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).
+Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
+$\directD{\rho}{\Gamma}{Y}(e)$. First, $\rho, Y$ specifies if we want to compute the whole gradient regarding the variables from $\Gamma$ or a subset of it.
+For a subset $\rho\subset \Gamma$, one chooses $Y$ to be the projection function sending a variable 
+$x_i:G$ of $\Gamma$ to $x_i$ if it belongs to $\rho$ and to $0_G$ otherwise.
+In particular, we take $Y=Id_\Gamma$ to compute the whole gradient.
+Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
+and we need to initialize the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
+which we initialize at $1$ to run the backpropagation. 
+All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.
 
@@
 Ground types are defined inductively by 
-$$G::= \RR \mid G\times \ldots \times G \mid A[G]$$
+$$G::= \RR \mid G\times \ldots \times G \mid \Array{\RR}{n}$$
 
@@
     $0_{G_1\times \ldots \times G_n}$ & $\defeq$ &  $< O_{G_1},\ldots, 0_{G_n} >$ \\
-    $0_{\Array{\reals}{n}}$& $\defeq$ & !ZerosLike(n)! \\
+    $0_{\Array{\RR}{n}}$& $\defeq$ & !ZerosLike(n)! \\
     $a\widehat{+}_\RR b$ & $\defeq$ & $ a+b$ \\
     $(a_1,\ldots,a_n)\widehat{+}_{G_1\times\ldots\times G_n}(b_1,\ldots,b_n)$ & $\defeq$ & $(a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n)$ \\
-    $A\widehat{+}_{A[\RR]}B $ & $\defeq$ & !map2! + $A$ $B$ 
+    $A\widehat{+}_{\Array{\RR}{n}}B $ & $\defeq$ & !map2! + $A$ $B$ 
 \end{tabular}
@@
 \begin{lemma}[Typing $\directD{\rho}{\Gamma}{Y}$]
-    If $\Gamma \vdash$ !e: A!, then $\Gamma$!,Y: !$\Gamma\to \rho \vdash \directD{\rho}{\Gamma}{Y}$!(e): !$\directD{\rho}{\Gamma}{Y}$!(A)!.
+    If $\Gamma \vdash$ !e: A!, then $\Gamma$!,Y:!$\Gamma\to \rho \vdash \directD{\rho}{\Gamma}{Y}$!(e): !$\directD{\rho}{\Gamma}{Y}$!(A)!.
 \end{lemma}
+\begin{proof}
+By induction on derivation of $\Gamma\vdash$!e:A!.
+\end{proof}
 
@@
 
-Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
-$\directD{\rho}{\Gamma}{Y}(e)$. First, $\rho, Y$ specifies if we want to compute the whole gradient regarding the variables from $\Gamma$ or a subset of it.
-For a subset $\rho\subset \Gamma$, one chooses $Y$ to be the projection function sending a variable 
-$x_i:G$ of $\Gamma$ to $x_i$ if it belongs to $\rho$ and to $0_G$ otherwise.
-In particular, we take $Y=Id_\Gamma$ to compute the whole gradient.
-Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
-and we need to initialize the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
-which we initialize at $1$ to run the backpropagation. 
-All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.
-
 Following the insight from Section~\ref{subsec:insights}, 
-we don't want to keep all the lambda abstractions as this is costly. 
+we do not want to keep all the lambda abstractions as this is costly. 
 The transformation is designed in such a way that all the lambda abstractions are given arguments,  
@@
 More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
-where the !ei! are !xi! except for at most $k$ (independant of $n$) terms.
+where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
 In fact we have $k=2$, except for the !map2! case which requires a more detailed analysis (see Section~\ref{sec:complexity}).
 
-Using inlining as given in Figure~\ref{fig:optim}, this allows us to rewrite
+Using inlining as given in Figure~\ref{fig:optim} allows us to rewrite
 
@@
 Using the invariant above, most of the !e!$_{i}$ are variables. 
-Without loss of generality, assume the onlt terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
-we can use forward subsitution on the other !e!$_{i}$ and we obtain
+Without loss of generality, assume the only terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
+we can use forward substitution on the other !e!$_{i}$ and we obtain
 
@@
 \end{example}
+
+Admittedly, the transformation presented in this section may be hard to read, and it is not straightforward to show its correctness directly. 
+Next, we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).

### 2021-07-07 rad-icfp-2021/correctness.tex
writing section 5

@@
 
-It is standard to give a denotational semantics of a first-order language in a Cartesian category.
-An alternative approach consists of giving semantics in a multicategory.
 A multicategory generalizes a category by allowing multimorphisms, that is, morphisms from a list of objects to an object.
-Most categorical structures from category theory can be phrases similarly in multicategories.
+Most categorical structures from category theory can be phrased similarly in multicategories.
+It is standard to give a denotational semantics of a first-order language in a Cartesian category, and alternatively in a multicategory.
 
-In terms of semantics, a term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$ in a multicategory.
-Substitution is still interpreted as composition.
-
-One can consider a syntactic model for a language, which consists in a free multicategory.
-Our source and target languages induce syntactic multicategories as follows.
+A term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$ in a multicategory.
+Substitution is interpreted as composition. We first consider a syntactic model for a language, which consists of a 
+free multicategy on some base types and primitives. Our source and target languages induce syntactic multicategories as follows.
 
@@
 
-As is standard, $\SynSource$ satisfies the following universal property: 
-for every Cartesian multicategory $\catC$ with arrays map and reduce,
-and every object $F(\reals)\in\catC$ and morphisms $F(\underline{c})\in\catC(1;F(\reals))$, 
-$F(op1)\in\catC(F(\reals);F(\reals))$, $F(op2)\in\catC(F(\reals),F(\reals);F(\reals))$, there is a unique
-multifunctor $F:\SynSource\to\catC$ respecting the interpretation and preserving all the structure.
+$\SynSource$ satisfies the following universal property: 
+for every Cartesian multicategory $\catC$,
+and every object $F(\reals)\in\catC$, morphisms $F(\underline{c})\in\catC(1;F(\reals))$, 
+$F(op1)\in\catC(F(\reals);F(\reals))$, $F(op2)\in\catC(F(\reals),F(\reals);F(\reals))$, 
+there is a unique multifunctor $F:\SynSource\to\catC$ respecting the interpretation and preserving all the categorical structure.
 
-We can, for instance, use this to give a simple semantics of Source 
+This allows us to give a simple semantics of Source 
 in the multicategory of Cartesian spaces and smooth maps between them.
@@
 \begin{definition}[$\CartSp$]
-Let $\CartSp$ be the Cartesian category whose objects are Euclidean spaces
-and whose morphisms are smooth functions between Euclidean spaces.
-TODO:it's not a multicategory yet
+Let $\CartSp$ be the Cartesian multicategory whose objects are Euclidean spaces
+and whose morphisms $[A_1,\ldots,A_n]\to B$ are smooth functions $A_1\times\ldots A_n \to B$.
 
-We can interpret Source in $\CartSp$ as follows.
+We interpret Source in $\CartSp$ as follows. A context $\Gamma=\{x_1:A_1,\ldots,x_n:A_n\}$ is interpreted as the product $\prod_{1\leq i \leq n}\sem{A_i}$.
+Well typed terms $\Gamma\vdash$!e: !$A$ are interpreted as functions $\sem{\Gamma}\to\sem{A}$.
+
     \begin{tabular}{r c l}
@@
     \end{tabular}
-A context $\Gamma=\{x_1:A_1,\ldots,x_n:A_n\}$ is interpreted as the product $\prod_{1\leq i \leq n}\sem{A_i}$.
-Well typed terms $\Gamma\vdash$!e: !$A$ are interpreted as functions $\sem{\Gamma}\to\sem{A}$. This is
-routinely defined by induction on the structure of typing derivations. Constants
-!c! $\reals$ are interpreted as constant functions; and the primitives operations !op1,op2!
-are interpreted by composing with the corresponding functions, which are smooth. 
-Variables are interpreted as $\sem{x_i}(\rho)\defeq \rho_i$. 
-The remaining constructions are interpreted as follows.
     \begin{tabular}{r c l}
-    $\seml$!map2 (x,y.e1) e2 e3!$\semr$ &=& TODO\\
-    $\seml$!reduce (x,y.e1) e2 e3!$\semr$ &=& TODO
+        !$\seml$op1$\semr$! & $\defeq$ & op1: $\reals\to\reals$ \\
+        !$\seml$op2$\semr$! & $\defeq$ & op2: $\reals\times \reals\to\reals$ \\
+        !$\seml$c$\semr$! & $\defeq$ & $c\in\RR$ \\
+        !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & \\
+        !$\seml$reduce (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$! & $\defeq$ &
     \end{tabular}
+
+Variables are interpreted as projections, let binding uses composition in the multicategory. 
+Pairs are interpreted using the Cartesian structure of the multicategory.
 \end{definition}
 
-TODO: what about semantics Target?
+We can similarly interpret Target in a multicategory of smooth-like spaces and functions. 
+However, Target is higher-order but $\CartSp$ is not Cartesian Closed. 
+Instead, we can for instance interpret Target in the category of Diffeological spaces, 
+as in \cite{huot2020correctness}. 
+Diffeological spaces (\cite{iglesias2013diffeology}) are a conservative extension of $\CartSp$. 
+The key idea will be that a higher order function is called smooth if it sends smooth functions to smooth functions, meaning that we can never use it
+to build first order functions that are not smooth.
+
+\begin{definition}
+	A \emph{diffeological space} $(X,\plots{X})$ consists of a set $X$ together with, for each $n$ and each open subset $U$ of $\RR^n$,  a set $\plots{X}^U\subseteq [U\to X]$ of functions, called \emph{plots}, such that
+	\begin{itemize}
+	 	\item all constant functions are plots;
+	 	\item if $f:V\to U$ is a smooth function and $p\in\plots{X}^U$, then $f;p\in\plots{X}^V$;
+     \item if $\seq[i\in I]{p_i\in\plots{X}^{U_i}}$ is a compatible family of plots $(x\in U_i\cap U_j\Rightarrow p_i(x)=p_j(x))$
+     and $\seq[i\in I]{U_i}$ covers $U$,
+     then the gluing $p:U\to X:x\in U_i\mapsto p_i(x)$ is a plot.
+	 \end{itemize} 
+\end{definition}
+We call a function $f:X\to Y$ between diffeological spaces \emph{smooth} if, for all plots
+$p\in\plots{X}^U$, we have that $p;f\in \plots{Y}^U$. We write $\Diff(X,Y)$ for the set of smooth maps from $X$ to $Y$. 
+Smooth functions compose, and so we have a category $\Diff$ of diffeological spaces and smooth functions.
+
+A diffeological space is thus a set equipped with structure.
+Many constructions of sets carry over straightforwardly to diffeological spaces.
+
+\begin{example}[Cartesian diffeologies]\label{ex:cartesian-diffeologies}
+Each open subset $U$ of $\RR^n$ can be given the structure of a diffeological space by taking all the
+smooth functions $V\to U$ as $\plots{U}^V$.
+It is easily seen that smooth functions from $V\to U$ in the traditional sense coincide with
+smooth functions in the sense of diffeological spaces.
+Thus diffeological spaces have a profound relationship with ordinary calculus.
+\end{example}
+
+\begin{example}[Product diffeologies]
+Given a family $\seq[i\in I]{X_i}$ of diffeological spaces,
+we can equip the product $\prod_{i\in I}X_i$ of sets with the
+\emph{product diffeology} in which $U$-plots are precisely the functions
+of the form $\seq[i\in I]{p_i}$ for $p_i\in\plots{X_i}^U$.  
+\end{example}
+
+\begin{example}[Functional diffeology]
+We can equip the set $\Diff(X,Y)$ of smooth functions between diffeological spaces with the \emph{functional diffeology}
+in which $U$-plots consist of functions $f:U\to \Diff(X,Y)$ such that 
+$(u,x)\mapsto f(u)(x)$ is an element of $\Diff(U\times X, Y)$.
+\end{example}
+
+We can thus interpret function types !$\seml$A -> B$\semr$! = $\Diff$($\seml$!A!$\semr$,$\seml$!B!$\semr$).
 
@@
 One main reason for introducing UNF is to have a better handle over
-the computation flow of the term, similar to what ANF or CPS provide. 
+the computation flow of the term, similarly to what the ANF or CPS transformations do.
 A convenient categorical setting for this is to use string diagrams.
@@
 Similarly to categories and multicategories, 
-most categorical abstractions used for the semantics of languages are easily adapted to them.
+most categorical construtions used for the semantics of functional languages 
+have equivalent in concategories.
 
-In particular, one can form a syntactic concategory, 
+In particular, one can form a syntactic concategory on some base types and primitives, 
 which satisfies a similar universal property as syntactic multicategories.
-Two such examples are of particular interest for us, 
-because they will allow us to interpret Source and Target UNF, 
-but also explain the UNF and $UNF^{-1}$ transformations.
+Two partiular examples are of interest for us, as they will allow us to interpret Source and Target UNF
+but they will also help us explain the UNF and $UNF^{-1}$ transformations.
 
 \begin{definition}[$\ConcatS$]
-Let $\ConcatS$ the syntactic Cartesian concategory whose types are those of Source and 
-with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$ $map2_{x,y.e}$
-and $reduce_{x,y.e1;e2}$.
-TODO: actually I just need projection and a representing object for pairs, it might be weaker than Cartesian? edit: I think I need Cartesian.
+Let $\ConcatS$ be the syntactic Cartesian concategory whose types are those of Source and 
+with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$, 
+TODO: $map2_{x,y.e}$and $reduce_{x,y.e1;e2}$.
 \end{definition}
@@
 Let $\ConcatT$ be the syntactic Cartesian concategory whose types are those of Target and 
-with primitives given by 
+whose primitives are given by 
 $op1: \reals \to \reals$, 
@@
 $J^Tmap2_{x,y.e}: \Array{\reals}{n} \to \Array{\reals}{n}, \Array{\reals}{n}$, 
-$J^Treduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
-TODO: not too sure, should I say what a Cartesian concat is? do I need $J^Tproj$?
+$J^Treduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$ 
 \end{definition}
@@
 We can interpret Target UNF in $\ConcatT$ as follows. 
-The part common with Source UNF is interpreted in the same way as for Source UNF.
+The part common with Source UNF is interpreted in the same way as for the case of Source UNF.
 
@@
 
+In summary, we have the following picture
+
+\[
+\begin{tikzcd}
+    SourceUNF \ar[r,"\sem{-}"] & Concat1 \ar[r,"\Dsynrevsymbol_{\rho}"] & Concat2 \ar[r,"\UNFSymbol^{-1}"] & C_{Target} \\
+    Source \ar[r,"\UNFSymbol"] & C_{Source} \ar[u, hook] & TargetUNF \ar[u,"\sem{-}"] & Target \ar[u, hook] 
+\end{tikzcd}
+\]
+
 \subsection{Correctness theorem} % (fold)
@@
 First, we start from the correctness of $\Dsynrevsymbol_{\rho}$:Source UNF $\to$ Target UNF, which is easy to establish, 
-and then propagate this information to Source and Target via the $\UNFSymbol, \UNFSymbol^{-1}$ transformations.
+and then propagate this information to Source and Target via the $\UNFSymbol$ and $\UNFSymbol^{-1}$ transformations.
 
 \begin{proposition}[Correctness $\Dsynrevsymbol_{\rho}$]
-    For every term !$\reals^{\times n} \vdash$ e: $\reals^{\times n+1}$! in Source UNF,\\ 
-    $\pi_2 \seml \Dsynrevsymbol_{\rho}$!e!$\semr(x_1,\ldots x_n,Id_{\RR^n})$=$J^T_{(x_1,\ldots x_n)}\seml$!e!$\semr$.
+    For every term !$\reals^{\times n} \vdash$ e: $\reals^{\times n+1}$! in Source UNF,
+    \begin{center}
+        $\pi_2 \seml \Dsynrevsymbol_{\rho}$!e!$\semr(x_1,\ldots x_n,Id_{\RR^n})$=$J^T_{(x_1,\ldots x_n)}\seml$!e!$\semr$
+    \end{center} 
 \end{proposition}

### 2021-07-07 rad-icfp-2021/correctness.tex
some progress correctness

@@
 
-TODO: I think need to explain here the (co)monoid structure induced by $D$
+TODO: I think need to explain here the (co)monoid structure induced by $\Dsynrevsymbol_{\rho}$
 
@@
     $\seml$!J!$^T$!var!$_{T;i} \semr$ &=& TODO \\
-    $\seml$!J!$^T$!op!$_{T;n} \semr$ &=&  \\
-    $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& \\
-    $\seml$!J!$^T$!proj!$_{T1;T2;T3} \semr$ &=& \\
-    $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& \\
-    $\seml$!J!$^T$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
-    $\seml$!<e1, e2>!$\semr$  &=& !<!$\seml$!e1!$\semr$, $\seml$!e2!$\semr$!>! \\
-    $\seml$!e1!$\icomp$!e2!$\semr$ &=&  \\
+    $\seml$!J!$^T$!op!$_{T;n} \semr$ &=& $id_T\concatcomp J^Top_n$ \\
+    $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& $id_T\concatcomp$!<$\pi_1,\pi_2$>! \\
+    $\seml$!J!$^T$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T_1}\concatcomp\widehat{0}_{T_2}\concatcomp id_{T_3}$ \\
+    $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& $id_T\concatcomp map2_{T;x,y.e}$ \\
+    $\seml$!J!$^T$!reduce!$_{T;x,y.e_1;e_2}\semr$  &=& $id_T\concatcomp reduce_{T;x,y.e_1;e_2}$ \\
+    $\seml$!<e$_1$, e$_2$>!$\semr$  &=& !<!$\seml$!e1!$\semr$, $\seml$!e2!$\semr$!>! \\
+    $\seml$!e$_1$!$\icomp$!e$_2$!$\semr$ &=& $\Lambda$($id_\Gamma\concatcomp\seml$!e2!$\semr$; $\Lambda^{-1}$($\seml$!e1!$\semr$)) \\
  \end{tabular}
-TODO: Do I need any kind of proposition saying that this makes sense?
-Might need to define on types as well to be clearer: It's an identity on types functor...
 
@@
 \begin{definition}[Mutlicat from concat]
-    A multicategory $\catC$ naturally defines a concategory with the same objects as $\catC$ and
-    with morphisms $A\to [B_1,\ldots,B_n]$ being $n$ morphisms $A\to Bi$ of $\catC$. 
+    TODO
 \end{definition}
@@
 \begin{definition}[Concat from multicat]
-    TODO
+    A multicategory $\catC$ naturally defines a concategory with the same objects as $\catC$ and
+    with morphisms $A\to [B_1,\ldots,B_n]$ being $n$ morphisms $A\to Bi$ of $\catC$. 
 \end{definition}
@@
 What remains to explain now is the reverse mode transformation between Source UNF and Target UNF.
-We construct a functor $D:\ConcatS\to\ConcatT$. 
+We construct a functor $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$. 
 This functor will easily be shown to compute reverse-mode derivatives.
-Because terms of Source UNF in $\ConcatS$, we observe the effect of $D$ on them
+Because terms of Source UNF in $\ConcatS$, we observe the effect of $\Dsynrevsymbol_{\rho}$ on them
 and show that it matches the syntactic $\Dsynrevsymbol_{\rho}$ 
@@
 
+The design of $\Dsynrevsymbol_{\rho}$ on UNF is inspired by the semantic one between concategories, 
+and it's routine to check that they match.
+
 \begin{proposition}[semantic of syntactic D matches D lax functor]
-    TODO
+    Given a well typed term !T$_1$ $\vdash$ e: T$_2$! in Source UNF, we have
+    \begin{center}
+        $\seml\Dsynrevsymbol_{\rho}$!e!$\semr$ = $\Dsynrevsymbol_{\rho}\seml$!e!$\semr$
+    \end{center}
 \end{proposition}
@@
 From there, the intuition for the Jacobian of a primitive in Target UNF is that it should be the Jacobian of
-the corresponding term in Target. This is easily checked for scalar operations.
+the corresponding term in Target. This is easily checked for scalar operations. For the more difficult cases, we have
 
 \begin{proposition}
+$\UNFSymbol^{-1}$ preserves the semantics of Jacobians of !map2! and !reduce!.
     \begin{center}
@@
 
-\begin{proof}
-    TODO
-\end{proof}
-
-By inspecting what that composition of transformations does on terms of Source, 
+By inspecting what that composition of transformations does on the terms of Source, 
 we show that this indeed gives computes the same as the transformation 
 $\directD{\rho}{\Gamma}{Y}$ from Section~\ref{sub:Macro for pure reverse mode transformation}. 
+This should not come as a surprise because the design of $\directD{\rho}{\Gamma}{Y}$ was in fact guided via 
+this decomposition and intermediate representation.
 
@@
 \end{proposition}
-
-\begin{proof}
-    TODO
-\end{proof}
 

### 2021-07-07 rad-icfp-2021/correctness.tex
fixing boundary issues

@@
     $\seml$!T1xT2!$\semr$ & $\defeq$& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
-    $\seml \Array{\reals}{n}\semr$ & $\defeq$ & $\prod_{1\leq i \leq n} \RR$ 
+    $\seml \Array{\reals}{n}\semr$ & $\defeq$ & $\prod_{1\leq i \leq n} \RR$ \\
+    !$\seml$op1$\semr$! & $\defeq$ & op1: $\reals\to\reals$ 
     \end{tabular}
     \begin{tabular}{r c l} 
-        !$\seml$op1$\semr$! & $\defeq$ & op1: $\reals\to\reals$ \\
         !$\seml$op2$\semr$! & $\defeq$ & op2: $\reals\times \reals\to\reals$ \\
         !$\seml$c$\semr$! & $\defeq$ & $c\in\RR$ \\
-        !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & \\
+        !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & TODO \\
         !$\seml$reduce (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$! & $\defeq$ &
@@
 \begin{definition}[Mutlicat from concat]
-    TODO
+    We define $\CSource$ to be the multicategory with the same objects as Source and 
+    $\CSource([A_1,\ldots,A_n],B)=\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B])$.
 \end{definition}
@@
     SourceUNF \ar[r,"\sem{-}"] & Concat1 \ar[r,"\Dsynrevsymbol_{\rho}"] & Concat2 \ar[r,"\UNFSymbol^{-1}"] & C_{Target} \\
-    Source \ar[r,"\UNFSymbol"] & C_{Source} \ar[u, hook] & TargetUNF \ar[u,"\sem{-}"] & Target \ar[u, hook] 
+    Source \ar[r,"\UNFSymbol"] & \CSource \ar[u, hook] & TargetUNF \ar[u,"\sem{-}"] & Target \ar[u, hook] 
 \end{tikzcd}

### 2021-07-07 rad-icfp-2021/figures/macro_diff.tex
fixing boundary issues

--- 2021-02-23 rad-icfp-2021/figures/macro_diff.tex
@@
     \begin{tabular}{r c l}
-    $\Dsynrevsymbol$(!var!$_{\Gamma,i}$) &=& !<var!$_{\Gamma,i}$!, J!$^T$!var!$_{\Gamma,i}$!>! \\
-    $\Dsynrevsymbol$(!op!$_{\Gamma,n}$) &=& !<op!$_{\Gamma,n}$!, J!$^T$!op!$_{\Gamma,n}$!>! \\ 
-    $\Dsynrevsymbol$(!e1;e2!) &=& $\Dsynrevsymbol$(!e1!);$\Dsynrevsymbol$(!e2!)\\ 
-    $\Dsynrevsymbol$(!map!$_{x.e}$) &=& !<map!$_{x.e}$!, J!$^T$!map!$_{x.e}$!>! \\ 
-    $\Dsynrevsymbol$(!map2!$_{x,y.e}$) &=& !<map2!$_{x,y.e}$!, J!$^T$!map2!$_{x,y.e}$!>! \\ 
-    $\Dsynrevsymbol$(!foldl!$_{x,y.e}$) &=& !<foldl!$_{x,y.e}$!, J!$^T$!foldl!$_{x,y.e}$!>! \\  
+        $\Dsynrevsymbol_{\rho}$(!A1,$\ldots$,An!) &=& !A1,$\ldots$,An,(A1x$\ldots$xAn)->!$\rho$
     \end{tabular}
+
+    \medskip
+    \begin{tabular}{r c l}
+    $\Dsynrevsymbol_{\rho}$(!var!$_{T;i}$) &=& $F$(!var$_{T;i}$, J$^T$var$_{T;i}$!) \\
+    $\Dsynrevsymbol_{\rho}$(!op!$_{\Gamma;n}$) &=& $F$(!op$_{T;n}$,J$^T$op$_{T;,n}$!) \\ 
+    $\Dsynrevsymbol_{\rho}$(!pair!$_{T;A\times B}$) &=& $F$(!pair$_{T;A\times B}$, J$^T$pair$_{T;A\times B}$!) \\
+    $\Dsynrevsymbol_{\rho}$(!proj!$_{T1;T2;T3}$) &=& $F$(!proj$_{T1;T2;T3}$,J$^T$proj$_{T1;T2;T3}$!) \\
+    $\Dsynrevsymbol_{\rho}$(!e1!$\comp$!e2!) &=& $\Dsynrevsymbol_{\rho}$(!e1!); $\Dsynrevsymbol_{\rho}$(!e2!)\\ 
+    $\Dsynrevsymbol_{\rho}$(!map2!$_{T;x,y.e}$) &=& $F$(!map2$_{T;x,y.e}$, J$^T$map2$_{T;x,y.e}$!) \\
+    $\Dsynrevsymbol_{\rho}$(!reduce!$_{T;x,y.e1;e2}$) &=& $F$(!reduce$_{T;x,y.e1;e2}$, J$^T$reduce$_{T;x,y.e1;e2}$!) \\
+    \end{tabular}\\
+    where !$F$(A,B)$\defeq$ <proj$_{T;\underline{T}->\rho;[]} \comp$ A, proj$_{[];T;\underline{T}->\rho} \icomp$ (proj$_{T;\underline{T}->\rho;[]}$ $\comp$ B)>!
     \caption{Reverse-mode differentiation from Source UNF to Target UNF}
-    \label{fig:diff_macro}    \end{figure*}
+    \label{fig:diff_macro}    
+\end{figure*}

### 2021-07-07 rad-icfp-2021/generalization.tex
minor
fixing boundary issues

@@
-\section{Generalizations}
+\section{Beyond the Source language}
 \label{sec:generalization}
@@
 \begin{tabular}{r c l}
-$\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
-&=& $\ldots$ as before up until !A3! \\
-&& !let B0 = map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A0 A) in!\\
-&& !let B1 = scanr !$1_\Gamma$ $\widehat{\times}$ !B0 in! \\
-&& !let B2 = reduce !$\widehat{+}$ $0_\Gamma$! B1 in! \\
-&& !<reduce (x,y.e1) y1 A, fun (x1,$\ldots$,xn,z) ->! \\
-&& !Y((x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!B2,0,map2 (a,b. a*b*z) A2 A3)!
+$\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e1) e$_2$ e$_3$!) 
+&=& $\ldots$ as before up until !A_$3$! \\
+&& !let B$_0$ = map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A$_0$ A) in!\\
+&& !in let B$_1$ = scanr $1_\Gamma$ $\widehat{\times}$ B$_0$ in! \\
+&& !let B$_2$ = reduce $\widehat{+}$ $0_\Gamma$ B$_1$ in! \\
+&& !<reduce (x,y.e1) y$_1$ A, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
+&& !Y((x$_1$,$\ldots$,x$_n$)$\widehat{+}$B$_2$,0,map2 (a,b. a*b*z) A$_2$ A$_3$)!
 \end{tabular}
@@
 
-\subsection{Adding more array operators}
+\subsection{More array operators}
 \label{sub:Adding more array operators}
@@
 \begin{tabular}{r c l}
-    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !foldl (x,y.e$_1$) e$_2$ e$_3$!) &=&
+    $\directD{\rho}{\Gamma}{Y}$(!foldl (x,y.e$_1$) e$_2$ e$_3$!) &=&
             !let v,Y$_1$ = $\directD{\rho}{\Gamma}{Y}$(e$_2$) in! \\
@@
 
-\subsection{Adding more scalar operators} % (fold)
+\subsection{Non-smooth scalar operators} % (fold)
 \label{sub:Adding more scalar operators}
@@
 
-\subsection{Adding conditionals} % (fold)
+\subsection{Conditionals} % (fold)
 \label{sub:Adding conditionals}
@@
 \begin{tabular}{r c l}
-$\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) &=& !if e1 then !$\directD{\rho}{\Gamma}{Y}$!(e2) else !$\directD{\rho}{\Gamma}{Y}$!(e3)! 
+$\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) &=& !if e$_1$ then $\directD{\rho}{\Gamma}{Y}$(e$_2$) else $\directD{\rho}{\Gamma}{Y}$(e$_3$)! 
 \end{tabular}
@@
 \begin{tabular}{r c l}
-    $\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) 
-    &=& !let b=e1 in!   \\
-    && !<if b then e2 else e3, fun (x1,$\ldots$,xn,z) ->! \\
-    && !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e2)(x1,$\ldots$,xn,z)+!\\
-    && \quad!(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e3)(x1,$\ldots$,xn,z))>!
+    $\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) 
+    &=& !let b=e$_1$ in!   \\
+    && !<if b then e$_2$ else e$_3$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
+    && !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e$_2$)(x$_1$,$\ldots$,x$_n$,z)+!\\
+    && \quad!(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e$_3$)(x$_1$,$\ldots$,x$_n$,z))>!
 \end{tabular}
@@
 \begin{tabular}{r c l}
-    $\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) 
-    &=& !let b=e1 in!   \\
-    && !<if b then e2 else e3, fun (x1,$\ldots$,xn,z) ->! \\
-    && !Y((x1,$\ldots$,xn)!$\widehat{+}$($\grad_\Gamma$!(e2)*b!$\widehat{+}\grad_\Gamma$!(e3)*(1-b))*z)>!\\
+    $\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) 
+    &=& !let b=e$_1$ in!   \\
+    && !<if b then e$_2$ else e$_3$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
+    && !Y((x$_1$,$\ldots$,x$_n$)!$\widehat{+}$($\grad_\Gamma$!(e$_2$)*b!$\widehat{+}\grad_\Gamma$!(e$_3$)*(1-b))*z)>!\\
 \end{tabular}
@@
 
-\subsection{Adding general arrays} % (fold)
+\subsection{General arrays} % (fold)
 \label{sub:Adding general arrays}
@@
 \begin{tabular}{r c l}
-    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !map2 (x,y.e1: G) e2 e3!) &=&  
-    !let A,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
-    && !let B,Y2 = !$\directD{\rho}{\Gamma,A}{Y1}$!(e3) in! \\
-    && !let C=!$\grad_{\Gamma}$!e1 in!\\
-    && !<map2 (x,y.e1) A B, fun (x1,$\ldots$,xn,Z) -> !\\
-    && !Y2( (x1,$\ldots$,xn)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
-    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x]!$\widehat{*}$!b) A Z,!\\
-    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e1)[a/x]!$\widehat{*}$!b) B Z )>!\\
+    $\directD{\rho}{\Gamma}{Y}$(!map2 (x,y.e$_1$: G) e$_2$ e$_3$!) &=&  
+    !let A,Y$_1$ = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
+    && !let B,Y$_2$ = !$\directD{\rho}{\Gamma,A}{Y1}$!(e$_3$) in! \\
+    && !let C=!$\grad_{\Gamma}$!e$_1$ in!\\
+    && !<map2 (x,y.e$_1$) A B, fun (x$_1$,$\ldots$,x$_n$,Z) -> !\\
+    && !Y$_2$( (x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
+    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x]!$\widehat{*}$!b) A Z,!\\
+    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e$_1$)[a/x]!$\widehat{*}$!b) B Z )>!\\
     \end{tabular}
@@
     && !let A$_{0}$ = shift1R (scanl (x,y.e$_{1}$) y$_{1}$ A) in! \\
-    && !let A$_{1}$ = shift1L (map2 (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A$_{0}$ A) in! \\
-    && !let A$_{2}$ = map2 (a,b.(!$\grad_{\{y\}}$!)e$_{1}$[a/x,b/y]) A$_{0}$ A in! \\
+    && !let A$_{1}$ = shift1L (map2! \\ 
+    && !      (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A$_{0}$ A) in! \\
+    && !let A$_{2}$ = map2! \\
+    && !      (a,b.(!$\grad_{\{y\}}$!)e$_{1}$[a/x,b/y]) A$_{0}$ A in! \\
     && !let A$_{3}$ = scanr $\widetilde{*}$ 1 A$_{1}$ in! \\

### 2021-07-07 rad-icfp-2021/purerad.tex
fixing boundary issues

@@
     !fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)!  $\transto$ \\
-    !fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{n-1}$,y$_n$ =e$_{n-1},$e$_1$ in (f$_{1}$[x$_{1}$/y$_{1}$],$\ldots$,f$_{n-2}[$x$_{n-2}$/y$_{n-2}]$,f$_{n-1}$,f$_{n}$)!
+    !fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{n-1}$,y$_n$ = e$_{n-1},$e$_1$ in!\\
+    \hspace{3cm}!(f$_{1}$[x$_{1}$/y$_{1}$],$\ldots$,f$_{n-2}[$x$_{n-2}$/y$_{n-2}]$,f$_{n-1}$,f$_{n}$)!
 \end{tabular}
+
 It means that it does not change the evaluation cost of most of the !f!$_{i}$ for $1\leq i \leq n-2$.
 The new evaluation cost is reduced to the sum of the cost of evaluating 
-the !f!$_{i}$ in addition to the cost of evaluating !e!$_{n-1}$ and !e!$_{n}$, O($n$) movement of variables.
+the !f!$_{i}$ in addition to the cost of evaluating !e!$_{n-1}$ and !e!$_{n}$, gaining O($n$) movement of variables.
 

### 2021-07-07 rad-icfp-2021/unf.tex
fixing boundary issues

@@
 \begin{example}
-    !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$! 
-    represents !x$_1$:$\reals$, x$_2$:$\reals$, x$_3$:$\reals$ $\vdash$ let (x$_1$,x$_2$,x$_3$,x$_4$)=(x$_1$,x$_2$,x$_3$,cos(x$_3$)) in (x$_1$,x$_2$,<x$_3$,x$_4$>)!.
-    $\comp$ can more generally be understood as a let binding of a tuple, and primitives only act on the last parts of the tuple, always returning the context.
+    The term in Source UNF !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$! 
+    intuitively represents a term \\
+    !x$_1$:$\reals$, x$_2$:$\reals$, x$_3$:$\reals$ $\vdash$ let (x$_1$,x$_2$,x$_3$,x$_4$)=(x$_1$,x$_2$,x$_3$,cos(x$_3$)) in (x$_1$,x$_2$,<x$_3$,x$_4$>)!.
+    The operator $\comp$ can more generally be understood as a let binding of a tuple, and primitives only act on the last parts of the tuple, always also returning their whole context.
 \end{example}
@@
 
-\subsection{transformations to and from UNF} % (fold)
+\subsection{Transformations to and from UNF} % (fold)
 \label{sub:transformations to and from UNF}
@@
 \begin{lemma}[Well typedness $\UNFSymbol$]
-    If !$\Gamma$=x$_1$:A$_{1}$,$\ldots$,x$_n$:A$_{n}$ $\vdash$ e:B! then
+    Let !$\Gamma$=x$_1$:A$_{1}$,$\ldots$,x$_n$:A$_{n}$ $\vdash$ e:B! be a term in Source. Then
+    \begin{center}
         !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ $\UNFSymbol$(e): A$_{1}$,$\ldots$,A$_{n}$,B!.
+    \end{center}
 \end{lemma}
@@
 Target UNF does not have variables. 
-A context type !T=[A$_1$,$\ldots$,A$_n$]! is transformed to a context !$\Gamma$=x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$!.
+A context type !T=[A$_1$,$\ldots$,A$_n$]! is transformed to a context \\
+!$\Gamma$=x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$!.
 We use this convention when definiting $\UNFSymbol^{-1}$. 
-For operators like !op$_{\Gamma;m}$!, we extend the notation above by saying that the canonical variables are 
+For operators like !op$_{\Gamma;m}$!, we extend the notation above by saying that the variables from the context are \\
 !x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$,x$_{n+1}$:$\reals$,$\ldots$,x$_{n+m}$:$\reals$!.

### 2021-07-07 rad-icfp-2021/figures/macro_diff.tex
.

@@
 \begin{figure*}[t]
+\begin{tabular}{|c|}
+\hline
     \begin{tabular}{r c l}
-        $\Dsynrevsymbol_{\rho}$(!A1,$\ldots$,An!) &=& !A1,$\ldots$,An,(A1x$\ldots$xAn)->!$\rho$
-    \end{tabular}
-
-    \medskip
-    \begin{tabular}{r c l}
+    $\Dsynrevsymbol_{\rho}$(!A1,$\ldots$,An!) &=& !A1,$\ldots$,An,(A1x$\ldots$xAn)->!$\rho$\\ \\
     $\Dsynrevsymbol_{\rho}$(!var!$_{T;i}$) &=& $F$(!var$_{T;i}$, J$^T$var$_{T;i}$!) \\
@@
     \end{tabular}\\
-    where !$F$(A,B)$\defeq$ <proj$_{T;\underline{T}->\rho;[]} \comp$ A, proj$_{[];T;\underline{T}->\rho} \icomp$ (proj$_{T;\underline{T}->\rho;[]}$ $\comp$ B)>!
+    where !$F$(A,B)$\defeq$ <proj$_{T;\underline{T}->\rho;[]} \comp$ A, proj$_{[];T;\underline{T}->\rho} \icomp$ (proj$_{T;\underline{T}->\rho;[]}$ $\comp$ B)>!\\\hline
+    \end{tabular}
     \caption{Reverse-mode differentiation from Source UNF to Target UNF}

### 2021-07-07 rad-icfp-2021/relwork.tex
.

@@
 Relatively recent work in AD (2008-2020):
-\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,wang2018demystifying,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}
+\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,lantern_icfp,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}
 

### 2021-07-07 rad-icfp-2021/relwork.tex
some related work

--- 2021-07-06 rad-icfp-2021/relwork.tex
@@
 \noindent \textbf{More array primitives}
-filter, flatten, gather, scanl, ...
-can be derived by hand once and then it's ok
+We focused on giving reverse derivatives for a small set of array primitives. 
+Other common primitives include filter, flatten, gather. 
+These functions can easily be added to our Source language once we provide a reverse derivative for them.
+This is reminiscent of hand-crafted derivatives present in large AD frameworks (usually hundreds), 
+and these can already be added in added in our language. 
+One only needs to make sure that the provided transpose Jacobian is correct.
 
 \noindent \textbf{Recursion}
-TODO
+It would be an interesting extension to our work to see if one can define efficient pure reverse derivatives on general recursion.
 
 \noindent \textbf{Higher derivatives}
-TODO
-
-\noindent \textbf{Sparse data and matrix representation}
-TODO
+There are recent approaches to computing higher derivatives using jets \cite{betancourt2018geometric} which generalise forward and reverse mode.
+A more standard approach is to iterate forward or reverse-mode. 
+Forward mode is standard and can easily be defined on our Target language. 
+As our optimization procedure removes lambdas, to iterate reverse-mode we would only need to compute reverse derivatives for the new primitives of Target, that is !scanl, scanr, shift1L, shift1R!.
 
@@
 \noindent \textbf{Usage of iteration mechanics}
-A immense effort in machine learning for the past decade has been in finding
+An immense effort in machine learning for the past decade has been in finding
 good architectures, to limit computational costs, 
 avoid vanishing and exploding gradients, 
-and have better building blocks of large complicated systems than traditional layers of a neural network.
+and have better building blocks for large complicated systems than traditional layers of a neural network.
 Different approaches 

### 2021-07-07 rad-icfp-2021/generalization.tex
minor

@@
 We note that adding conditionals does not break differentiability as long as there are no non-trivial primitives $\reals\to\BB$.
-Non smooth functions such as the Rectified Linear Unit (ReLU) in machine learning which can be defined as $reLU(x)\defeq max(0,x)$ 
+Non smooth functions such as the Rectified Linear Unit (ReLU) in machine learning which can be defined as $ReLU(x)\defeq max(0,x)$ 
 requires a non smooth primitive such as $>0: \reals\to\BB$. In several AD systems such as TensorFlow, 

### 2021-07-07 rad-icfp-2021/relwork.tex
minor

@@
 and more practical considerations of efficiency and implementations, 
-which often require a lot more care than is acknowledegd in more theoretical works.
+which often require a lot more care than is acknowledged in more theoretical works.
 
 An exception is \cite{abadi-plotkin2020} which gives a reverse-mode transformation that is proved correct and with the right complexity.
-This work is purely first-order, does not have arrays and uses a non standard semantics. 
+This work is purely first-order, does not have arrays and uses a non-standard semantics. 
 Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.

### 2021-07-07 rad-icfp-2021/generalization.tex
.

@@
 \label{sec:generalization}
+In this section, we generalize our language by relaxing the imposed restrictions.
+First, we allow free variables for the function argument of !reduce! in Section~\ref{sub:lift_reduce}. 
+Then we show how to support more array operations (Section~\ref{sub:lift_more_arr_op}), conditionals (Section~\ref{sub:lift_conditional}), and non-smooth scalar operators (Section~\ref{sub:lift_non_smooth}).
+Finally, we extend the language to support general arrays instead of supporting only $\Array{\reals}{n}$ in Section~\ref{sub:lift_gen_arr}.
 
 \subsection{Lifting the restriction on reduce}
-\label{sub:Lifting the restriction on reduce}
+\label{sub:lift_reduce}
 
@@
 !reduce (x,y.e) v A!. We need to add a term depending on $\grad_{\{xi\}}e$ to !yi! in the continuation.
-Writing the derivative becomes quite cumbersome, the notation is more compact using arrays of tuples. 
-Similarly to he monoid $\widehat{+},0_\Gamma$ defined in Section~\ref{sub:Macro for pure reverse mode transformation},
-we use $\widehat{\times},1_\Gamma$ for obvious extension of the monoid $(\reals,\times,1)$.
+As writing the derivative becomes quite cumbersome, we use a more compact notation using arrays of tuples. 
+Similarly to the monoid $\widehat{+},0_\Gamma$ defined in Section~\ref{sub:Macro for pure reverse mode transformation},
+we use $\widehat{\times},1_\Gamma$ for obvious extension of the monoid $(\reals,\times,1)$. The modified reverse-mode transformation is shown in Figure~\ref{fig:lift_reduce}.
 
 We can now compute the reverse-mode transformation as follows.
-\begin{center}
-\begin{tabular}{r c l}
+
+\begin{figure}
+
+\begin{small}
+\begin{tabular}{|r c l|}
+\hline
 $\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e1) e$_2$ e$_3$!) 
-&=& $\ldots$ as before up until !A_$3$! \\
-&& !let B$_0$ = map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A$_0$ A) in!\\
-&& !in let B$_1$ = scanr $1_\Gamma$ $\widehat{\times}$ B$_0$ in! \\
-&& !let B$_2$ = reduce $\widehat{+}$ $0_\Gamma$ B$_1$ in! \\
+&=& $\ldots$ same as Figure~\ref{fig:direct_diff_macro} up until !A_$3$! \\
+&& !let B$_0$=map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A$_0$ A) in!\\
+&& !let B$_1$=scanr $1_\Gamma$ $\widehat{\times}$ B$_0$ in! \\
+&& !let B$_2$=reduce $\widehat{+}$ $0_\Gamma$ B$_1$ in! \\
 && !<reduce (x,y.e1) y$_1$ A, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
-&& !Y((x$_1$,$\ldots$,x$_n$)$\widehat{+}$B$_2$,0,map2 (a,b. a*b*z) A$_2$ A$_3$)!
+&& !Y((x$_1$,$\ldots$,x$_n$)$\widehat{+}$B$_2$,0,map2 (a,b. a*b*z) A$_2$ A$_3$)! \\ \hline
 \end{tabular}
-\end{center}
+\end{small}
+\caption{The reverse-mode AD transformation of \codekw{reduce} without restrictions on its function argument.}
+\label{fig:lift_reduce}
+\end{figure}
 
 \subsection{More array operators}
-\label{sub:Adding more array operators}
+\label{sub:lift_more_arr_op}
 
@@
 of the intermediate computations and the final result.
-The reverse derivative of fold left is then
+The reverse derivative of !foldl! is then as shown in Figure~\ref{fig:foldl_map_trans}.
+In this figure, we have shown the translation of the !map! operator as well, which is very similar to !map2!.
 
+\begin{figure}
 \begin{center}
-\begin{tabular}{r c l}
+\begin{tabular}{|r c l|}
+\hline
     $\directD{\rho}{\Gamma}{Y}$(!foldl (x,y.e$_1$) e$_2$ e$_3$!) &=&
@@
             && !Y$_2$(x$_1$,$\ldots$,x$_n$,y$_1$,B)>! \\
-\end{tabular}
-\end{center}
-
-Another useful example is !map!, which is very similar to !map2!.
-
-\begin{center}
-    \begin{tabular}{r c l}
     $\directD{\rho}{\Gamma}{Y}$(!map (x.e$_{1}$) e$_{2}$!) &=&  
@@
             && !Y$_{1}$( (x$_{1}$,$\ldots$,x$_n$)$\widehat{+}$(G*(reduce + 0 Z)),!\\
-            && \quad\quad !map2 (a,b.($\grad_{\{x\}}$e$_{1}$)[a/x]*b) A Z )>!\\
+            && \quad\quad !map2 (a,b.($\grad_{\{x\}}$e$_{1}$)[a/x]*b) A Z )>!\\\hline
 \end{tabular}
 \end{center}
-
-\subsection{Non-smooth scalar operators} % (fold)
-\label{sub:Adding more scalar operators}
-
-We assumed the unary and binary operators were denoted by smooth functions $\RR^n\to\RR$. 
-There is no additional difficulty in considering operators which are partial functions 
-like division or operators which are not smooth at a point like square root.
-
-These functions are then given intentional derivatives which provide valid derivatives 
-on the domain of definition and differentiability of the operator. 
-These functions are well known to be the bete noire of AD \cite{griewank2008evaluating} 
-and we do not provide novel solutions to these.  
-Several recent work have shown how to give semantics to such operators in the context of AD \cite{vakar2020denotational,mazza2021automatic,sherman2021,lee2020correctness}.
+\caption{The reverse-mode AD transformation of \codekw{foldl} and \codekw{map} operators.}
+\label{fig:foldl_map_trans}
+\end{figure}
 
 \subsection{Conditionals} % (fold)
-\label{sub:Adding conditionals}
+\label{sub:lift_conditional}
 
+\smartpara{Non-Solution 1}
 Adding conditionals to the source language can be done easily, for instance via defining
@@
 
+\smartpara{Non-Solution 2}
 A slightly better option would be to define
@@
 but there is still a non-linear usage of the continuation variable !Y!.
-One solution, if we know we want to compute the whole gradient of the expression, is to define instead
+
+\smartpara{Solution}
+If we know we want to compute the whole gradient of the expression, we can define the translation as follows:
 
@@
 
+\subsection{Non-smooth scalar operators} % (fold)
+\label{sub:lift_non_smooth}
+
+We assumed the unary and binary operators were denoted by smooth functions $\RR^n\to\RR$. 
+There is no additional difficulty in considering operators which are partial functions 
+like division or operators which are not smooth at a point like square root.
+
+These functions are then given intentional derivatives which provide valid derivatives 
+on the domain of definition and differentiability of the operator. 
+These functions are well known to be the bete noire of AD \cite{griewank2008evaluating} 
+and we do not provide novel solutions to these.  
+Several recent work have shown how to give semantics to such operators in the context of AD \cite{vakar2020denotational,mazza2021automatic,sherman2021,lee2020correctness}.
+
 \subsection{General arrays} % (fold)
-\label{sub:Adding general arrays}
+\label{sub:lift_gen_arr}
 
@@
 $\widetilde{*}$ corresponds to matrix-matrix multiplication.
-Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!.
+Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!, as can be seen in Figure~\ref{fig:lift_gen_arr}.
 
+\begin{figure}
 \begin{center}
-\begin{tabular}{r c l}
+\begin{tabular}{|r c l|}
+\hline
     $\directD{\rho}{\Gamma}{Y}$(!map2 (x,y.e$_1$: G) e$_2$ e$_3$!) &=&  
@@
     && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e$_1$)[a/x]!$\widehat{*}$!b) B Z )>!\\
-    \end{tabular}
-    \end{center}
-
-\begin{center}
-\begin{tabular}{r c l}
     $\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e$_{1}$) e$_{2}$ e$_{3}$!) &=&
@@
     && !<reduce (x,y.e$_{1}$) y$_{1}$ A, fun (x$_{1}$,$\ldots$,x$_n$,z) ->! \\
-    && !Y$_{2}$(x$_{1}$,$\ldots$,x$_n$, map2 (x,y. x$\widehat{*}$(y$\widehat{*}$z)) A$_{2}$ A$_{3}$>! \\
+    && !Y$_{2}$(x$_{1}$,$\ldots$,x$_n$, map2 (x,y. x$\widehat{*}$(y$\widehat{*}$z)) A$_{2}$ A$_{3}$>! \\ \hline
 \end{tabular}
 \end{center}
+\caption{The reverse-mode AD transformation of \codekw{map2} and \codekw{reduce} for general arrays.}
+\label{fig:lift_gen_arr}
+\end{figure}
 

### 2021-07-07 rad-icfp-2021/complexity.tex
.
fixes from Sam's comments section 6

@@
-\section{Complexity}
+\section{Complexity analysis}
 \label{sec:complexity}
+
+In this section, we introduce a simple cost model on our language and show 
+that our reverse-mode transformation satisfies a version of the cheap gradient principle (Theorem \ref{thm:complexity}).
 
@@
 We assume the cost is divided into $4$ elementary measures, being the number of MOVES, ADDS, MULTS, and NLOPS.
-MOVES assumes a flat memory and represents moving a fixed size information  (e.g. 64 bits). 
+MOVES assumes a flat memory and represents moving fixed-size information (e.g. 64 bits). 
 ADDS represents the number of additions, 
@@
 
+We define the nesting of array operations $\NestA$ of a term !e! of Source by induction as follows.
+
+\begin{tabular}{r c l}
+    $\NestA$(!c!) &=& 0 \\
+    $\NestA$(!x!) &=& 0 \\
+    $\NestA$(!let x:A = e$_1$ in e$_2$:B!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)  \\ 
+    $\NestA$(!< e$_1$, e$_2$ >:AxB!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!) \\ 
+    $\NestA$($\pi_i$(!e!)) &=& $\NestA$(!e!)\\
+    $\NestA$(!e$_1$ op2 e$_2$!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)\\
+    $\NestA$(!op1 e!) &=& $\NestA$(!e!) \\
+    $\NestA$(!map2 (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
+    $\NestA$($\Gamma\vdash $ !reduce (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
+\end{tabular}
+
+\begin{theorem}
+    \label{thm:complexity}
+    Given a term !$\Gamma \vdash$ e: $\reals$! such that $\NestA$(!e!)$\leq$ $p$.
+    Denote by $G$ the term $\grad_\Gamma$!e! after inlining and forward substitution.
+    TODO: be more precise with "inlining and forward substitution".
+    Then !$\cost$(G) $\leq$ $4*2^{p}*\cost$(e)!.
+\end{theorem}
+
 The cheap gradient principle (see e.g. \cite{griewank2008evaluating}) for reverse-mode
@@
 
+\begin{proof}[Notes]
     As $\directD{\rho}{\Gamma}{Y}$ is defined by induction on programs, it suffices to show locally 
@@
 
-Combining the two previous points, we conclude that our reverse mode transformation satisfies a version of the cheap gradient principle.
-
-\begin{theorem}
-    Given a term !$\Gamma \vdash$ e: $\reals$! in Source with at most $p$ nested !map2! or !reduce!. 
-    Denote by $G$ the term $\grad_\Gamma$!e! after inlining and forward substitution.
-    Then !$\cost$(G) $\leq$ $4*2^{p}*\cost$(e)!.
-\end{theorem}
-
-\begin{proof}[notes]
-    The proof is by routine induction on !e!. 
+After that, the proof is by routine induction on !e!. 
 First, one computes the cost of $\directD{\rho}{\Gamma}{Y}$(!e!).
@@
 In Figure~\ref{fig:optim} we present a of optimizations. 
-As our language is purely functional and effect free, 
-we can use these optimizations aggressively. 
+These optimizations are all obviously valid according to the semantic model. 
+As our language is purely functional, we can use these extra optimizations aggressively. 
 A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
-After these optimizations, the output program is essentially a sequence of let bindings 
+After these optimizations, the output program is essentially a sequence of let-binding
 and resembles SSA-code.

### 2021-07-07 rad-icfp-2021/complexity.tex
typo

@@
 
-We define the nesting of array operations $\NestA$ of a term !e! of Source by induction as follows.
+We define the nesting of array operations $\NestA$ of a term !e! of Source by induction on !e! as follows.
 
@@
 First, one computes the cost of $\directD{\rho}{\Gamma}{Y}$(!e!).
-It has $N$ too many MOVES which are removed by inlining and forward substitution using the invariant that at each step, 
+It has $n$ too many MOVES which are removed by inlining and forward substitution using the invariant that at each step, 
 at most 2 !x!$_i$ in the continuation are not variables.

### 2021-07-08 rad-icfp-2021/purerad.tex
finishing fxing Sam's comments Sec3

@@
 we currently restrict to arrays of reals.
-We show how to lift these restrictions in Section~\ref{sec:generalization}.
+We show how to lift these restrictions and how to differentiate some other array operators in Section~\ref{sec:generalization}.
 
@@
 In more detail, for every unary scalar operation !op1!, 
-we assumed a given operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
-For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
+we assumed a given operator $\partial$!op1! whose semantics should be the derivative of !op1!, e.g. $\partial$!cos!=!sin!.
+Similarly, for every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2!, 
 respectively representing the first and second partial derivative of !op2!.
@@
 
-Following the insight from Section~\ref{subsec:insights}, 
-we do not want to keep all the lambda abstractions as this is costly. 
+As can be seen from the examples, $\directD{\rho}{\Gamma}{Y}$ introduces a lot of functions of several arguments.
+Following the insight from Section~\ref{subsec:insights}, we do not want to keep all these costly lambda abstractions.
 The transformation is designed in such a way that all the lambda abstractions are given arguments,  
-and we can use partial evaluation to do beta-reduce all these lambda abstractions.
+and we can use partial evaluation to beta-reduce all these lambda abstractions.
 
@@
 where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
-In fact we have $k=2$, except for the !map2! case which requires a more detailed analysis (see Section~\ref{sec:complexity}).
-
-Using inlining as given in Figure~\ref{fig:optim} allows us to rewrite
+In fact we have $k=2$, except for the !map2! and !reduce! cases which require a more detailed analysis (see Section~\ref{sec:complexity}).
+We can first use inlining as given in Figure~\ref{fig:optim}.
+Using the invariant above, most of the !e!$_{i}$ below are variables. 
+Without loss of generality, assume the only terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
+We can then use forward substitution on the other !e!$_{i}$.
+In brief:
 
@@
 $\transto$ \\
-!fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)!
-\end{tabular}
-
-Using the invariant above, most of the !e!$_{i}$ are variables. 
-Without loss of generality, assume the only terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
-we can use forward substitution on the other !e!$_{i}$ and we obtain
-
-\begin{tabular}{l}
 !fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)! $\transto$ \\
@@
 
-It means that it does not change the evaluation cost of most of the !f!$_{i}$ for $1\leq i \leq n-2$.
+This rewriting does not change the evaluation cost of the !f!$_{i}$ for $1\leq i \leq n-2$.
 The new evaluation cost is reduced to the sum of the cost of evaluating 

### 2021-07-08 rad-icfp-2021/unf.tex
.
fixing Sam's comments sec4

@@
 It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
-The grammar of Target UNF is given in Figure~\ref{fig:unf_source_grammar} and its type system in Figure~\ref{fig:target_unf_typesystem}.
+The grammar of Target UNF is given in Figure~\ref{fig:unf_target_grammar} and its type system in Figure~\ref{fig:target_unf_typesystem}.
 
@@
 The second part returns the continuation term pre-composed by the transpose Jacobian of the primitive. 
+There is a sharp contrast between the simplicity of $\Dsynrevsymbol_{\rho}$ compared to $\directD{\rho}{\Gamma}{Y}$ from Figure~\ref{fig:direct_diff_macro}.
+
+\begin{lemma}[Well typedness $\Dsynrevsymbol_{\rho}$]
+    Let !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$! be a term in Source UNF. \\
+    Then 
+    % \begin{center}
+        !A$_{1}$,$\ldots$,A$_{n}$,A$_{1}\times\ldots\times$ A$_{n}$->$\rho$ $\vdash$ $\Dsynrevsymbol_{\rho}$(e): B$_{1}$,$\ldots$,B$_{m}$,B$_{1}\times\ldots\times$ B$_{m}$->$\rho$!
+    % \end{center}
+\end{lemma}
+\begin{proof}
+    By induction on derivation of !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$!.
+\end{proof}
 
@@
 \end{lemma}
+\begin{proof}
+    By induction on derivation of $\Gamma\vdash$!e:B!.
+\end{proof}
 
@@
 We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
-Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of ANF transformation. 
-$\UNFSymbol^{-1}$ does not preserve a typing property as nicely as $\UNFSymbol$, 
-because it treats primitives with and without a $J$ differently.
+Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of the ANF transformation \cite{sabry1993reasoning}. 
+$\UNFSymbol^{-1}$ does not have as a simple typing property as $\UNFSymbol$, 
+because it treats primitives with and without a $J$ differently and should be performed after $\Dsynrevsymbol_{\rho}$.
+
+\begin{lemma}[Well typedness $\UNFSymbol^{-1}$]
+    Let !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$! be a term in Source UNF. Then
+    \begin{center}
+        !x$_1$:A$_{1}$,$\ldots$,x$_n$:A$_{n}$,x$_{n+1}$:A$_{1}\times\ldots\times$ A$_{n}$->$\rho$ $\vdash$ $\UNFSymbol^{-1}$($\Dsynrevsymbol_{\rho}$(e)): B$_m$ $\times$ (B$_{1}\times\ldots\times$ B$_{m}$->$\rho$) !.
+    \end{center}
+\end{lemma}
+\begin{proof}
+    By induction on derivation of !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$!.
+\end{proof}
 

### 2021-07-08 rad-icfp-2021/unf.tex
more fixes sec4

@@
-\section{UNF}
+\section{Unary Normal Form}
 \label{sec:unf}
@@
 Following the intuition highlighted in Section~\ref{subsec:insights}, 
-we present a new language which we call unary form (UNF). 
+we present a new language which we call Unary Normal Form (UNF). 
 There is a Source UNF for the Source language and a Target UNF for the Target language.
 Just as Target is an extension of Source, Target UNF is an extension of Source UNF.
+They serve as intermediate representation our the reverse-mode compilation pipeline (see Figure~\ref{fig:outline}).
 
@@
 
-By introducing UNF, we are trying to decouple some of the problems.
+By introducing UNF, we are decoupling some of the problems.
 UNF introduces the call-by-value evaluation of the arguments before evaluating the arguments and a good management of the environment.
@@
 
+A judgment in Source UNF is a triple !(T$_1$,e,T$_2$)! where !T$_1$! and !T$_2$! are lists of types of Source, and !e! is an expression of the Source UNF.
 The typing rules are detailed in Figure~\ref{fig:source_unf_typesystem}.
 $\reals^{\times n}$ is a notation for $\reals,\ldots,\reals$ with $n$ factors $\reals$. 
+When !T$_1$! and !T$_2$! are lists of types and !A! a type, the operator !,! denotes snoc in !T$_1$,A! and denotes append in !T$_1$,T$_2$!.
 

### 2021-07-08 rad-icfp-2021/complexity.tex
minor
several minor fixes

@@
 
-\begin{proof}[Notes]
+\begin{proof}[Proof Sketch]
     As $\directD{\rho}{\Gamma}{Y}$ is defined by induction on programs, it suffices to show locally 
@@
 
+\input{figures/optims}
+
 \subsection{Optimizations} % (fold)
@@
 A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
-After these optimizations, the output program is essentially a sequence of let-binding
+After these optimizations, the output program is essentially a sequence of let bindings 
 and resembles SSA-code.
-
-\input{figures/optims}

### 2021-07-08 rad-icfp-2021/figures/macro_diff.tex
several minor fixes

@@
     \begin{tabular}{r c l}
-    $\Dsynrevsymbol_{\rho}$(!A1,$\ldots$,An!) &=& !A1,$\ldots$,An,(A1x$\ldots$xAn)->!$\rho$\\ \\
+    $\Dsynrevsymbol_{\rho}$(!A$_1$,$\ldots$,A$_n$!) &=& !A$_1$,$\ldots$,A$_n$,(A$_1\times\ldots\times$A$_n$)->$\rho$!\\ \\
     $\Dsynrevsymbol_{\rho}$(!var!$_{T;i}$) &=& $F$(!var$_{T;i}$, J$^T$var$_{T;i}$!) \\
-    $\Dsynrevsymbol_{\rho}$(!op!$_{\Gamma;n}$) &=& $F$(!op$_{T;n}$,J$^T$op$_{T;,n}$!) \\ 
+    $\Dsynrevsymbol_{\rho}$(!op!$_{\Gamma;n}$) &=& $F$(!op$_{T;n}$,J$^T$op$_{T;n}$!) \\ 
     $\Dsynrevsymbol_{\rho}$(!pair!$_{T;A\times B}$) &=& $F$(!pair$_{T;A\times B}$, J$^T$pair$_{T;A\times B}$!) \\
-    $\Dsynrevsymbol_{\rho}$(!proj!$_{T1;T2;T3}$) &=& $F$(!proj$_{T1;T2;T3}$,J$^T$proj$_{T1;T2;T3}$!) \\
-    $\Dsynrevsymbol_{\rho}$(!e1!$\comp$!e2!) &=& $\Dsynrevsymbol_{\rho}$(!e1!); $\Dsynrevsymbol_{\rho}$(!e2!)\\ 
+    $\Dsynrevsymbol_{\rho}$(!proj!$_{T_1;T_2;T_3}$) &=& $F$(!proj$_{T_1;T_2;T_3}$,J$^T$proj$_{T1;T2;T3}$!) \\
+    $\Dsynrevsymbol_{\rho}$(!e$_1\comp$e$_2$!) &=& $\Dsynrevsymbol_{\rho}$(!e$_1$!)$\comp$ $\Dsynrevsymbol_{\rho}$(!e$_2$!)\\ 
     $\Dsynrevsymbol_{\rho}$(!map2!$_{T;x,y.e}$) &=& $F$(!map2$_{T;x,y.e}$, J$^T$map2$_{T;x,y.e}$!) \\
-    $\Dsynrevsymbol_{\rho}$(!reduce!$_{T;x,y.e1;e2}$) &=& $F$(!reduce$_{T;x,y.e1;e2}$, J$^T$reduce$_{T;x,y.e1;e2}$!) \\
+    $\Dsynrevsymbol_{\rho}$(!reduce!$_{T;x,y.e_1;e_2}$) &=& $F$(!reduce$_{T;x,y.e_1;e_2}$, J$^T$reduce$_{T;x,y.e_1;e_2}$!) \\
     \end{tabular}\\

### 2021-07-08 rad-icfp-2021/generalization.tex
several minor fixes

@@
 
-We now show how to generalize our reverse-mode transformation to be defined on arrays over any ground type $G$.
+We now show how to generalize our reverse-mode transformation to be defined on arrays over any ground type !G!.
 That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.
 
-A ground type $G$ is interpreted as an Euclidean space $A$. 
+A ground type !G! is interpreted as an Euclidean space $A$. 
 It is in particular a real vector space.
-Similarly, a ground context $\Gamma=x1:G1,\ldots,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
-The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}Ai)\otimes A$
-where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.
+Similarly, a ground context !$\Gamma$=x$_1$:G$_1$,$\ldots$,x$_n$:G$_n$! is interpreted as $\bigoplus_{1\leq i\leq n}A_i$, where $A_i\defeq\seml$(!G$_i$!)$\semr$.
+The denotation of the gradient of a term !$\Gamma \vdash$ e: G! at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}A_i)\otimes A$
+where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}A_i\otimes A$.
 
@@
     $\reals \otimes A$ & $\defeq$ & $A$ \\
-    $(A1 \times \ldots \times An)\otimes A$ & $\defeq$ & $(A1\otimes A) \times \ldots \times (An \otimes A)$ \\
-    $\Array{A1}{n} \otimes A$ & $\defeq$ & $\Array{A1\otimes A}{n}$
+    $(A_1 \times \ldots \times A_n)\otimes A$ & $\defeq$ & $(A_1\otimes A) \times \ldots \times (A_n \otimes A)$ \\
+    $\Array{A_1}{n} \otimes A$ & $\defeq$ & $\Array{A_1\otimes A}{n}$
 \end{tabular}
 
-With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldots\times An$ as expected.
+With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A_1\times\ldots\times A_n$ as expected.
 For !map2!, we need a generalization of $*:\reals\to\reals$. 
-If !e1: A!, then $\grad_{\{x\}}$!e1: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
+If !e$_1$: A!, then $\grad_{\{x\}}$!e$_1$: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
 If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.

### 2021-07-08 rad-icfp-2021/unf.tex
several minor fixes

@@
     intuitively represents a term \\
-    !x$_1$:$\reals$, x$_2$:$\reals$, x$_3$:$\reals$ $\vdash$ let (x$_1$,x$_2$,x$_3$,x$_4$)=(x$_1$,x$_2$,x$_3$,cos(x$_3$)) in (x$_1$,x$_2$,<x$_3$,x$_4$>)!.
+    !x$_1$:$\reals$,x$_2$:$\reals$,x$_3$:$\reals$ $\vdash$ let (x$_1$,x$_2$,x$_3$,x$_4$)=(x$_1$,x$_2$,x$_3$,cos(x$_3$)) in (x$_1$,x$_2$,<x$_3$,x$_4$>)!.\\
     The operator $\comp$ can more generally be understood as a let binding of a tuple, and primitives only act on the last parts of the tuple, always also returning their whole context.
@@
 
+\begin{example}
+    Continuing with our previous example !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$!, we have
+
+    \begin{tabular}{r c l}
+        !$\Dsynrevsymbol_{\rho}$(cos$_{\reals,\reals}\comp$ pair$_{\reals,\reals; \reals\times\reals}$)!
+        &=& !<> $\comp$! \\
+        &&  !<> !
+    \end{tabular}
+\end{example}
+
 \begin{lemma}[Well typedness $\Dsynrevsymbol_{\rho}$]
     Let !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$! be a term in Source UNF. \\
-    Then 
-    % \begin{center}
-        !A$_{1}$,$\ldots$,A$_{n}$,A$_{1}\times\ldots\times$ A$_{n}$->$\rho$ $\vdash$ $\Dsynrevsymbol_{\rho}$(e): B$_{1}$,$\ldots$,B$_{m}$,B$_{1}\times\ldots\times$ B$_{m}$->$\rho$!
-    % \end{center}
+    Then  !A$_{1}$,$\ldots$,A$_{n}$,A$_{1}\times\ldots\times$ A$_{n}$->$\rho$ $\vdash$ $\Dsynrevsymbol_{\rho}$(e): B$_{1}$,$\ldots$,B$_{m}$,B$_{1}\times\ldots\times$ B$_{m}$->$\rho$!.
 \end{lemma}
@@
 $\UNFSymbol$ sequentializes a term, mimicking a left-to-right call-by-value evaluation.
-Because of this, we we're trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
+Because of this, when we are trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
 When we need to pass the result of type !A! of !e$_{1}$! through !e$_{2}$! which does not take it as an input, we can transform !e$_{2}$! to accept and pass it.
@@
 Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of the ANF transformation \cite{sabry1993reasoning}.
+
+\begin{example}
+    TODO
+\end{example}
+
+Target UNF does not have variables. 
+A context type !T=[A$_1$,$\ldots$,A$_n$]! is transformed to a context \\
+!$\Gamma$=x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$!.
+We use this convention when definiting $\UNFSymbol^{-1}$. 
+For operators like !op$_{\Gamma;m}$!, we extend the notation above by saying that the variables from the context are \\
+!x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$,x$_{n+1}$:$\reals$,$\ldots$,x$_{n+m}$:$\reals$!.
+
+\begin{example}
+    TODO
+\end{example}
+
 $\UNFSymbol^{-1}$ does not have as a simple typing property as $\UNFSymbol$, 
@@
 
-Target UNF does not have variables. 
-A context type !T=[A$_1$,$\ldots$,A$_n$]! is transformed to a context \\
-!$\Gamma$=x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$!.
-We use this convention when definiting $\UNFSymbol^{-1}$. 
-For operators like !op$_{\Gamma;m}$!, we extend the notation above by saying that the variables from the context are \\
-!x$_1$:A$_1$,$\ldots$,x$_n$:A$_n$,x$_{n+1}$:$\reals$,$\ldots$,x$_{n+m}$:$\reals$!.
-
 \input{figures/unf_to_target}

### 2021-07-08 rad-icfp-2021/complexity.tex
minor

@@
 
-We define the nesting of array operations $\NestA$ of a term !e! of Source by induction on !e! as follows.
+We define the Nesting of Array Operations $\NestA$ of a term !e! of Source by induction on !e! as follows.
 
@@
     $\NestA$(!map2 (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
-    $\NestA$($\Gamma\vdash $ !reduce (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
+    $\NestA$(!reduce (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
 \end{tabular}

### 2021-07-08 rad-icfp-2021/complexity.tex
reorganizing sections 5 and 6

@@
 that our reverse-mode transformation satisfies a version of the cheap gradient principle (Theorem \ref{thm:complexity}).
+
+\subsection{Partial evaluation and optimization} % (fold)
+\label{sub:Partial evaluation and optimization}
+
+As can be seen from the examples, $\directD{\rho}{\Gamma}{Y}$ introduces a lot of functions of several arguments.
+Following the insight from Section~\ref{subsec:insights}, we do not want to keep all these costly lambda abstractions.
+The transformation is designed in such a way that all the lambda abstractions are given arguments,  
+and we can use partial evaluation to beta-reduce all these lambda abstractions.
+
+By inspection in Figure~\ref{fig:direct_diff_macro}, 
+we see that there is a linear usage of  each continuation $Y$ 
+and that it is always applied to almost the identity. 
+More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
+where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
+In fact we have $k=2$, except for the !map2! and !reduce! cases which require a more detailed analysis (see Section~\ref{sec:complexity}).
+We can first use inlining as given in Figure~\ref{fig:optim}.
+Using the invariant above, most of the !e!$_{i}$ below are variables. 
+Without loss of generality, assume the only terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
+We can then use forward substitution on the other !e!$_{i}$.
+In brief:
+
+\begin{tabular}{l}
+!fun (x$_{1}$,$\ldots$,x$_n$) -> (fun (y$_{1}$,$\ldots$,y$_n$) -> (f$_{1}$,$\ldots$,f$_n$))(e$_{1}$,$\ldots$,e$_n$)! 
+$\transto$ \\
+!fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)! $\transto$ \\
+!fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{n-1}$,y$_n$ = e$_{n-1},$e$_1$ in!\\
+\hspace{3cm}!(f$_{1}$[x$_{1}$/y$_{1}$],$\ldots$,f$_{n-2}[$x$_{n-2}$/y$_{n-2}]$,f$_{n-1}$,f$_{n}$)!
+\end{tabular}
+
+This rewriting does not change the evaluation cost of the !f!$_{i}$ for $1\leq i \leq n-2$.
+The new evaluation cost is reduced to the sum of the cost of evaluating 
+the !f!$_{i}$ in addition to the cost of evaluating !e!$_{n-1}$ and !e!$_{n}$, gaining O($n$) movement of variables.
+
+\begin{example}
+    After forward-substitution and inlining the inner !Y!$_{i}$, the gradient of the terms from the introduction reduces to
+
+    \begin{tabular}{c l}
+        & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
+        =& !let w$_1$,Y$_1$= <x$_1$ * x$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y(y$_{1}$+x$_2$*z,y$_{2}$+x$_1$*z,y$_{3}$) > in! \\
+        & !let w$_{2}$,Y$_{2}$= <w$_{1}$ * x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{1}$(y$_{1}$+w$_1$*z,y$_{2}$,y$_{3}$,y$_{4}$+x$_{1}$*z) > in! \\
+        & !let y,Y$_{3}$= <w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{2}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
+        & !<y, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{3}$(y$_{1}$,y$_{2}$,y$_{3}$,0,z) >!
+    \end{tabular}
+
+After another simplification step, we obtain
+
+\begin{tabular}{c l}
+    & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
+    =& !let w$_1$= x$_1$ * x$_{2}$ in let w$_{2}$= w$_{1}$ * x$_{1}$ in! \\
+        & !<w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> let y'$_{1}$=y$_{1}$+w$_1$*z in !\\
+        & \quad\quad\quad!let z'=y$_4$+x$_1$*z in Y(y'+x$_2$*z',y$_{2}$+x$_1$*z',y$_{3}$) >!
+\end{tabular}
+
+Similarly, for the gradient of !prod(A)! we obtain
+
+\begin{tabular}{c l}
+        & $\grad_\Gamma$(!prod(A)!) \\
+        & !let A$_{0}$= shift1R (scanl * 1 A) in!\\
+        & !let A$_{1}$= shift1L (map2 (a,b.b) A$_{0}$ A) in!\\
+        & !let A$_{2}$= map2 (a,b.a) A$_{0}$ A in!\\
+        & !let A$_{3}$= scanr * 1 A$_{1}$ in!\\
+        & !<prod(A), fun (X,z) -> Y(X+map2 (a,b. a*z*b) A$_{2}$ A$_{3}$)>! 
+    \end{tabular}
+\end{example}
+
+Though this transformation after the inlining above already has the right complexity, 
+its purity allows us to make the most out of generic optimizations.  
+A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.
+
+\begin{example}
+As shown in appendix \ref{sub:gradintro}, the optimizations from Figure~\ref{fig:optim} 
+are sufficient to show that the gradients of the terms of the introduction reduce to the following.
+
+\begin{tabular}{{r c l}}
+    $\nabla_A$!prod(A)! &=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))!\\
+    $\nabla_A$!sum(A)! &=& !map (x -> 1) A!\\
+    $\nabla_A$!dot(A,B)! &=& !B! 
+\end{tabular}
+\end{example}
 

### 2021-07-08 rad-icfp-2021/purerad.tex
reorganizing sections 5 and 6

@@
 
-\subsection{Partial evaluation and optimization} % (fold)
-\label{sub:Partial evaluation and optimization}
-
-As can be seen from the examples, $\directD{\rho}{\Gamma}{Y}$ introduces a lot of functions of several arguments.
-Following the insight from Section~\ref{subsec:insights}, we do not want to keep all these costly lambda abstractions.
-The transformation is designed in such a way that all the lambda abstractions are given arguments,  
-and we can use partial evaluation to beta-reduce all these lambda abstractions.
-
-By inspection in Figure~\ref{fig:direct_diff_macro}, 
-we see that there is a linear usage of  each continuation $Y$ 
-and that it is always applied to almost the identity. 
-More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
-where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
-In fact we have $k=2$, except for the !map2! and !reduce! cases which require a more detailed analysis (see Section~\ref{sec:complexity}).
-We can first use inlining as given in Figure~\ref{fig:optim}.
-Using the invariant above, most of the !e!$_{i}$ below are variables. 
-Without loss of generality, assume the only terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
-We can then use forward substitution on the other !e!$_{i}$.
-In brief:
-
-\begin{tabular}{l}
-!fun (x$_{1}$,$\ldots$,x$_n$) -> (fun (y$_{1}$,$\ldots$,y$_n$) -> (f$_{1}$,$\ldots$,f$_n$))(e$_{1}$,$\ldots$,e$_n$)! 
-$\transto$ \\
-!fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)! $\transto$ \\
-!fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{n-1}$,y$_n$ = e$_{n-1},$e$_1$ in!\\
-\hspace{3cm}!(f$_{1}$[x$_{1}$/y$_{1}$],$\ldots$,f$_{n-2}[$x$_{n-2}$/y$_{n-2}]$,f$_{n-1}$,f$_{n}$)!
-\end{tabular}
-
-This rewriting does not change the evaluation cost of the !f!$_{i}$ for $1\leq i \leq n-2$.
-The new evaluation cost is reduced to the sum of the cost of evaluating 
-the !f!$_{i}$ in addition to the cost of evaluating !e!$_{n-1}$ and !e!$_{n}$, gaining O($n$) movement of variables.
-
-\begin{example}
-    After forward-substitution and inlining the inner !Y!$_{i}$, the gradient of the terms from the introduction reduces to
-
-    \begin{tabular}{c l}
-        & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
-        =& !let w$_1$,Y$_1$= <x$_1$ * x$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y(y$_{1}$+x$_2$*z,y$_{2}$+x$_1$*z,y$_{3}$) > in! \\
-        & !let w$_{2}$,Y$_{2}$= <w$_{1}$ * x$_{1}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{1}$(y$_{1}$+w$_1$*z,y$_{2}$,y$_{3}$,y$_{4}$+x$_{1}$*z) > in! \\
-        & !let y,Y$_{3}$= <w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$,z) -> Y$_{2}$(y$_{1}$,y$_{2}$,y$_{3}$,y$_{4}$+z)> in! \\
-        & !<y, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> Y$_{3}$(y$_{1}$,y$_{2}$,y$_{3}$,0,z) >!
-    \end{tabular}
-
-After another simplification step, we obtain
-
-\begin{tabular}{c l}
-    & $\grad_\Gamma$(!let w$_{1}$ = x$_{1}$ * x$_{2}$ in let w$_{2}$ = w$_{1}$ * x$_{1}$ in w$_{2}$!) \\
-    =& !let w$_1$= x$_1$ * x$_{2}$ in let w$_{2}$= w$_{1}$ * x$_{1}$ in! \\
-        & !<w$_{2}$, fun (y$_{1}$,y$_{2}$,y$_{3}$,z) -> let y'$_{1}$=y$_{1}$+w$_1$*z in !\\
-        & \quad\quad\quad!let z'=y$_4$+x$_1$*z in Y(y'+x$_2$*z',y$_{2}$+x$_1$*z',y$_{3}$) >!
-\end{tabular}
-
-Similarly, for the gradient of !prod(A)! we obtain
-
-\begin{tabular}{c l}
-        & $\grad_\Gamma$(!prod(A)!) \\
-        & !let A$_{0}$= shift1R (scanl * 1 A) in!\\
-        & !let A$_{1}$= shift1L (map2 (a,b.b) A$_{0}$ A) in!\\
-        & !let A$_{2}$= map2 (a,b.a) A$_{0}$ A in!\\
-        & !let A$_{3}$= scanr * 1 A$_{1}$ in!\\
-        & !<prod(A), fun (X,z) -> Y(X+map2 (a,b. a*z*b) A$_{2}$ A$_{3}$)>! 
-    \end{tabular}
-\end{example}
-
-Though this transformation after the inlining above already has the right complexity, 
-its purity allows us to make the most out of generic optimizations.  
-A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.
-
-\begin{example}
-As shown in appendix \ref{sub:gradintro}, the optimizations from Figure~\ref{fig:optim} 
-are sufficient to show that the gradients of the terms of the introduction reduce to the following.
-
-\begin{tabular}{{r c l}}
-    $\nabla_A$!prod(A)! &=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))!\\
-    $\nabla_A$!sum(A)! &=& !map (x -> 1) A!\\
-    $\nabla_A$!dot(A,B)! &=& !B! 
-\end{tabular}
-\end{example}
-
 Admittedly, the transformation presented in this section may be hard to read, and it is not straightforward to show its correctness directly. 
-Next, we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).
+Next, we decompose this transformation into three simpler steps via a novel intermediate representation.

### 2021-07-08 rad-icfp-2021/complexity.tex
minor

@@
 
-In this section, we introduce a simple cost model on our language and show 
+In this section, we introduce a simple cost model for our language and show 
 that our reverse-mode transformation satisfies a version of the cheap gradient principle (Theorem \ref{thm:complexity}).

### 2021-07-08 rad-icfp-2021/complexity.tex
rewriting complexity section

--- 2021-07-08 rad-icfp-2021/complexity.tex
@@
 
-In this section, we introduce a simple cost model on our language and show 
-that our reverse-mode transformation satisfies a version of the cheap gradient principle (Theorem \ref{thm:complexity}).
+In this section, we introduce a simple cost model for our language and show 
+that, after partial evaluation, our reverse-mode transformation satisfies a version of the cheap gradient principle (Theorem \ref{thm:complexity}).
 
@@
 
-By inspection in Figure~\ref{fig:direct_diff_macro}, 
-we see that there is a linear usage of  each continuation $Y$ 
-and that it is always applied to almost the identity. 
+\begin{lemma}
+    Let $\Gamma \vdash$!e: A! be a term in Source. 
+    Every variable !Y! in $\directD{\rho}{\Gamma}{Y}$(!e!) which is not of ground type is used exactly once.
+\end{lemma}
+
+\begin{proof}
+    By induction on the judgment of !e!. We inspect each rule in Figure~\ref{fig:direct_diff_macro}.
+\end{proof}
+
+From the previous lemma, we see that there is a linear usage of  each continuation $Y$. 
+This is one key property which has been ensured by $\UNFSymbol$.
+We also note that, apart from the !map2! case, this continuation variable is always applied to almost the identity. 
 More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
 where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
-In fact we have $k=2$, except for the !map2! and !reduce! cases which require a more detailed analysis (see Section~\ref{sec:complexity}).
-We can first use inlining as given in Figure~\ref{fig:optim}.
+In fact we have $k=2$, except for the !map2! and !reduce!.
+
+We can first use inlining, the first optimization rule given in Figure~\ref{fig:optim}.
 Using the invariant above, most of the !e!$_{i}$ below are variables. 
 Without loss of generality, assume the only terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
-We can then use forward substitution on the other !e!$_{i}$.
+We can then use forward substitution, the third optimization rule in Figure~\ref{fig:optim}, on the other !e!$_{i}$.
 In brief:
@@
 
-Though this transformation after the inlining above already has the right complexity, 
-its purity allows us to make the most out of generic optimizations.  
-A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.
+We call this optimization step partial evaluation in the rest of the complexity section. 
+We have the following result.
 
-\begin{example}
-As shown in appendix \ref{sub:gradintro}, the optimizations from Figure~\ref{fig:optim} 
-are sufficient to show that the gradients of the terms of the introduction reduce to the following.
+\begin{lemma}
+    \label{lem:noHO}
+    Let $\Gamma \vdash$!e: A! be a term in Source. 
+    After the partial evaluation step, $\directD{\rho}{\Gamma}{Y}$(!e!) does not have lambda abstractions.
+    Its only variable (bound and free) which does not have a ground type is !Y!.
+\end{lemma}
 
-\begin{tabular}{{r c l}}
-    $\nabla_A$!prod(A)! &=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))!\\
-    $\nabla_A$!sum(A)! &=& !map (x -> 1) A!\\
-    $\nabla_A$!dot(A,B)! &=& !B! 
-\end{tabular}
-\end{example}
+\begin{proof}
+    By induction on the judgment of !e!. 
+    We inspect each rule in Figure~\ref{fig:direct_diff_macro} 
+    and note that the partial evaluation step precisely allows us to conclude the inductive step.
+\end{proof}
 
@@
 So the cost of !map (x.e)! on an array of size $n$ will be $n*\cost$(!e!).
-The cost function extends compositionally to the whole language in the obvious way.
-For instance we have !$\cost$(let x=e$_1$ in e$_2$)!= !$\cost$(e$_1$)+$\cost$(e$_2$)!.
-TODO: for let binding and same in map, do I need to add some MOVES?
+Following Lemma \ref{lem:noHO}, we do not need our cost model to deal with higher-order variables and lambda abstractions.
+We can thus restrict our attention to the subset of the Target language which does not contain lambdas, applications or variables which are not of ground type.
+The cost function extends compositionally to the restricted Target language. 
+It is given in Figure~\ref{fig:costmodel}.
+
+\input{figures/costmodel}
 
@@
 
+We can now phrase our main complexity theorem.
+
 \begin{theorem}
@@
     Given a term !$\Gamma \vdash$ e: $\reals$! such that $\NestA$(!e!)$\leq$ $p$.
-    Denote by $G$ the term $\grad_\Gamma$!e! after inlining and forward substitution.
-    TODO: be more precise with "inlining and forward substitution".
+    Denote by $G$ the term $\grad_\Gamma$!e! after the partial evaluation step from Section~\ref{sub:Partial evaluation and optimization}.
     Then !$\cost$(G) $\leq$ $4*2^{p}*\cost$(e)!.
@@
 A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
-After these optimizations, the output program is essentially a sequence of let bindings 
-and resembles SSA-code.
+After these optimizations, the output program is essentially a sequence of let-bindings 
+and resembles SSA-code \cite{cytron1989efficient}.
+
+Even though our reverse-mode transformation has the right complexity after the partial evaluation step, 
+the constant factor is of huge importance in practice. 
+The purity of our transformation allows us to make the most out of generic optimizations.  
+A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.
+In addition, hand-crafted efficient derivatives and more optimizations can easily be added to our language.
+
+\begin{example}
+    As shown in appendix \ref{sub:gradintro}, the optimizations from Figure~\ref{fig:optim} 
+    are sufficient to show that the gradients of the terms of the introduction reduce to the following.
+
+    \begin{tabular}{{r c l}}
+        $\nabla_A$!prod(A)! &=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))!\\
+        $\nabla_A$!sum(A)! &=& !map (x -> 1) A!\\
+        $\nabla_A$!dot(A,B)! &=& !B! 
+    \end{tabular}
+\end{example}

### 2021-07-08 rad-icfp-2021/intro.tex
fixing boundary issues
reorganizing sections 5 and 6
reworked the introduction a bit

@@
 
-In this paper, we define a compilation scheme for a purely functional, provably correct, and efficient reverse-mode AD. 
-To do so, we go start from a simple and general representation, where we can relatively easily prove correctness of reverse-mode.
-Then, we define the UNF representation inspired by SSA~\cite{cytron1989efficient} and ANF, that leads to more efficient implementations.
+In this paper, we define a purely functional (without references or control mechanisms such as state monads), denotationally correct, and provably efficient reverse-mode AD. 
+To do so, we define the Unary Normal Form (UNF) representation inspired by PROPs \cite{hackney2015category} and compilation to categories \cite{elliott2017compiling}.
+We can easily define and prove correctness of reverse-mode on this representation. 
+The whole reverse-mode transformation is obtained by compiling the language to this intermediate representation, applying the simpler reverse-mode transformation, and compiling again to the original language.
+After standard optimizations, the output program looks like SSA~\cite{cytron1989efficient} or ANF~\cite{sabry1993reasoning}, which leads to more efficient implementations.
 
@@
     \end{center}   
-For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.
+This is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.
 This term can be further optimized via constant propagation and algebraic simplifications to give
@@
 We introduce an alternative and easier to follow compilation pipeline for efficient reverse-mode AD (bottom of Figure~\ref{fig:outline}).
+\item We prove the complexity guarantees for the reverse ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
 \item We proof the correctness of our transformations (top/bottom parts of Figure~\ref{fig:outline}) by defining the denotational semantics of our languages using multicategories and concategories (Section~\ref{sec:correctness}).
-\item We prove the complexity guarantees for the reverse ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
 \end{itemize}
 
-Next, we recall rudiments of automatic differentiation, forward and reverse-mode differentiation
+Next, we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
 

### 2021-07-08 rad-icfp-2021/background.tex
minor changes section2

@@
 to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
-But while this is a helpful intuition, the formalization of infinitesimals can be intricate, 
-and the development in this paper is focussed on the elementary formulation in~\eqref{eqn:dualnumber}.
+But while this is a helpful intuition, the formalization of infinitesimals can be intricate.
 
@@
 The problem with the previous approach shows up when one wants to compute the full gradient of a function $\RR^n\to\RR$, for a large $n$. 
-Forward-mode only computes one directional derivative, for instance one partial derivative. This implies $n$ passes must be performed through the forward derivative to compute the whole gradient.
+Forward-mode only computes one directional derivative, for instance one partial derivative. 
+This implies $n$ passes must be performed through the forward derivative to compute the whole gradient.
 By using the symmetry in the chain rule, there is a way to compute the whole gradient faster, and this method is reverse-mode differentiation.
@@
 Because the computation flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
-(TODO: paragraph below from Matthijs' paper, need to rewrite.)
-Reverse-mode AD is only well-understood as a source-code transformation 
-(also called define-then-run style AD) on limited programming languages. 
+Reverse-mode AD is only well-understood as a source-code transformation on limited programming languages. 
 Typically, its implementations on more expressive languages that have features such as higher-order functions and conditionals
-make use of define-by-run approaches. 
-These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
+make use of define-by-run approaches. These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
 first-order program is left, and then they evaluate this new program \cite{carpenter2015stan,paszke2017automatic}. 
 Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. 
-As such, these AD libraries need to be implemented using carefully, manually optimized code, that for example does not contain any common subexpressions. 
-This implementation process is precarious and labour intensive. 
-Further, some whole-program optimizations that a compiler would detect go entirely unused in such systems.
+As such, the implementation process is precarious and labour intensive as these AD libraries need to be implemented using carefully, manually optimized code.
+In addition, some whole-program optimizations that a compiler would detect go entirely unused in such systems.
 
@@
 Following \cite{pearlmutter2008reverse}, there is a simple way to define an inefficient yet purely functional reverse-mode transformation for first-order programs.
-We review a slight modification of the transformation we presented, which is also better explained on an example.
+We review a slight modification of the transformation they presented, which is also better explained through an example.
 
@@
 
-This method is quite inefficient still. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
+This method is still quite inefficient. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
 Second, each $J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.
 
-The imperative version of reverse mode for a binary operator $op(x,y)$ is something like $x'+= \partial_1op(x,y);y'+= \partial_2op(x,y)$. 
-It is for these cases that mutation is usually key. 
+On the other hand, the imperative version of reverse mode for a binary operator $op(x,y)$ adds $x'+= \partial_1op(x,y);y'+= \partial_2op(x,y)$, 
+where $\partial_iop$ are the partial derivatives of $op$. 
+Mutation is usually key for these cases.
 If we see our term as a directed graph, reverse mode needs to backpropagate from the end of the graph to the starting nodes via every path.
@@
 One of the key simple ideas that we used was to transform every operator into a unary one. 
-This essentially trivialises the computation flow to a line. 
-Even if the starting program was a straight line program, 
-having non-unary operators was a source of inefficiency which justified mutation in the first place.
+This essentially trivializes the computation flow to a line. 
+Even if the starting program was a straight-line program, 
+having non-unary operators was a source of inefficiency and justified the use of mutation in the first place.
 By returning every variable every time, the problem of using a variable several times does not need to be dealt with via mutation. 
-This simple idea of transforming a program into essentially a straight line is what our new intermediate representation unary-form does.
+This simple idea of transforming a program into essentially a straight line is what our new intermediate representation UNF allows.
 
-\subsection{Insights for efficient reverse-mode}\
+\subsection{Insights for efficient reverse-mode}
 \label{subsec:insights}
@@
 the row is zero except for at most two indices (one for unary operators).
-This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,\ldots,y(n+i)).(y1,\ldots,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
+This means we can use a more compact representation $J^T\sem{op(x_i,x_j)}:=\lambda (y_1,\ldots,y_{n+i}).(y_1,\ldots,y_{n+i})+[i]\partial_1op(x_i,x_j)+[j]\partial_2op(x_i,x_j)$, 
 where $[k]$ means that the element is added at the $k$-th index of the tuple.

### 2021-07-08 rad-icfp-2021/generalization.tex
minor
added a paragraph 7.2

@@
 
+\smartpara{Recipe for adding array operators} To add more array operators to our language, 
+one doesn't need to go through UNF and most of the backend of our work can be used as a black box.
+The non-smooth variables of the operator (typically booleans or integers) should be considered as external the language, similarly to the way we treat the index $n$ of arrays.
+If the operator has type $A\to B$, then one should provide its transpose Jacobian, a term $B\to A$. 
+Such operators can often be unrolled to first-order programs. To check correctness of the given Jacobian, it suffices to 
+check that the semantics of the transpose Jacobian matches the Jacobian of the unrolled program. 
+To ensure the complexity guarantee of the whole transformation, one needs to check that there a linear usage of the continuation variable $Y$ and that the cost of the 
+proposed Jacobian is at most $k$ times the complexity of the operator.
+
 \subsection{Conditionals} % (fold)
@@
 
-This time, there is a linear usage of the continuation variable !Y! (and of the variables !x1,$\ldots$xn! too).
-
-We note that adding conditionals does not break differentiability as long as there are no non-trivial primitives $\reals\to\BB$.
+This time, there is a linear usage of the continuation variable !Y!. We note that adding conditionals does not break differentiability as long as there are no non-trivial primitives $\reals\to\BB$.
 Non smooth functions such as the Rectified Linear Unit (ReLU) in machine learning which can be defined as $ReLU(x)\defeq max(0,x)$ 

### 2021-07-08 rad-icfp-2021/generalization.tex
.

--- 2021-07-08 rad-icfp-2021/generalization.tex
@@
 Then we show how to support more array operations (Section~\ref{sub:lift_more_arr_op}), conditionals (Section~\ref{sub:lift_conditional}), and non-smooth scalar operators (Section~\ref{sub:lift_non_smooth}).
-Finally, we extend the language to support general arrays instead of supporting only $\Array{\reals}{n}$ in Section~\ref{sub:lift_gen_arr}.
+Afterwards, we extend the language to support general arrays instead of supporting only $\Array{\reals}{n}$ in Section~\ref{sub:lift_gen_arr}.
+Finally, we provide a recipe for adding more constructs to the language in Section~\ref{sub:lift_recipe}.
 
@@
 More generally, there is growing interest in tensor calculus \cite{liao2019differentiable,bernstein2020differentiating}.
+
+\subsection{Recipe for Adding More Constructs}
+\label{sub:lift_recipe}
+To add more constructs to our language, 
+there is no need to go through UNF; most of the backend of our work can be used as a black box.
+The non-smooth variables of the operator (typically booleans or integers) should be considered as external to the language, similarly to the way we treat the index $n$ of arrays.
+
+\smartpara{Reverse AD}
+If the operator has type $A\to B$, then one should provide its transpose Jacobian, a term $B\to A$. 
+Such operators can often be unrolled to first-order programs.
+
+\smartpara{Correctness}
+To check correctness of the given Jacobian, it suffices to 
+check that the semantics of the transpose Jacobian matches the Jacobian of the unrolled program.
+
+\smartpara{Complexity}
+To ensure the complexity guarantee of the whole transformation, one needs to check that there a linear usage of the continuation variable $Y$ and that the cost of the 
+proposed Jacobian is at most $k$ times the complexity of the operator.

### 2021-07-08 rad-icfp-2021/correctness.tex
added a paragraph 7.2
some progress correctness

@@
 Substitution is interpreted as composition. We first consider a syntactic model for a language, which consists of a 
-free multicategy on some base types and primitives. Our source and target languages induce syntactic multicategories as follows.
+free multicategory on some base types and primitives. Our source and target languages induce syntactic multicategories as follows.
 
@@
         !$\seml$c$\semr$! & $\defeq$ & $c\in\RR$ \\
-        !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & TODO \\
-        !$\seml$reduce (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$! & $\defeq$ &
+        !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>;swap;$(\seml$ e$_1\semr)^{\times n}$! \\
+        !$\seml$reduce (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$! & $\defeq$ <> &
     \end{tabular}
+where $\Delta_n$ is the $n$-copy map !<id,$\ldots$,id>!.
+The semantics for !reduce (x,y.e$_1$) e$_2$ e$_3$!
+is given by first !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>!, following a swap isomorphism  $\Gamma^n \times \RR \times \RR^n \to \reals \times (\reals \times \Gamma)^n$.
+Finally, we apply $\seml$ e$_1\semr \times id$ $n$ times, where the identity is of the obvious type at each stage.
 
-Variables are interpreted as projections, let binding uses composition in the multicategory. 
+Variables are interpreted as projections $\pi_i$, let binding as composition in the multicategory. 
 Pairs are interpreted using the Cartesian structure of the multicategory.
@@
 We can similarly interpret Target in a multicategory of smooth-like spaces and functions. 
-However, Target is higher-order but $\CartSp$ is not Cartesian Closed. 
-Instead, we can for instance interpret Target in the category of Diffeological spaces, 
+However, Target is higher-order and $\CartSp$ is not Cartesian Closed. 
+Instead, we can interpret Target in the category of Diffeological spaces, 
 as in \cite{huot2020correctness}. 
 Diffeological spaces (\cite{iglesias2013diffeology}) are a conservative extension of $\CartSp$. 
-The key idea will be that a higher order function is called smooth if it sends smooth functions to smooth functions, meaning that we can never use it
+The key idea will be that a higher-order function is called smooth if it sends smooth functions to smooth functions, meaning that we can never use it
 to build first order functions that are not smooth.
@@
 Many constructions of sets carry over straightforwardly to diffeological spaces.
-
-\begin{example}[Cartesian diffeologies]\label{ex:cartesian-diffeologies}
-Each open subset $U$ of $\RR^n$ can be given the structure of a diffeological space by taking all the
-smooth functions $V\to U$ as $\plots{U}^V$.
-It is easily seen that smooth functions from $V\to U$ in the traditional sense coincide with
-smooth functions in the sense of diffeological spaces.
-Thus diffeological spaces have a profound relationship with ordinary calculus.
-\end{example}
-
-\begin{example}[Product diffeologies]
-Given a family $\seq[i\in I]{X_i}$ of diffeological spaces,
+For instance, given a family $\seq[i\in I]{X_i}$ of diffeological spaces,
 we can equip the product $\prod_{i\in I}X_i$ of sets with the
@@
 of the form $\seq[i\in I]{p_i}$ for $p_i\in\plots{X_i}^U$.  
-\end{example}
-
-\begin{example}[Functional diffeology]
-We can equip the set $\Diff(X,Y)$ of smooth functions between diffeological spaces with the \emph{functional diffeology}
+Cartesian spaces $\RR^n$ can be given the structure of a diffeological space by taking all the
+smooth functions $U\to \RR^n$ as $\plots{\RR^n}^V$. We can equip the set $\Diff(X,Y)$ of smooth functions between diffeological spaces with the \emph{functional diffeology}
 in which $U$-plots consist of functions $f:U\to \Diff(X,Y)$ such that 
 $(u,x)\mapsto f(u)(x)$ is an element of $\Diff(U\times X, Y)$.
-\end{example}
-
 We can thus interpret function types !$\seml$A -> B$\semr$! = $\Diff$($\seml$!A!$\semr$,$\seml$!B!$\semr$).
@@
 which satisfies a similar universal property as syntactic multicategories.
-Two partiular examples are of interest for us, as they will allow us to interpret Source and Target UNF
+Two particular examples are of interest for us, as they will allow us to interpret Source and Target UNF
 but they will also help us explain the UNF and $UNF^{-1}$ transformations.
@@
 with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$, 
-TODO: $map2_{x,y.e}$and $reduce_{x,y.e1;e2}$.
+$map2_{x,y.e}: \Array{\reals}{n},\Array{\reals}{n}\to \Array{\reals}{n}$ and $reduce_{x,y.e1;e2}:\Array{\reals}{n}\to \reals$.
 \end{definition}
@@
 One may notice that the syntax of $\ConcatS$ is somewhere in between the syntax of Source and of Source UNF.
-
-\begin{notation}[Morphisms in a Cartesian concategory]
-    TODO: projection, unique map, $\concatcomp$, swap, pair, id, ...
-\end{notation}
-
+Given two morphisms !e$_1$:T$_1$->T$_2$! and !e$_2$:T$_3$->T$_4$!, we denote by !e$_1 \concatcomp$ e$_2$:T$_1$,T$_3$->T$_2$,T$_4$!.
+We denote by $u_T$ the unique morphism from !T! to the terminal object $[]$.
 We can interpret Source UNF in $\ConcatS$ as follows.
@@
    $\seml$!pair!$_{T;A\times B} \semr$ &=& $pair_{T;A\times B}$ \\
-   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T1}\concatcomp 1_{T2}\concatcomp id_{T3}$\\
+   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T1}\concatcomp u_{T2}\concatcomp id_{T3}$\\
    $\seml$!e1!$\comp$!e2!$\semr$  &=& $\seml$!e1!$\semr \comp \seml$!e2!$\semr$ \\
@@
 
-TODO: I think need to explain here the (co)monoid structure induced by $\Dsynrevsymbol_{\rho}$
-
 \begin{tabular}{l c l}
@@
     We define $\CSource$ to be the multicategory with the same objects as Source and 
-    $\CSource([A_1,\ldots,A_n],B)=\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B])$.
+    $\CSource([A_1,\ldots,A_n],B)=\{f\in\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B], \forall i.f;\pi_{A_i}=id_{A_i})$.
+    The composition of $f_i:\underline{A}_i\to B_i$ with $g:\underline{B}\to C$ is given 
+    $f_1\concatcomp\ldots f_n;(u_{\underline{B}_1}\concatcomp A_1 \concatcomp \ldots \concatcomp u_{\underline{B}_n}\concatcomp A_n);g$.
+    In other words each term $f_i$ forgets about its output $\underline{B}_i$, and then we use the composition in the concategory.
 \end{definition}
+
+We can interpret Source in $\CSource$ as follows.
+TODO
 
@@
     A multicategory $\catC$ naturally defines a concategory with the same objects as $\catC$ and
-    with morphisms $A\to [B_1,\ldots,B_n]$ being $n$ morphisms $A\to Bi$ of $\catC$. 
+    with morphisms $\underline{A}\to [B_1,\ldots,B_n]$ being $n$ morphisms $\underline{A}\to B_i$ of $\catC$. 
 \end{definition}
 
-\begin{proposition}[Initiality of concat2 gives $UNF^-1$]
+This allows to form a concategory $\CTarget$ from the syntactic multicategory for Target.
+We can interpret $\ConcatT$ in $\CTarget$ as follows.
 TODO
-\end{proposition}
+
+This interpretation is in essence $\UNFSymbol^{-1}$. 
+The difference is that to preserve typing, the semantic $\UNFSymbol^{-1}$
+sends a non Jacobian primitive to a tuple, as in the example \ref{}. 
+This is highly inefficient and the the syntactic $\UNFSymbol^{-1}$ 
+additionally projects to the last element, where the non trivial information of the term is.
 
@@
 
-\begin{definition}[D as a lax functor]
-    TODO
+\begin{definition}[$\Dsynrevsymbol_{\rho}$ as a lax functor]
+    We define $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$ as follows.
+    $\Dsynrevsymbol_{\rho}$(![A$_1$,$\ldots$,A$_n$])=[A$_1$,$\ldots$,A$_n$,A$_1\times\ldots\times$A$_n\to \rho$]!.
 \end{definition}
@@
 
-Combining the the previous propositions, we have shown that $\directD{\rho}{\Gamma}{Y}$ is correct.
+Combining the previous propositions, we have shown that $\directD{\rho}{\Gamma}{Y}$ is correct.
 

### 2021-07-08 rad-icfp-2021/unf.tex
minor
reorganizing sections 5 and 6
cost model figure
more improvement correctness

@@
 Every primitive is indexed by a list of types from the source language, 
-which corresponds to a context being carried. 
+which corresponds to a context being carried. We will often elide the brackets ![!,!]!.
 Every constant, unary, or binary operator from Source has a corresponding $n$-ary operator !op$_{T,n}$! in UNF.
@@
 \begin{example}
+    \label{exm:unf}
     The term in Source UNF !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$! 
@@
 
-    \begin{tabular}{r c l}
-        !$\Dsynrevsymbol_{\rho}$(cos$_{\reals,\reals}\comp$ pair$_{\reals,\reals; \reals\times\reals}$)!
-        &=& !<> $\comp$! \\
-        &&  !<> !
+    \begin{tabular}{l}
+        !$\reals$,$\reals$,$\reals$,$\reals\times\reals\times\reals$->$\rho \vdash \Dsynrevsymbol_{\rho}$(cos$_{\reals,\reals}\comp$ pair$_{\reals,\reals; \reals\times\reals}$): $\reals$,$\reals$,$\reals \times \reals$,$(\reals\times\reals\times(\reals\times\reals))$->$\rho$!
+\\
+        = !<proj$_{\reals^3;\reals^3->\rho;[]} \comp$ cos$_{\reals,\reals}$, proj$_{[];\reals^3;\reals^3->\rho} \icomp$ (proj$_{\reals^3;\reals^3->\rho;[]}$ $\comp$ J$^T$cos$_{\reals,\reals}$)> $\comp$!
+\\
+        !<proj$_{\reals^4;\reals^4->\rho;[]} \comp$ pair$_{\reals,\reals; \reals\times\reals}$, proj$_{[];\reals^4;\reals^4->\rho} \icomp$ (proj$_{\reals^4;\reals^4->\rho;[]}$ $\comp$ J$^T$pair$_{\reals,\reals; \reals\times\reals}$)>!
     \end{tabular}
@@
 
-% TODO: we could merge everything of the form op;proj into just op by changing the type of op.
-
 \input{figures/source_to_unf}
@@
 Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of the ANF transformation \cite{sabry1993reasoning}.
-
-\begin{example}
-    TODO
-\end{example}
 
@@
 \begin{example}
-    TODO
+    Continuing again with our previous example !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$!, 
+    we have, after removing a few unnecessary let-bindings:
+
+    \begin{tabular}{l}
+        !x$_1$,x$_2$,x$_3$:$\reals$,x$_4$:$\reals^3$->$\rho$ $\vdash$ $\UNFSymbol^{-1}$($\Dsynrevsymbol_{\rho}$(cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$): $(\reals\times\reals) \times (\reals\times\reals\times(\reals\times\reals))$->$\rho$!
+\\
+        = !TODO!
+    \end{tabular}
 \end{example}

### 2021-07-08 rad-icfp-2021/correctness.tex

@@
 
-We can interpret Source in $\CSource$ as follows.
-TODO
+We can interpret Source in $\CSource$ as follows. The functor is an identity on types. 
+On morphisms, we interpret them as morphisms in the concategory as follows.
+The terminal map is interpreted as the identity. The operators !op1,op2! by themselves.
+Crucially, the semantics of a !let! becomes composition in the concategory. 
+The semantics of a variable is the pairing of the identity with a projection.
+The operation !map2 (x,y.e)! is interpreted as the pairing of the identity and itself, and !reduce (x,y.e$_1$) $e_2$! as itself.
+
+Interpretating Source in $\CSource$ allows us to see terms of Source as morphisms in the concategory $\ConcatS$.
+This allows us to compare to semantics of terms of Source and of Source UNF.
+The following proposition can be shown by induction on the structure of the terms.
 
 \begin{proposition}[construction above gives UNF]
-    TODO
+    Let !$\Gamma \vdash$ e:A! be a term in Source. 
+    Seen as morphisms in $\CSource$, we have !$\seml$e$\semr$=$\seml\UNFSymbol$(e)$\semr$!.
 \end{proposition}
@@
 This allows to form a concategory $\CTarget$ from the syntactic multicategory for Target.
-We can interpret $\ConcatT$ in $\CTarget$ as follows.
-TODO
+We can interpret $\ConcatT$ in $\CTarget$. 
+The functor is identity on objects, sends operations !op1,op2! to themselves. 
+It sends Jacobian operations to terms of Target as given by $\UNFSymbol^{-1}$ in Figure~\ref{fig:unf_to_target}
 
@@
 The difference is that to preserve typing, the semantic $\UNFSymbol^{-1}$
-sends a non Jacobian primitive to a tuple, as in the example \ref{}. 
+sends a non Jacobian primitive to a tuple, as in the Example \ref{exm:unf}. 
 This is highly inefficient and the the syntactic $\UNFSymbol^{-1}$ 
@@
     $\Dsynrevsymbol_{\rho}$(![A$_1$,$\ldots$,A$_n$])=[A$_1$,$\ldots$,A$_n$,A$_1\times\ldots\times$A$_n\to \rho$]!.
+    TODO
 \end{definition}

### 2021-07-08 rad-icfp-2021/purerad.tex
.

@@
 It consists of a first-order functional language with arrays and a few typical second-order array operations. 
-The types !T1,T2! and terms !e1,e2! are given in Figure~\ref{fig:source_grammar}.
+The types !T1,T2!, terms !e1,e2!, and typing rules are given in Figure~\ref{fig:source_grammar}.
 We have included a minimal set of array operations for the sake of illustration,  it is not hard to add more.
@@
 
-The typing rules are given in Figure~\ref{fig:source_typesystem}. 
 For scalar operations, we assume a given set of operations, including $+$ and $*$. 
@@
 We show how to lift these restrictions and how to differentiate some other array operators in Section~\ref{sec:generalization}.
-
-\input{figures/source_type_system}
 
@@
 
-The types and terms are presented in Figure~\ref{fig:target_grammar}.
+The grammar for types and terms along with the type system of the target language are presented in Figure~\ref{fig:target_grammar}.
 Our lambda abstractions take $n$ arguments as we are not interested here with partial applications. 
@@
 and let expressions binding such lambda abstractions. We only need the target language to be second-order.
-The type system for the extended grammar of the target language is presented in Figure~\ref{fig:target_typesystem}.
 
 \input{figures/target_grammar}
-
-\input{figures/target_type_system}
 

### 2021-07-08 rad-icfp-2021/unf.tex
.

@@
 A judgment in Source UNF is a triple !(T$_1$,e,T$_2$)! where !T$_1$! and !T$_2$! are lists of types of Source, and !e! is an expression of the Source UNF.
-The typing rules are detailed in Figure~\ref{fig:source_unf_typesystem}.
+The typing rules are detailed in Figure~\ref{fig:unf_source_grammar}.
 $\reals^{\times n}$ is a notation for $\reals,\ldots,\reals$ with $n$ factors $\reals$. 
 When !T$_1$! and !T$_2$! are lists of types and !A! a type, the operator !,! denotes snoc in !T$_1$,A! and denotes append in !T$_1$,T$_2$!.
-
-\input{figures/source_unf_typesystem}
 
@@
 It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
-The grammar of Target UNF is given in Figure~\ref{fig:unf_target_grammar} and its type system in Figure~\ref{fig:target_unf_typesystem}.
+The grammar and type system of target UNF is given in Figure~\ref{fig:unf_target_grammar}.
 
 \input{figures/target_unf}
-
-\input{figures/target_unf_typesystem}
 

### 2021-07-08 rad-icfp-2021/relwork.tex
minor
gaining space section 8-9

@@
 
-\noindent \textbf{More dynamic language}
-JAX \cite{bradbury2020jax,frostig2018compiling} is a modern framework that provides state-of-the-art automatic differentiation. 
-One feature that makes it popular is that it uses JIT compilation, 
-which seems to be the best avenue to help best ML practitioners as it allows them 
-to quickly experiment but also benefit from the full power of static optimizations.
-We are interested in seeing how our work could sit in a JIT compiler.
+% \noindent \textbf{More dynamic language}
+% JAX \cite{bradbury2020jax,frostig2018compiling} is a modern framework that provides state-of-the-art automatic differentiation. 
+% One feature that makes it popular is that it uses JIT compilation, 
+% which seems to be the best avenue to help best ML practitioners as it allows them 
+% to quickly experiment but also benefit from the full power of static optimizations.
+% We are interested in seeing how our work could sit in a JIT compiler.
 
-\noindent \textbf{More optimizations.}
-As pointed in \ref{sub:Adding general arrays} it would be interesting to have a more principled representation 
-of tensors which could lead to more optimizations. 
-There is always a tension between dynamicity and the possibility for static optimizations 
-and we believe our language offers a good tradeoff, but it will be interesting to tailor it 
-further for specific applications.
+% \noindent \textbf{More optimizations.}
+% As pointed in \ref{sub:Adding general arrays} it would be interesting to have a more principled representation 
+% of tensors which could lead to more optimizations. 
+% There is always a tension between dynamicity and the possibility for static optimizations 
+% and we believe our language offers a good tradeoff, but it will be interesting to tailor it 
+% further for specific applications.
 
-\noindent \textbf{Cost function and memory accesses}
-Our cost model \ref{sub:costModel} is simple but has limits. 
-In particular, it does not capture well cost in term of
-parallel computing, and memory management (cache optimization is one of best way to optimize such linear algebra intensive computations).
-For instance, in the earlier days of automatic differentiation, 
-advanced checkpointing techniques were used to avoid storing all the intermediate values just to save memory 
-and try avoiding saturation of RAM. 
-These problems are harder to analyze and, eventually, it seems to come down to a lot of experimenting and benchmarking.
+% \noindent \textbf{Cost function and memory accesses}
+% Our cost model \ref{sub:costModel} is simple but has limits. 
+% In particular, it does not capture well cost in term of
+% parallel computing, and memory management (cache optimization is one of best way to optimize such linear algebra intensive computations).
+% For instance, in the earlier days of automatic differentiation, 
+% advanced checkpointing techniques were used to avoid storing all the intermediate values just to save memory 
+% and try avoiding saturation of RAM. 
+% These problems are harder to analyze and, eventually, it seems to come down to a lot of experimenting and benchmarking.
 
-\noindent \textbf{Parallelism and GPU}
-We tried to locally preserve parallelism. Having a pure transformation should help with found parallelism. 
-It would be interesting to implement on transformations and see how they perform on GPU.
-It is well-known that the addition of references provoked by the standard reverse-mode transformation
-restricts a lot of optimizations and it is not hard to find examples in C where the compiler does not detect simple optimizations 
-because of the references introduced by reverse-mode.
+% \noindent \textbf{Parallelism and GPU}
+% We tried to locally preserve parallelism. Having a pure transformation should help with found parallelism. 
+% It would be interesting to implement on transformations and see how they perform on GPU.
+% It is well-known that the addition of references provoked by the standard reverse-mode transformation
+% restricts a lot of optimizations and it is not hard to find examples in C where the compiler does not detect simple optimizations 
+% because of the references introduced by reverse-mode.
 
@@
 
-\noindent \textbf{Recursion}
-It would be an interesting extension to our work to see if one can define efficient pure reverse derivatives on general recursion.
+% \noindent \textbf{Recursion}
+% It would be an interesting extension to our work to see if one can define efficient pure reverse derivatives on general recursion.
 
-\noindent \textbf{Higher derivatives}
-There are recent approaches to computing higher derivatives using jets \cite{betancourt2018geometric} which generalise forward and reverse mode.
-A more standard approach is to iterate forward or reverse-mode. 
-Forward mode is standard and can easily be defined on our Target language. 
-As our optimization procedure removes lambdas, to iterate reverse-mode we would only need to compute reverse derivatives for the new primitives of Target, that is !scanl, scanr, shift1L, shift1R!.
+% \noindent \textbf{Higher derivatives}
+% There are recent approaches to computing higher derivatives using jets \cite{betancourt2018geometric} which generalise forward and reverse mode.
+% A more standard approach is to iterate forward or reverse-mode. 
+% Forward mode is standard and can easily be defined on our Target language. 
+% As our optimization procedure removes lambdas, to iterate reverse-mode we would only need to compute reverse derivatives for the new primitives of Target, that is !scanl, scanr, shift1L, shift1R!.
 
-\noindent \textbf{Further applications}
-Reverse-mode differentiation is notoriously difficult to grasp, implement, and optimize.
-We hope this work could help develop a new perspective on pure reverse-mode automatic differentiation.
-Other applications include the use of fancier datatypes e.g. dependent types, 
-in probabilistic programming, e.g. for HMC or Variational Inference. 
-It is also of interest to see how to extend this work with other techniques which 
-have had renewed interest such as implicit differentiation \cite{blondel2021efficient,lorraine2020optimizing}.
+% \noindent \textbf{Further applications}
+% Reverse-mode differentiation is notoriously difficult to grasp, implement, and optimize.
+% We hope this work could help develop a new perspective on pure reverse-mode automatic differentiation.
+% Other applications include the use of fancier datatypes e.g. dependent types, 
+% in probabilistic programming, e.g. for HMC or Variational Inference. 
+% It is also of interest to see how to extend this work with other techniques which 
+% have had renewed interest such as implicit differentiation \cite{blondel2021efficient,lorraine2020optimizing}.
 
@@
 \label{sec:related_work}
+
+\noindent \textbf{Correctness of AD in functional languages.}
+Several recent work \cite{huot2020correctness,vakar2020reverse,vakar2020denotational,brunel2019backpropagation,barthe2020versatility,mazza2021automatic,ee2020correctness} have focused on correctness of AD in a purely functional setting,
+often leaving efficiency on the side, especially for reverse-mode differentiation. 
+We see our work as a complement and a first bridge between these works 
+and more practical considerations of efficiency and implementations, 
+which often require a lot more care than is acknowledged in more theoretical works.
+
+An exception is \cite{abadi-plotkin2020} which gives a reverse-mode transformation that is proved correct and with the right complexity.
+This work is purely first-order, does not have arrays and uses a non-standard semantics. 
+Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.
+They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
+Instead, they follow the principle of unrolling for recursive calls and evaluating the conditional before differentiating, 
+which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
+This is a well-known limit in TensorFlow \cite{abadi2016tensorflow} which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.
 
@@
 which is usually derived by hand and one main goal of these papers.
-
-\noindent \textbf{Correctness of AD in functional languages.}
-Several recent work \cite{huot2020correctness,vakar2020reverse,vakar2020denotational,brunel2019backpropagation,barthe2020versatility,mazza2021automatic,ee2020correctness} have focused on correctness of AD in a purely functional setting,
-often leaving efficiency on the side, especially for reverse-mode differentiation. 
-We see our work as a complement and a first bridge between these works 
-and more practical considerations of efficiency and implementations, 
-which often require a lot more care than is acknowledged in more theoretical works.
-
-An exception is \cite{abadi-plotkin2020} which gives a reverse-mode transformation that is proved correct and with the right complexity.
-This work is purely first-order, does not have arrays and uses a non-standard semantics. 
-Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.
-They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
-Instead, they follow the principle of unrolling for recursive calls and evaluating the conditional before differentiating, 
-which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
-This is a well-known limit in TensorFlow \cite{abadi2016tensorflow} which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.
 

### 2021-07-08 rad-icfp-2021/correctness.tex
added Sam's intro section 6

@@
 
-This section provides a categorical analysis of our language and transformations.
-The main result at the end of section is the correctness of the reverse mode transformation from Section~\ref{sub:Macro for pure reverse mode transformation}:
+We now explain that the correctness of the three steps of the $\directD{\rho}{\Gamma}{Y}$ transformation from Section~\ref{sub:Macro for pure reverse mode transformation} can be understood in terms of translating between different standard categorical languages,
+and verified by working in a well-known category for differentiability: diffeological spaces.
+
+In general, categories with products (and more generally monoidal categories) can be equivalently presented in the following three ways. 
+Although the three styles of presentation are categorically equivalent, the choice of presentation affects the internal language syntax.
+\begin{itemize}
+    \item categories, where each arrow has one source and one target. These are most familiar but don’t directly match with a programming syntax (except the categorical abstract machine). 
+    \item multicategories, where each arrow has a list of source objects !A$_1,\ldots,$A$_n$! and one target !B!, thought of as a map !A$_1$ $\times\ldots\times$A$_n\to$B!. 
+    These are very close to the syntax usually used in typed programming languages \cite{ambek1968deductive,staton2013universal}, 
+    thinking of an arrow as a typed term !x$_1$:A$_1,\ldots,$x$_n$:A$_n$ $\vdash$ B!.
+    \item concategories, aka coloured props \cite{bonchi2015full,fong2019backprop}, where each arrow has a list of source objects !A$_1\ldots$A$_n$!
+and a list of target objects !B$_1\ldots$B$_m$!, thought of as a map !A$_1\times\ldots\times$A$_m \to$ B$_1\times\ldots\times$B$_m$!.
+These match the syntax of the UNF language.
+\end{itemize}
+
+Informally, notice that it makes sense to talk about the opposite of a category or a concategory, but not the opposite of a multicategory. 
+For roughly this reason, concategories are a more natural place to consider reverse derivatives than multicategories, 
+even though they are syntactically less familiar. 
+The $\directD{\rho}{\Gamma}{Y}$ translation passes from the multicategory for the source syntax to the concategory for UNF, where reverse-mode differentiation is easier, and then back to the multicategory for the target syntax.
+In this way the translations between Source/Target and Source UNF/Target UNF are merely changing perspective from multicategories to concategories.
+We work with a specific multicategory and concategory built from diffeological spaces and smooth maps, so that we can verify the construction of reverse derivatives. 
+Since diffeological spaces support products, they can be presented as a category, a multicategory and a concategory. 
+In abstract mathematics the difference between these presentations is almost trivial. 
+But in terms of programming syntax, the difference between presentations has a big effect, as can be seen from the difference between the complex translation in Section~\ref{sec:simplediff} and the simple translations in Section~\ref{sec:unf}, which are much more readily verified.
 
@@
      For every $\Gamma \vdash$ !e:!$\reals$, we have 
-     $\sem{\grad_\Gamma e}= \grad_\Gamma\sem{e}$.
+     $\sem{\grad_\Gamma e}= \grad_\Gamma\sem{e}$ in the category of diffeological spaces.
  \end{theorem}
-where $\sem{-}$ is an appropriate denotational semantics of smooth functions.
 

### 2021-07-08 rad-icfp-2021/correctness.tex
fixing some refs

@@
     \item multicategories, where each arrow has a list of source objects !A$_1,\ldots,$A$_n$! and one target !B!, thought of as a map !A$_1$ $\times\ldots\times$A$_n\to$B!. 
-    These are very close to the syntax usually used in typed programming languages \cite{ambek1968deductive,staton2013universal}, 
+    These are very close to the syntax usually used in typed programming languages \cite{lambek1968deductive,staton2013universal}, 
     thinking of an arrow as a typed term !x$_1$:A$_1,\ldots,$x$_n$:A$_n$ $\vdash$ B!.

### 2021-07-08 rad-icfp-2021/relwork.tex
fixing some refs

@@
 \noindent \textbf{Correctness of AD in functional languages.}
-Several recent work \cite{huot2020correctness,vakar2020reverse,vakar2020denotational,brunel2019backpropagation,barthe2020versatility,mazza2021automatic,ee2020correctness} have focused on correctness of AD in a purely functional setting,
+Several recent work \cite{huot2020correctness,vakar2020reverse,vakar2020denotational,brunel2019backpropagation,barthe2020versatility,mazza2021automatic,lee2020correctness} have focused on correctness of AD in a purely functional setting,
 often leaving efficiency on the side, especially for reverse-mode differentiation. 
@@
 TODO
-\cite{ee2020correctness}: non diff function
+\cite{lee2020correctness}: non diff function
 \cite{sherman2021}: non diff HO functions

### 2021-07-08 rad-icfp-2021/background.tex
.

@@
 
+The idea of automatic differentiation (AD) is to systematically
+transform a differentiable function $g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives.
 An intuition for $h$ is often given in terms of dual numbers.
@@
 to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
-But while this is a helpful intuition, the formalization of infinitesimals can be intricate.
 
-The reader may also notice that $h$ encodes all the partial derivatives of
-$g$. For example, 
-if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
-$h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
-and similarly 
-$h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
-And conversely, if $g$ is differentiable in each argument, then
-a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
-combinations of partial derivatives:
-
-\begin{center}
-    \begin{tabular}{r c l}
-        $h(x_1,x_1',x_2,x_2')$ &= & $(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))$
-    \end{tabular}
-\end{center}
-
-In summary, the idea of differentiation with dual numbers is 
-to transform a differentiable function
-$g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives.
-We packaged this up in~\eqref{eqn:dualnumber} as a sort-of invariant which is useful for building derivatives of compound functions $\RR\to\RR$ in a compositional way.
-The idea of automatic differentiation is to perform this transformation at the source code level.
-
-The main two ways in which it is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
+The main two ways in which AD is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
 Our approach focuses on source code transformation which is better fitted for compilation and optimizations.
 
-\subsection{Reverse-mode differentiation}
+% An intuition for $h$ is often given in terms of dual numbers.
+% The transformed function operates on pairs of numbers, $(x,x')$, and it is common
+% to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
+% But while this is a helpful intuition, the formalization of infinitesimals can be intricate.
+
+% The reader may also notice that $h$ encodes all the partial derivatives of
+% $g$. For example, 
+% if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
+% $h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
+% and similarly 
+% $h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
+% And conversely, if $g$ is differentiable in each argument, then
+% a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
+% combinations of partial derivatives:
+
+% \begin{center}
+%     \begin{tabular}{r c l}
+%         $h(x_1,x_1',x_2,x_2')$ &= & $(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))$
+%     \end{tabular}
+% \end{center}
+
+% In summary, the idea of differentiation with dual numbers is 
+% to transform a differentiable function
+% $g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives.
+We packaged this up in~\eqref{eqn:dualnumber} as a sort-of invariant which is useful for building derivatives of compound functions $\RR\to\RR$ in a compositional way.
+% The idea of automatic differentiation is to perform this transformation at the source code level.
+
+% The main two ways in which it is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
+% Our approach focuses on source code transformation which is better fitted for compilation and optimizations.
+
+\subsection{Reverse-mode AD}
 
@@
 
-\subsection{Purely functional inefficient reverse-mode}
+\subsection{Inefficiency of purely functional reverse-mode AD}
 
@@
 
-To actually reverse the order of computation needed for this transpose of jacobians, we use a sort of simple continuation.
+To actually reverse the order of computation needed for this transpose of jacobians, we use a sort of simple continuation;
 $f_i$ is turned into $\Dsynrevsymbol{f_i}:=<f_i, \lambda Y. Y\circ J^Tf_i>$ where $Y:\RR^{n+i-1}\to \RR^n$. 
-We recover compositionality by noting that $<f_{i+1}(f_i), (\lambda Y. Y\circ J^Tf_{i+1}>)(\lambda Y. Y\circ J^Tf_i)>$ reduces to
+We recover compositionality by noting that $<f_{i+1}(f_i), (\lambda Y. Y\circ J^Tf_{i+1})(\lambda Y. Y\circ J^Tf_i)>$ reduces to
 $<f_{i+1}\circ f_i, \lambda Y. Y\circ J^Tf_i \circ J^Tf_{i+1}>$, and thus by induction we can obtain $<f, \lambda Y. YJ^Tf>$.
@@
 
-This method is still quite inefficient. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
-Second, each $J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.
+This purely functional implementation has the following issues in terms of efficiency:
 
-On the other hand, the imperative version of reverse mode for a binary operator $op(x,y)$ adds $x'+= \partial_1op(x,y);y'+= \partial_2op(x,y)$, 
+\smartpara{Issue 1}
+If we see the term as a directed graph, reverse mode needs to backpropagate from the end of the graph to the starting nodes via every path.
+However, it is hard to keep track of all these information in parallel in a functional way.
+Mutation is usually key for these cases; 
+the imperative version of reverse mode for a binary operator $op(x,y)$ adds $x'+= \partial_1op(x,y);y'+= \partial_2op(x,y)$, 
 where $\partial_iop$ are the partial derivatives of $op$.
-Mutation is usually key for these cases.
-If we see our term as a directed graph, reverse mode needs to backpropagate from the end of the graph to the starting nodes via every path.
-It's hard to keep track of all these information in parallel in a functional way.
+
+\smartpara{Issue 2} 
+Each $J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.
+
+\smartpara{Issue 3}
+We have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
+
+\subsection{Insights for efficient purely functional reverse-mode AD}
+\label{subsec:insights}
+Overall we use the following three insights to solve the inefficiency associated with the purely functional implementations of reverse-mode AD.
+
+\smartpara{Insight 1}
 One of the key simple ideas that we used was to transform every operator into a unary one. 
@@
 
-\subsection{Insights for efficient reverse-mode}
-\label{subsec:insights}
-
+\smartpara{Insight 2} 
 If we look at $J^Tf_{i+1}$, we notice that this function is almost the identity, except at the last row. 
@@
 
-We know in advance that such jacobian functions are going to be applied one to another and we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
+\smartpara{Insight 3}
+We know in advance that the jacobian functions are going to be applied one to another and we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
 Because each function is almost the identity, we obtain a lot of substitutions of the form $[x/y]$ where both $x$ and $y$ are variables. 

### 2021-07-08 rad-icfp-2021/background.tex
Minor revisions

--- 2021-07-05 rad-icfp-2021/background.tex
@@
 Suppose given a function $f=f_n\circ\ldots\circ f_1 :\RR^n\to\RR$. 
-The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(\ldots(Jf_1v))\ldots)$ for a direction $v\in\RR^n$. 
-Reverse-mode, on the other hand, computes $(Jf)^Tv =J^Tf_1(J^Tf_{2}(\ldots(J^Tf_nv))\ldots)$ for a vector $v\in\RR$.
+The simplest way to see this mathematically, is first to say that forward mode essentially computes $(\J f)v=\J f_n(\J f_{n-1}(\ldots(\J f_1v))\ldots)$ for a direction $v\in\RR^n$. 
+Reverse-mode, on the other hand, computes $(\J f)^Tv = \J^Tf_1(\J^T f_{2}(\ldots(\J^T f_nv))\ldots)$ for a vector $v\in\RR$.
 In particular, taking $v=1$ computes the gradient of $f$.
@@
 
-Let the term we want to differentiate be $x1:\RR,\ldots,xn:\RR\vdash exp(cos(xi))$.
-Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
-Instead of considering these as operations from $\RR\to\RR$, we consider them as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
+Let the term we want to differentiate be $x_1:\RR,\ldots,x_n:\RR\vdash \exp(\cos(x_i))$.
+Following the chain rule, we need the jacobian matrices of $\cos$ at $x_i$ and of $\exp$ at $\cos(x_i)$. 
+Instead of considering these as operations from $\RR\to\RR$, we consider them as functions from the whole context. So $\cos$ and $\exp$ are seen as functions $\RR^n\to\RR$.
 However, by simply doing this we lose compositionality. 
-So $cos$ will also return its context and is now seen as a function $\sem{cos}:\RR^n\to\RR^{n+1}$.
-Similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
-We thus obtain $\sem{exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $J\sem{cos} \in Mat_{n,n+1}$, $J\sem{exp} \in Mat_{n+1,n+2}$ also compose nicely.
+So $\cos$ will also return its context and is now seen as a function $\sem{\cos}:\RR^n\to\RR^{n+1}$.
+Similarly for $\exp$, but also taking the return value of $\sem{\cos}$ as an extra argument, the one it will actually use and not simply return. 
+We thus obtain $\sem{\exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $\J\sem{\cos} \in Mat_{n,n+1}$, $\J\sem{\exp} \in Mat_{n+1,n+2}$ also compose nicely.
 The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
@@
     \begin{tabular}{r c l}
-        $\nabla f$ &=& $J^Tf(0,\ldots,0,1)=J^Tf_1(J^Tf_{2}(\ldots(J^Tf_m(0,\ldots,0,1))\ldots)$
+        $\nabla f$ &=& $\J^Tf(0,\ldots,0,1)= \J^Tf_1(J^Tf_{2}(\ldots(\J^Tf_m(0,\ldots,0,1))\ldots)$
     \end{tabular}
@@
 To actually reverse the order of computation needed for this transpose of jacobians, we use a sort of simple continuation.
-$f_i$ is turned into $\Dsynrevsymbol{f_i}:=<f_i, \lambda Y. Y\circ J^Tf_i>$ where $Y:\RR^{n+i-1}\to \RR^n$. 
-We recover compositionality by noting that $<f_{i+1}(f_i), (\lambda Y. Y\circ J^Tf_{i+1}>)(\lambda Y. Y\circ J^Tf_i)>$ reduces to
-$<f_{i+1}\circ f_i, \lambda Y. Y\circ J^Tf_i \circ J^Tf_{i+1}>$, and thus by induction we can obtain $<f, \lambda Y. YJ^Tf>$.
+$f_i$ is turned into $\Dsynrevsymbol{f_i}:= \langle f_i, \lambda Y. Y\circ \J^Tf_i \rangle$ where $Y:\RR^{n+i-1}\to \RR^n$. 
+We recover compositionality by noting that $\langle f_{i+1}(f_i), (\lambda Y. Y\circ \J^Tf_{i+1}>)(\lambda Y. Y\circ \J^Tf_i)\rangle$ reduces to
+$\langle f_{i+1}\circ f_i, \lambda Y. Y\circ \J^Tf_i \circ \J^Tf_{i+1} \rangle$, and thus by induction we can obtain $\langle f, \lambda Y. Y\J^Tf \rangle$.
 By applying the identity continuation $\RR^n\to\RR^n$ on the second component then the result to $(0,\ldots,0,1)$, 
@@
 This method is quite inefficient still. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
-Second, each $J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.
+Second, each $\J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.
 
@@
 One of the key simple ideas that we used was to transform every operator into a unary one. 
-This essentially trivialises the computation flow to a line. 
+This essentially trivializes the computation flow to a line. 
 Even if the starting program was a straight line program, 
@@
 
-If we look at $J^Tf_{i+1}$, we notice that this function is almost the identity, except at the last row. 
-Even on the last row, if the original term was a unary or binary operator like $cos, exp, +, *$, 
+If we look at $\J^Tf_{i+1}$, we notice that this function is almost the identity, except at the last row. 
+Even on the last row, if the original term was a unary or binary operator like $\cos, \exp, +, *$, 
 the row is zero except for at most two indices (one for unary operators).
-This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,\ldots,y(n+i)).(y1,\ldots,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
+This means we can use a more compact representation $\J^T\sem{op(xi,xj)}:=\lambda (y_1,\ldots,y_{n+i}).(y_1,\ldots,y_{n+i})+[i]\partial_1op(x_i, x_j)+[j]\partial_2op(x_i, x_j)$, 
 where $[k]$ means that the element is added at the $k$-th index of the tuple.
@@
 Because each function is almost the identity, we obtain a lot of substitutions of the form $[x/y]$ where both $x$ and $y$ are variables. 
-This allows us to drastically reduce the size of $J^Tf$. In fact, for simple programs, this is basically enough to get an efficient purely functional reverse derivative transformation.
+This allows us to drastically reduce the size of $\J^Tf$. In fact, for simple programs, this is basically enough to get an efficient purely functional reverse derivative transformation.
 We develop this idea further for a richer language.

### 2021-07-08 rad-icfp-2021/abstract.tex
.

--- 2021-07-03 rad-icfp-2021/abstract.tex
@@
 \begin{abstract}
-    Deep learning is moving towards increasingly sophisticated optimization objectives that employ
-    tensors and operations on tensors.
-    Reverse-mode differentiation is used for optimization, 
-    but it does not preserve purity of the underlying program and makes it notoriously harder to optimize.
-    In a purely functional setting, most array operations can be expressed conveniently using map and fold.
-    We present a differentiable programming language which is the first to deliver on efficient, pure and parallel-friendly reverse-mode differentiation on
-    an expressive array language. 
+    Reverse-mode differentiation is used for optimization, but it does not preserve purity of the underlying program which makes it notoriously harder to optimize.
+    We present a reverse-mode differentiation on a purely functional language with array operations.  
+    It is the first one to deliver a provably efficient, purely functional and denotationnally correct reverse-mode differentiation.
     We show that our transformation is semantically correct and verifies the cheap gradient principle. 
-    Our transformation is presented as a compilation scheme through a novel intermediate representation that we call `unary form`.
+    Inspired by PROPs and compilation to categories, we introduce a novel intermediate representation that we call `unary form'.
+    Our reverse-mode transformation is factored as a compilation scheme through this intermediate representation.
     We obtain provably efficient gradients by performing general partial evaluation optimizations after our reverse-mode transformation, as opposed to manually derived ones.

### 2021-07-08 rad-icfp-2021/intro.tex
some fixes

@@
 % An efficient purely functional reverse-mode transformation would be more easily explainable, optimizable, and reusable.
-
-% TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
-% all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g.
-in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress.
-% For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.
 
@@
     \end{center}   
-This is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.
+This is very close to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.
 This term can be further optimized via constant propagation and algebraic simplifications to give
@@
 %     \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
-%     \item several additions to the source language (\S\ref{sec:generalization}) TODO:should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.
+%     \item several additions to the source language (\S\ref{sec:generalization})
 % \end{itemize}

### 2021-07-08 rad-icfp-2021/complexity.tex
some fixes
gaining some space

@@
 
+\begin{center}
 \begin{tabular}{r c l}
@@
     $\NestA$(!x!) &=& 0 \\
+\end{tabular}
+    \begin{tabular}{r c l}
+    $\NestA$($\pi_i$(!e!)) &=& $\NestA$(!e!)\\
+    $\NestA$(!op1 e!) &=& $\NestA$(!e!) \\
+\end{tabular}
+\end{center}
+
+\begin{center}
+\begin{tabular}{r c l}
     $\NestA$(!let x:A = e$_1$ in e$_2$:B!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)  \\ 
     $\NestA$(!< e$_1$, e$_2$ >:AxB!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!) \\ 
-    $\NestA$($\pi_i$(!e!)) &=& $\NestA$(!e!)\\
     $\NestA$(!e$_1$ op2 e$_2$!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)\\
-    $\NestA$(!op1 e!) &=& $\NestA$(!e!) \\
     $\NestA$(!map2 (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
@@
 \end{tabular}
+\end{center}
 
@@
     Denote by $G$ the term $\grad_\Gamma$!e! after the partial evaluation step from Section~\ref{sub:Partial evaluation and optimization}.
-    Then !$\cost$(G) $\leq$ $4*2^{p}*\cost$(e)!.
+    Then !$\cost$(G) $\leq$ $4*3^{p}*\cost$(e)!.
 \end{theorem}

### 2021-07-08 rad-icfp-2021/correctness.tex
gaining some space

@@
     $\seml \Array{\reals}{n}\semr$ & $\defeq$ & $\prod_{1\leq i \leq n} \RR$ \\
-    !$\seml$op1$\semr$! & $\defeq$ & op1: $\reals\to\reals$ 
     \end{tabular}
     \begin{tabular}{r c l} 
+        !$\seml$op1$\semr$! & $\defeq$ & op1: $\reals\to\reals$ \\
         !$\seml$op2$\semr$! & $\defeq$ & op2: $\reals\times \reals\to\reals$ \\
         !$\seml$c$\semr$! & $\defeq$ & $c\in\RR$ \\
-        !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>;swap;$(\seml$ e$_1\semr)^{\times n}$! \\
-        !$\seml$reduce (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$! & $\defeq$ <> &
     \end{tabular}
+
+    \begin{tabular}{r c l} 
+    !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>;swap;$(\seml$ e$_1\semr)^{\times n}$!
+    \end{tabular}
+
 where $\Delta_n$ is the $n$-copy map !<id,$\ldots$,id>!.
@@
    $\seml$!pair!$_{T;A\times B} \semr$ &=& $pair_{T;A\times B}$ \\
-   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T1}\concatcomp u_{T2}\concatcomp id_{T3}$\\
+   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T1}\concatcomp u_{T2}\concatcomp id_{T3}$
+    \end{tabular}
+   \begin{tabular}{l c l}
    $\seml$!e1!$\comp$!e2!$\semr$  &=& $\seml$!e1!$\semr \comp \seml$!e2!$\semr$ \\
@@
    $\seml$!map2!$_{T;x,y.e}\semr$  &=& $id_\Gamma\concatcomp map2_{x,y.e}$ \\
-   $\seml$!reduce!$_{T;x,y.e;e}\semr$ &=& $id_\Gamma\concatcomp reduce_{x,y.e1;e2}$ \\
+   $\seml$!reduce!$_{T;x,y.e;e}\semr$ &=& $id_\Gamma\concatcomp reduce_{x,y.e1;e2}$
 \end{tabular}
@@
     $\seml$!J!$^T$!var!$_{T;i} \semr$ &=& TODO \\
+    $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& $id_T\concatcomp map2_{T;x,y.e}$ \\
+    $\seml$!e$_1$!$\icomp$!e$_2$!$\semr$ &=& $\Lambda$($id_\Gamma\concatcomp\seml$!e2!$\semr$; $\Lambda^{-1}$($\seml$!e1!$\semr$)) \\
+    $\seml$!J!$^T$!reduce!$_{T;x,y.e_1;e_2}\semr$  &=& $id_T\concatcomp reduce_{T;x,y.e_1;e_2}$ \\
+    $\seml$!J!$^T$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T_1}\concatcomp\widehat{0}_{T_2}\concatcomp id_{T_3}$ 
+    \end{tabular}
+    \begin{tabular}{l c l}
     $\seml$!J!$^T$!op!$_{T;n} \semr$ &=& $id_T\concatcomp J^Top_n$ \\
+    $\seml$!<e$_1$, e$_2$>!$\semr$  &=& !<!$\seml$!e1!$\semr$, $\seml$!e2!$\semr$!>! \\
     $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& $id_T\concatcomp$!<$\pi_1,\pi_2$>! \\
-    $\seml$!J!$^T$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T_1}\concatcomp\widehat{0}_{T_2}\concatcomp id_{T_3}$ \\
-    $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& $id_T\concatcomp map2_{T;x,y.e}$ \\
-    $\seml$!J!$^T$!reduce!$_{T;x,y.e_1;e_2}\semr$  &=& $id_T\concatcomp reduce_{T;x,y.e_1;e_2}$ \\
-    $\seml$!<e$_1$, e$_2$>!$\semr$  &=& !<!$\seml$!e1!$\semr$, $\seml$!e2!$\semr$!>! \\
-    $\seml$!e$_1$!$\icomp$!e$_2$!$\semr$ &=& $\Lambda$($id_\Gamma\concatcomp\seml$!e2!$\semr$; $\Lambda^{-1}$($\seml$!e1!$\semr$)) \\
  \end{tabular}

### 2021-07-08 rad-icfp-2021/appendix1.tex
finishing some examples
writing section 5.4
some fixes
putting some sections in appendix

@@
 
+ \label{sub:Reverse derivative of array operations}
+
 We now prove that the reverse mode transformation is correct on array operations. 
-The proofs consist in showing that unrolling the computation before differentiation 
-and then differentiating gives the same as differentiation and then unrolling.
+% The proofs consist in showing that unrolling the computation before differentiation 
+% and then differentiating gives the same as differentiation and then unrolling.
 
@@
 
-\begin{proof}
-    TODO
-\end{proof}
+% \begin{proof}
+% Let's call !t! the term !map2 (x,.y) A B!. 
+% By chosing !Z! to a hot vector at i, call it !Z$_i$!, we're back to showing the result for the term $\Gamma \vdash$!e[a$_i$/x,b$_i$/y]: $\reals$!.
+% In $\directD{\rho}{\Gamma}{Y}$(!t!)(!Z$_i$!), the term !G*(reduce + 0 Z)! reduces to !G!. 
+% This means 
+% \end{proof}
 
@@
 
+\subsection{Adding more array operators}
+\label{sub:lift_more_arr_op}
+
+There are two main differences with fold left !foldl! compared to !reduce!. 
+First, in !foldl (x,y.e) v A! the starting accumulation element !v! is not a unit for !(x,y.e)!,
+so it will have non-trivial derivatives in general and we need to account for that.
+Second, we will need a more general !scanl! which allows !(x,y.e)! as a function argument. 
+In other words, we need the general scan left computing the intermediate values of a fold left.
+This should be a different primitive but we will still call it !scanl! in this section.
+
+Finally, the former point implies we need a few more array manipulations.
+These can be elegantly dealt with by changing the semantics of !scanl! and !scanr!. 
+Let's assume that they now return a pair of an array of size $n$ 
+of the intermediate computations and the final result.
+The reverse derivative of !foldl! is then as shown in Figure~\ref{fig:foldl_map_trans}.
+In this figure, we have shown the translation of the !map! operator as well, which is very similar to !map2!.
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{|r c l|}
+\hline
+    $\directD{\rho}{\Gamma}{Y}$(!foldl (x,y.e$_1$) e$_2$ e$_3$!) &=&
+            !let v,Y$_1$ = $\directD{\rho}{\Gamma}{Y}$(e$_2$) in! \\
+            && !let A,Y$_2$ = $\directD{\rho}{\Gamma,v}{Y_1}$(e$_3$) in! \\
+            && !let A$_0$,r$_1$ = (scanl (x,y.e$_1$) v A) in! \\
+            && !let A$_1$ = map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x,b/y]) A$_0$ A! \\
+            && !let A$_2$ = map2 (a,b.(!$\grad_{\{y\}}$!)e$_1$[a/x,b/y]) A$_0$ A! \\
+            && !let r$_2$, A$_3$ = scanr * 1 A$_1$! \\
+            && !<r$_1$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
+            && !let y$_1$,B = r$_2$*z ,map2 (x,y. x*y*z) A$_2$ A$_3$ in! \\
+            && !Y$_2$(x$_1$,$\ldots$,x$_n$,y$_1$,B)>! \\
+    $\directD{\rho}{\Gamma}{Y}$(!map (x.e$_{1}$) e$_{2}$!) &=&  
+            !let A,Y$_{1}$ = $\directD{\rho}{\Gamma}{Y}$(e$_{2}$) in! \\
+            && !let G=$\grad_{\Gamma}$ e$_{1}$ in!\\
+            && !<map (x.e$_{1}$) A, fun (x$_{1}$,$\ldots$,x$_n$,Z) -> !\\
+            && !Y$_{1}$( (x$_{1}$,$\ldots$,x$_n$)$\widehat{+}$(G*(reduce + 0 Z)),!\\
+            && \quad\quad !map2 (a,b.($\grad_{\{x\}}$e$_{1}$)[a/x]*b) A Z )>!\\\hline
+\end{tabular}
+\end{center}
+\caption{The reverse-mode AD transformation of \codekw{foldl} and \codekw{map} operators.}
+\label{fig:foldl_map_trans}
+\end{figure}
+
+\subsection{Adding non-smooth scalar operators} % (fold)
+\label{sub:lift_non_smooth}
+
+We assumed the unary and binary operators were denoted by smooth functions $\RR^n\to\RR$. 
+There is no additional difficulty in considering operators which are partial functions 
+like division or operators which are not smooth at a point like square root.
+
+These functions are then given intentional derivatives which provide valid derivatives 
+on the domain of definition and differentiability of the operator. 
+These functions are well known to be the bete noire of AD \cite{griewank2008evaluating} 
+and we do not provide novel solutions to these.  
+Several recent work have shown how to give semantics to such operators in the context of AD \cite{vakar2020denotational,mazza2021automatic,sherman2021,lee2020correctness}.
+
 \subsection{gradient from the introduction}
@@
 Finding these constant arrays is key to a lot of optimizations that leverage the ring algebraic structure of the reals to arrays.
-
-TODO: add ZerosLike ONesLike notation officially somewhere.
 

### 2021-07-08 rad-icfp-2021/generalization.tex
putting some sections in appendix

@@
 \label{fig:lift_reduce}
-\end{figure}
-
-\subsection{More array operators}
-\label{sub:lift_more_arr_op}
-
-There are two main differences with fold left !foldl! compared to !reduce!. 
-First, in !foldl (x,y.e) v A! the starting accumulation element !v! is not a unit for !(x,y.e)!,
-so it will have non-trivial derivatives in general and we need to account for that.
-Second, we will need a more general !scanl! which allows !(x,y.e)! as a function argument. 
-In other words, we need the general scan left computing the intermediate values of a fold left.
-This should be a different primitive but we will still call it !scanl! in this section.
-
-Finally, the former point implies we need a few more array manipulations.
-These can be elegantly dealt with by changing the semantics of !scanl! and !scanr!. 
-Let's assume that they now return a pair of an array of size $n$ 
-of the intermediate computations and the final result.
-The reverse derivative of !foldl! is then as shown in Figure~\ref{fig:foldl_map_trans}.
-In this figure, we have shown the translation of the !map! operator as well, which is very similar to !map2!.
-
-\begin{figure}
-\begin{center}
-\begin{tabular}{|r c l|}
-\hline
-    $\directD{\rho}{\Gamma}{Y}$(!foldl (x,y.e$_1$) e$_2$ e$_3$!) &=&
-            !let v,Y$_1$ = $\directD{\rho}{\Gamma}{Y}$(e$_2$) in! \\
-            && !let A,Y$_2$ = $\directD{\rho}{\Gamma,v}{Y_1}$(e$_3$) in! \\
-            && !let A$_0$,r$_1$ = (scanl (x,y.e$_1$) v A) in! \\
-            && !let A$_1$ = map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x,b/y]) A$_0$ A! \\
-            && !let A$_2$ = map2 (a,b.(!$\grad_{\{y\}}$!)e$_1$[a/x,b/y]) A$_0$ A! \\
-            && !let r$_2$, A$_3$ = scanr * 1 A$_1$! \\
-            && !<r$_1$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
-            && !let y$_1$,B = r$_2$*z ,map2 (x,y. x*y*z) A$_2$ A$_3$ in! \\
-            && !Y$_2$(x$_1$,$\ldots$,x$_n$,y$_1$,B)>! \\
-    $\directD{\rho}{\Gamma}{Y}$(!map (x.e$_{1}$) e$_{2}$!) &=&  
-            !let A,Y$_{1}$ = $\directD{\rho}{\Gamma}{Y}$(e$_{2}$) in! \\
-            && !let G=$\grad_{\Gamma}$ e$_{1}$ in!\\
-            && !<map (x.e$_{1}$) A, fun (x$_{1}$,$\ldots$,x$_n$,Z) -> !\\
-            && !Y$_{1}$( (x$_{1}$,$\ldots$,x$_n$)$\widehat{+}$(G*(reduce + 0 Z)),!\\
-            && \quad\quad !map2 (a,b.($\grad_{\{x\}}$e$_{1}$)[a/x]*b) A Z )>!\\\hline
-\end{tabular}
-\end{center}
-\caption{The reverse-mode AD transformation of \codekw{foldl} and \codekw{map} operators.}
-\label{fig:foldl_map_trans}
 \end{figure}
@@
 the condition is evaluated before differentiation is applied, and so conditionals are never directly differentiated.
-
-\subsection{Non-smooth scalar operators} % (fold)
-\label{sub:lift_non_smooth}
-
-We assumed the unary and binary operators were denoted by smooth functions $\RR^n\to\RR$. 
-There is no additional difficulty in considering operators which are partial functions 
-like division or operators which are not smooth at a point like square root.
-
-These functions are then given intentional derivatives which provide valid derivatives 
-on the domain of definition and differentiability of the operator. 
-These functions are well known to be the bete noire of AD \cite{griewank2008evaluating} 
-and we do not provide novel solutions to these.  
-Several recent work have shown how to give semantics to such operators in the context of AD \cite{vakar2020denotational,mazza2021automatic,sherman2021,lee2020correctness}.
 

### 2021-07-08 rad-icfp-2021/relwork.tex
rewrote conclusion a bit

@@
 \noindent \textbf{Higher-order functions}
+TODO
 Matthijs'work, use of defunctionalization, question efficiency and usefulness, 
@@
 
-We have introduced a transformation on programs to compute provably efficient (\ref{sec:complexity}) 
-gradients via reverse derivatives in a purely functional way (\ref{sub:Macro for pure reverse mode transformation})
-on a simple yet expressive language with functions on arrays (\ref{sub:sourcelang}), 
-combined with standard functional optimizations (\ref{fig:optim}).  
-We introduced a novel intermediate representation, unary form (\ref{sec:unf}) 
+We have introduced a transformation on programs to compute provably efficient (\S\ref{sec:complexity}) 
+gradients via reverse derivatives in a purely functional way (\S\ref{sub:Macro for pure reverse mode transformation})
+on a simple yet expressive language with functions on arrays (\S\ref{sub:sourcelang}), 
+combined with standard functional optimizations (\S\ref{fig:optim}).  
+We introduced a novel intermediate representation, Unary Normal Form (\S\ref{sec:unf}) 
 to decompose our transformation into simpler transformations.
-We gave denotational semantics to our languages and our transformations (\ref{sec:correctness}) 
-and showed correctness of the reverse-mode transformation (\ref{sub:Correctness theorem}).
-We showed (\ref{sec:generalization}) how to lift some of the restrictions that
-we introduced on arrays and how to extend our approach to a richer language with more primitive operations and conditionals.
+We gave denotational semantics to our languages and our transformations (\S\ref{sec:correctness}), 
+and showed correctness of the reverse-mode translation (\S\ref{sub:Correctness theorem}).
+We showed (\S\ref{sec:generalization}) how to lift the restrictions that
+we introduced on arrays and how to extend our approach with other constructs such as conditionals.
 \clearpage

### 2021-07-08 rad-icfp-2021/appendix1.tex
.

@@
 
-\subsection{gradient from the introduction}
+\subsection{General arrays} % (fold)
+\label{sub:lift_gen_arr}
+
+We now show how to generalize our reverse-mode transformation to be defined on arrays over any ground type !G!.
+That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.
+
+A ground type !G! is interpreted as an Euclidean space $A$. 
+It is in particular a real vector space.
+Similarly, a ground context !$\Gamma$=x$_1$:G$_1$,$\ldots$,x$_n$:G$_n$! is interpreted as $\bigoplus_{1\leq i\leq n}A_i$, where $A_i\defeq\seml$(!G$_i$!)$\semr$.
+The denotation of the gradient of a term !$\Gamma \vdash$ e: G! at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}A_i)\otimes A$
+where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}A_i\otimes A$.
+
+We can define $\otimes$ on the types of our language inductively by
+
+\begin{tabular}{r c l}
+    $\reals \otimes A$ & $\defeq$ & $A$ \\
+    $(A_1 \times \ldots \times A_n)\otimes A$ & $\defeq$ & $(A_1\otimes A) \times \ldots \times (A_n \otimes A)$ \\
+    $\Array{A_1}{n} \otimes A$ & $\defeq$ & $\Array{A_1\otimes A}{n}$
+\end{tabular}
+
+With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A_1\times\ldots\times A_n$ as expected.
+For !map2!, we need a generalization of $*:\reals\to\reals$. 
+If !e$_1$: A!, then $\grad_{\{x\}}$!e$_1$: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
+If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.
+Similarly, for !A$_{3}$! in !reduce! we need a new primitive $\widetilde{*}:(A\otimes A)\times(A\otimes A) \to (A\otimes A)$.
+$\widetilde{*}$ corresponds to matrix-matrix multiplication.
+Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!, as can be seen in Figure~\ref{fig:lift_gen_arr}.
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{|r c l|}
+\hline
+    $\directD{\rho}{\Gamma}{Y}$(!map2 (x,y.e$_1$: G) e$_2$ e$_3$!) &=&  
+    !let A,Y$_1$ = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
+    && !let B,Y$_2$ = !$\directD{\rho}{\Gamma,A}{Y1}$!(e$_3$) in! \\
+    && !let C=!$\grad_{\Gamma}$!e$_1$ in!\\
+    && !<map2 (x,y.e$_1$) A B, fun (x$_1$,$\ldots$,x$_n$,Z) -> !\\
+    && !Y$_2$( (x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
+    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x]!$\widehat{*}$!b) A Z,!\\
+    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e$_1$)[a/x]!$\widehat{*}$!b) B Z )>!\\
+    $\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e$_{1}$) e$_{2}$ e$_{3}$!) &=&
+    !let y$_{1}$,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
+    && !let A,Y$_{2}$ = !$\directD{\rho}{\Gamma,y_1}{Y_1}$!(e$_{3}$) in! \\
+    && !let A$_{0}$ = shift1R (scanl (x,y.e$_{1}$) y$_{1}$ A) in! \\
+    && !let A$_{1}$ = shift1L (map2! \\ 
+    && !      (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A$_{0}$ A) in! \\
+    && !let A$_{2}$ = map2! \\
+    && !      (a,b.(!$\grad_{\{y\}}$!)e$_{1}$[a/x,b/y]) A$_{0}$ A in! \\
+    && !let A$_{3}$ = scanr $\widetilde{*}$ 1 A$_{1}$ in! \\
+    && !<reduce (x,y.e$_{1}$) y$_{1}$ A, fun (x$_{1}$,$\ldots$,x$_n$,z) ->! \\
+    && !Y$_{2}$(x$_{1}$,$\ldots$,x$_n$, map2 (x,y. x$\widehat{*}$(y$\widehat{*}$z)) A$_{2}$ A$_{3}$>! \\ \hline
+\end{tabular}
+\end{center}
+\caption{The reverse-mode AD transformation of \codekw{map2} and \codekw{reduce} for general arrays.}
+\label{fig:lift_gen_arr}
+\end{figure}
+
+Evidently, one can combine all the generalizations from the previous subsections.
+Even though this transformation has the correct complexity, it is open for future research to find
+even better representations to allow for more optimizations. 
+In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
+and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.
+More generally, there is growing interest in tensor calculus \cite{liao2019differentiable,bernstein2020differentiating}.
+
+\subsection{Gradient from the introduction}
 \label{sub:gradintro}

### 2021-07-08 rad-icfp-2021/generalization.tex
.

@@
 First, we allow free variables for the function argument of !reduce! in Section~\ref{sub:lift_reduce}. 
-Then we show how to support more array operations (Section~\ref{sub:lift_more_arr_op}), conditionals (Section~\ref{sub:lift_conditional}), and non-smooth scalar operators (Section~\ref{sub:lift_non_smooth}).
-Afterwards, we extend the language to support general arrays instead of supporting only $\Array{\reals}{n}$ in Section~\ref{sub:lift_gen_arr}.
+Then we show how to support conditionals in Section~\ref{sub:lift_conditional}.
+% Then we show how to support more array operations (Section~\ref{sub:lift_more_arr_op}), conditionals (Section~\ref{sub:lift_conditional}), and non-smooth scalar operators (Section~\ref{sub:lift_non_smooth}).
+% Afterwards, we extend the language to support general arrays instead of supporting only $\Array{\reals}{n}$ in Section~\ref{sub:lift_gen_arr}. 
 Finally, we provide a recipe for adding more constructs to the language in Section~\ref{sub:lift_recipe}.
+Further generalizations for more array operations, non-smooth scalar operators, and general array support can be found in the supplementary materials.
 
@@
 
-\subsection{General arrays} % (fold)
-\label{sub:lift_gen_arr}
-
-We now show how to generalize our reverse-mode transformation to be defined on arrays over any ground type !G!.
-That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.
-
-A ground type !G! is interpreted as an Euclidean space $A$. 
-It is in particular a real vector space.
-Similarly, a ground context !$\Gamma$=x$_1$:G$_1$,$\ldots$,x$_n$:G$_n$! is interpreted as $\bigoplus_{1\leq i\leq n}A_i$, where $A_i\defeq\seml$(!G$_i$!)$\semr$.
-The denotation of the gradient of a term !$\Gamma \vdash$ e: G! at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}A_i)\otimes A$
-where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}A_i\otimes A$.
-
-We can define $\otimes$ on the types of our language inductively by
-
-\begin{tabular}{r c l}
-    $\reals \otimes A$ & $\defeq$ & $A$ \\
-    $(A_1 \times \ldots \times A_n)\otimes A$ & $\defeq$ & $(A_1\otimes A) \times \ldots \times (A_n \otimes A)$ \\
-    $\Array{A_1}{n} \otimes A$ & $\defeq$ & $\Array{A_1\otimes A}{n}$
-\end{tabular}
-
-With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A_1\times\ldots\times A_n$ as expected.
-For !map2!, we need a generalization of $*:\reals\to\reals$. 
-If !e$_1$: A!, then $\grad_{\{x\}}$!e$_1$: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
-If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.
-Similarly, for !A$_{3}$! in !reduce! we need a new primitive $\widetilde{*}:(A\otimes A)\times(A\otimes A) \to (A\otimes A)$.
-$\widetilde{*}$ corresponds to matrix-matrix multiplication.
-Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!, as can be seen in Figure~\ref{fig:lift_gen_arr}.
-
-\begin{figure}
-\begin{center}
-\begin{tabular}{|r c l|}
-\hline
-    $\directD{\rho}{\Gamma}{Y}$(!map2 (x,y.e$_1$: G) e$_2$ e$_3$!) &=&  
-    !let A,Y$_1$ = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
-    && !let B,Y$_2$ = !$\directD{\rho}{\Gamma,A}{Y1}$!(e$_3$) in! \\
-    && !let C=!$\grad_{\Gamma}$!e$_1$ in!\\
-    && !<map2 (x,y.e$_1$) A B, fun (x$_1$,$\ldots$,x$_n$,Z) -> !\\
-    && !Y$_2$( (x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
-    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x]!$\widehat{*}$!b) A Z,!\\
-    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e$_1$)[a/x]!$\widehat{*}$!b) B Z )>!\\
-    $\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e$_{1}$) e$_{2}$ e$_{3}$!) &=&
-    !let y$_{1}$,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
-    && !let A,Y$_{2}$ = !$\directD{\rho}{\Gamma,y_1}{Y_1}$!(e$_{3}$) in! \\
-    && !let A$_{0}$ = shift1R (scanl (x,y.e$_{1}$) y$_{1}$ A) in! \\
-    && !let A$_{1}$ = shift1L (map2! \\ 
-    && !      (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A$_{0}$ A) in! \\
-    && !let A$_{2}$ = map2! \\
-    && !      (a,b.(!$\grad_{\{y\}}$!)e$_{1}$[a/x,b/y]) A$_{0}$ A in! \\
-    && !let A$_{3}$ = scanr $\widetilde{*}$ 1 A$_{1}$ in! \\
-    && !<reduce (x,y.e$_{1}$) y$_{1}$ A, fun (x$_{1}$,$\ldots$,x$_n$,z) ->! \\
-    && !Y$_{2}$(x$_{1}$,$\ldots$,x$_n$, map2 (x,y. x$\widehat{*}$(y$\widehat{*}$z)) A$_{2}$ A$_{3}$>! \\ \hline
-\end{tabular}
-\end{center}
-\caption{The reverse-mode AD transformation of \codekw{map2} and \codekw{reduce} for general arrays.}
-\label{fig:lift_gen_arr}
-\end{figure}
-
-Evidently, one can combine all the generalizations from the previous subsections.
-Even though this transformation has the correct complexity, it is open for future research to find
-even better representations to allow for more optimizations. 
-In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
-and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.
-More generally, there is growing interest in tensor calculus \cite{liao2019differentiable,bernstein2020differentiating}.
-
 \subsection{Recipe for Adding More Constructs}

### 2021-07-08 rad-icfp-2021/relwork.tex
minor

@@
 \noindent \textbf{Higher-order functions}
-TODO
-Matthijs'work, use of defunctionalization, question efficiency and usefulness, 
-interested in functions like root of function or minimal of a function. Implicit function theorem
 Some recent work \cite{vakar2020reverse,sherman2021} present reverse-mode in a higher-order language.
@@
 and \cite{sherman2021} uses sophisticated higher-order primitives such as root finding, max, argmax or integral. 
-Their work focuses on computable reals which is hard to compare in terms of efficiency 
-with our more standard approach of AD.
+Their work focuses on computable reals which is hard to compare in terms of efficiency with our more standard approach of AD.
+It is however quite difficult to prove a complexity result in the higher-order setting. 
+In addition, standard techniques of defunctionalization struggle when higher-order is combined with recursion.
 
-\noindent \textbf{More array primitives}
+\noindent \textbf{Array primitives}
 We focused on giving reverse derivatives for a small set of array primitives. 
@@
 and these can already be added in added in our language. 
-One only needs to make sure that the provided transpose Jacobian is correct.
+As shown in \ref{sub:lift_recipe}, one mostly only needs to make sure that the provided transpose Jacobian is correct.
 
@@
 % 	https://github.com/apple/swift/blob/master/docs/DifferentiableProgramming.md#approaches-to-automatic-differentiation
-
-TODO: understand related work sentence from Diff. Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
-    For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
-    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.
-
-TODO: should talk a bit about the recent paper: reverse AD in Dex \cite{paszke2021getting}
 
@@
 \cite{shaikhha2019efficient}: similar approach but here more complicated for reverse mode, would be interesting to investigate more loop fusions here.
+\cite{paszke2021getting}
+Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
+    For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
+    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.
 

### 2021-07-09 rad-icfp-2021/correctness.tex
fixing some todos

@@
 \begin{tabular}{l c l}
-    $\seml$!J!$^T$!var!$_{T;i} \semr$ &=& TODO \\
+    $\seml$!J!$^T$!var!$_{T;i} \semr$ &=& $\pi_T \widehat{+}(0,\pi_A,0)$ \\
     $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& $id_T\concatcomp map2_{T;x,y.e}$ \\
@@
  \end{tabular}
+ where !T,A! $\vdash$ (0,$\pi_A$,0): !T! and the projection $\pi_A$ lands in the $i$-th element of the list !T!.
 
@@
 \begin{definition}[$\Dsynrevsymbol_{\rho}$ as a lax functor]
-    We define $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$ as follows.
-    $\Dsynrevsymbol_{\rho}$(![A$_1$,$\ldots$,A$_n$])=[A$_1$,$\ldots$,A$_n$,A$_1\times\ldots\times$A$_n\to \rho$]!.
-    TODO
+    We define $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$ as follows.\\
+    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_n$])=[A$_1$,$\ldots$,A$_n$,A$_1\times\ldots\times$A$_n\to \rho$]!.
+    !$\Dsynrevsymbol_{\rho}$(op1)=<op1,$\pi_{last}\circ$ J$^T$op1>!,\\ 
+    !$\Dsynrevsymbol_{\rho}$(op2)=<op2,$\pi_{last}\circ$ J$^T$op2>!,\\
+    !$\Dsynrevsymbol_{\rho}$(map2 (x,y.e))=<map2 (x,y.e),$\pi_{last}\circ$ J$^T$map2 (x,y.e)>!,\\
+    !$\Dsynrevsymbol_{\rho}$(reduce$_{x,y.e_1,e_2}$)=<reduce$_{x,y.e_1,e_2}$,$\pi_{last}\circ$ J$^T$reduce$_{x,y.e_1,e_2}$>!, 
+    where $\pi_{last}$ is the projection to the last element (the continuation variable).
+    It naturally extends on sequential composition. It does not automatically extend to a multi-functor between concategories because \\
+    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$] $\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$]$\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.
+    However there is a map !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])$\to \Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$])!.
+    Internally as a lambda term, it is given by \\
+    $(x_1,\ldots,x_k,Y_1,(x_{k+1},\ldots,x_n,Y_s)\mapsto (x_1,\ldots,x_n,\lambda(y_1,\ldots,y_n)\to Y_1(y_1,\ldots,y_k)\widehat{+}Y_2(y_{k+1},\ldots,y_n)$.
+    Here, $\widehat{+}$ is the reverse-derivative of the copy map which is known to be fanout. 
+    It is not surprising to see it appear as $\Dsynrevsymbol_{\rho}$ is a semantic functor, and it does not need to be efficient. 
 \end{definition}
 
-The design of $\Dsynrevsymbol_{\rho}$ on UNF is inspired by the semantic one between concategories, 
+The design of the syntactic $\Dsynrevsymbol_{\rho}$ on the UNF language is inspired by the semantic one between concategories, 
 and it's routine to check that they match.

### 2021-07-09 rad-icfp-2021/appendix1.tex
some proofs in appendix

@@
 
-% \begin{proof}
-% Let's call !t! the term !map2 (x,.y) A B!. 
-% By chosing !Z! to a hot vector at i, call it !Z$_i$!, we're back to showing the result for the term $\Gamma \vdash$!e[a$_i$/x,b$_i$/y]: $\reals$!.
-% In $\directD{\rho}{\Gamma}{Y}$(!t!)(!Z$_i$!), the term !G*(reduce + 0 Z)! reduces to !G!. 
-% This means 
-% \end{proof}
+\begin{proof}
+Let us denote by !t! the term !map2 (x,y.e) A B!. Without loss of generality, assume !A!=$[a_1,\ldots,a_n]$ and !B!=$[b_1,\ldots,b_n]$.
+By chosing !Z! to a hot vector at i, call it !Z$_i$!, we're back to showing the result for the term $\Gamma \vdash$!e[a$_i$/x,b$_i$/y]: $\reals$!.
+In !$\directD{\rho}{\Gamma}{Y}$(t)(Z$_i$)!, the term !G = (map2 * Z (map2 (a,b.($\grad_\Gamma$e$_{1}$)[a/x.b/y]) A B))! 
+reduces to $\frac{\partial e}{\partial x_i}$[a$_i$/x,b$_i$/y]. As !a$_i$,b$_i$! are independant of $x_i$, 
+this term is equal to $\frac{\partial e[a_i/x,b_i/y]}{\partial x_i}$, as expected.
+Similarly, the term !map2 * (map2 (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A B) Z! reduces to $\frac{\partial e}{\partial x}$[a$_i$/x,b$_i$/y].
+As !a$_i$! is a variable, and independant of !b$_i$!, this is equal to $\frac{\partial e[a_i/x,b_i/y]}{\partial a_i}$.
+\end{proof}
 
@@
 \begin{proof}
-    TODO
+    The notation for the general case are cumbersome and non-insightful. 
+    We will exemplify the proof on the case of an array of size 3. We use infix notation for the binary operation $e$.
+    The output term unrolls to $v_3\defeq ((v ~e ~a_1)~ e~ a_2) ~e~ a_3$, where $v$ is the unit of $e$.
+    By the lemma above we know that the derivative w.r.t $v$ is 0, and focus on the partial derivatives w.r.t $a_i$.
+    Write $v_1\defeq(v~ e~ a_1)$, $v_2 \defeq v_1 ~e ~a_2$.
+    By inspection we have
+    \begin{itemize}
+        \item $\frac{\partial v_3}{\partial a_3}=\frac{\partial e}{\partial x_2}(v_2,a_3)$
+        \item $\frac{\partial v_3}{\partial a_2}=\frac{\partial v_3}{\partial x_1}\times\frac{\partial v_2}{\partial x_2}=\frac{\partial e}{\partial x_1}(v_2,a_3)\times \frac{\partial e}{\partial x_2}(v_1,a_2)$
+        \item $\frac{\partial v_3}{\partial a_1}=\frac{\partial v_3}{\partial x_1}\times\frac{\partial v_2}{\partial x_2}\times\frac{\partial v_1}{\partial x_2}=\frac{\partial e}{\partial x_1}(v_2,a_3)\times \frac{\partial e}{\partial x_1}(v_1,a_2)\times\frac{\partial e}{\partial x_2}(v,a_1)$
+    \end{itemize}
+    We thus need the following intermediate results
+    \begin{itemize}
+        \item $A_0=[v,v_1,v_2]$=!shift1R(scanl v e A)!
+        \item $A_1 =[\grad_{\{x_1\}}e(v_1,a_2),\grad_{\{x_1\}}e(v_2,a_3)]=$ !shift1L (map2 (a,b.$\grad_{\{x_1\}}$e(a/x,b/y)) $A_0$ A)!
+        \item $A_2 =[\grad_{\{x_2\}}e(v,a_1),\grad_{\{x_2\}}e(v_1,a_2),\grad_{\{x_2\}}e(v_2,a_3)]=$ !map2 (a,b.$\grad_{\{x_2\}}$e(a/x,b/y)) $A_0$ A!
+        \item $A_3 =[\frac{\partial e}{\partial x_1}(v_2,a_3)\times \frac{\partial e}{\partial x_1}(v_1,a_2),\frac{\partial e}{\partial x_1}(v_2,a_3),1]=$ !scanr 1 * $A_1$!
+    \end{itemize}
+    And we return $A_4=$ !map2 $A_2$ $A_3$!.
 \end{proof}

### 2021-07-09 rad-icfp-2021/appendix1.tex
fixing proof appendix

@@
 \end{tabular}
+
  \end{proof}
@@
     & !let A$_0$ = shift1R (scanl + 0 A) in! & \\
-    & !let A$_1$ = shift1L (map2 (x,y -> 1) A$_0$ A) in! & \\ 
+    & !let A$_1$ = shift1L! & \\
+    & \quad\quad !(map2 (x,y -> 1) A$_0$ A) in! & \\ 
     & !let A$_2$ = map2 (x,y -> 1) A$_0$ A in! & \\
@@
     $\stackrel{\beta-reduction}{=}$  & !let A$_0$ = shift1R (scanl + 0 A) in! & \\
-    & !let A$_1$ = shift1L (map2 (x,y -> 1) A$_0$ A) in! & !A$_1$=OnesLike(shift1L(A))! \\
+    & !let A$_1$ = shift1L! & !A$_1$=OnesLike(shift1L(A))! \\
+    & \quad\quad !(map2 (x,y -> 1) A$_0$ A) in! & \\
     & !let A$_2$ = map2 (x,y -> 1) A$_0$ A in! & !A$_2$=OnesLike(A)!\\
@@
     & !map2 * A$_3$ A$_2$! & \\
-    = & !let A$_3$ = scanr * 1 OnesLike(shift1L(A)) in! & !A$_3$=OnesLike(A)!\\
+    = & !let A$_3$ = scanr * 1! & !A$_3$=OnesLike(A)!\\
+    &  \quad\quad!OnesLike(shift1L(A)) in! & \\
     & !map2 * A$_3$ OnesLike(A)! & forward substitution \\
@@
     \end{tabular}
+
 \end{proof}
@@
 Let's call this term !Y!.
-For convenience, let us also rewrite !map2 * B (map (x -> 1) A)! 
+For convenience, let us also rewrite \\
+!map2 * B (map (x -> 1) A)! 
 as !let C = map (x -> 1) A in map2 * B C!.
@@
     $\defeq$ & \Big(!fun z ->! & \\ 
-    & !let A$_0$ = shift1R (scanl + 0 A) in!  \hspace{4cm}\rdelim\}{5}{3mm}[\parbox{40mm}{same reduction\\as previously}]\\
+    & !let A$_0$ = shift1R (scanl + 0 A) in! &\\
     & !let A$_1$ = shift1L (map2 (x,y -> 1) A$_0$ A) in! & \\ 
@@
     & !Y(map2 (a,b -> a*b*z) A$_3$ A$_2$)! & \\
-    & \Big)(ZerosLike(A),1)\\
-    = & !Y(OnesLike(A))! & \\
+    & \Big)!(ZerosLike(A),1)!\\
+    = & !Y(OnesLike(A))! & (same reduction as previously)\\
     = & !let C$_1$ = map2 (a,b -> b) A B in! & !C$_1$ = B! \\
     & !let C$_2$ = map2 (a,b -> a) A B in! & !C$_2$ = A! \\
-    & !(map2 * C$_1$ OnesLike(A), map2 * C$_2$ OnesLike(A))!  &\\
+    & !(map2 * C$_1$ OnesLike(A),! &\\ 
+    & !map2 * C$_2$ OnesLike(A))!  &\\
     = & !let C$_1$=B in! & forward substitution \\
@@
     & !map2 * C$_2$ OnesLike(A))! & !C$_2$! \\
-    = & !(map2 * B OnesLike(A), map2 * A OnesLike(A))! \\
+    = & !(map2 * B OnesLike(A),! & \\
+    & !map2 * A OnesLike(A))! & \\
     = & !(B, A)!

### 2021-07-09 rad-icfp-2021/correctness.tex
some small fixes

@@
 
-\begin{definition}[Mutlicat from concat]
-    We define $\CSource$ to be the multicategory with the same objects as Source and 
+\begin{definition}[Mutlicategory from concat]
+    We define $\CSource$ to be the multicategory with the same objects as Source and\\ 
     $\CSource([A_1,\ldots,A_n],B)=\{f\in\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B], \forall i.f;\pi_{A_i}=id_{A_i})$.
     The composition of $f_i:\underline{A}_i\to B_i$ with $g:\underline{B}\to C$ is given 
-    $f_1\concatcomp\ldots f_n;(u_{\underline{B}_1}\concatcomp A_1 \concatcomp \ldots \concatcomp u_{\underline{B}_n}\concatcomp A_n);g$.
+    $f_1\concatcomp\ldots\concatcomp f_n;(u_{\underline{B}_1}\concatcomp A_1 \concatcomp \ldots \concatcomp u_{\underline{B}_n}\concatcomp A_n);g$.
     In other words each term $f_i$ forgets about its output $\underline{B}_i$, and then we use the composition in the concategory.
@@
     where $\pi_{last}$ is the projection to the last element (the continuation variable).
-    It naturally extends on sequential composition. It does not automatically extend to a multi-functor between concategories because \\
-    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$] $\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$]$\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.
+    It naturally extends on sequential composition. 
+    It does not automatically extend to a multi-functor between concategories because \\
+    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]!{$\concatcomp$}![A$_{k+1}$,$\ldots$,A$_n$] $\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp\Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.\\
     However there is a map !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])$\to \Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$])!.

### 2021-07-09 rad-icfp-2021/relwork.tex
some small fixes

@@
 They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
-Instead, they follow the principle of unrolling for recursive calls and evaluating the conditional before differentiating, 
+Instead, they follow the principle of unrolling the recursive calls and evaluating the conditional before differentiating, 
 which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
-This is a well-known limit in TensorFlow \cite{abadi2016tensorflow} which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.
+This is a known limitation in TensorFlow \cite{abadi2016tensorflow}. %which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.
 
@@
 
-Relatively recent work in AD (2008-2020):
-\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,lantern_icfp,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}
+% Relatively recent work in AD (2008-2020):
+% \cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,lantern_icfp,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}
 

### 2021-07-09 rad-icfp-2021/relwork.tex
improving related work

--- 2021-07-08 rad-icfp-2021/relwork.tex
@@
 
-An exception is \cite{abadi-plotkin2020} which gives a reverse-mode transformation that is proved correct and with the right complexity.
-This work is purely first-order, does not have arrays and uses a non-standard semantics. 
-Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.
-They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
-Instead, they follow the principle of unrolling for recursive calls and evaluating the conditional before differentiating, 
-which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
-This is a well-known limit in TensorFlow \cite{abadi2016tensorflow} which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.
+% An exception is \cite{abadi-plotkin2020}.
+% This work is purely first-order, does not have arrays and uses a non-standard semantics. 
+% Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.
+% They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
+% Instead, they follow the principle of unrolling the recursive calls and evaluating the conditional before differentiating, 
+% which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
+% This is a known limitation in TensorFlow \cite{abadi2016tensorflow}. %which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.
 
@@
 which require extremely efficient tensor computations and automatic differentiation for backpropagation, 
-combining array languages and automatic differentiation (tensor calculus) in the best 
-fitted intermediate representation for optimizations is of key interest and active research.
-\cite{bernstein2020differentiating} 
-TODO
-\cite{laue2018computing,laue2020simple}
+combining array languages and automatic differentiation (tensor calculus) in the best-fitted 
+intermediate representation for optimizations is of key interest and active research. 
+See for instance the recent work by \cite{bernstein2020differentiating,laue2018computing,laue2020simple}.
+Advanced array programming is considered an orthogonal problem to AD and we focused our work on the latter.
 
@@
 
-TODO
-\cite{lee2020correctness}: non diff function
-\cite{sherman2021}: non diff HO functions
-\cite{pearlmutter2008reverse}: based on this
-\cite{elliott2018simple}: bit similar because using CT as useful language, but we take impl and opti more seriously.
-\cite{shaikhha2019efficient}: similar approach but here more complicated for reverse mode, would be interesting to investigate more loop fusions here.
-\cite{paszke2021getting}
-Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
-    For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
-    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.
-
-Relatively recent work in AD (2008-2020):
-\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,lantern_icfp,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}
-
-Good recent surveys: \cite{van2018automatic,baydin2017automatic}
+Table~\ref{tbl:relwork} does not reflect all the aspects of AD. For instance \cite{lee2020correctness} studies in more detail non differentiability, 
+and \cite{sherman2021} the differentiability of highly-non trivial higher-order functions. 
+Our work is somewhat close in spirit to the idea of \cite{elliott2018simple} of compiling to categories.
+The idea of using closures as back-propagators is receiving recent attention, as is highlighted in \cite{vytiniotis2019differentiable},
+these ideas are used in Julia Zygote \cite{innes2019zygote}, Swift AD \cite{wei2018first}, and recently in \cite{paszke2021getting}.
+These seem closer to using control mechanims than having purely functional reverse-derivatives as far as we could tell.
+Other aspects of AD still are discussed in recent surveys \cite{van2018automatic,baydin2017automatic}.
 

### 2021-07-09 rad-icfp-2021/background.tex
Minor fixes
minor changes section2

@@
 
-% \subsection{Rudiments of differentiation and dual numbers.}
-% Recall that the derivative of a function $f:\RR\to \RR$, if it exists, is a function
-% $\nabla f:\RR\to \RR$ such that $\nabla f(x_0)=\frac {\dif f(x)}{\dif x}(x_0)$ is the gradient of $f$ at $x_0$.
-
-% To find $\nabla f$ in a compositional way, two generalizations are reasonable:
-% \begin{itemize}
-% \item We need both $f$ and $\nabla f$ when calculating $\nabla (f;g)$
-% of a composition $f;g$, using the chain rule, so we are really interested in the pair $(f,\nabla f):\RR\to \RR\times \RR$;
-% \item In building $f$ we will need to consider functions of multiple arguments, such as $+:\RR^2\to \RR$, and these functions should propagate derivatives.
-% \end{itemize}
-% Thus we are more generally interested in transforming a function $g:\RR^n\to \RR$ into a function
-% $h:(\RR\times \RR)^n\to \RR\times \RR$ in such a way that for any
-% $f_1\dots f_n:\RR\to\RR$, 
-% \begin{equation}
-%   \label{eqn:dualnumber}
-%   (f_1,\nabla f_1,\dots, f_n,\nabla f_n);h
-%   =
-%   ((f_1,\dots, f_n);g,\nabla ((f_1, \dots, f_n);g))\text.
-% \end{equation}
-
-% An intuition for $h$ is often given in terms of dual numbers.
-% The transformed function operates on pairs of numbers, $(x,x')$, and it is common
-% to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
-% But while this is a helpful intuition, the formalization of infinitesimals can be intricate, 
-% and the development in this paper is focussed on the elementary formulation in~\eqref{eqn:dualnumber}.
-
-% The reader may also notice that $h$ encodes all the partial derivatives of
-% $g$. For example, 
-% if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
-% $h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
-% and similarly 
-% $h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
-% And conversely, if $g$ is differentiable in each argument, then
-% a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
-% combinations of partial derivatives:
-% \[\textstyle h(x_1,x_1',x_2,x_2')=(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))\text.\]
-
-% In summary, the idea of differentiation with dual numbers is 
-% to transform a differentiable function
-% $g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives.
-We packaged this up in~\eqref{eqn:dualnumber} as a sort-of invariant which is useful for building derivatives of compound functions $\RR\to\RR$ in a compositional way.
-% The idea of automatic differentiation is to perform this transformation at the source code level.
-
-% The two main ways in which it is performed in practice is by operator overloading \cite{} or by source code transformation. We focus our work on the latter.
-
-% \subsection{Reverse-mode AD}
-
-% AD comes in two main flavors: forward mode and reverse mode. 
-% Forward mode computes a directional derivative of the Jacobian while reverse-mode computes a directional derivative of the transpose Jaobian. 
-% This has a significant impact in practice as this means reverse mode can compute the whole gradient of a function $f:\RR^n\to \RR$ in one pass, whereas forward mode needs $n$ passes. 
-% Assuming that both have about the same complexity, this can have a huge impact in practice. 
-% Such functions are expremely commonly seen as loss functions in the context of machine learning for instance.
-
-% Reverse-mode computes the derivatives backward, using the symmetry of the chain rule.
-% This means the compute flow is reverted after the first pass to compute the value of the function and its intermediate values. 
-% These values appear in the chain rule so should be stored or recomputed.
-% These things make reverse-mode harder to implement efficiently, and especially harder to analyze and optimize. 
-% Traditional implementations use references.
-% For instance, let 
-%  \begin{align*}
-%      \trm\defeq &\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
-%  \end{align*}
-%  The imperative version of the reverse differentiation of $\trm$ is
-% 	\begin{align*}
-% 		&w_1=\var_1 * \var_2 \\
-% 		&w_2=w_1 * \var_1 \\
-%         &\var_1':=0~;~\var_2':=0 \\
-%         &w_1':=0 \\
-% 		&w_2':=1 \\
-% 		&w_1'+= \var_1 * w_2'; \var_1'+= w_1*w_2' \\
-% 		&\var_1'+= \var_2 * w_1'; \var_2'+=\var_1 * w_1'\\
-% 		&\text{return }(\var_1',\var_2')  
-% 	\end{align*}
-% The tangent part for every variable from the context $\var_i'$ and every intermediate variable $w_i'$ is initialized at $0$, 
-% except for the return variable $w_2$ for which  $w_2':=1$.
-
-% As noted in \cite{pearlmutter2008reverse}, in a functional language, one can have a purely functional reverse-mode using continuations. 
-% This representation is notoriously inefficient, and the existing implementations of reverse mode in a functional setting have followed the lines of the imperative version above \cite{pearlmutter2008reverse, wang2018demystifying, baydin2016diffsharp}.
-
-% In the next section, we first introduce a very simple first-order language. 
-% Then we define reverse-mode on this language as a program transformation. 
-% This first transformation is quite inefficient and we how we can use standard optimizations on functional languages to obtain a provably efficient pure program.
-% The key idea behind that will be generalized and explored in detail in the rest of the paper is that of turning every operator into a unary one.     
-% The remaining of the paper will generalize this idea and explore some consequences of this new approach.
-
 \subsection{Rudiments of forward-mode AD and dual numbers}
 
-Recall that the derivative of a function $f:\RR\to \RR$, if it exists, is a function
-$\nabla f:\RR\to \RR$ such that $\nabla f(x_0)=\frac {\dif f(x)}{\dif x}(x_0)$ is the gradient of $f$ at $x_0$.
+To find the gradient $\nabla f$ of a function $f:\RR\to \RR$ in a compositional way, need both $f$ and $\nabla f$ when calculating $\nabla (f;g)$.
+We are more thus generally interested in transforming a function $f:\RR^n\to \RR$ into a function
+$g:(\RR\times \RR)^n\to \RR\times \RR$ in such a way that for any
+$f_1\dots f_n:\RR\to\RR$, 
+\begin{center}
+    $(f_1,\nabla f_1,\dots, f_n,\nabla f_n);g = ((f_1,\dots, f_n);f,\nabla ((f_1, \dots, f_n);f))$.
+\end{center}
 
-To find $\nabla f$ in a compositional way, two generalizations are reasonable:
-\begin{itemize}
-\item We need both $f$ and $\nabla f$ when calculating $\nabla (f;g)$
-of a composition $f;g$, using the chain rule, so we are really interested in the pair $(f,\nabla f):\RR\to \RR\times \RR$;
-\item In building $f$ we will need to consider functions of multiple arguments, such as $+:\RR^2\to \RR$, and these functions should propagate derivatives.
-\end{itemize}
-Thus we are more generally interested in transforming a function $g:\RR^n\to \RR$ into a function
-$h:(\RR\times \RR)^n\to \RR\times \RR$ in such a way that for any
-$f_1\dots f_n:\RR\to\RR$, 
-\begin{equation}
-  \label{eqn:dualnumber}
-  (f_1,\nabla f_1,\dots, f_n,\nabla f_n);h
-  =
-  ((f_1,\dots, f_n);g,\nabla ((f_1, \dots, f_n);g))\text.
-\end{equation}
-
-The idea of automatic differentiation (AD) is to systematically
-transform a differentiable function $g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives.
-An intuition for $h$ is often given in terms of dual numbers.
-The transformed function operates on pairs of numbers, $(x,x')$, and it is common
+The idea of AD is to systematically transform a differentiable function $f:\RR^n\to \RR$ to a function $g:\RR^{2n}\to \RR^2$ which captures~$f$ and all its partial derivatives.
+An intuition for $g$ is often given in terms of dual numbers. The transformed function operates on pairs of numbers, $(x,x')$, and it is common
 to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
-
-The main two ways in which AD is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
+The main two ways in which AD is performed in practice is by operator overloading or by source code transformation (see e.g.
+\cite{griewank2008evaluating} Chapter 6).
 Our approach focuses on source code transformation which is better fitted for compilation and optimizations.
 
-% An intuition for $h$ is often given in terms of dual numbers.
-% The transformed function operates on pairs of numbers, $(x,x')$, and it is common
-% to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
-% But while this is a helpful intuition, the formalization of infinitesimals can be intricate.
-
-% The reader may also notice that $h$ encodes all the partial derivatives of
-% $g$. For example, 
-% if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
-% $h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
-% and similarly 
-% $h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
-% And conversely, if $g$ is differentiable in each argument, then
-% a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
-% combinations of partial derivatives:
-
-% \begin{center}
-%     \begin{tabular}{r c l}
-%         $h(x_1,x_1',x_2,x_2')$ &= & $(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))$
-%     \end{tabular}
-% \end{center}
-
-% In summary, the idea of differentiation with dual numbers is 
-% to transform a differentiable function
-% $g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives.
-We packaged this up in~\eqref{eqn:dualnumber} as a sort-of invariant which is useful for building derivatives of compound functions $\RR\to\RR$ in a compositional way.
-% The idea of automatic differentiation is to perform this transformation at the source code level.
-
-% The main two ways in which it is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
-% Our approach focuses on source code transformation which is better fitted for compilation and optimizations.
-
-\subsection{Reverse-mode AD}
+\subsection{Reverse-mode Automatic Differentiation}
 
@@
 Suppose given a function $f=f_n\circ\ldots\circ f_1 :\RR^n\to\RR$. 
-The simplest way to see this mathematically, is first to say that forward mode essentially computes $(\J f)v=\J f_n(\J f_{n-1}(\ldots(\J f_1v))\ldots)$ for a direction $v\in\RR^n$. 
+Mathematically, forward mode essentially computes $(\J f)v=\J f_n(\J f_{n-1}(\ldots(\J f_1v))\ldots)$ for a direction $v\in\RR^n$. 
 Reverse-mode, on the other hand, computes $(\J f)^Tv = \J^Tf_1(\J^T f_{2}(\ldots(\J^T f_nv))\ldots)$ for a vector $v\in\RR$.
@@
 
-Because the computation flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
+Because the computation flow of the function is reversed, the actual implementation of reverse-mode is way quite tricky. 
 Reverse-mode AD is only well-understood as a source-code transformation on limited programming languages. 
@@
 first-order program is left, and then they evaluate this new program \cite{carpenter2015stan,paszke2017automatic}. 
-Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. 
-As such, the implementation process is precarious and labour intensive as these AD libraries need to be implemented using carefully, manually optimized code.
-In addition, some whole-program optimizations that a compiler would detect go entirely unused in such systems.
+Such approaches have the severe drawback that the obtained code cannot benefit from existing optimizing compilers.
+As such, the implementation process is tedious and labor-intensive as these AD libraries need to be implemented using carefully, manually optimized code.
+In addition, some whole-program optimizations that a compiler would detect are completely missed.
 
@@
 Following \cite{pearlmutter2008reverse}, there is a simple way to define an inefficient yet purely functional reverse-mode transformation for first-order programs.
-We review a slight modification of the transformation they presented, which is also better explained through an example.
+We review a slight modification of their transformation, which is also better explained through an example.
 
-Let the term we want to differentiate be $x_1:\RR,\ldots,x_n:\RR\vdash \exp(\cos(x_i))$.
-Following the chain rule, we need the jacobian matrices of $\cos$ at $x_i$ and of $\exp$ at $\cos(x_i)$. 
+Let us consider the term $x_1:\RR,\ldots,x_n:\RR\vdash \exp(\cos(x_i))$.
+To compute its gradient, following the chain rule, we need the jacobian matrices of $\cos$ at $x_i$ and of $\exp$ at $\cos(x_i)$. 
 Instead of considering these as operations from $\RR\to\RR$, we consider them as functions from the whole context. So $\cos$ and $\exp$ are seen as functions $\RR^n\to\RR$.
 However, by simply doing this we lose compositionality. 
-So $\cos$ will also return its context and is now seen as a function $\sem{\cos}:\RR^n\to\RR^{n+1}$.
-Similarly for $\exp$, but also taking the return value of $\sem{\cos}$ as an extra argument, the one it will actually use and not simply return. 
-We thus obtain $\sem{\exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $\J\sem{\cos} \in Mat_{n,n+1}$, $\J\sem{\exp} \in Mat_{n+1,n+2}$ also compose nicely.
-The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
+So we modify $\cos$ to also return its context. It is now seen as a function $\sem{\cos}:\RR^n\to\RR^{n+1}$.
+Similarly, $\exp$ is transformed. It also needs to take the return value of $\sem{\cos}$ as an extra argument, the one it will actually use and not simply return. 
+We thus obtain $\sem{\exp}:\RR^{n+1}\to\RR^{n+2}$. Now the jacobians matrices $\J\sem{\cos} \in Mat_{n,n+1}$, $\J\sem{\exp} \in Mat_{n+1,n+2}$ compose nicely.
+The same can be done for binary operators and let bindings.
 This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ\ldots\circ f_1$. 
@@
 
-To actually reverse the order of computation needed for this transpose of jacobians, we use a sort of simple continuation;
+To actually reverse the order of computation needed for this transpose of jacobians, we use a simple continuation;
 $f_i$ is turned into $\Dsynrevsymbol{f_i}:=<f_i, \lambda Y. Y\circ J^Tf_i>$ where $Y:\RR^{n+i-1}\to \RR^n$. 
@@
 $<f_{i+1}\circ f_i, \lambda Y. Y\circ J^Tf_i \circ J^Tf_{i+1}>$, and thus by induction we can obtain $<f, \lambda Y. YJ^Tf>$.
-By applying the identity continuation $\RR^n\to\RR^n$ on the second component then the result to $(0,\ldots,0,1)$, 
-we have obtained a purely functional to compute $\nabla f$.
+By applying the identity continuation $\RR^n\to\RR^n$ on the second component and then the result to $(0,\ldots,0,1)$, 
+we have obtained a purely functional way to compute $\nabla f$.
 
@@
 We develop this idea further for a richer language.
+
+\pagebreak

### 2021-07-09 rad-icfp-2021/generalization.tex
fixing map2
minor changes section2

@@
 && !<reduce (x,y.e1) y$_1$ A, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
-&& !Y((x$_1$,$\ldots$,x$_n$)$\widehat{+}$B$_2$,0,map2 (a,b. a*b*z) A$_2$ A$_3$)! \\ \hline
+&& !Y((x$_1$,$\ldots$,x$_n$)$\widehat{+}(z*$B$_2$),0,map2 (a,b. a*b*z) A$_2$ A$_3$)! \\ \hline
 \end{tabular}
@@
 
-It is slightly better because both derivatives of !e2! and !e3! are put together, and this might unlock some optimizations, 
+Now both derivatives of !e2! and !e3! are put together, and this might unlock some optimizations, 
 but there is still a non-linear usage of the continuation variable !Y!.
@@
 Non smooth functions such as the Rectified Linear Unit (ReLU) in machine learning which can be defined as $ReLU(x)\defeq max(0,x)$ 
-requires a non smooth primitive such as $>0: \reals\to\BB$. In several AD systems such as TensorFlow, 
+requires a non-smooth primitive such as $>0: \reals\to\BB$. In several AD systems such as TensorFlow, 
 the condition is evaluated before differentiation is applied, and so conditionals are never directly differentiated.
@@
 \smartpara{Correctness}
-To check correctness of the given Jacobian, it suffices to 
+To check the correctness of the given Jacobian, it suffices to 
 check that the semantics of the transpose Jacobian matches the Jacobian of the unrolled program.

### 2021-07-09 rad-icfp-2021/complexity.tex
fixing map2
minor correctness section

@@
 If we try to show that the cheap gradient principle holds by induction on the term, it fails on the !map2!.
-When differentiating !map2!, there are two series of calls to (sub)gradients of !e$_1$!. 
+When differentiating !map2!, there are three series of calls to (sub)gradients of !e$_1$!. 
 The induction hypothesis is then too weak to conclude. 
@@
 \begin{example}
-    As shown in appendix \ref{sub:gradintro}, the optimizations from Figure~\ref{fig:optim} 
+    As shown in the supplementary material,  
+    %appendix \ref{sub:gradintro}, 
+    the optimizations from Figure~\ref{fig:optim} 
     are sufficient to show that the gradients of the terms of the introduction reduce to the following.

### 2021-07-09 rad-icfp-2021/correctness.tex
minor correctness section

@@
 
- \begin{theorem}
-     For every $\Gamma \vdash$ !e:!$\reals$, we have 
-     $\sem{\grad_\Gamma e}= \grad_\Gamma\sem{e}$ in the category of diffeological spaces.
- \end{theorem}
-
 \subsection{Denotational semantics} % (fold)
@@
 
-A term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$ in a multicategory.
+A term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted in a multicategory as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$.
 Substitution is interpreted as composition. We first consider a syntactic model for a language, which consists of a 
@@
 
-We similarly define $\SynTarget$, the syntactic multicategory for Target.
+We similarly define $\SynTarget$, the syntactic multicategory on the target language.
 
@@
 
-    \begin{tabular}{r c l} 
-    !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>;swap;$(\seml$ e$_1\semr)^{\times n}$!
-    \end{tabular}
-
-where $\Delta_n$ is the $n$-copy map !<id,$\ldots$,id>!.
-The semantics for !reduce (x,y.e$_1$) e$_2$ e$_3$!
-is given by first !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>!, following a swap isomorphism  $\Gamma^n \times \RR \times \RR^n \to \reals \times (\reals \times \Gamma)^n$.
-Finally, we apply $\seml$ e$_1\semr \times id$ $n$ times, where the identity is of the obvious type at each stage.
-
 Variables are interpreted as projections $\pi_i$, let binding as composition in the multicategory. 
 Pairs are interpreted using the Cartesian structure of the multicategory.
+This interpretation extends to !map2 (x,y.e$_1$) e$_2$ e$_3$!. It is given by
+!<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>;swap;$(\seml$ e$_1\semr)^{\times n}$!
+where $\Delta_n$ is the $n$-copy map !<id,$\ldots$,id>!, and !swap! a permutation.
+Similarly, the semantics for !reduce (x,y.e$_1$) e$_2$ e$_3$!
+is then given by first !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>!, followed by a perumutation $\Gamma^n \times \RR \times \RR^n \to \reals \times (\reals \times \Gamma)^n$.
+Finally, we apply $\seml$ e$_1\semr \times id$ $n$ times, where the identity is of the obvious type at each stage.
 \end{definition}
 
-We can similarly interpret Target in a multicategory of smooth-like spaces and functions. 
+We can similarly interpret the target language in a multicategory of smooth-like spaces and functions. 
 However, Target is higher-order and $\CartSp$ is not Cartesian Closed. 
@@
 One main reason for introducing UNF is to have a better handle over
-the computation flow of the term, similarly to what the ANF or CPS transformations do.
+the computation flow of the term, in the same vein as the ANF or CPS transformations.
 A convenient categorical setting for this is to use string diagrams.
 To better fit the standard denotational semantics of languages, 
-we can instead use concategories.
-
-Concategories are a generalization of categories and multicategories in that 
-they allow morphisms both from and to a list of objects. 
-Similarly to categories and multicategories, 
-most categorical construtions used for the semantics of functional languages 
+we use concategories instead. Similarly to categories and multicategories, 
+most categorical constructions used for the semantics of functional languages 
 have equivalent in concategories.
@@
 which satisfies a similar universal property as syntactic multicategories.
-Two particular examples are of interest for us, as they will allow us to interpret Source and Target UNF
-but they will also help us explain the UNF and $UNF^{-1}$ transformations.
+Two particular examples are of interest to us, as they will allow us to interpret Source UNF and Target UNF
+but they will also help us explain the $\UNFSymbol$ and $UNF^{-1}$ transformations.
 
@@
 
-This is proved in Appendix \ref{sub:Reverse derivative of array operations}.
+This is proved in the supplementary material.
+%Appendix \ref{sub:Reverse derivative of array operations}.
 

### 2021-07-09 rad-icfp-2021/correctness.tex
minor correctness section

@@
 One may notice that the syntax of $\ConcatS$ is somewhere in between the syntax of Source and of Source UNF.
-Given two morphisms !e$_1$:T$_1$->T$_2$! and !e$_2$:T$_3$->T$_4$!, we denote by !e$_1 \concatcomp$ e$_2$:T$_1$,T$_3$->T$_2$,T$_4$!.
+Given two morphisms !e$_1$:T$_1$->T$_2$! and !e$_2$:T$_3$->T$_4$!, we denote by !e$_1 \concatcomp$ e$_2$:T$_1$,T$_3$->T$_2$,T$_4$! 
+their parallel composition.
 We denote by $u_T$ the unique morphism from !T! to the terminal object $[]$.
+$pair_{T;A\times B}$ is the canonical isomorphism pairing !A,B! into !A$\times$B! in the context !T!.
 We can interpret Source UNF in $\ConcatS$ as follows.
@@
 this gives us a way to compare terms in Source with terms in Source UNF.
-This comparison precisely gives the UNF transformation.
+This comparison of morphisms of $\ConcatS$ gives the $\UNFSymbol$ transformation.
 
@@
 
-Interpretating Source in $\CSource$ allows us to see terms of Source as morphisms in the concategory $\ConcatS$.
-This allows us to compare to semantics of terms of Source and of Source UNF.
+Interpretating Source in $\CSource$ allows us to see terms of Source as morphisms in the concategory $\ConcatS$, 
+and to compare them to terms of Source UNF which are already interpreted in $\CSource$.
 The following proposition can be shown by induction on the structure of the terms.
@@
     Let !$\Gamma \vdash$ e:A! be a term in Source. 
-    Seen as morphisms in $\CSource$, we have !$\seml$e$\semr$=$\seml\UNFSymbol$(e)$\semr$!.
+    Seen as morphisms in $\CSource$, !$\seml$e$\semr$=$\seml\UNFSymbol$(e)$\semr$!.
 \end{proposition}
 
-In a similar vein, we form a concategory from the syntactic multicategory for Target.
-We then use the universal property of $\ConcatT$ to construct a functor from $\ConcatT$ to this concategory.
-This allows us to compare the terms of Target UNF and the terms of Target. 
-$UNF^-1$ arises in this way.
+Somewhat dually, we form a concategory from the syntactic multicategory for Target.
+Then we use the universal property of $\ConcatT$ to construct a functor from $\ConcatT$ to this concategory.
+This allows us to compare the terms of Target UNF and the terms of Target, and $\UNFSymbol^{-1}$ arises in this way.
 
@@
 
-This allows to form a concategory $\CTarget$ from the syntactic multicategory for Target.
+We thus consider the concategory $\CTarget$ from the syntactic multicategory for Target.
 We can interpret $\ConcatT$ in $\CTarget$. 
-The functor is identity on objects, sends operations !op1,op2! to themselves. 
-It sends Jacobian operations to terms of Target as given by $\UNFSymbol^{-1}$ in Figure~\ref{fig:unf_to_target}
+The functor is identity-on-objects, sends operations !op1,op2! to themselves. 
+It sends Jacobian operations to terms of Target as given by $\UNFSymbol^{-1}$ in Figure~\ref{fig:unf_to_target}.
 
@@
 The difference is that to preserve typing, the semantic $\UNFSymbol^{-1}$
-sends a non Jacobian primitive to a tuple, as in the Example \ref{exm:unf}. 
-This is highly inefficient and the the syntactic $\UNFSymbol^{-1}$ 
-additionally projects to the last element, where the non trivial information of the term is.
+sends a non-Jacobian primitive to a tuple, as in the Example \ref{exm:unf}. 
+This is highly inefficient, and the the syntactic $\UNFSymbol^{-1}$ 
+additionally projects to the last element, where the non-trivial information of the term is.
 
@@
 We construct a functor $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$. 
-This functor will easily be shown to compute reverse-mode derivatives.
+This functor is easily shown to compute reverse-mode derivatives.
 Because terms of Source UNF in $\ConcatS$, we observe the effect of $\Dsynrevsymbol_{\rho}$ on them
@@
     !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]!{$\concatcomp$}![A$_{k+1}$,$\ldots$,A$_n$] $\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp\Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.\\
-    However there is a map !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])$\to \Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$])!.
-    Internally as a lambda term, it is given by \\
+    Still, there is a map !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])$\to \Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$])!.
+    Internally, as a lambda term, it is given by \\
     $(x_1,\ldots,x_k,Y_1,(x_{k+1},\ldots,x_n,Y_s)\mapsto (x_1,\ldots,x_n,\lambda(y_1,\ldots,y_n)\to Y_1(y_1,\ldots,y_k)\widehat{+}Y_2(y_{k+1},\ldots,y_n)$.
@@
 
-In summary, we have the following picture
+In summary, we have the following picture (not a proper categorical diagram):
 
@@
 
-First, we start from the correctness of $\Dsynrevsymbol_{\rho}$:Source UNF $\to$ Target UNF, which is easy to establish, 
+First, we start from the correctness of the syntactic $\Dsynrevsymbol_{\rho}$: Source UNF $\to$ Target UNF, which is easy to establish, 
 and then propagate this information to Source and Target via the $\UNFSymbol$ and $\UNFSymbol^{-1}$ transformations. 
+The semantics brackets $\sem{-}$ in this section are in diffeological spaces.
 
@@
 From there, the intuition for the Jacobian of a primitive in Target UNF is that it should be the Jacobian of
-the corresponding term in Target. This is easily checked for scalar operations. For the more difficult cases, we have
+the corresponding term in Target. This is easily checked for scalar operations. 
+For the non-trivial cases, we have
 

### 2021-07-09 rad-icfp-2021/relwork.tex
minor changes section2

@@
 In addition, standard techniques of defunctionalization struggle when higher-order is combined with recursion.
+Our reverse-mode transformation does not support recursion, 
+so it is currently always possible to partially evaluate a higher-order program to our Source language, seen as intermediate representation,
+then perform our reverse-mode transformation.
 
@@
 \noindent \textbf{Correctness of AD in functional languages.}
-Several recent work \cite{huot2020correctness,vakar2020reverse,vakar2020denotational,brunel2019backpropagation,barthe2020versatility,mazza2021automatic,lee2020correctness} have focused on correctness of AD in a purely functional setting,
+Several recent work \cite{huot2020correctness,vakar2020reverse,vakar2020denotational,brunel2019backpropagation,barthe2020versatility,mazza2021automatic,lee2020correctness,abadi-plotkin2020} have focused on correctness of AD in a purely functional setting,
 often leaving efficiency on the side, especially for reverse-mode differentiation. 
 We see our work as a complement and a first bridge between these works 
-and more practical considerations of efficiency and implementations, 
+and more practical considerations of efficiency, 
 which often require a lot more care than is acknowledged in more theoretical works.
@@
 
-\noindent \textbf{Usage of iteration mechanics}
+\noindent \textbf{Usage of iteration mechanics.}
 An immense effort in machine learning for the past decade has been in finding
@@
 and have better building blocks for large complicated systems than traditional layers of a neural network.
-Different approaches 
-(Dynamic neural networks \cite{jin2017manipulability,wu2016deep}, 
+Different approaches such as Dynamic neural networks \cite{jin2017manipulability,wu2016deep}, 
 Recursive NN \cite{socher2011parsing,biancofiore2017recursive}, 
@@
 Dynamic Recursive NN \cite{guo2019dynamic}, 
-Top-down Tree LSTM \cite{zhang2015top}, 
-Recursion in DNN \cite{jeong2018improving}) 
+Top-down Tree LSTM \cite{zhang2015top}, and
+Recursion in DNN \cite{jeong2018improving}
 have found that recursive data structures such as trees are good candidates.
-We have emphasized here on differentiating fold left on arrays for efficiency, 
+We have emphasized here on differentiating fold-based recursion on arrays for efficiency, 
 but one should be able to adapt this on any algebraic data type. 
-It will be interesting to see if and how we recover efficient backpropagation on the proposed architectures 
+It will be interesting to see if and how we recover efficient purely functional backpropagation (as opposed to the imperative version of~\cite{lantern_icfp}) on the proposed architectures 
 which is usually derived by hand and one main goal of these papers.
 
-\noindent \textbf{Array Languages and AD}
-
+\noindent \textbf{Array Languages and AD.}
 Given the enormous computation needs for state-of-the-art large scale machine learning applications, 
@@
 combining array languages and automatic differentiation (tensor calculus) in the best-fitted 
-intermediate representation for optimizations is of key interest and active research. 
-See for instance the recent work by \cite{bernstein2020differentiating,laue2018computing,laue2020simple}.
-Advanced array programming is considered an orthogonal problem to AD and we focused our work on the latter.
+intermediate representation for optimizations is of key interest and active research~\cite{bernstein2020differentiating,laue2018computing,laue2020simple}. 
+Advanced array programming is considered an orthogonal problem to AD and we focused our work on the differentiation aspect.
 
 \noindent \textbf{Comparison to other recent papers.}
-
 Table~\ref{tbl:relwork} does not reflect all the aspects of AD. For instance \cite{lee2020correctness} studies in more detail non differentiability, 
-and \cite{sherman2021} the differentiability of highly-non trivial higher-order functions. 
+and \cite{sherman2021} the differentiability of highly non-trivial higher-order and partial functions. 
 Our work is somewhat close in spirit to the idea of \cite{elliott2018simple} of compiling to categories.
-The idea of using closures as back-propagators is receiving recent attention, as is highlighted in \cite{vytiniotis2019differentiable},
-these ideas are used in Julia Zygote \cite{innes2019zygote}, Swift AD \cite{wei2018first}, and recently in \cite{paszke2021getting}.
-These seem closer to using control mechanims than having purely functional reverse-derivatives as far as we could tell.
-Other aspects of AD still are discussed in recent surveys \cite{van2018automatic,baydin2017automatic}.
+The idea of using closures as back-propagators is receiving recent attention, as is highlighted in \cite{vytiniotis2019differentiable,lantern_icfp}.
+These ideas are used in Julia Zygote \cite{innes2019zygote}, Swift AD \cite{wei2018first}, and recently in \cite{paszke2021getting}.
+These seem closer to using control mechanims than having purely functional reverse-derivatives.
+Other aspects of AD are discussed in recent surveys \cite{van2018automatic,baydin2017automatic}.
 

### 2021-07-09 rad-icfp-2021/purerad.tex
minor

@@
 Ground types are defined inductively by
-$$G::= \RR \mid G\times \ldots \times G \mid \Array{\RR}{n}$$
+\begin{center}
+    $G::= \RR \mid G\times \ldots \times G \mid \Array{\RR}{n}$
+\end{center}
 
@@
     $0_{G_1\times \ldots \times G_n}$ & $\defeq$ &  $< O_{G_1},\ldots, 0_{G_n} >$ \\
-    $0_{\Array{\RR}{n}}$& $\defeq$ & !ZerosLike(n)! \\
+    $0_{\Array{\RR}{n}}$& $\defeq$ & !ZerosLike(n)!
+\end{tabular}
+\begin{tabular}{l c l}
     $a\widehat{+}_\RR b$ & $\defeq$ & $ a+b$ \\
-    $(a_1,\ldots,a_n)\widehat{+}_{G_1\times\ldots\times G_n}(b_1,\ldots,b_n)$ & $\defeq$ & $(a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n)$ \\
+    $(a_1,\ldots,a_n)\widehat{+}_{G_1\times\ldots\times G_n}$ & $\defeq$ & $(a_1\widehat{+}_{G_1}b_1,\ldots,a_n\widehat{+}_{G_n}b_n)$ \\
+    $(b_1,\ldots,b_n)$ && \\
     $A\widehat{+}_{\Array{\RR}{n}}B $ & $\defeq$ & !map2! + $A$ $B$ 
@@
 
-We have the following typing lemma for $\directD{\rho}{\Gamma}{Y}$.
+We have the following typing lemma for $\directD{\rho}{\Gamma}{Y}$, routinely proved by induction on derivation of $\Gamma\vdash$!e:A!.
 \begin{lemma}[Typing $\directD{\rho}{\Gamma}{Y}$]
@@
 \end{lemma}
-\begin{proof}
-By induction on derivation of $\Gamma\vdash$!e:A!.
-\end{proof}
 

### 2021-07-09 rad-icfp-2021/generalization.tex
.
minor

@@
 \end{small}
+\vspace{-0.4cm}
 \caption{The reverse-mode AD transformation of \codekw{reduce} without restrictions on its function argument.}
+\vspace{-0.4cm}
 \label{fig:lift_reduce}
@@
 
-\begin{tabular}{r c l}
-    $\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) 
-    &=& !let b=e$_1$ in!   \\
-    && !<if b then e$_2$ else e$_3$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
-    && !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e$_2$)(x$_1$,$\ldots$,x$_n$,z)+!\\
-    && \quad!(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e$_3$)(x$_1$,$\ldots$,x$_n$,z))>!
+\begin{tabular}{l}
+    $\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) \\
+    !let b=e$_1$ in <if b then e$_2$ else e$_3$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
+    !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e$_2$)(x$_1$,$\ldots$,x$_n$,z)+(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e$_3$)(x$_1$,$\ldots$,x$_n$,z))>!
 \end{tabular}
@@
 
-\begin{tabular}{r c l}
-    $\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) 
-    &=& !let b=e$_1$ in!   \\
-    && !<if b then e$_2$ else e$_3$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
-    && !Y((x$_1$,$\ldots$,x$_n$)!$\widehat{+}$($\grad_\Gamma$!(e$_2$)*b!$\widehat{+}\grad_\Gamma$!(e$_3$)*(1-b))*z)>!\\
+\begin{tabular}{l}
+    !$\directD{\rho}{\Gamma}{Y}$(if e$_1$ then e$_2$ else e$_3$) = let b=e$_1$ in <if b then e$_2$ else e$_3$,! \\
+    !fun (x$_1$,$\ldots$,x$_n$,z) -> Y((x$_1$,$\ldots$,x$_n$)!$\widehat{+}$($\grad_\Gamma$!(e$_2$)*b!$\widehat{+}\grad_\Gamma$!(e$_3$)*(1-b))*z)>!\\
 \end{tabular}

### 2021-07-09 rad-icfp-2021/correctness.tex
.

@@
     $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& $id_T\concatcomp$!<$\pi_1,\pi_2$>! \\
- \end{tabular}
+ \end{tabular}\\
  where !T,A! $\vdash$ (0,$\pi_A$,0): !T! and the projection $\pi_A$ lands in the $i$-th element of the list !T!.
@@
 
-Somewhat dually, we form a concategory from the syntactic multicategory for Target.
+Dually, we form a concategory from the syntactic multicategory for Target.
 Then we use the universal property of $\ConcatT$ to construct a functor from $\ConcatT$ to this concategory.
@@
 sends a non-Jacobian primitive to a tuple, as in the Example \ref{exm:unf}. 
-This is highly inefficient, and the the syntactic $\UNFSymbol^{-1}$ 
+This is highly inefficient, and the syntactic $\UNFSymbol^{-1}$ 
 additionally projects to the last element, where the non-trivial information of the term is.

### 2021-07-09 rad-icfp-2021/correctness.tex
.

@@
     We define $\CSource$ to be the multicategory with the same objects as Source and\\ 
-    $\CSource([A_1,\ldots,A_n],B)=\{f\in\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B], \forall i.f;\pi_{A_i}=id_{A_i})$.
+    $\CSource([A_1,\ldots,A_n],B)=\{f\in\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B], \forall i.f;\pi_{A_i}=id_{A_i})$\\
     The composition of $f_i:\underline{A}_i\to B_i$ with $g:\underline{B}\to C$ is given 

### 2021-07-09 rad-icfp-2021/relwork.tex

@@
 
-We have introduced a transformation on programs to compute provably efficient (\S\ref{sec:complexity}) 
+We introduced a transformation on programs to compute provably efficient (\S\ref{sec:complexity}) 
 gradients via reverse derivatives in a purely functional way (\S\ref{sub:Macro for pure reverse mode transformation})
@@
 We introduced a novel intermediate representation, Unary Normal Form (\S\ref{sec:unf}) 
-to decompose our transformation into simpler transformations.
-We gave denotational semantics to our languages and our transformations (\S\ref{sec:correctness}), 
-and showed correctness of the reverse-mode translation (\S\ref{sub:Correctness theorem}).
+to decompose our translation into simpler ones.
+We gave denotational semantics to our languages (\S\ref{sec:correctness}), 
+and proved correctness of the reverse-mode translation (\S\ref{sub:Correctness theorem}).
 We showed (\S\ref{sec:generalization}) how to lift the restrictions that

### 2021-07-09 rad-icfp-2021/abstract.tex
minor

@@
 \begin{abstract}
-    Reverse-mode differentiation is used for optimization, but it does not preserve purity of the underlying program which makes it notoriously harder to optimize.
+    Reverse-mode differentiation is used for optimization, but it introduces references which break the purity of the underlying program, making them notoriously harder to optimize.
     We present a reverse-mode differentiation on a purely functional language with array operations.  
-    It is the first one to deliver a provably efficient, purely functional and denotationnally correct reverse-mode differentiation.
+    It is the first one to deliver a provably efficient, purely functional, and denotationally correct reverse-mode differentiation.
     We show that our transformation is semantically correct and verifies the cheap gradient principle. 
@@
     We obtain provably efficient gradients by performing general partial evaluation optimizations after our reverse-mode transformation, as opposed to manually derived ones.
-    For simple first-order programs, the obtained output programs resemble static single assignment (SSA) code. 
+    For simple first-order programs, the obtained output programs resemble static-single-assignment (SSA) code. 
     We emphasize the modularity of our approach and show how our language can easily be enriched with more optimized primitives, as required for some speedups in practice.

### 2021-07-09 rad-icfp-2021/intro.tex
minor

@@
 In this paper, we define a purely functional (without references or control mechanisms such as state monads), denotationally correct, and provably efficient reverse-mode AD. 
-To do so, we define the Unary Normal Form (UNF) representation inspired by PROPs \cite{hackney2015category} and compilation to categories \cite{elliott2017compiling}.
+To do so, we define the Unary Normal Form (UNF) representation inspired by PROPs \cite{maclane1965categorical} %\cite{hackney2015category} 
+and compilation to categories \cite{elliott2017compiling}.
 We can easily define and prove correctness of reverse-mode on this representation. 

### 2021-07-09 rad-icfp-2021/correctness.tex
minor

@@
 The $\directD{\rho}{\Gamma}{Y}$ translation passes from the multicategory for the source syntax to the concategory for UNF, where reverse-mode differentiation is easier, and then back to the multicategory for the target syntax.
-In this way the translations between Source/Target and Source UNF/Target UNF are merely changing perspective from multicategories to concategories.
+In this way, the translations between Source/Target and Source UNF/Target UNF are merely changing perspective from multicategories to concategories.
 We work with a specific multicategory and concategory built from diffeological spaces and smooth maps, so that we can verify the construction of reverse derivatives. 
 Since diffeological spaces support products, they can be presented as a category, a multicategory and a concategory. 
-In abstract mathematics the difference between these presentations is almost trivial. 
+Mathematically, the difference between these presentations is almost trivial. 
 But in terms of programming syntax, the difference between presentations has a big effect, as can be seen from the difference between the complex translation in Section~\ref{sec:simplediff} and the simple translations in Section~\ref{sec:unf}, which are much more readily verified.
@@
 The key idea will be that a higher-order function is called smooth if it sends smooth functions to smooth functions, meaning that we can never use it
-to build first order functions that are not smooth.
+to build first-order functions that are not smooth.
 
@@
 To better fit the standard denotational semantics of languages, 
-we use concategories instead. Similarly to categories and multicategories, 
+we use concategories instead. Similar to the settings of categories and multicategories, 
 most categorical constructions used for the semantics of functional languages 
@@
 The terminal map is interpreted as the identity. The operators !op1,op2! by themselves.
-Crucially, the semantics of a !let! becomes composition in the concategory. 
+Crucially, the semantics of a !let! is simple composition in the concategory. 
 The semantics of a variable is the pairing of the identity with a projection.
@@
 
-Interpretating Source in $\CSource$ allows us to see terms of Source as morphisms in the concategory $\ConcatS$, 
+Interpreting Source in $\CSource$ allows us to see terms of Source as morphisms in the concategory $\ConcatS$, 
 and to compare them to terms of Source UNF which are already interpreted in $\CSource$.
@@
 By inspecting what that composition of transformations does on the terms of Source, 
-we show that this indeed gives computes the same as the transformation 
+we show that this indeed computes the same as the transformation 
 $\directD{\rho}{\Gamma}{Y}$ from Section~\ref{sub:Macro for pure reverse mode transformation}. 

### 2021-07-09 rad-icfp-2021/notes/main.tex
minor

@@
  Semantically, $\UNFSymbol(e1) \widehat{;} \UNFSymbol(e2)$ is a macro for $\UNFSymbol(e1) ; \UNFSymbol(\widetilde{e2}); \sigma$ where $\Gamma,\Delta_1,A\vdash \widetilde{e2}:B$ is the weakening of $e2$, and $\sigma: \Gamma\times\Delta_1\times A \times \Delta_2 \times B \to \Gamma\times\Delta_1\times \Delta_2 \times A\times B$ is the obvious permutation with this type.
-The weakening and permutation are slightly changed in the obvious way for let binding to make the thing typecheck.
+The weakening and permutation are slightly changed in the obvious way for let binding to make the thing to type-check.
 
- TODO: I probably need to add permutations $\sigma$ in the language of UNF then.
+ : I probably need to add permutations $\sigma$ in the language of UNF then.
 
@@
 
-Because UNF inspects the code of its agument, it is not invariant under $\beta,\eta$ and so cannot be directly expressed as a structure preserving functor from the syntactic category of the source language modulo $\beta,\eta$ to another category.
+Because UNF inspects the code of its agument, it is not invariant under $\beta,\eta$ and so cannot be directly expressed as a structure-preserving functor from the syntactic category of the source language modulo $\beta,\eta$ to another category.
 

### 2021-07-09 rad-icfp-2021/abstract.tex
Typos and fixes.

@@
 \begin{abstract}
-    Reverse-mode differentiation is used for optimization, but it introduces references which break the purity of the underlying program, making them notoriously harder to optimize.
+    Reverse-mode differentiation is used for optimization, but it introduces references, which break the purity of the underlying programs, making them notoriously harder to optimize.
     We present a reverse-mode differentiation on a purely functional language with array operations.  

### 2021-07-09 rad-icfp-2021/intro.tex
Typos and fixes.

--- 2021-07-08 rad-icfp-2021/intro.tex
@@
 Deep learning is moving towards increasingly sophisticated optimization objectives that employ tensors and operations on tensors.
-Reverse-mode Automatic Differentiation (AD) is a technique to automatically compute the gradient of objective functions of form $\RR^n\to\RR$.
+Reverse-mode Automatic Differentiation (AD) is a technique to automatically compute the gradient of objective functions of the form $\RR^n\to\RR$.
 Such functions appear a lot in practice: for instance, as loss functions in machine learning.
 
-In order to reach the efficiency of the usual imperative version of reverse-mode, the transformations even in functional languages~\cite{pearlmutter2008reverse} resort to references.
+In order to reach the efficiency of the usual imperative version of reverse-mode, the transformations, even in functional languages~\cite{pearlmutter2008reverse}, resort to references.
 The lack of purity in reverse-mode is known to make it significantly harder to optimize and parallelize. 
@@
 In this paper, we define a purely functional (without references or control mechanisms such as state monads), denotationally correct, and provably efficient reverse-mode AD. 
-To do so, we define the Unary Normal Form (UNF) representation inspired by PROPs \cite{hackney2015category} and compilation to categories \cite{elliott2017compiling}.
+To do so, we define the Unary Normal Form (UNF) representation inspired by PROPs~\cite{maclane1965categorical} %\cite{hackney2015category} 
+and compilation to categories~\cite{elliott2017compiling}.
 We can easily define and prove correctness of reverse-mode on this representation. 
-The whole reverse-mode transformation is obtained by compiling the language to this intermediate representation, applying the simpler reverse-mode transformation, and compiling again to the original language.
+The whole reverse-mode transformation is obtained by compiling the language to this Intermediate Representation (IR), applying the simpler reverse-mode transformation, and compiling again to the original language.
 After standard optimizations, the output program looks like SSA~\cite{cytron1989efficient} or ANF~\cite{sabry1993reasoning}, which leads to more efficient implementations.
@@
 
-We introduce the general idea of efficient reverse-mode in a functional setting the next examples.
+We introduce the general idea of efficient reverse-mode in a functional setting through the following examples.
 % We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.

### 2021-07-09 rad-icfp-2021/intro.tex
Typos and fixes.

@@
 \end{center}
-The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm. 
+The part !(0,0,0,0,1)! corresponds to initializing the tangent variables in the imperative reverse-mode algorithm. 
 After some general partial evaluation techniques that will be detailed further in the paper, we obtain:
@@
 This is very close to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.
+
+\begin{absolutelynopagebreak}
 This term can be further optimized via constant propagation and algebraic simplifications to give
@@
     \end{center}
+\end{absolutelynopagebreak}
+
 \end{example}

### 2021-07-09 rad-icfp-2021/intro.tex
Typos and fixes.

@@
 where !reduce! is a known fold-left operator for which the function argument is associative. 
-It is notably faster to execute than a fold-left as it is parallel friendly.
+It is notably faster to execute than a fold-left as it is parallel-friendly.
 
-The gradient of each of these expressions w.r.t. !A! will be:
+The gradient of each of these expressions with respect to !A! is:
 \begin{center}
@@
 \end{center}
-where:
+where
 \begin{itemize}
@@
 \end{itemize}
-These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
+These gradients are a few examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
 \end{example}
@@
 Inspired by this insight and following IRs such as SSA and ANF, we introduce a novel IR, UNF (Section~\ref{sec:unf})).
-We introduce an alternative and easier to follow compilation pipeline for efficient reverse-mode AD (bottom of Figure~\ref{fig:outline}).
-\item We prove the complexity guarantees for the reverse ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
-\item We proof the correctness of our transformations (top/bottom parts of Figure~\ref{fig:outline}) by defining the denotational semantics of our languages using multicategories and concategories (Section~\ref{sec:correctness}).
+We introduce an alternative and easier-to-follow compilation pipeline for efficient reverse-mode AD (bottom of Figure~\ref{fig:outline}).
+\item We prove complexity guarantees for the reverse-ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
+\item We prove the correctness of our transformations (top/bottom parts of Figure~\ref{fig:outline}) by defining a denotational semantics of our languages using multicategories and concategories (Section~\ref{sec:correctness}).
 \end{itemize}

### 2021-07-09 rad-icfp-2021/background.tex
Typos and fixes.

@@
 
-To find the gradient $\nabla f$ of a function $f:\RR\to \RR$ in a compositional way, need both $f$ and $\nabla f$ when calculating $\nabla (f;g)$.
-We are more thus generally interested in transforming a function $f:\RR^n\to \RR$ into a function
-$g:(\RR\times \RR)^n\to \RR\times \RR$ in such a way that for any
-$f_1\dots f_n:\RR\to\RR$, 
+To find the gradient $\nabla f$ of a function $f:\RR\to \RR$ compositionally, need both $f$ and $\nabla f$ when calculating $\nabla (f;g)$.
+This is the reason why we are more generally interested in transforming a function $f:\RR^n\to \RR$ into a function
+$g:(\RR\times \RR)^n\to \RR\times \RR$ in such a way that for every
+$f_1, \dots, f_n:\RR\to\RR$, 
 \begin{center}
@@
 
-The idea of AD is to systematically transform a differentiable function $f:\RR^n\to \RR$ to a function $g:\RR^{2n}\to \RR^2$ which captures~$f$ and all its partial derivatives.
+The idea of AD is to systematically transform a differentiable function $f:\RR^n\to \RR$ into a function $g:\RR^{2n}\to \RR^2$ which captures~$f$ and all its partial derivatives.
 An intuition for $g$ is often given in terms of dual numbers. The transformed function operates on pairs of numbers, $(x,x')$, and it is common

### 2021-07-09 rad-icfp-2021/purerad.tex
minor

@@
 and the provided initial value should be a unit of the binary operation.
-It is a well-known parallel friendly construct. 
+It is a well-known parallel-friendly construct. 
 For the sake of simplicity in the presentation, the bound function in !reduce! 
@@
 we currently restrict to arrays of reals.
-We show how to lift these restrictions and how to differentiate some other array operators in Section~\ref{sec:generalization}.
+We show how to lift these restrictions and how to differentiate some other array operators in Section~\ref{sec:generalization} 
+and in the supplementary material.
 

### 2021-07-09 rad-icfp-2021/correctness.tex
Concat temporary fix

@@
 \begin{itemize}
-    \item categories, where each arrow has one source and one target. These are most familiar but don’t directly match with a programming syntax (except the categorical abstract machine). 
+    \item categories, where each arrow has one source and one target. These are most familiar but do not directly match with a programming syntax (except the categorical abstract machine). 
     \item multicategories, where each arrow has a list of source objects !A$_1,\ldots,$A$_n$! and one target !B!, thought of as a map !A$_1$ $\times\ldots\times$A$_n\to$B!. 
@@
     It does not automatically extend to a multi-functor between concategories because \\
-    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]!{$\concatcomp$}![A$_{k+1}$,$\ldots$,A$_n$] $\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp\Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.\\
+    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$]) $\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp\Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.\\
     Still, there is a map !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])$\to \Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$])!.

### 2021-09-06 rad-icfp-2021/rebuttal_POPL.txt
starting fixing things after POPL reviews

Rebuttal:

We are very grateful to the anonymous reviewers for their detailed feedback. 
If we are given the chance for a revised version, all the minor comments and typos raised by the reviewers have been addressed.

For reviewer2's question:

For reviewer3's questions:

Reviewer 1:

Paper summary
This unary form is like Forth with typed stacks;
A Jacobian amounts to a function from stack-gradients to stack-gradients, 
which can be composed by partial evaluation as long as control flow is static.

Weaknesses
Although the paper is organized so that earlier material motivates later material, insufficient explanation makes understanding much earlier material require understanding later material. For instance:
Understanding Section 3 requires understanding Section 4 (I recommend simply dropping Section 3).
Understanding Section 4 requires understanding Section 6 (I recommend stating the invariant maintained by Figure 8 formally up front).
Understanding Section 5.1 requires understanding Section 5.4 (I recommend merging the material).
Understanding the second line on page 8 (below Figure 4) requires knowing the notation \mathit{Id}_\GammaId 
Γ
​
  (which is introduced in Figure 4, which is not referred to in the text above) and the notation 0_\Gamma0 
Γ
​
  (which is introduced below in Notation 1).
More broadly speaking, understanding this paper requires a prior understanding of reverse-mode AD to an extent that limits its audience severely at POPL.
The paper does not firmly establish the languages that its transformations move among. Examples:

Figure 4, bottom two rows, (x:A[R]^n)(x:A[R] 
n
 ) uses the unbound variable xx
To type-check e_1;e_2e 
1
​
 ;e 
2
​
  in Source UNF, Figure 6 requires T,A \vdash e_2:T,A,BT,A⊢e 
2
​
 :T,A,B, but Figure 9 uses proj and pair after ;;, and the typing rules for proj and pair in Figure 6 do not match T,A \vdash e_2:T,A,BT,A⊢e 
2
​
 :T,A,B.
Figure 9 says "constant seen as a 0-ary operator", which should be stated when defining Source UNF. Typical programming languages only have a finite number of operators.
Page 20 (between Definitions 6.4 and 6.5) interpret a \tilde{;} 
;
~
​
  construct in Source UNF, but there is no such construct in Figure 6.
Although the paper couches semantics and correctness in category theory, it boils down to the sentence "This functor is easily shown to compute reverse-mode derivatives" on page 21, yet the paper never formalizes what it means to "compute reverse-mode derivatives". Ultimately the source language in this paper is so "limited" (Section 2.2) that it doesn't advance our understanding of reverse-mode AD much.

Page 1 "PROPs": what are they?
Page 1 "often complicated heuristics with no guarantees are used": Give concrete examples of those heuristics in a footnote.
Page 2, Example 1.1: The prime ' 
′is used both for dual (w) and for update (y). Only use the prime ' 
′for one of these two purposes; choose a different symbol for the other purpose. Also, rename y1,y2,y3,y4,y5 to Dx1,Dx2,Dx3,Dw1,Dw2 (or use some other letter instead of D).
Refer to Table 1 in the text nearby.
Figure 1 and elsewhere: don't use math mode for italics.
Page 2 "The gradient of each of these expressions with respect to A is:": Clarify whether your transformation turns the 3 lines of code displayed above into the 3 lines of code displayed below.
Page 4, first sentence of Section 2.1, "need both" -> "we need both". The "To" at the beginning of this sentence is confusing: even if our goal is not "to find the gradient ∇f of a function f:R→R compositionally", don't we still "need both f and ∇f when calculating ∇(f;g)"?
Page 4, Section 2.2, define J.
Page 5, "lose compositionality": Clarify what is lost. Compositionality means that the meaning of a complex expression is determined by its structure and the meanings of its constituents.
Page 5, after "by noting that", f_{i+1}(f_i)f 
i+1
​
 (f 
i
​
 ) should be f_{i+1} \circ f_if 
i+1
​
 ∘f 
i
​
 , )()( should be )\circ()∘(, and on the next line, YJ^TfYJ 
T
 f should be Y\circ J^TfY∘J 
T
 f. On the next line, "applying the identity continuation Rn→Rn on the second component" should be "applying the second component to the identity continuation Rn→Rn".

Page 7, "We only need the target language to be second-order.": The target language can probably stay first-order if you treat functions at the meta-level of the transformation.
Page 9, "The idea is that ρ represents the return type of the derivative part" needs to be explained more clearly, with formalization and examples.
Figure 7, where is the "term pairing" e ::= <e,e> construct of target UNF used? I don't see it used in Figure 8.
Page 16, above Theorem 5.4, in the displayed equations, pairs of braces are missing after "max" (perhaps you forgot to escape your braces with backslashes).
Page 19, after "This interpretation extends to", the extra spaces before e are strange, and some e2s seem like they should be e3s. These problems make this paragraph hard to understand.
Page 20 "somewhere in between the syntax" is too vague.
Page 21, the second-to-last paragraph needs to make the following phrases concrete by referring to Example 4.1: "non-Jacobian primitive"? "a tuple"? "highly inefficient"? "projects to the last element"? "non-trivial information of the term"?
Section 7.2 discusses modifications to the composed transformation of Figure 5, but it should be easier to understand (and prove correct and efficient) to modify the decomposed transformations of Section 4 instead.
Page 23, "makes everything harder to optimize" is too broad.
Page 23, define "non-linear usage".
Page 24, "compute the whole gradient of the expression" is unclear. Give examples where we do and examples where we don't want to compute the whole gradient of the expression. It seems that there are common cases of conditionals that your solution does not handle, and that motivates the use of mutable cells allocated at run time.
Show line numbers in draft manuscripts to ease commenting. Number all displayed equations.
Spell check.

Reviewer 2:

Paper summary

Strengths
A unique aspect of the work is to consider performance aspects and to perform complexity analysis where the reverse-mode transformation is still pure.

Weaknesses
The complexity analysis does not consider any parallelism as part of the cost model. Although perhaps reasonable as a first step, the ability to parallelize AD is extremely important in practice.
The work does not include any implementation, neither as an explicit implementation for empirical evaluation, nor a mechanized approach when proving correctness or complexity. Hence, the practical performance implications can only be considered in theory.

Figure 1 shows a descriptive outline of the compilation strategy and an overview of the paper. As shown in the top of the figure and explained in Section 3, there is a direct translation from the source language to the target language. As stated in the middle of page 9, it is neither trivial to understand this approach, nor to prove it correct. And, this is why you describe the second chain via UNF? But, why do you then describe this direct translation in the first place? What does it add, compared to the second chain?
I found the presentation of UNF slightly more complicated than what I think it needs to be. Specifically, when you first introduce the source and target UNF languages (Figures 6 and 7), it is not obvious what the semantics of UNF actually is. It is first in the latter part of the section it becomes clearer, when the transformation from source to source UNF is described. Hence, this section would be more readable if the order is changed, and where the source UNF is introduced together with the transformation from the source language. A more detailed example would also be valuable.
The re-translation step (non-strict transformation UNF^{-1}) is not completely clear. In particular, the reintroduction of variable names, as briefly discussed at the end of page 13.
It would be good if the discussion of how to extend the language further could be elaborated a bit more. For instance, currently the the source language contains product types, but not sum types. How would such an extension affect the approach? Is it direct, and in such a case, why?
On page 24, it says that your approach does not support recursion, which is also clear from the rest of the paper. But, in table 1, the column for "This Paper" shows recursion half filled in. It says in the caption that it means "not supporting general recursion". Is the rational to fill it in for this case, that it can use operations like "reduce" on arrays? It would be good to be clearer regarding this already in the introduction.

Questions for the response period
Why do you present the direct translation (Section 3)? What does it add, compared to performing the translation and differentiation via UNF?

Reviewer 3:

Notable features of the approach include:
The core language studied has no form of recursion, and no conditionals. The paper briefly discusses extensions (in particular with conditionals) at the end, but still banning primitive functions from reals to booleans which generate non-smooth functions.

Weaknesses
Quite a few recent papers propose program transformations for reverse-mode AD of purely functional programs. See for instance [Vákár 2021] which also has arrays, motivates the transformation categorically and validates it via denotational semantics in diffeological spaces. [Mazza and Pagani 2021] do it for PCF, with full support of recursion and conditionals; and there are others.
In contrast, the language considered here is minimalistic. Table 1 seems to claim a partial support for recursion, but it is hard to see what that corresponds to as there is no recursion or iteration in the language. I guess it is map/reduce? The table also claims that conditionals are supported, but this is only briefly sketched in Section 7.2, and this is without primitives R→ B such as >0 making conditionals useful.
The UNF syntax and conversions seems more like a tool than an end in itself, though it does make differentiation more intuitive.
So aside from efficiency, I am not sure which aspect of this work to put forward with respect to existing literature. The paper should also be more upfront about the fact that the main original contribution, with respect to existing work, is a proof of the cheap gradient principle.
Explanations are a bit lacking and assume familiarity which gradients, Jacobians etc.

Comments for author
Section 1.1: the example is hard to follow, because no typing and/or explanation for the components is given, and this appears before the introduction to AD in Section 2.
Section 2.1: this is titled "Rudiments of forward-mode AD and dual numbers", but it only seems to introduce the objective of AD, not forward-mode.
Section 2.2: "The problem with the previous approach shows up when one wants to compute the full gradient of a function R n → R, for a large n. Forward-mode only computes one directional derivative..." -> this sounds weird since we have not introduced forward-mode.
p.4: "Typically ... completely missed" -> this largely ignores the recent literature on the topic.
p.5: the brackets <, > make the spacing really awkward. Please use \langle, \rangle. Plus, pairing of functions is written (f, g) before (in Section 2.1).
p.5: "x′+ = ∂1op(x, y); y′+ = ∂2op(x, y)" -> spacing makes it really hard to parse.
p.5: "We have to carry a continuation and β-reduce a lot of higher-order functions." -> this is a bit imprecise. Why is that a problem?
p.6: "call-by-value" -> sounds a bit strange for a first-order language without calls (nor operational semantics).
p.6: "The types T1 , T2 , terms e1 , e2" -> why not "types T, terms e"? Besides, sometimes indices appear sometimes as subscripts, and sometimes not. This happens inconsistently throughout the whole paper.
Figure 3: the translation also involves partial derivatives of operations, which do not appear here.
Figure 4: "∇Γ e" -> inconsistent, sometimes this is ∇Γ(e), sometimes not.
Figure 4: in the clauses ZerosLike(n) and OnesLike(n), x is unbound.
p.8: reverse-mode AD is not actually defined! Well it is defined in Figure 5, two pages later, and Figure 5 is not referred to before p.12.
Lemma 3.2: but we still do not have the definition!
p.9: "UNF introduces the call-by-value evaluation of the arguments before evaluating the arguments and a good management of the environment" -> ?
Figure 5: missing '(' in the clause for let
p.11: "cos_{R, R}" -> doesn't match the grammar of terms. Should it be "cos_{R, R; 2}"?
Figure 7: the grammar for types implies that one cannot form e.g. [R, R] -> R (as we only add lists of types from target). Is that intended?
Figure 7: what is [A -> B]?
The arrow appears sometimes as ->, sometimes as →
p.13: "Doing UNF followed by UNF^{−1} performs some version of the ANF transformation" -> this sounds weird since later on it is said that UNF^{-1} has no general typing preservation property and should be performed after AD.
Lemma 5.1: "Every variable ... is used exactly once" -> in what sense? Only one occurrence (so not stable under beta-reduction)? Linear typing?
p.15: "This rewriting ... movement of variables" -> this does not make sense before the cost model is defined.
Figure 12: suddenly conditionals appear. Why?
p.17: "The induction hypothesis is then too weak to conclude" -> but you have not said what the IH is.
p.19: "modulo the ηβ-laws" -> undefined, and Source does not have functions. Do you mean ηβ for products?
p.19: "SynSource satisfies the following universal property" -> what about arrays? reduce/map? let?
p.19: "⟨∆n,,[[e2]],[[e2]]⟩; swap;([[e1]])×n" -> why ",,"? Also, spacing is off.
p.20: I was not aware of "concategories". I don't think it is a well-known notion. There should be a reference, including derived notions such as cartesian concategories which are used in the paper without much introduction.
p.20: "explain the UNF and UNF^{−1} transformations." -> font inconsistent.
p.20: "We denote by ... in the context T" -> font for T inconsistent.
Please give the typing of [[-]]!
p.21: Notations \Lambda and \Lambda^{-1} not introduced.
Definition 6.6: "Mutlicategory"
What is \underline{-}? What are u_{\underline{B}_i}? A "++" is split in two "+" around the line.
p.21: "We can interpret Source in C_{Source} as follows" -> is this an interpretation or a functor? A functor from SynSource to C_{Source}?
There are inconsistencies in the font of 'e' below.
The naming convention of categorical structure SynSource/C_{Source} etc is not enlightening. Besides, SynSource is not used beyond p.19.
p.21: "Seen as morphisms in C_{Source}" -> what is seen as a morphism in C_{Source}?
p.22: the diagram is not very clear. Where is AD from Source to Target? Proposition 6.10 corresponds to the commutation of which square?

Questions for the response period
Why does the language not include a primitive to create an array? Doesn't that limit expressivity?

Lemma 5.1: Every variable ... is used exactly once" -> in what sense? Only one occurrence (so not stable under beta-reduction)? Linear typing?

### 2021-09-06 rad-icfp-2021/complexity.tex
starting fixing things after POPL reviews

@@
 and we can use partial evaluation to beta-reduce all these lambda abstractions. 
+By inspecting each rule in Figure~\ref{fig:direct_diff_macro}, this allows us to prove by induction on the judgment of !e!:
 
@@
     Let $\Gamma \vdash$!e: A! be a term in Source. 
-    Every variable !Y! in $\directD{\rho}{\Gamma}{Y}$(!e!) which is not of ground type is used exactly once.
+    Every variable !Y! in $\directD{\rho}{\Gamma}{Y}$(!e!) which is not of ground type has exactly one occurrence in the term.
 \end{lemma}
 
-\begin{proof}
-    By induction on the judgment of !e!. We inspect each rule in Figure~\ref{fig:direct_diff_macro}.
-\end{proof}
-
-From the previous lemma, we see that there is a linear usage of  each continuation $Y$. 
-This is one key property which has been ensured by $\UNFSymbol$.
+From the previous lemma, we see that there is a linear usage of each continuation variable $Y$. 
+This is one key property ensured by $\UNFSymbol$.
 We also note that, apart from the !map2! case, this continuation variable is always applied to almost the identity. 
@@
     \label{lem:noHO}
-    Let $\Gamma \vdash$!e: A! be a term in Source. 
+    Let $\Gamma \vdash$ !e:A! be a term in Source. 
     After the partial evaluation step, $\directD{\rho}{\Gamma}{Y}$(!e!) does not have lambda abstractions.
@@
 The gives a complexity function $\cost$ valued in $\RR^4$. 
-For instance, we have
+For primitive operations, we have for instance
 
 \begin{tabular}{ll}
-    $\cost(*)=(3,0,1,0)$ & $\cost(+)=(3,1,0,0)$\\
-    $\cost(c)=(1,0,0,0)$ & $\cost($!sin!$)=(2,0,0,1)$
+    $\cost$(!*!)=$(3,0,1,0)$ & $\cost($!+!$)=(3,1,0,0)$\\
+    $\cost$(!c!)=$(1,0,0,0)$ & $\cost($!sin!$)=(2,0,0,1)$
 \end{tabular}
 
+More generally $\cost($!op1!$)=(2,0,0,1)$ for the other unary operations.
 For simplicity, we will not address any parallelism in our cost model. 
@@
 \begin{tabular}{r c l}
-    $\NestA$(!c!) &=& 0 \\
-    $\NestA$(!x!) &=& 0 \\
-\end{tabular}
-    \begin{tabular}{r c l}
-    $\NestA$($\pi_i$(!e!)) &=& $\NestA$(!e!)\\
-    $\NestA$(!op1 e!) &=& $\NestA$(!e!) \\
-\end{tabular}
-\end{center}
-
-\begin{center}
-\begin{tabular}{r c l}
-    $\NestA$(!let x:A = e$_1$ in e$_2$:B!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)  \\ 
-    $\NestA$(!< e$_1$, e$_2$ >:AxB!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!) \\ 
-    $\NestA$(!e$_1$ op2 e$_2$!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)\\
-    $\NestA$(!map2 (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
-    $\NestA$(!reduce (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
+    $\NestA$(!c!), $\NestA$(!x!) &=& 0 \\
+    $\NestA$($\pi_i$(!e!)), $\NestA$(!op1 e!) &=& $\NestA$(!e!) \\
+    $\NestA$(!let x:A = e$_1$ in e$_2$:B!) &=& max($\NestA$(!e$_1$!), $\NestA$(!e$_2$!))  \\ 
+    $\NestA$(!< e$_1$, e$_2$ >:AxB!) &=& max($\NestA$(!e$_1$!), $\NestA$(!e$_2$!)) \\ 
+    $\NestA$(!e$_1$ op2 e$_2$!) &=& max($\NestA$(!e$_1$!), $\NestA$(!e$_2$!))\\
+    $\NestA$(!map2 (x,y.e$_1$) e$_2$ e$_3$!) &=& max(1+$\NestA$(!e$_1$!), $\NestA$(!e$_2$!), $\NestA$(!e$_3$!)) \\
+    $\NestA$(!reduce (x,y.e$_1$) e$_2$ e$_3$!) &=& max(1+$\NestA$(!e$_1$!), $\NestA$(!e$_2$!), $\NestA$(!e$_3$!))
 \end{tabular}
@@
 There is a first restriction that prevents our transformation from satisfying the cheap gradient principle.
-If we try to show that the cheap gradient principle holds by induction on the term, it fails on the !map2!.
+If we try to show that the cheap gradient principle holds by induction on the term, it fails on !map2!.
 When differentiating !map2!, there are three series of calls to (sub)gradients of !e$_1$!. 
@@
 
-In Figure~\ref{fig:optim} we present a of optimizations. 
+In Figure~\ref{fig:optim} we present a list of optimizations. 
 These optimizations are all obviously valid according to the semantic model. 
@@
 The purity of our transformation allows us to make the most out of generic optimizations.  
-A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.
 In addition, hand-crafted efficient derivatives and more optimizations can easily be added to our language.

### 2021-09-06 rad-icfp-2021/correctness.tex
Indices fix
starting fixing things after POPL reviews

@@
     $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& $id_T\concatcomp map2_{T;x,y.e}$ \\
-    $\seml$!e$_1$!$\icomp$!e$_2$!$\semr$ &=& $\Lambda$($id_\Gamma\concatcomp\seml$!e2!$\semr$; $\Lambda^{-1}$($\seml$!e1!$\semr$)) \\
+    $\seml$!e$_1$!$\icomp$!e$_2$!$\semr$ &=& $\Lambda$($id_\Gamma\concatcomp\seml$!e$_2$!$\semr$; $\Lambda^{-1}$($\seml$!e$_1$!$\semr$)) \\
     $\seml$!J!$^T$!reduce!$_{T;x,y.e_1;e_2}\semr$  &=& $id_T\concatcomp reduce_{T;x,y.e_1;e_2}$ \\
@@
     $\seml$!J!$^T$!op!$_{T;n} \semr$ &=& $id_T\concatcomp J^Top_n$ \\
-    $\seml$!<e$_1$, e$_2$>!$\semr$  &=& !<!$\seml$!e1!$\semr$, $\seml$!e2!$\semr$!>! \\
+    $\seml$!<e$_1$, e$_2$>!$\semr$  &=& !<!$\seml$!e$_1$!$\semr$, $\seml$!e$_2$!$\semr$!>! \\
     $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& $id_T\concatcomp$!<$\pi_1,\pi_2$>! \\
@@
 The difference is that to preserve typing, the semantic $\UNFSymbol^{-1}$
-sends a non-Jacobian primitive to a tuple, as in the Example \ref{exm:unf}. 
+sends a non-Jacobian primitive to a tuple, as in Example \ref{exm:unf}. 
 This is highly inefficient, and the syntactic $\UNFSymbol^{-1}$ 
@@
 We construct a functor $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$. 
-This functor is easily shown to compute reverse-mode derivatives.
-Because terms of Source UNF in $\ConcatS$, we observe the effect of $\Dsynrevsymbol_{\rho}$ on them
+This functor computes reverse-mode derivatives. \MH{define more precisely}
+Because terms of Source UNF are interpreted in $\ConcatS$, we observe the effect of $\Dsynrevsymbol_{\rho}$ on them
 and show that it matches the syntactic $\Dsynrevsymbol_{\rho}$ 
@@
     where $\pi_{last}$ is the projection to the last element (the continuation variable).
-    It naturally extends on sequential composition. 
-    It does not automatically extend to a multi-functor between concategories because \\
+    It naturally extends to sequential composition. 
+    It does not automatically extend to a multi-functor \MH{define} between concategories because \\
     !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$]) $\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp\Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.\\
@@
     $(x_1,\ldots,x_k,Y_1,(x_{k+1},\ldots,x_n,Y_s)\mapsto (x_1,\ldots,x_n,\lambda(y_1,\ldots,y_n)\to Y_1(y_1,\ldots,y_k)\widehat{+}Y_2(y_{k+1},\ldots,y_n)$.
-    Here, $\widehat{+}$ is the reverse-derivative of the copy map which is known to be fanout. 
+    Here, $\widehat{+}$ is the reverse-derivative of the copy map, which is known to be fanout \MH{define}. 
     It is not surprising to see it appear as $\Dsynrevsymbol_{\rho}$ is a semantic functor, and it does not need to be efficient. 

### 2021-09-06 rad-icfp-2021/generalization.tex
starting fixing things after POPL reviews

@@
 Similarly to the monoid $\widehat{+},0_\Gamma$ defined in Section~\ref{sub:Macro for pure reverse mode transformation},
-we use $\widehat{\times},1_\Gamma$ for obvious extension of the monoid $(\reals,\times,1)$. The modified reverse-mode transformation is shown in Figure~\ref{fig:lift_reduce}.
-
-We can now compute the reverse-mode transformation as follows.
+we use $\widehat{\times},1_\Gamma$ for obvious extension of the monoid $(\reals,\times,1)$. 
+The modified reverse-mode transformation is shown in Figure~\ref{fig:lift_reduce}.
 
@@
 \begin{tabular}{l}
-    $\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) \\
+    $\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) = \\
     !let b=e$_1$ in <if b then e$_2$ else e$_3$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
@@
 there is no need to go through UNF; most of the backend of our work can be used as a black box.
-The non-smooth variables of the operator (typically booleans or integers) should be considered as external to the language, similarly to the way we treat the index $n$ of arrays.
+Non-smooth variables (typically booleans or integers) should be considered as external to the language, similarly to the way we treat indices $n$ of arrays.
 

### 2021-09-06 rad-icfp-2021/relwork.tex
starting fixing things after POPL reviews

@@
 This is reminiscent of hand-crafted derivatives present in large AD frameworks (usually hundreds), 
-and these can already be added in added in our language. 
+and these can already be added in our language. 
 As shown in \ref{sub:lift_recipe}, one mostly only needs to make sure that the provided transpose Jacobian is correct.

### 2021-09-06 rad-icfp-2021/unf.tex
saving more space
removed unfinished example
starting fixing things after POPL reviews

@@
 Just as Target is an extension of Source, Target UNF is an extension of Source UNF.
-They serve as intermediate representation our the reverse-mode compilation pipeline (see Figure~\ref{fig:outline}).
+They serve as intermediate representations in the reverse-mode compilation pipeline (see Figure~\ref{fig:outline}).
 
@@
 
-The Target UNF is an extension of Source UNF. 
+Target UNF is an extension of Source UNF. 
 Its types are lists of types of Target. 
@@
 This weakening is required when we transform a non-unary operator such as $+$ or !pair!, as we transform the arguments in order and we need to keep all of their results.
-We write this new term $(\widetilde{e_{2}})_{A}$ and it defined by induction on !e$_{2}$! as follows. 
+We write this new term $(\widetilde{e_{2}})_{A}$. It is defined by induction on !e$_{2}$! as follows. 
 We assume given a type !A! that needs to be passed, and drop the index !A!.
@@
     $\widetilde{var_{T;i}}$  &=& $var_{T,A;i}$ \\
-    $\widetilde{op_{T;n}}$  &=& $op_{T,A;n}$ \\§
+    $\widetilde{op_{T;n}}$  &=& $op_{T,A;n}$ \\
     $\widetilde{pair_{T;B\times C}}$  &=& $pair_{T,A;B\times C}$ \\
-    $\widetilde{proj_{T_1;T_2;T_3}}$  &=& $proj_{T_1;T_2;T_3,A}$ \\
+    $\widetilde{proj_{T_1;T_2;T_3}}$  &=& $proj_{T_1;T_2;T_3,A}$ 
+\end{tabular}
+\begin{tabular}{r c l}
     $\widetilde{e_1 \comp e_2}$  &=& $\widetilde{e_1}\comp\widetilde{e_2}$ \\
     $\widetilde{map2_{T;x,y.e}}$  &=& $map2_{T,A;x,y.e}$ \\
-    $\widetilde{reduce_{T;x,y.e_1;e_2}}$ &=& $reduce_{T,A;x,y.e_1;e_2}$ \\
+    $\widetilde{reduce_{T;x,y.e_1;e_2}}$ &=& $reduce_{T,A;x,y.e_1;e_2}$
 \end{tabular}
@@
 
-\begin{example}
-    Continuing again with our previous example !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$!, 
-    we have, after removing a few unnecessary let-bindings:
+% \begin{example}
+%     Continuing again with our previous example !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$!, 
+%     we have, after removing a few unnecessary let-bindings:
 
-    \begin{tabular}{l}
-        !x$_1$,x$_2$,x$_3$:$\reals$,x$_4$:$\reals^3$->$\rho$ $\vdash$ $\UNFSymbol^{-1}$($\Dsynrevsymbol_{\rho}$(cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$): $(\reals\times\reals) \times (\reals\times\reals\times(\reals\times\reals))$->$\rho$!
+%     \begin{tabular}{l}
+%         !x$_1$,x$_2$,x$_3$:$\reals$,x$_4$:$\reals^3$->$\rho$ $\vdash$ $\UNFSymbol^{-1}$($\Dsynrevsymbol_{\rho}$(cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$): $(\reals\times\reals) \times (\reals\times\reals\times(\reals\times\reals))$->$\rho$!
 \\
-        = !TODO!
-    \end{tabular}
-\end{example}
+%         = !TODO!
+%     \end{tabular}
+% \end{example}
 
